### 2024-06-20 ms_figure_1_overview.Rmd
Add Figure 1 report
Co-authored-by: Michael B Stadler <stadler.michael@gmail.com>
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>

---
title: "Figure 1 - overview"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
        self_contained: true
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    rdsrnaseq: "data/rnaseq_del.rds"
    dbdtxt: "data/TF_DBD_annotation.txt"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    baitclass: "data/ipms_bait_class.txt"
    idmap: "data/id_mapping_table.txt"
    tagcntfile: "data/taggedTFs_counts.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
options(scipen = 6)
```

# Parameter values

```{r params}
params
```

# Load required packages and helper functions

```{r load-packages, class.source="fold-show"}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(ggplot2)
    library(cowplot)
    library(einprot)
    library(dplyr)
    library(forcats)
    library(jsonlite)
    library(ComplexHeatmap)
    library(ggrepel)
    library(colorspace)
})

source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/ipms_params.R")
source("params/mapping_functions.R")
```

# Read data

We load two `SingleCellExperiment` objects, containing data and results from the 
low- and high-salt IPMS experiments, respectively.

```{r load-data, class.source="fold-show"}
sce150 <- readRDS(params$rds150)
sce500 <- readRDS(params$rds500)
```

We extract the full list of baits, as well as load a classification in terms of 
the number of interaction partners.

```{r, class.source="fold-show"}
## Feature ID mapping table
idmap <- read.delim(params$idmap)

## Experiment names
allexps <- unique(sub("(ID[0-9]+|SPB4482)_([^_]*)_?.*", "\\2", colnames(sce150)))
allexps <- allexps[allexps != "untagged"]

## Corresponding bait protein names
stopifnot(all(allexps %in% idmap$bait))
allbaits <- idmap$unique_einprot_id[match(allexps, idmap$bait)]

baitclass <- read.delim(params$baitclass)
stopifnot(all(.capitalize(allexps) %in% baitclass$Gene_name))
table(class150 = baitclass$class, 
      class500 = baitclass$class500, useNA = "ifany")
```

In the following set of IPs, we identified the bait as not pulled down in the 
low-salt condition.

```{r define-nobait}
(nobait_150 <- baitclass$Gene_name[baitclass$class == "nobait"])
```

Next, we load a `SummarizedExperiment` with quantifications from RNA-seq data.

```{r, class.source="fold-show"}
rnaseq <- readRDS(params$rdsrnaseq)
stopifnot(all(allbaits %in% rownames(rnaseq)))
```

We also load the information about the DNA-binding domain family for each 
transcription factor.

```{r load-dbd, class.source="fold-show"}
dbd <- read.delim(params$dbdtxt) |>
    dplyr::group_by(DBD_class, Conserved_human) |>
    dplyr::tally() |>
    dplyr::mutate(cons = ifelse(Conserved_human, "Conserved in human",
                                "Not conserved in human")) |>
    dplyr::group_by(DBD_class) |>
    dplyr::mutate(totn = sum(n)) |>
    dplyr::ungroup()
```

Here, we load the filtered peaks identified from ChIP-seq experiments, with
the information about which baits showed enrichments for each peak.

```{r load-peaks, class.source="fold-show"}
peaks <- read.csv(params$peakcsv, row.names = 1)
enrcolumns <- grep("^is_enr_in[.]", colnames(peaks))
peakenr <- as.matrix(peaks[, enrcolumns])
colnames(peakenr) <- sub("^is_enr_in[.]", "", colnames(peakenr))
nbrPeaks <- colSums(peakenr)
nbrPeaksSpec <- colSums(peakenr[peaks$peaktype == "specific peaks", ])
```

Finally, we load the number of ChIP-seq fragments in input samples that did not
map to the wild-type genome sequence, but to one of the tagged transcription
factor constructs.

```{r load-tagcnt, class.source="fold-show"}
tagcnt <- as.matrix(read.csv(file = params$tagcntfile, row.names = 1))
```

# Preprocessing

## Number of interactors in IP-MS for each TF

We summarize the number of observed proteins pulled down for each TF in 
the IP-MS data (low and high salt conditions), excluding the bait itself. 
A protein is considered pulled down if the adjusted p-value in a comparison 
against the large complement group (see manuscript for details) is below 
`r adjpThreshold` and the log-fold change exceeds `r log2fcThreshold`.

```{r calc-nbr-interactors-by-tf}
#| message: false
#| warning: false

tc150 <- .getTestCols(sce150, adjp_cutoff = adjpThreshold,
                      logfc_cutoff = log2fcThreshold)
tc500 <- .getTestCols(sce500, adjp_cutoff = adjpThreshold,
                      logfc_cutoff = log2fcThreshold)

## Get matrix indicating whether the proteins pass the interaction threshold 
## in the experiments
interactors150 <- tc150$interactor
bait_idx150 <- cbind(match(.getProteinNameFromComparison(colnames(interactors150)), 
                           rownames(interactors150)), 
                     seq_len(ncol(interactors150)))
interactors150[bait_idx150] <- NA
colnames(interactors150) <- .getOrigBaitNameFromComparison(colnames(interactors150),
                                                           idmap = idmap)
interactors150 <- lapply(structure(colnames(interactors150), names = colnames(interactors150)), 
                         function(cn) {
                             rownames(interactors150)[which(interactors150[, cn])]
                         })

interactors500 <- tc500$interactor
bait_idx500 <- cbind(match(.getProteinNameFromComparison(colnames(interactors500)), 
                           rownames(interactors500)), 
                     seq_len(ncol(interactors500)))
interactors500[bait_idx500] <- NA
colnames(interactors500) <- .getOrigBaitNameFromComparison(colnames(interactors500),
                                                           idmap = idmap)
interactors500 <- lapply(structure(colnames(interactors500), names = colnames(interactors500)), 
                         function(cn) {
                             rownames(interactors500)[which(interactors500[, cn])]
                         })
## For this figure, we only care about the interactors that are _retained_ in 500 mM
## (not new ones that are detected e.g. due to the different complements)
for (nm in names(interactors500)) {
    interactors500[[nm]] <- intersect(interactors500[[nm]], interactors150[[nm]])
}
intdf <- dplyr::bind_rows(
    data.frame(exp = "150 mM NaCl IP-MS",
               id = names(interactors150),
               nbrInteractions = vapply(interactors150, length, NA_real_),
               inHighSalt = names(interactors150) %in% names(interactors500)),
    data.frame(exp = "Retained in 500 mM NaCl IP-MS",
               id = names(interactors500),
               nbrInteractions = vapply(interactors500, length, NA_real_),
               inHighSalt = TRUE)
) |>
    dplyr::left_join(baitclass, by = c("id" = "Gene_name")) |>
    dplyr::mutate(idorig = id) |>
    dplyr::mutate(id = .getProteinFromOrigBait(idorig,
                                               idmap = idmap)) |>
    dplyr::mutate(id = replace(id, idorig == "Untagged", "Untagged"))
```

# Facts and numbers - ChIP-seq

```{r calc-bound-sites, class.source="fold-show"}
## How many TFs did we analyze by ChIP-seq?
is_TF <- colnames(peakenr) != "Untagged"
sum(is_TF)

## How many unique peak regions did we identify?
nrow(peaks)

## How many binding events (any TF) do we have in total?
sum(peakenr[, is_TF])

## How many TFs have at least one peak?
summary(colSums(peakenr)[is_TF] > 0)
colnames(peakenr)[is_TF & colSums(peakenr) == 0]

## What is the range of numbers of peaks over TFs?
summary(colSums(peakenr)[is_TF])
summary(colSums(peakenr)[is_TF & colSums(peakenr) > 0]) # without zero-peak TFs

## What fraction of TFs is bound at a given site?
## ... for all peaks
summary(rowMeans(peakenr[, is_TF]))

## ... for peaks by type
table(peaks$peaktype)
summary(rowSums(peakenr[peaks$peaktype == "common peaks (ubiquitous)", is_TF]))
summary(rowSums(peakenr[peaks$peaktype == "common peaks (frequent)", is_TF]))
summary(rowSums(peakenr[peaks$peaktype == "specific peaks", is_TF]))
```

# Facts and numbers - IP-MS

```{r numbers-ipms, class.source="fold-show"}
## Number of NCBI Conserved Domains shown in the figure (excluding "Other")
length(setdiff(unique(dbd$DBD_class), "Other"))

## Percentage of domains conserved in human
round(100 * sum(dbd$n[dbd$Conserved_human]) / sum(dbd$n), 2)

## Number of analyzed TFs
## ... 150 mM NaCl
length(setdiff(unique(sce150$Gene_name), "untagged"))

## ... 500 mM NaCl
length(setdiff(unique(sce500$Gene_name), "untagged"))
```

# Figure 1

## Number of TFs for each DBD family

The figure below shows the number of transcription factors belonging to each 
DNA-binding domain family, further stratified by whether or not they have a 
human ortholog.

```{r plot-dbd}
#| fig.width: 6
#| fig.height: 4

(gg_dbd <- ggplot(dbd, aes(x = fct_reorder(DBD_class, totn, .desc = TRUE), y = n, 
                           fill = cons)) + 
     geom_col() + 
     scale_fill_manual(values = c("Conserved in human" = lighten(main_colors[1], 0.3),
                                  "Not conserved in human" = main_colors[1]), 
                       name = "") + 
     labs(x = "DBD family (NCBI Conserved Domains)",
          y = "Number of\ntranscription factors") + 
     theme_cowplot() + 
     theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
           legend.position = "inside",
           legend.position.inside = c(0.55, 0.9)))
```

## WT RNA-seq TPM values for TFs

Next, we extract the average expression (in TPM - transcripts per million) for 
each transcription factor across three wild-type samples.

```{r plot-tpm-tfs}
#| fig.width: 6
#| fig.height: 4

ave_tpm <- data.frame(
    bait = allbaits, 
    bait_pulled = factor(
        ifelse(.capitalize(allexps) %in% nobait_150, 
               "Bait not pulled down (150 mM NaCl IP-MS)",
               "Bait pulled down (150 mM NaCl IP-MS)"),
        levels = c("Bait pulled down (150 mM NaCl IP-MS)",
                   "Bait not pulled down (150 mM NaCl IP-MS)")),
    tpm = rowMeans(assay(rnaseq, "abundance")[allbaits, 
                                              rnaseq$Group_name == "wt_WT_mRNA"])
)
gg_tpm <- ggplot(ave_tpm, aes(x = fct_reorder(bait, tpm, .desc = TRUE), y = tpm, 
                              color = bait_pulled)) + 
    geom_hline(yintercept = 0, linetype = "dashed", color = na_color) +
    geom_point(size = 1) + 
    theme_cowplot() + 
    theme(legend.position = "inside", legend.position.inside = c(0.15, 0.9)) +
    scale_color_manual(values = c("Bait pulled down (150 mM NaCl IP-MS)" = pt_color,
                                  "Bait not pulled down (150 mM NaCl IP-MS)" = na_color),
                       name = "") + 
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
    labs(x = "Transcription factors",
         y = "Transcription factor\ngene expression (TPM)")
(gg_tpm + theme(axis.text.x = element_blank(), 
                axis.ticks.x = element_blank()))
```

## Number of interactors in IP-MS for each TF

Here, we plot the number of interactors for each TF in the 150 mM NaCl 
condition, as well as the number of interactors retained in the 500 mM NaCl
condition.

```{r nbr-interactors-by-tf}
#| fig.width: 8
#| fig.height: 5
#| message: false
#| warning: false

idOrder <- intdf |>
    dplyr::filter(exp == "150 mM NaCl IP-MS") |>
    dplyr::arrange(desc(nbrInteractions)) |>
    dplyr::pull(id)
(gg_nbrint <- ggplot(intdf |>
                         dplyr::filter(!id %in% nobait_150 & 
                                           id != "Untagged") |>
                         dplyr::filter(inHighSalt),
                     aes(x = fct_relevel(id, idOrder),
                         y = nbrInteractions)) +
        geom_hline(yintercept = 0, linetype = "dashed", color = na_color) + 
        geom_line(color = na_color) + 
        geom_point(shape = 21, aes(fill = exp, color = exp, size = exp)) +
        scale_fill_manual(values = c(`150 mM NaCl IP-MS` = main_colors[5], 
                                     `Retained in 500 mM NaCl IP-MS` = pt_color), 
                          name = "") + 
        scale_color_manual(values = c(`150 mM NaCl IP-MS` = main_colors[5], 
                                      `Retained in 500 mM NaCl IP-MS` = "transparent"),
                           name = "") + 
        scale_size_manual(values = c(`150 mM NaCl IP-MS` = 3, 
                                     `Retained in 500 mM NaCl IP-MS` = 2),
                          name = "") + 
        theme_cowplot() + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1,
                                         size = 9, vjust = 0.5), 
              legend.position = c(0.35, 0.9)) + 
        labs(x = "", y = "Number of interactors") + 
        coord_cartesian(ylim = c(0, NA)))
```

```{r}
(gg_nbrint_all <- ggplot(
    intdf |>
        dplyr::filter(exp == "150 mM NaCl IP-MS" & 
                          id != "Untagged"),
    aes(x = fct_relevel(id, levels(fct_reorder(.capitalize(ave_tpm$bait),
                                               ave_tpm$tpm, .desc = TRUE))), 
        y = nbrInteractions)) + 
     geom_hline(yintercept = 0, linetype = "dashed", colour = na_color) +
     geom_line() + 
     geom_point(shape = 21, size = 3, colour = main_colors[5],
                fill = main_colors[5]) +
     theme_cowplot() + 
     theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), 
           legend.position = "bottom") + 
     labs(x = "", y = "Number of interactors") + 
     coord_cartesian(ylim = c(0, NA)))
```

## Number of peaks for each TF

```{r nbr-peaks-by-tf-main}
#| fig.width: 8
#| fig.height: 5
#| warning: false

## Number of peaks per TF (unfiltered)
pd0 <- intdf |>
    dplyr::filter(exp == "150 mM NaCl IP-MS" & id != "Untagged") |>
    dplyr::mutate(nbrPeaks = nbrPeaks[id],
                  nbrPeaksSpec = nbrPeaksSpec[id]) |>
    dplyr::mutate(chipped = !is.na(nbrPeaks))

## Only chipped TFs, in long format
pd <- pd0 |>
    dplyr::filter(chipped) |>
    dplyr::mutate(id = fct_reorder(factor(id), .x = nbrPeaks,
                                   .fun = function(x) x, .desc = TRUE)) |>
    tidyr::pivot_longer(cols = c("nbrPeaks", "nbrPeaksSpec"),
                        names_to = "peakType",
                        values_to = "peakNumber") |>
    dplyr::mutate(peakType = fct_relabel(
        peakType,
        function(x) c(nbrPeaks = "All ChIP peaks",
                      nbrPeaksSpec = "Specific ChIP peaks")[x]))

(gg_nbrpeaks_all_narrow_by_npeaks <- ggplot(pd |> dplyr::filter(chipped),
    aes(x = id, y = peakNumber)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = na_color) +
    geom_line(color = na_color) +
    geom_point(mapping = aes(color = peakType, fill = peakType, size = peakType), shape = 21) +
    scale_fill_manual(values = c(`All ChIP peaks` = main_colors[5],
                                 `Specific ChIP peaks` = pt_color), name = "") +
    scale_color_manual(values = c(`All ChIP peaks` = main_colors[5],
                                  `Specific ChIP peaks` = pt_color), name = "") +
    scale_size_manual(values = c(`All ChIP peaks` = 1.75,
                                 `Specific ChIP peaks` = 1),
                      name = "") +
    theme_cowplot() + 
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          legend.position = c(0.55, 0.9)) + 
    labs(x = "Transcription factors",
         y = "Number of enriched peaks\n(IP/input > 2)") + 
    coord_cartesian(ylim = c(0, NA)))
```

```{r nbr-peaks-by-tf-supp}
#| fig.width: 8
#| fig.height: 5
#| warning: false

(gg_nbrpeaks_all_wide <- ggplot(pd0,
    aes(x = fct_relevel(id, levels(fct_reorder(.capitalize(ave_tpm$bait), ave_tpm$tpm, .desc = TRUE))))) + 
    geom_hline(yintercept = 0, linetype = "dashed", color = na_color) +
    geom_segment(
        mapping = aes(y = nbrPeaks, yend = nbrPeaksSpec,
                      xend = fct_relevel(id, levels(fct_reorder(.capitalize(ave_tpm$bait), ave_tpm$tpm, .desc = TRUE)))),
        color = na_color) +
    geom_point(mapping = aes(y = nbrPeaks),
               shape = 21, size = 3.5, fill = main_colors[5], color = main_colors[5]) +
    geom_point(mapping = aes(y = nbrPeaksSpec),
               shape = 21, size = 2.5, fill = pt_color) +
    geom_tile(data = pd0 |> dplyr::filter(!chipped),
              aes(x = fct_relevel(id, levels(fct_reorder(.capitalize(ave_tpm$bait), ave_tpm$tpm, .desc = TRUE))),
                  y = I(0)),
              width = 0.75, height = Inf, fill = na_color) +
    geom_point(aes(x = 3, y = 400), size = 3.5, fill = main_colors[5], color = main_colors[5], shape = 21) +
    geom_point(aes(x = 3, y = 375), size = 2.5, fill = pt_color, shape = 21) +
    geom_tile(aes(x = 3, y = 350), width = 2, height = 15, fill = na_color) +
    annotate("text", x = 5, y = c(400, 375, 350), hjust = 0, vjust = 0.5,
             label = c("All ChIP peaks", "Specific ChIP peaks", "No ChIP data"),
             size = 4.2, fontface = "plain") +
    theme_cowplot() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
    labs(x = "", y = "Number of peaks (IP/input > 2)") + 
    coord_cartesian(ylim = c(0, NA)))
```

## Sequence evidence for tagged transcription factors

In order to validate the genomic sequences of tagged transcription factors
in the engineered strains, we used all reads from ChIP input samples that
did not map to the *S. pombe* reference (wildtype) genome.

For each sample, the unmapped reads were aligned against all tagged
transcription factor sequences. The resulting  alignment counts are summarized
in the `tagcnt` matrix of `r nrow(tagcnt)` rows by `r ncol(tagcnt)` columns.
The rows correspond to tagged transcription factor sequences, and the columns
to *S. pombe* strains in which a transcription factor has been tagged.
The values correspond to the number of read pairs from ChIP input samples of
each strain, that mapped unambigously to one of the expected tagged
transcription factor constructs.

Tagged transcription factor sequences and *S. pombe* strains are ordered
in the same way, expect for an additional column on the right side
corresponding to the untagged parent strain (`r colnames(tagcnt)[ncol(tagcnt)]`).

To account for differences in sequencing depth, the counts are normalized
by dividing through the total number of read pairs that mapped to each 
tagged transcription factor sequence, and then visualized as a heatmap.

```{r plot-taggedTFs}
#| fig.width: 6
#| fig.height: 6

# define heatmap dimensions
w_taggedTFs <- 130  # total heatmap width (mm)
h_taggedTFs <- 130  # total heatmap height (mm)
fs <- 5  # small font size
fl <- 10  # large font size

# normalize (percent of reads)
perc <- sweep(tagcnt, 1, rowSums(tagcnt), "/") * 100

rng <- range(perc)
hm5 <- Heatmap(matrix = perc, name = "Percent of reads mapped to\ntagged transcription factor ",
               cluster_rows = FALSE, cluster_columns = FALSE,
               layer_fun = function(j, i, x, y, width, height, fill) {
                   grid.rect(x = x, y = y, width = width, height = height, 
                             gp = gpar(col = "grey95", fill = NA))
               },
               column_title = "Strains",
               row_title = "Tagged transcription factor",
               row_title_gp = gpar(fontsize = fl),
               row_names_gp = gpar(fontsize = fs),
               column_title_gp = gpar(fontsize = fl),
               column_names_gp = gpar(fontsize = fs),
               col = circlize::colorRamp2(breaks = seq(rng[1], rng[2], length.out = 64),
                                          colors = rev(hcl.colors(64, "Purples"))),
               heatmap_legend_param = list(labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(20, "mm"),
                                           title_position = "lefttop",
                                           title_gp = gpar(fontsize = fl)),
               heatmap_width = unit(w_taggedTFs, "mm"), # total heatmap width
               heatmap_height = unit(h_taggedTFs, "mm"), # total heatmap height
               use_raster = TRUE)

hm_taggedTFs <- grid.grabExpr(
    hm5 <- draw(hm5, heatmap_legend_side = "bottom"),
    width = w_taggedTFs, height = h_taggedTFs
)

plot_grid(hm_taggedTFs)
```

## Put together

```{r Fig1}
#| fig.width: 12
#| fig.height: 17.3
#| warning: false

fig1c <- ggdraw() + 
    draw_image("schematics/Fig1C_placeholder_screen_overview_Arial.png")

cowplot::plot_grid(
    cowplot::plot_grid(
        gg_dbd, 
        gg_tpm + theme(axis.text.x = element_blank(), 
                       axis.ticks.x = element_blank()),
        nrow = 1,
        labels = c("A", "B"),
        align = "vh",
        axis = "b"
    ),
    fig1c,
    cowplot::plot_grid(
        gg_nbrint,
        gg_nbrpeaks_all_narrow_by_npeaks,
        nrow = 1, 
        labels = c("D", "E"),
        align = "vh", 
        axis = "b"
    ),
    nrow = 3,
    labels = c(NA, "C", NA),
    align = "v",
    axis = "b",
    rel_heights = c(50, 73, 50)
)
```

## Supplementary figure

```{r suppFig1}
#| fig.width: 15
#| fig.height: 15
#| warning: false
#| message: false

cowplot::plot_grid(
    gg_tpm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
        labs(x = "") + geom_point(size = 3), 
    gg_nbrint_all,
    gg_nbrpeaks_all_wide,
    ncol = 1, 
    align = "v", 
    axis = "l",
    labels = c("A", "B", "C"),
    rel_heights = c(1, 1, 1)
)
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-06-21 params/get_testres_function.R
add hyper-parameter files
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

@@ -1,0 +1,41 @@
+## Get adjusted p-value and logFC matrices from SCE object
+.getTestCols <- function(sce, adjp_cutoff = 0.0005, logfc_cutoff = 1) {
+    rd <- rowData(sce)
+    ## Get t-statistics and adjusted p-values from all comparisons
+    tstat_cols <- setdiff(grep("\\.t$", colnames(rd), value = TRUE),
+                          c(grep("highsaltcompl", colnames(rd), value = TRUE),
+                            grep("atf1_150_tube|Atf1_150_tube", colnames(rd), value = TRUE)))
+    adjp_cols <- setdiff(grep("adj.P.Val", colnames(rd), value = TRUE),
+                         c(grep("highsaltcompl", colnames(rd), value = TRUE),
+                           grep("atf1_150_tube|Atf1_150_tube", colnames(rd), value = TRUE)))
+    pval_cols <- setdiff(grep("P.Value", colnames(rd), value = TRUE),
+                         c(grep("highsaltcompl", colnames(rd), value = TRUE),
+                           grep("atf1_150_tube|Atf1_150_tube", colnames(rd), value = TRUE)))
+    logfc_cols <- setdiff(grep("logFC", colnames(rd), value = TRUE),
+                          c(grep("highsaltcompl", colnames(rd), value = TRUE),
+                            grep("se.logFC", colnames(rd), value = TRUE),
+                            grep("atf1_150_tube|Atf1_150_tube", colnames(rd), value = TRUE)))
+
+    tstat <- rd[, tstat_cols]
+    adjp <- rd[, adjp_cols]
+    pval <- rd[, pval_cols]
+    logfc <- rd[, logfc_cols]
+
+    stopifnot(all(sub("adj.P.Val", "t", colnames(adjp)) == colnames(tstat)))
+    stopifnot(all(sub("adj.P.Val", "logFC", colnames(adjp)) == colnames(logfc)))
+    stopifnot(all(sub("adj.P.Val", "P.Value", colnames(adjp)) == colnames(pval)))
+
+    tstat <- as.matrix(tstat)
+    adjp <- as.matrix(adjp)
+    pval <- as.matrix(pval)
+    logfc <- as.matrix(logfc)
+
+    ## Define interactors based on the provided cutoffs
+    interactor <- adjp <= adjp_cutoff &
+        logfc > logfc_cutoff
+    colnames(interactor) <- sub("\\.adj.P.Val", "", colnames(interactor))
+
[…]

### 2024-06-21 params/ipms_params.R

@@ -1,0 +1,8 @@
+## Parameters related to IP-MS data
+
+## Significance thresholds for interaction experiments
+adjpThreshold <- 0.0002
+log2fcThreshold <- 1
+
+## Less stringent thresholds
+adjpThresholdLoose <- 0.01

### 2024-06-21 params/mapping_functions.R

@@ -1,0 +1,75 @@
+.capitalize <- function(x) {
+    paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
+}
+
+## Get the protein name (einprotID) corresponding to a given comparison
+.getProteinNameFromComparison <- function(x) {
+    ## x should be of the form sre1_150_plate_vs_highsaltcompl_sre1_150_plate.P.Value
+    ## returns sre1
+    .capitalize(sub("([^_]*)_[^_]*_([^_]*)_.*", "\\1", x))
+}
+
+## Get the bait name (original ID) corresponding to a given comparison
+.getOrigBaitNameFromComparison <- function(x, idmap) {
+    ## x should be of the form sre1_150_plate_vs_highsaltcompl_sre1_150_plate.P.Value
+    ## returns the original bait/experiment name for sre1
+    i <- sub("([^_]*)_[^_]*_([^_]*)_.*", "\\1", x)
+    ifelse(.capitalize(i) == "Untagged", "Untagged",
+           .capitalize(idmap$bait[match(i, .capitalize(idmap$unique_einprot_id))]))
+}
+
+## Simplify the comparison ID to just bait_{plate|tube}
+.getSimplifiedComparison <- function(x) {
+    ## x should be of the form sre1_150_plate_vs_highsaltcompl_sre1_150_plate.P.Value
+    ## returns sre1_plate
+    sub("([^_]*)_[^_]*_([^_]*)_.*", paste0("\\1", "_", "\\2"), x)
+}
+
+## Get the PomBase ID corresponding to a given comparison
+.getPomBaseIdFromComparison <- function(x, idmap) {
+    ## x should be of the form sre1_150_plate_vs_highsaltcompl_sre1_150_plate.P.Value
+    ## returns SPBC19C2.09
+    i <- sub("([^_]*)_[^_]*_([^_]*)_.*", "\\1", x)
+    ifelse(.capitalize(i) == "Untagged", "Untagged",
+           idmap$gene_stable_id[match(i, .capitalize(idmap$unique_einprot_id))])
+}
+
+## Get the UniProt ID corresponding to a given comparison
+.getUniProtIdFromComparison <- function(x, idmap) {
+    ## x should be of the form sre1_150_plate_vs_highsaltcompl_sre1_150_plate.P.Value
+    ## returns SPBC19C2.09
[…]

### 2024-06-21 params/plot_settings.R

@@ -1,0 +1,18 @@
+## Colors
+## general color map (IBM colors)
+main_colors <- c("#648FFF", "#785EF0", "#DC267F", "#FE6100", "#FFB000")
+na_color <- "gray70"
+bg_color <- "gray90"
+pt_color <- "black"
+## Binary heatmaps
+binary_heatmap_colors <- c("FALSE" = "white", "TRUE" = "darkblue")
+enrichment_heatmap_colors <- rev(hcl.colors(9, "RdBu"))
+## Complexes
+complex_colors <- c(SAGA = "#aa30aa", NuA4 = "#46c8f5")
+## Peak set colors
+peakset_colors <- c("Common (ubiquitous)" = main_colors[1],
+                    "Common peaks (ubiquitous)" = main_colors[1],
+                    "Common (frequent)" = main_colors[3],
+                    "Common peaks (frequent)" = main_colors[3],
+                    "Specific" = main_colors[5],
+                    "Specific peaks" = main_colors[5])

### 2024-06-21 params/plottracks_function.R

@@ -1,0 +1,124 @@
+# center a vector by dividing through a quantile (typically the median)
+# use a pseudocount to prevent division by zero
+.center_robust <- function(x, pscnt = 0.1, relquant = 0.5) {
+    xc <- quantile(x + pscnt, prob = relquant)
+    (x + pscnt) / xc
+}
+
+# smooth a vector using a running mean over windows of k_smooth elements
+.smooth_runmean <- function(x, k_smooth) {
+    require(S4Vectors, quietly = TRUE)
+    as.numeric(runmean(x = as(x, "Rle"),
+                       k = k_smooth, endrule = "constant"))
+}
+
+# plot genome-browser like tracks given a data.frame and gene names
+plottracks <- function(df, gns, cols = NULL, k_smooth = 101,
+                       lwd = 0.75, window_width = 13000, font_size = 14) {
+    stopifnot(exprs = {
+        require(tidyverse, quietly = TRUE)
+        require(cowplot, quietly = TRUE)
+        all(c("gene_name", "start", "end", "strand",
+              "window_start", "window_end", "cpm_chip",
+              "cpm_input") %in% colnames(df))
+        window_width <= min(df |>
+                                group_by(gene_name) |>
+                                mutate(w = window_end - window_start + 1) |>
+                                ungroup() |>
+                                pull(w))
+        all(gns %in% df$gene_name)
+        is.null(cols) || all(gns %in% names(cols))
+    })
+
+    # define colours
+    if (is.null(cols)) {
+        cols <- hcl.colors(n = length(gns), palette = "Dark3")
+        names(cols) <- gns
+    }
+    cols <- c(structure(unname(cols[gns]), names = paste0(gns, "_rel_chip")),
+              structure(rep("gray", length(gns)), names = paste0(gns, "_rel_input")))
+
+
+    # filter df, smooth and sub-sample
+    pd <- df |>
+        filter(gene_name %in% gns) |>
+        group_by(gene_name) |>
+        mutate(gene_name = factor(gene_name, levels = gns),
+               pos = (seq(min(window_start), max(window_end), by = 1) -
+                          ifelse(strand == "+", start, end)) *
+                   ifelse(strand == "+", 1, -1),
+               smooth_chip = .smooth_runmean(cpm_chip, k_smooth),
+               smooth_input = .smooth_runmean(cpm_input, k_smooth),
+               rel_chip = .center_robust(smooth_chip, pscnt = 0.001),
[…]

### 2024-06-21 ms_figure_5_networks.Rmd
add figure 5
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "Figure 5 - networks"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    coords150: "data/visnetwork_coords_150_new.csv"
    rds500: "data/ipms_500_sce.rds"
    coords500: "data/visnetwork_coords_500_new.csv"
    rds150untag: "data/ipms_150_untag_sce.rds"
    interactors: "data/Gene_list_TFinteractors_Pombase20240226.txt"
    idmap: "data/id_mapping_table.txt"
    rad24_rad25_tbl: "data/FigS5_Rad24_Rad25_interactingTFs_list_motifs.txt"
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
```

# Parameter values

```{r params}
params
```

# Load required packages and helper functions

```{r load-packages, class.source="fold-show"}
suppressPackageStartupMessages({
    library(visNetwork)
    library(SingleCellExperiment)
    library(dplyr)
    library(ggplot2)
    library(igraph)
    library(shiny)
    library(cowplot)
    library(ggplotify)
    library(ggraph)
    library(latex2exp)
    library(tidygraph)
    library(ggrepel)
    library(ComplexHeatmap)
    library(flextable)
    library(kableExtra)
    library(tidyr)
})

source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/mapping_functions.R")
```

# Settings

```{r}
w_trad <-  84 # heatmap body width (mm)
h_trad <- 178 # total heatmap height (mm)

fs <- 6  # small font size
fl <- 10 # large font size
ft <- 12 # title font size
```

# Read data

Here we load `SingleCellExperiment` objects corresponding to the low- and 
high-salt IP-MS experiments, as well as an object corresponding to the low-salt
experiments, where each condition was compared to the untagged controls 
rather than the large complement group as in the other two objects.

```{r load-data, class.source="fold-show"}
sce150 <- readRDS(params$rds150)
sce500 <- readRDS(params$rds500)
sce150untag <- readRDS(params$rds150untag)
```

We also load a table with mappings between different types of feature 
identifiers, and a table with known, annotated physical interactions 
(obtained from PomBase).

```{r load-idmap, class.source="fold-show"}
idmap <- read.delim(params$idmap)

interactors <- read.delim(params$interactors) |>
    dplyr::select(Systematic.ID, Gene.name, Physical.interactors) |>
    tidyr::separate_rows(Physical.interactors, sep = ",") |>
    dplyr::mutate(Gene.name = .capitalize(Gene.name),
                  Physical.interactors = .capitalize(Physical.interactors))
```

# Helper function

```{r}
## Overwrite the get_legend function from cowplot temporarily, 
## since the cowplot one doesn't work with ggplot2 3.5.0
## see https://github.com/wilkelab/cowplot/issues/202

get_legend <- function(plot, legend = NULL) {
    gt <- ggplotGrob(plot)

    pattern <- "guide-box"
    if (!is.null(legend)) {
        pattern <- paste0(pattern, "-", legend)
    }

    indices <- grep(pattern, gt$layout$name)

    not_empty <- !vapply(
        gt$grobs[indices], 
        inherits, what = "zeroGrob", 
        FUN.VALUE = logical(1)
    )
    indices <- indices[not_empty]

    if (length(indices) > 0) {
        return(gt$grobs[[indices[1]]])
    }
    return(NULL)
}
```

# Calculate pairwise protein similarities

To define the similarity between two proteins, the following procedure is used: 
Let $t_{ij}$ and $padj_{ij}$ denote the t-statistic and adjusted p-value for 
protein $i$ in comparison $j$, respectively. Then define 
$$\tilde{t}_{ij} = \left\{\begin{array}{cl}
    0 & \quad \textrm{if } t_{ij} <= 0 \textrm{ or } padj_{ij} > 0.1 \\
    t_{ij} & \quad \textrm{otherwise}
  \end{array}\right.$$

The similarity between proteins $i$ and $k$ is then defined by 
$$sim_{ik} = \frac{\sum_{j=1}^N\tilde{t}_{ij}\tilde{t}_{kj}}{max\left(\sum_{j=1}^N\tilde{t}_{ij}, \sum_{j=1}^N\tilde{t}_{kj}\right)}$$

or, if written in another way:
$$sim_{ik} = min\left(\frac{\sum_{j=1}^N\tilde{t}_{ij}\tilde{t}_{kj}}{\sum_{j=1}^N\tilde{t}_{ij}},\frac{\sum_{j=1}^N\tilde{t}_{ij}\tilde{t}_{kj}}{\sum_{j=1}^N\tilde{t}_{kj}}\right)$$
it can be interpreted as the smallest of two weighted means.

The following helper function extracts t-statistics and adjusted p-values 
and calculates the truncated values mentioned above ($\tilde{t}$).

```{r, class.source="fold-show"}
## Extract t-statistics and adjusted p-values and threshold
getTruncT <- function(sce, adjpThr = 0.1) {
    ## Extract adjusted p-values and log-fold changes
    tc <- .getTestCols(sce)

    adjpvals <- tc$adjp
    tstats <- tc$tstat
    colnames(adjpvals) <- .getProteinNameFromComparison(colnames(adjpvals))
    colnames(tstats) <- .getProteinNameFromComparison(colnames(tstats))
    adjpvals <- as.matrix(adjpvals[, grep("untagged|Untagged", colnames(adjpvals), invert = TRUE)])
    tstats <- as.matrix(tstats[, grep("untagged|Untagged", colnames(tstats), invert = TRUE)])
    stopifnot(colnames(adjpvals) == colnames(tstats),
              rownames(adjpvals) == rownames(tstats))

    ## Need to pass a relatively loose significance threshold in at least one IP
    ## (don't want things to be similar just because they are weakly but consistently
    ## changing in many IPs)
    tstats[adjpvals > adjpThr] <- 0
    tstats[tstats <= 0] <- 0
    tstats[is.na(tstats)] <- 0

    tstats
}
```

We also define a function to calculate the similarities between each pair 
of proteins.

```{r, class.source="fold-show"}
## Helper function to get matrix of similarities
simsFromResults <- function(sce, adjpThr = 0.1) {

    tstats <- getTruncT(sce = sce, adjpThr = adjpThr)

    simil <- tstats %*% t(tstats)
    simil0 <- simil
    for (i in seq_len(nrow(simil))) {
        for (j in seq(i, ncol(simil))) {
            simil[i, j] <- simil0[i, j] / max(sum(tstats[i, ]), sum(tstats[j, ]))
            simil[j, i] <- simil0[j, i] / max(sum(tstats[i, ]), sum(tstats[j, ]))
        }
    }
    simil[is.na(simil)] <- 0

    ## Don't need self-connections
    diag(simil) <- 0

    stopifnot(rownames(simil) == colnames(simil))

    list(simil = simil, tstats = tstats, TFs = colnames(tstats))
}
```

To illustrate, let's calculate the similarity for a specific example pair 
(Atf1 and Pcr1):

```{r, class.source="fold-show"}
tmp <- getTruncT(sce = sce150, adjpThr = 0.1)[c("Atf1", "Pcr1"), ]
round(tmp[, colSums(tmp) != 0], 1)
```

$$sim_{\textrm{atf1,pcr1}} = \min\left(\frac{3\cdot2.3+2.4\cdot2.5+12.2\cdot7.2+8.4\cdot6.9}{2.3+2.5+7.2+6.9},\frac{3\cdot2.3+2.4\cdot2.5+12.2\cdot7.2+8.4\cdot6.9}{3+2.4+12.2+8.4}\right) = min\left(8.4,6.1\right) = 6.1$$

```{r, class.source="fold-show"}
simsFromResults(sce150[c("Atf1", "Pcr1"), ], adjpThr = 0.1)$simil
```

Next, we calculate similarities for all pairs of proteins in each of the 
data sets.

```{r get-similarities, class.source="fold-show"}
simil150 <- simsFromResults(sce150)
simil500 <- simsFromResults(sce500)
maxSimil <- round(max(max(simil150$simil, na.rm = TRUE),
                      max(simil500$simil, na.rm = TRUE)), 1)
```

# Generate networks

Given the matrix of similarities calculated above, we create a graph by first 
thresholding the similarities (setting all values below a given threshold to 0), 
and then considering the remaining (non-zero) values in the matrix as 
representing edges in the graph. Thus, two proteins in the graph are connected 
if their enrichment patterns across the IP experiments are similar. 
Only graph components with at least two proteins (i.e., no singletons) are 
displayed.

```{r make-networks}
createNetwork <- function(similList, threshold, known_interactors, settings = "igraph") {
    ## Get similarity matrix and apply threshold
    simil <- similList$simil
    tfs <- similList$TFs
    simil[simil < threshold] <- 0
    simil[is.na(simil)] <- 0

    ## Create graph
    gr <- graph_from_adjacency_matrix(simil, weighted = TRUE)
    gr <- as.undirected(gr, mode = "collapse", edge.attr.comb = list(weight = "mean"))

    ## Remove singletons
    comps <- components(gr)
    too_small <- which(table(comps$membership) < 2)
    keep <- V(gr)[!(comps$membership %in% too_small)]
    gr <- induced_subgraph(gr, keep)

    ## Set attributes for vertices and edges
    ## Need col2hex here since visNetwork doesn't understand all R colors (e.g. gray70)
    V(gr)$color <- gplots::col2hex(bg_color)
    V(gr)$color[names(V(gr)) %in% tfs] <- main_colors[5]
    V(gr)$vertexclass <- "other"
    V(gr)$vertexclass[names(V(gr)) %in% tfs] <- "Bait"

    if (settings == "visNetwork") {
        V(gr)$shape <- "triangle"
        V(gr)$size <- 15
        V(gr)$label.cex = 2.08
    } else if (settings == "igraph") {
        V(gr)$shape <- "circle"
    } else {
        stop("Unknown 'settings' parameter")
    }

    V(gr)$shape[names(V(gr)) %in% tfs] <- "square"
    E(gr)$value <- E(gr)$weight
    E(gr)$width <- E(gr)$weight / max(simil) * 1.5

    ## Indicate annotated physical interactions
    eL <- as_edgelist(gr)
    for (i in seq_along(E(gr))) {
        v1 <- eL[i, 1]
        v2 <- eL[i, 2]
        tmp <- known_interactors |>
            dplyr::filter(((Systematic.ID == v1 | Gene.name == v1) & 
                               Physical.interactors == v2) | 
                              ((Systematic.ID == v2 | Gene.name == v2) & 
                                   Physical.interactors == v1))
        if (nrow(tmp) == 0) {
            ## Novel interaction
            ## One of the elements is a transcription factor -> color orange
            if (v1 %in% tfs || v2 %in% tfs) {
                E(gr)[i]$color <- main_colors[3]
                E(gr)[i]$edgeclass <- "Unannotated TF interaction"
            } else{
                E(gr)[i]$color <- gplots::col2hex(na_color)
                E(gr)[i]$edgeclass <- "Other"
            }
        } else {
            E(gr)[i]$color <- gplots::col2hex(na_color)
            E(gr)[i]$edgeclass <- "Other"
        }
    }
    E(gr)$edgeclass <- factor(E(gr)$edgeclass, 
                              levels = c("Unannotated TF interaction", 
                                         "Other"))
    gr
}
```

```{r}
#| echo: false
#| eval: false

## Code for shiny app to compare thresholds and generate coordinate csv file
shinyAppThresholds <- function(similList, known_interactors) {
    server <- function(input, output) {
        output$visnetwork <- renderVisNetwork({
            gr <- createNetwork(similList, threshold = input$threshold, 
                                known_interactors = known_interactors, 
                                settings = "visNetwork")
            visNetwork::visIgraph(gr)
        })

        # Get position info
        observeEvent(input$store_position, {
            visNetworkProxy("visnetwork") |> visGetPositions()
        })

        output$coordinates <- renderPrint({
            nodes_positions()
        })

        ## Format positions
        nodes_positions <- reactive({
            positions <- input$visnetwork_positions
            if (!is.null(positions)){
                nodes_positions <- do.call("rbind", lapply(positions, function(x){data.frame(x = x$x, y = x$y)}))
                nodes_positions$id <- names(positions)
                nodes_positions
            } else {
                NULL
            }
        })

        output$downloadCsv <- shiny::downloadHandler(
            filename = function() {
                "graph_coordinates.csv"
            },
            content = function(file) {
                utils::write.table(nodes_positions(), file,
                                   row.names = FALSE, sep = ",",
                                   quote = FALSE)
            }
        )
    }

    ui <- fluidPage(
        sliderInput("threshold", label = "Score threshold", min = 0, max = maxSimil,
                    value = 6, step = 0.01),
        visNetworkOutput("visnetwork", width = "100%", height = "1000px"),
        verbatimTextOutput("coordinates"),
        actionButton("store_position", "Store positions"),
        downloadButton("downloadCsv", "Download")
    )

    shinyApp(ui = ui, server = server)
}
shiny::runApp(shinyAppThresholds(simil150, interactors))
shiny::runApp(shinyAppThresholds(simil500, interactors))
```

Next, we create networks for the low- and high-salt experiments.

```{r}
gr150 <- createNetwork(similList = simil150, threshold = 6, 
                       known_interactors = interactors, settings = "igraph")
gr500 <- createNetwork(similList = simil500, threshold = 6, 
                       known_interactors = interactors, settings = "igraph")
```

# Figure 5

```{r}
#| fig.width: 12
#| fig.height: 5

## Create small example similarity matrix and graph, for three selected proteins
poi <- c("Atf1", "Pcr1", "Arp8")
simil150example <- simsFromResults(sce150[poi, ])
gr150example <- createNetwork(similList = simil150example, threshold = 4.5,
                              known_interactors = interactors, 
                              settings = "igraph")

## General formula
## See https://uliniemann.com/blog/2022-02-21-math-annotations-in-ggplot2-with-latex2exp/
eqn1 <- data.frame(x = 0, y = 0, 
                   label = r"( $sim_{ik} = min\left(\frac{\sum_{j=1}^N\tilde{t}_{ij}\tilde{t}_{kj}}{\sum_{j=1}^N\tilde{t}_{ij}},\frac{\sum_{j=1}^N\tilde{t}_{ij}\tilde{t}_{kj}}{\sum_{j=1}^N\tilde{t}_{kj}}\right)$ )")

ggeqn1 <- ggplot(eqn1) +
    geom_text(aes(x = x, y = y, 
                  label = TeX(label, output = "character"), 
                  color = label), 
              parse = TRUE,
              key_glyph = draw_key_blank,
              size = 16/.pt) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_reverse(expand = c(0, 0)) + 
    scale_color_manual(values = "transparent", labels = TeX(eqn1$label,
                                                            italic = TRUE), 
                       name = "") + 
    guides(color = guide_legend(label.position = "top")) + 
    theme_void() + 
    theme(legend.text = element_text(size = 11))

## Specific example (atf1, pcr1)
eqn2 <- data.frame(x = 0, y = 0, 
                   label = r"( $sim_{Atf1,Pcr1} = min\left(\frac{3 \cdot 2.3 + 2.4 \cdot 2.5 + 12.2 \cdot 7.2 + 8.4 \cdot 6.9}{2.3 + 2.5 + 7.2 + 6.9},\frac{3 \cdot 2.3 + 2.4 \cdot 2.5 + 12.2 \cdot 7.2 + 8.4 \cdot 6.9}{3 + 2.4 + 12.2 + 8.4}\right) = min\left(8.4, 6.1\right) = 6.1$ )")
# eqn2 <- data.frame(x = 0, y = 0, 
#                    label = r"( $sim_{\textrm{Atf1,Pcr1}} = \min\left(\frac{3 \cdot 2.3 + 2.4 \cdot 2.5 + 12.2 \cdot 7.2 + 8.4 \cdot 6.9}{2.3 + 2.5 + 7.2 + 6.9},\frac{3 \cdot 2.3 + 2.4 \cdot 2.5 + 12.2 \cdot 7.2 + 8.4 \cdot 6.9}{3 + 2.4 + 12.2 + 8.4}\right) = min\left(8.4, 6.1\right) = 6.1$ )")

ggeqn2 <- ggplot(eqn2) +
    geom_text(aes(x = x, y = y, 
                  label = TeX(label, output = "character")), 
              parse = TRUE, 
              size = 11/.pt) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_reverse(expand = c(0, 0)) + 
    theme_void()

## Truncated t-statistics
tstats_vis <- data.frame(simil150example$tstats) |>
    tibble::rownames_to_column("Protein") |>
    tidyr::pivot_longer(names_to = "IP", values_to = "Value", -Protein) |>
    dplyr::mutate(Protein = factor(Protein, levels = poi))

## Reorder IPs for better visualization
enr <- unique(tstats_vis$IP[tstats_vis$Value > 0])
notenr <- setdiff(tstats_vis$IP, enr)
allips <- c(notenr, enr)
intlen <- round(length(notenr) / (length(enr) + 1))
idx <- seq(from = intlen, by = intlen, length.out = length(enr))
levs <- allips[order(c(seq_along(notenr), idx + 0.5))]
tstats_vis$IP <- factor(tstats_vis$IP, levels = levs)

## Plot truncated tstats across all IPs and label relevant points
labels <- tstats_vis |>
    dplyr::filter(Value > 0) |>
    dplyr::pull(IP) |>
    unique()

tplot <- ggplot(tstats_vis, aes(x = IP, y = Value, color = Protein, fill = Protein)) + 
        geom_col() + 
        geom_text(
            aes(label = ifelse(IP %in% labels & Value > 0, round(Value, digits = 1), "")),
            fontface ="plain", color = "black", size = 2.5, nudge_y = 2) +
        scale_color_manual(values = c(main_colors[1], main_colors[5], "black")) +
        scale_fill_manual(values = c(main_colors[1], main_colors[5], "black")) +
        scale_y_continuous(expand = c(0, 1)) + 
        facet_wrap(~ Protein, ncol = 1) +
        labs(x = "IP", y = "Thresholded mod. t-stat.") + 
        theme_cowplot(font_size = 10) + 
        theme(legend.position = "none", 
              axis.text.x = element_blank(), 
              axis.ticks.x = element_blank(), 
              strip.background = element_rect(fill = NA, colour = NA), 
              strip.text = element_text(hjust = 0, size = 10, face = "bold"))

## Similarity matrix
similsub <- simil150example$simil[poi, poi] |>
    as.data.frame() |>
    tibble::rownames_to_column("row") |>
    tidyr::pivot_longer(names_to = "column", values_to = "Similarity", -row) |>
    dplyr::mutate(Similarity = round(Similarity, digits = 1),
                  row = factor(row, levels = rev(poi)),
                  column = factor(column, levels = poi)) |>
    dplyr::mutate(simlab = ifelse(Similarity == 0, "", Similarity)) |>
    dplyr::mutate(Similarity = ifelse(Similarity == 0, NA, Similarity))
similplot <- ggplot(similsub, aes(x = column, y = row, fill = Similarity)) + 
    geom_tile() + 
    geom_text(aes(label = simlab, color = Similarity > 4), 
              fontface = "bold") +
    scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "black")) + 
    scale_fill_gradient(low = "white", high = binary_heatmap_colors["TRUE"], 
                    na.value = na_color, 
                    limits = range(similsub$Similarity, na.rm = TRUE)) + 
    theme_cowplot() + 
    theme(axis.line = element_blank(), 
          axis.ticks = element_blank(),
          legend.position = "bottom",
          legend.title = element_text(size = 10, vjust = 0.82),
          legend.margin = margin(0, 0, 0, 0),
          legend.box.spacing = unit(0, "pt")) + 
    labs(x = "", y = "") + 
    guides(color = "none")

## Induced graph
ggr150example <- ggraph(as_tbl_graph(gr150example), 
                        layout = cbind(x = c(-300, 800), y = c(800, -300)), 
                        weights = weight) + 
    geom_edge_link(aes(color = edgeclass, width = weight)) +
    scale_edge_color_manual(values = c(`Unannotated TF interaction` = main_colors[3], 
                                       Other = gplots::col2hex(na_color)),
                            name = "") + 
    scale_edge_width(range = c(1.2, 1.9), guide = "none") + 
    geom_node_point(size = 4, aes(fill = vertexclass, shape = vertexclass), 
                    color = "black", stroke = 0.1) + 
    scale_shape_manual(values = c(other = 21, Bait = 22), name = "") + 
    scale_fill_manual(values = c(other = gplots::col2hex(bg_color), Bait = main_colors[5]),
                      name = "") + 
    geom_node_text(aes(label = name), size = 3.5, repel = TRUE, 
                   nudge_x = 0, nudge_y = -5) + 
    theme_graph(base_family = "sans", plot_margin = margin(30, 10, 40, 10)) + 
    guides(fill = guide_legend(override.aes = list(size = 4, alpha = 1))) + 
    coord_cartesian(clip = "off")

## Put together
fig5a <- cowplot::plot_grid(
    cowplot::plot_grid(
        tplot, 
        get_legend(ggeqn1, "right"), 
        similplot + theme(legend.position = "bottom"), 
        ggr150example + theme(legend.position = "none"), 
        align = "v", axis = "t", 
        labels = c("A", "", "", ""), nrow = 1, 
        rel_widths = c(1.1, 0.9, 1, 0.5)
    ),
    ggeqn2, ncol = 1, labels = "", 
    rel_heights = c(3, 1)
)
```

## Put together

```{r Fig5}
#| fig.width: 10
#| fig.height: 10

coords150 <- read.csv(params$coords150)
coords500 <- read.csv(params$coords500)
coords500$y <- -coords500$y

.plotGraph <- function(grph, coords, nudge_y) {
    ggraph(as_tbl_graph(grph), 
                 layout = coords, 
                 weights = weight) + 
    geom_edge_link(aes(color = edgeclass, width = weight)) +
    scale_edge_color_manual(values = c(`Unannotated TF interaction` = main_colors[3], 
                                       Other = gplots::col2hex(na_color)),
                            breaks = "Unannotated TF interaction",
                            name = "") + 
    scale_edge_width(range = c(0.2, 0.9), guide = "none") + 
    geom_node_point(size = 2, aes(fill = vertexclass, shape = vertexclass), 
                    color = "black", stroke = 0.1) + 
    scale_shape_manual(values = c(other = 21, Bait = 22), name = "",
                       breaks = "Bait") + 
    scale_fill_manual(values = c(other = gplots::col2hex(bg_color), Bait = main_colors[5]),
                      name = "", breaks = "Bait") + 
    geom_node_text(aes(label = name), size = 2.5, repel = FALSE, 
                   nudge_x = 0, nudge_y = nudge_y) + 
    theme_graph(base_family = "sans", plot_margin = margin(10, 10, 10, 10)) + 
    guides(fill = guide_legend(override.aes = list(size = 4, alpha = 1)))
}
ggr150 <- .plotGraph(grph = gr150, coords = as.matrix(coords150[, c("x", "y")]),
                     nudge_y = -45)
ggr500 <- .plotGraph(grph = gr500, coords = as.matrix(coords500[, c("x", "y")]),
                     nudge_y = -35)

cowplot::plot_grid(
    fig5a, 
    cowplot::plot_grid(
        cowplot::plot_grid(
            ggr150 + theme(legend.position = "none"), 
            ggr500 + theme(legend.position = "none"), 
            nrow = 1, labels = c("B", "C")
        ),
        get_legend(ggr150 + theme(legend.position = "bottom"), "bottom"),
        nrow = 2, labels = c("", ""), rel_heights = c(1, 0.25)
    ), ncol = 1, rel_heights = c(1, 2)
)
```

## Supplementary figure

We extract all IPs where either `Rad24` or `Rad25` are pulled down, 
using a relatively lenient adjusted p-value cutoff.

```{r suppFig5}
#| fig.width: 9
#| fig.height: 10

## Get all t-statistics and visualize
res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
rad <- res150untag$tstat[c("Rad24", "Rad25"), ]
colnames(rad) <- .getProteinNameFromComparison(colnames(rad))
rad <- t(rad)
hmrad <- Heatmap(rad, 
                 cluster_rows = TRUE, 
                 cluster_columns = FALSE, 
                 col = circlize::colorRamp2(
                     breaks = seq(1.5, 13.0, length.out = 64),
                     colors = colorRampPalette(
                         binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
                 border = TRUE, 
                 border_gp = gpar(lwd = 0.5),
                 name = "Mod.\nt-stat.",
                 na_col = binary_heatmap_colors["FALSE"],
                 column_names_gp = gpar(fontsize = fl),
                 row_names_gp = gpar(fontsize = fs),
                 show_row_names = TRUE, 
                 show_column_names = TRUE, 
                 use_raster = FALSE,
                 show_heatmap_legend = TRUE, 
                 width = unit(w_trad / 1.5, "mm"), 
                 heatmap_height = unit(1.3 * h_trad, "mm"),
                 column_title_gp = gpar(fontsize = ft),
                 heatmap_legend_param = list(
                     title_gp = gpar(fontsize = fl),
                     border = "gray10",
                     border_gp = gpar(lwd = 0.5)
                 ))
set.seed(42L)
hm_rad <- grid.grabExpr(
    hmrad <- draw(hmrad, merge_legend = TRUE,
                 heatmap_legend_side = "right",
                 annotation_legend_side = "right"),
    width = w_trad / 1.5, height = 1.3 * h_trad
)

## List of experiments where rad24 and/or rad25 were pulled down
comps <- colnames(res150untag$interactor)[colSums(
    res150untag$interactor[c("Rad24", "Rad25"), ], na.rm = TRUE) > 0]
explist <- .getUniProtIdFromComparison(comps,idmap = idmap)
cat(explist)
gnames <- .getProteinNameFromComparison(comps)
cat(gnames)

## Read table
set_flextable_defaults(font.size = 8)
rad24_rad25_tbl <- read.delim(params$rad24_rad25_tbl)[, 1:2] |>
    setNames(c("TF", "R-x-x-[S/T]-x-P")) |>
    dplyr::mutate(TF = .capitalize(TF))
stopifnot(all(gnames %in% rad24_rad25_tbl$TF))
ft <- flextable(rad24_rad25_tbl) |>
  width(width =  c(0.5, 1.25)) |>
  set_table_properties(layout = "fixed")
ggtbl <- ggplot() +
    theme_void() +
    annotation_custom(grid::rasterGrob(as_raster(ft)),
                      xmin = -Inf, xmax = Inf, ymin = 0, ymax = 1)

data.frame(`Number of TFs interacting with at least one 14-3-3 protein` = 
               nrow(rad24_rad25_tbl),
           `Number of such TFs with at least one optimal 14-3-3 binding motif` = 
               sum(rad24_rad25_tbl$`R-x-x-[S/T]-x-P` != ""),
           check.names = FALSE) |>
    pivot_longer(cols = everything()) |>
    kbl(col.names = NULL) |>
    kable_styling()

plot_grid(hm_rad, ggtbl, nrow = 1, align = "h", axis = "b", 
          rel_widths = c(2, 2.5), labels = c("A", "B"))
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-06-21 ms_figure_6_heterodimer.Rmd
add figure 6
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "Figure 6 - heterodimer"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds500: "data/ipms_500_sce.rds"
    rdsrnaseq_del: "data/rnaseq_del.rds"
    rdsrnaseq_ko: "data/rnaseq_ko.rds"
    dist_from_ne: "data/Fig6_lacR_quantification_distance_radius.txt"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    idmap: "data/id_mapping_table.txt"
    rnatracks_ko: "data/genome_browser_tracks_rnaseq_ko.csv.gz"
    chiptracks_ko: "data/genome_browser_tracks_chipseq_ko.csv.gz"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
```

# Parameter values

```{r params}
params
```

# Load required packages and helper functions

```{r load-packages, class.source="fold-show"}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(ggplot2)
    library(cowplot)
    library(dplyr)
    library(ggrepel)
    library(colorspace)
    library(DESeq2)
    library(ggh4x)
    library(ggtext)
    library(patchwork)
    library(ggsignif)
    library(kableExtra)
    library(scales)
    library(broom)
})

source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/mapping_functions.R")
```

# Read data

We load a `SingleCellExperiment` object, containing data and results from the 
high-salt IP-MS experiment.

```{r load-data, class.source="fold-show"}
sce500 <- readRDS(params$rds500)
```

Next, we load two `SummarizedExperiment` objects with quantifications 
from RNA-seq experiments.

```{r, class.source="fold-show"}
rnaseq_del <- readRDS(params$rdsrnaseq_del)
rnaseq_ko <- readRDS(params$rdsrnaseq_ko)
idmap <- read.delim(params$idmap)
```

# Figure 6

## Zone quantification

```{r}
## Read data with distances from nuclear envelope + cell radius and 
## calculate relative distance from NE
dist_ne <- read.delim(params$dist_from_ne) |>
    group_by(Condition, Replicate) |>
    mutate(rel_distance_from_ne = Distance_from_nuc_envelope / Radius_of_cell) |>
    mutate(Condition = ifelse(Condition == "ntu1D", "ntu1\U141E", 
                              ifelse(Condition == "ntu2D", "ntu2\U141E", Condition))) |>
    mutate(SampleID = factor(paste0(Condition, "\n", Replicate),
                             levels = c("wt\nrep1", "wt\nrep2", "ntu1\U141E\nrep1",
                                        "ntu1\U141E\nrep2", "ntu2\U141E\nrep1",
                                        "ntu2\U141E\nrep2")))

## Remove one cell where Distance was slightly larger than the Radius
# table(dist_ne$rel_distance_from_ne > 1)
dist_ne <- dist_ne |>
    filter(rel_distance_from_ne >= 0 & rel_distance_from_ne <= 1) |>
    dplyr::mutate(Condition = factor(Condition, 
                                     levels = c("wt", "ntu1\U141E", "ntu2\U141E")),
                  Replicate = factor(Replicate, 
                                     levels = c("rep1", "rep2")))

## Get zone classification
zone_thresholds <- 1 - c(1, sqrt(2), sqrt(3)) / sqrt(3)
zone_byrep <- dist_ne |>
    mutate(Zone = ifelse(rel_distance_from_ne > zone_thresholds[1], "Zone III", 
                         ifelse(rel_distance_from_ne > zone_thresholds[2], 
                                "Zone II", "Zone I"))) |>
    group_by(Condition, Replicate, Zone) |>
    tally() |>
    group_by(Condition, Replicate) |>
    mutate(fraction_by_zone = n / sum(n)) |>
    mutate(Zone = factor(Zone, levels = c("Zone I", "Zone II", "Zone III"))) |>
    ungroup()
zone_byrep |>
    mutate(fraction_by_zone = label_percent(accuracy = 0.01)(fraction_by_zone)) |>
    rename(`Number of cells` = n, 
           `Percentage of cells` = fraction_by_zone) |>
    kbl() |>
    kable_styling()

## Average zone distribution over replicates
zone_avg <- zone_byrep |> 
    dplyr::group_by(Zone, Condition) |> 
    dplyr::summarize(avg_fraction_by_zone = mean(fraction_by_zone), 
                     sd_fraction_by_zone = sd(fraction_by_zone), 
                     .groups = "drop")
```

```{r}
## Fraction of cells in Zone I
zone_byrep |> filter(Zone == "Zone I") |>
    group_by(Zone, Condition) |>
    summarize(average_percentage = label_percent(accuracy = 0.01)(mean(fraction_by_zone)), 
              individual_percentages = paste(label_percent(accuracy = 0.01)(fraction_by_zone), 
                                             collapse = ","), 
              .groups = "drop") |>
    rename(`Average percentage` = average_percentage,
           `Percentages for replicates` = individual_percentages) |>
    kbl() |>
    kable_styling()

## Fraction of cells in Zone III
zone_byrep |> filter(Zone == "Zone III") |>
    group_by(Zone, Condition) |>
    summarize(average_percentage = label_percent(accuracy = 0.01)(mean(fraction_by_zone)), 
              individual_percentages = paste(label_percent(accuracy = 0.01)(fraction_by_zone), 
                                             collapse = ","), 
              .groups = "drop") |>
    rename(`Average percentage` = average_percentage,
           `Percentages for replicates` = individual_percentages) |>
    kbl() |>
    kable_styling()
```

### Plot relative distance to perimeter in all conditions

```{r}
#| warning: false

## Number of cells per condition
table(dist_ne$Replicate, dist_ne$Condition)

get_xlabs <- function(x) {
    vapply(x, function(xx) sub("\n", "<br>", paste0("*", xx, "*", "<br>(n = ", 
                                                    length(which(paste0(dist_ne$Condition, "\n", 
                                                                        dist_ne$Replicate) == xx)), ")"), fixed = TRUE), "")
}

## Calculate p-values (average values per sample followed by linear model)
dps <- dist_ne |>
    group_by(SampleID, Condition, Replicate) |>
    summarize(mean_dist = mean(rel_distance_from_ne),
              median_dist = median(rel_distance_from_ne),
              .groups = "drop")
(aggreg <- summary(lm(median_dist ~ Condition, data = dps)) |> tidy())

## wt vs ntu1D
p_ntu1 <- paste0("p==", signif(aggreg$p.value[aggreg$term == "Conditionntu1\U141E"], 3))

## wt vs ntu2D
p_ntu2 <- paste0("p==", signif(aggreg$p.value[aggreg$term == "Conditionntu2\U141E"], 3))

(p_rel_dis <- ggplot(dist_ne, 
                     aes(x = SampleID, y = rel_distance_from_ne, 
                         fill = Condition)) + 
        geom_violin(trim = TRUE, bounds = c(0, 1)) +
        geom_signif(annotations = c(p_ntu1, p_ntu2),
                    xmin = c(1.5, 1.5), xmax = c(3.5, 5.5), 
                    y = c(1.05, 1.15),
                    tip_length = 0, step_increase = 0.1, parse = TRUE) + 
        geom_boxplot(width = 0.1, fill = "white") +
        labs(title = "Relative distance from nuclear envelope",
             x = "", y = "Distance from NE relative to radius") + 
        scale_fill_manual(values = c(na_color, main_colors[5], main_colors[1]),
                          name = "") + 
        scale_x_discrete(labels = get_xlabs) +
        scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1.2)) + 
        theme_cowplot() + 
        theme(axis.text.x = element_markdown(),
              legend.text = element_text(face = "italic")))
```

### Plot zone distribution by condition

```{r}
#| warning: false

(p_zone <- ggplot(zone_avg, aes(fill = Zone, y = 100 * avg_fraction_by_zone, 
                                x = Condition)) + 
     geom_bar(position = "stack", stat = "identity") +
     labs(title = "Relative zone distribution\nincl. both replicates", 
          x = "", y = "Relative distribution [%]") +
     scale_fill_manual(values = c(main_colors[1], main_colors[3], main_colors[5]),
                       name = "") + 
     scale_x_discrete(labels = get_xlabs) + 
     theme_cowplot() + 
     theme(axis.text.x = element_markdown()))
```

## Ntu1/Ntu2 500 mM NaCl IP-MS

Here we extract and visualize the t-statistics for all proteins in the 
ntu1 and ntu2 high-salt IP-MS experiments. In each case the condition of 
interest is compared to a broad complement group consisting of all other 
high-salt IP-MS samples.

```{r tstat-500-ntu1-ntu2}
#| fig.width: 4
#| fig.height: 4
#| warning: false

## Extract t-statistics for the two comparisons
rdhigh <- rowData(sce500)
contrasts_high <- c(
    "Ntu1_500_plate_vs_compl_Ntu1_500_plate.t", 
    "Ntu2_500_plate_vs_compl_Ntu2_500_plate.t")
high <- as.data.frame(rdhigh[, c("einprotId", contrasts_high)]) |>
    setNames(c("einprotId", c("Ntu1", "Ntu2")))

## Plot t-statistics
(gg_ipdimer <- ggplot(high, aes(x = Ntu1, y = Ntu2)) +
    geom_abline(data = high |> 
                    dplyr::filter(einprotId %in% c("Ntu1", "Ntu2")), 
                aes(slope = mean(Ntu2 / Ntu1), intercept = 0), 
                linetype = "dashed", color = "grey50") + 
    geom_text_repel(data = high |> 
                        dplyr::filter(einprotId %in% c("Ntu1", "Ntu2") | 
                                          Ntu2 > 5),
                    aes(label = einprotId), min.segment.length = 0) + 
    geom_point(alpha = 0.25) + 
    geom_point(data = high |> 
                   dplyr::filter(einprotId %in% c("Ntu1", "Ntu2")) |>
                   dplyr::mutate(type = "Baits"),
               mapping = aes(fill = type), color = main_colors[3],
               shape = 21, alpha = 1, size = 2) + 
    scale_fill_manual(values = c(Baits = main_colors[3])) + 
    theme_cowplot() + 
    labs(x = "500 mM NaCl mod. t-stat. (Ntu1)", 
         y = "500 mM NaCl mod. t-stat. (Ntu2)") + 
    theme(strip.text = element_blank(),
          legend.position = "inside",
           legend.position.inside = c(0.8, 0.1)) + 
    guides(fill = guide_legend(nrow = 1, byrow = TRUE, title = "")))
```

## RNA-seq expression levels for genes around the tna1 locus

```{r}
#| fig.width: 8
#| fig.height: 6

wt_tpm <- assay(rnaseq_del[, rnaseq_del$Group_name == "wt_WT_mRNA"], "abundance")
genepos <- rowRanges(rnaseq_del)
mcols(genepos) <- NULL
genepos <- as.data.frame(genepos)
stopifnot(rownames(wt_tpm) == rownames(genepos))
genepos$ave_tpm <- rowMeans(wt_tpm)
genepos$id <- rownames(genepos)

chrom <- genepos["tna1", "seqnames"]
startpos <- genepos["tna1", "start"]
endpos <- genepos["tna1", "end"]
windowsize <- 30000

(genomeplot_tpm <- 
        ggplot(genepos |> 
                   dplyr::filter(seqnames == chrom & 
                                     start > startpos - windowsize & 
                                     end < endpos + windowsize), 
               aes(x = start / 1e6, y = ave_tpm, label = id)) +
        geom_hline(yintercept = 0, color = na_color) + 
        geom_point(size = 3, alpha = 0.8, aes(color = id == "tna1")) +
        scale_color_manual(values = c(`TRUE` = main_colors[4], `FALSE` = pt_color)) + 
        geom_text_repel(min.segment.length = 0, max.overlaps = Inf, size = 3, 
                        force = 3) + 
        labs(x = "Genome position (Mb)", y = "Gene expression (TPM)") + 
        scale_y_continuous(expand = expansion(mult = c(0, 0), add = c(200, 50))) + 
        theme_cowplot() + 
        theme(legend.position = "none"))
```

## RNA-seq differential expression analysis - KO experiments

```{r rnaseq-de}
#| fig.width: 15
#| fig.height: 4
#| message: false
#| warning: false

## Filter lowly expressed genes and create DESeqDataSet
rnaseqfilt_ko <- rnaseq_ko[rowSums(assay(rnaseq_ko, "counts") >= 5) >= 2, ]
dds <- DESeq2::DESeqDataSet(rnaseqfilt_ko, design = ~ MappedGene)
dds <- DESeq2::estimateSizeFactors(dds)

## Filter genes with high inferential variance
highInfRV <- 10^(-0.75)
dds <- dds[rowData(dds)$meanInfRV < highInfRV, ]

## Perform differential expression analysis
dds <- DESeq2::DESeq(dds)
groups <- setdiff(dds$MappedGene, "wt")
names(groups) <- groups
resList <- lapply(groups, function(gr) {
    DESeq2::results(dds, contrast = c("MappedGene", gr, "wt"), 
                    alpha = 0.05)
})

## Create MA plots
rngY <- range(sapply(resList, function(rl) rl$log2FoldChange))
rngX <- range(sapply(resList, function(rl) log10(rl$baseMean)))
maplots <- lapply(structure(names(resList), names = names(resList)), function(nm) {
    ggplot(as.data.frame(resList[[nm]]), 
           aes(x = log10(baseMean), y = log2FoldChange)) + 
        geom_point(mapping = aes(color = abs(log2FoldChange) > 3 & 
                                     padj < 0.001), alpha = 0.25) + 
        scale_color_manual(values = c(`TRUE` = main_colors[4], `FALSE` = pt_color)) + 
        coord_cartesian(xlim = rngX, ylim = rngY) + 
        geom_text_repel(data = as.data.frame(resList[[nm]]) |>
                            tibble::rownames_to_column("gene") |>
                            dplyr::filter(gene == nm | 
                                              (log2FoldChange > 4.5 & padj < 0.05)), 
                        aes(label = gene), 
                        min.segment.length = 0, max.overlaps = Inf) + 
        theme_cowplot() + 
        labs(x = "Mean expression", y = "log2(fold-change)", 
             title = paste0(nm, "\U141E")) + 
        theme(legend.position = "none",
              plot.title = element_text(face = "italic"))
})

## Correlation between log2 fold changes for ntu1/ntu2
stopifnot(all(rownames(resList$ntu1) == rownames(resList$ntu2)))
dfcorr <- data.frame(gene = rownames(resList$ntu1), 
                     ntu1 = resList$ntu1$log2FoldChange,
                     ntu2 = resList$ntu2$log2FoldChange)
corrplot <- list(
    ggplot(dfcorr, aes(x = ntu1, y = ntu2)) + 
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", 
                    color = "grey50") + 
        geom_point(alpha = 0.25) + 
        geom_text_repel(data = dfcorr |>
                            dplyr::filter(gene %in% c("tna1", "SPCC576.17c")), 
                        aes(label = gene), 
                        min.segment.length = 0, max.overlaps = Inf) + 
        labs(x = "*ntu1\U141E* log2(fold-change)",
             y = "*ntu2\U141E* log2(fold-change)") + 
        theme_cowplot() + 
        theme(axis.title = element_markdown())
)

cowplot::plot_grid(
    plotlist = maplots, nrow = 1, align = "hv", axis = "b"
)
```

## Coverage tracks

```{r}
#| fig.width: 8
#| fig.height: 8
#| message: false
#| warning: false

covs_rna <- read.csv(params$rnatracks_ko) |>
    left_join(data.frame(sample = colnames(rnaseq_ko),
                         libsize = colSums(assay(rnaseq_ko, "counts")))) |>
    mutate(cpm = score / libsize * 1e6)
covs_chip <- read.csv(params$chiptracks_ko)
gtf <- rtracklayer::import(params$gtf)

goi <- c("med6", "SPAC1002.16c", "urg2", "urg3", "SPCC576.17c", "SPNCRNA.1263")
gtf <- subset(gtf, (gene_id %in% goi | gene_name %in% goi) & type == "gene") |>
    as.data.frame() |>
    left_join(idmap, by = join_by("gene_id" == "gene_stable_id")) |>
    dplyr::select(seqnames, start, end, strand, unique_einprot_id) |>
    mutate(y = ifelse(unique_einprot_id == "SPCC576.17c", 0, 1)) |>
    mutate(start2 = ifelse(strand == "+", start, end),
           end2 = ifelse(strand == "+", end, start)) |>
    mutate(unique_einprot_id_label = ifelse(grepl("^SP", unique_einprot_id), 
                                            unique_einprot_id, 
                                            paste0("*", unique_einprot_id, "\U207A*")))

## Create data frames for plotting
plotdf_chip1 <- covs_chip |>
    mutate(sample = sub("_rep[0-9]", "", sample)) |>
    mutate(exptype = stringr::str_extract(sample, "ChIP|IN")) |>
    mutate(pair = sub("_ChIP|_IN", "", sample)) |>
    mutate(flag = .capitalize(sub("_*TAG", "", sub(".*_KO_", "", pair)))) |>
    mutate(label = ifelse(pair == "ntu1_KO_ntu2TAG", "*ntu1\U141E*",
                          ifelse(pair == "ntu1_TAG", "*ntu2\U207A*",
                                 ifelse(pair == "ntu2_TAG", "*ntu1\U207A*",
                                        ifelse(pair == "ntu2_KO_ntu1TAG", "*ntu2\U141E*", 
                                               ""))))) |>
    mutate(cpm = score) |>
    mutate(pos = start) |>
    group_by(flag, seqnames) |>
    mutate(cpmnorm = cpm / max(abs(cpm))) |>
    ungroup() |>
    filter(pair %in% c("ntu2_TAG", "ntu2_KO_ntu1TAG", 
                       "ntu1_TAG", "ntu1_KO_ntu2TAG")) |>
    mutate(pair = factor(pair, levels = c("ntu1_TAG", "ntu2_KO_ntu1TAG",
                                          "ntu2_TAG", "ntu1_KO_ntu2TAG")))
plotdf_rna <- covs_rna |>
    filter(sample %in% c("SPBC530.08_KO_mRNA_rep1",
                         "SPBC16G5.16_KO_mRNA_rep1",
                         "wt_KO_mRNA_rep1")) |>
    mutate(sample = sub("_KO_mRNA_rep[0-9]", "", sample)) |>
    mutate(sample = replace(sample, sample == "SPBC16G5.16", "ntu1\U141E")) |>
    mutate(sample = replace(sample, sample == "SPBC530.08", "ntu2\U141E")) |>
    mutate(gene = sample) |>
    group_by(seqnames) |>
    mutate(cpmnorm = cpm / max(abs(cpm))) |>
    ungroup() |>
    mutate(sample = factor(sample, 
                           levels = c("wt", "ntu1\U141E", "ntu2\U141E")))
plotdf_label <- covs_rna |>
    filter(sample == "wt_KO_mRNA_rep1") |>
    mutate(cpmnorm = 0)

.get_empty_label <- function(x) {
    c("ntu1_TAG" = "", "ntu2_KO_ntu1TAG" = "",
      "ntu2_TAG" = "", "ntu1_KO_ntu2TAG" = "")
}
.add_space_to_label <- function(x) {
    paste(x, "    ")
}

maxIII <- max(plotdf_label$pos[plotdf_label$seqnames == "III"])
g1c <- ggplot(plotdf_label,
              aes(x = pos, y = cpmnorm)) + 
    geom_line(color = "white") + 
    geom_segment(x = maxIII - 2000, xend = maxIII,
                 y = 0, yend = 0, color = "black") + 
    geom_text(data = data.frame(seqnames = "III",
                                x = maxIII - 1000,
                                y = 0,
                                label = "2kb"),
              aes(x = x, y = y, label = label),
              vjust = -1) + 
    facet_grid(~ seqnames, scales = "free_x", space = "free_x")
g1a <- ggplot(plotdf_chip1 |>
                  mutate(seqnames = paste0("chr", seqnames)), 
              aes(x = pos, y = cpmnorm)) + 
    geom_area(aes(group = exptype, fill = interaction(exptype, flag))) + 
    geom_richtext(data = plotdf_chip1 |> 
                      mutate(seqnames = paste0("chr", seqnames)) |>
                      select(flag, pair, seqnames, pos, label) |>
                      filter(seqnames == "chrI") |>
                      filter(pos == min(pos)),
                  aes(x = pos, y = 0.65, label = label),
                  fill = NA, label.color = NA,
                  label.padding = grid::unit(rep(0, 4), "pt"),
                  hjust = 0, size = 4) + 
    scale_fill_manual(values = c(`ChIP.Ntu1` = main_colors[5], 
                                 `IN.Ntu1` = na_color,
                                 `ChIP.Ntu2` = main_colors[1], 
                                 `IN.Ntu2` = na_color)) + 
    facet_nested(flag + pair ~ seqnames, scale = "free_x", space = "free_x",
                 switch = "y",
                 labeller = labeller(pair = .get_empty_label, 
                                     flag = .add_space_to_label),
                 nest_line = element_line(),
                 strip = strip_nested(size = "variable")) + 
    scale_y_continuous(limits = c(0, 0.9), expand = c(0, 0))
g1b <- ggplot(plotdf_rna, 
              aes(x = pos, y = cpmnorm)) + 
    geom_area(aes(group = strand, fill = gene)) + 
    scale_fill_manual(values = c(wt = "black", "ntu1\U141E" = main_colors[5], 
                                 "ntu2\U141E" = main_colors[1])) + 
    facet_grid(sample ~ seqnames, scale = "free_x", space = "free_x",
               switch = "y") + 
    scale_y_continuous(limits = c(-0.9, 0.3), expand = c(0, 0))
g2 <- ggplot(gtf, aes(x = start2, xend = end2, y = y, yend = y, 
                      color = unique_einprot_id %in% c("tna1", "SPCC576.17c"), 
                      label = unique_einprot_id_label)) + 
    geom_segment(arrow = grid::arrow(length = unit(0.05, "inches")), 
                 linewidth = 2) + 
    geom_richtext(
        aes(x = (start2 + end2) / 2, y = y - 0.5),
        fill = NA, label.color = NA,
        label.padding = grid::unit(rep(0, 4), "pt"),
        size = 4
    ) + 
    # geom_text(aes(x = (start2 + end2) / 2, y = y - 0.5), 
    #           size.unit = "in", size = 0.1) + 
    scale_color_manual(values = c(`FALSE` = "black", `TRUE` = main_colors[4])) + 
    facet_grid(~ seqnames, scales = "free_x", space = "free_x") + 
    ylim(-0.75, 1.5)
gcov <- cowplot::plot_grid(
    g1c + theme_void() + theme(strip.text = element_blank()),
    g1a + theme_void() + theme(legend.position = "none", 
                               strip.text.y = element_text(size = 12),
                               strip.text.x = element_text(size = 12, hjust = 0.95)), 
    NULL,
    g1b + theme_void() + theme(strip.text.y = element_text(size = 12, 
                                                           face = "italic"), 
                               legend.position = "none",
                               strip.text.x = element_blank()), 
    g2 + theme_void() + 
        theme(legend.position = "none",
              strip.text = element_blank()), 
    ncol = 1, rel_heights = c(0.75, 2, 0.5, 2, 1),
    align = "v", axis = "lr")
(gcov2 <- cowplot::plot_grid(
    g1c + theme_bw() + theme(axis.text.x = element_text(size = 5)),
    g1a + theme_bw() + theme(legend.position = "none",
                             axis.text.x = element_text(size = 5)), 
    g1b + theme_bw() + theme(legend.position = "none",
                             axis.text.x = element_text(size = 5)), 
    g2 + theme_bw() + 
        theme(legend.position = "none",
              axis.text.x = element_text(size = 5)), 
    ncol = 1, rel_heights = c(0.75, 2, 2, 1),
    align = "v", axis = "lr"))
```

## Put together

```{r Fig6}
#| fig.width: 13
#| fig.height: 14
#| warning: false

fig6b <- ggdraw() + 
    draw_image("schematics/Fig6B_Nattou_protein_domains_cartoon_Arial.png")
fig6e <- ggdraw() + 
    draw_image("schematics/Fig6E_Nattou_lacR_cartoon_images_Arial.png")
fig6f <- ggdraw() + 
    draw_image("schematics/Fig6F_Nattou_lacR_microscopy_images_Arial.png")

fig6c <- wrap_plots(maplots, ncol = 1) | corrplot[[1]]

cowplot::plot_grid(
    cowplot::plot_grid(
        gg_ipdimer, 
        fig6c,
        nrow = 1, 
        rel_widths = c(1, 2), 
        labels = c("A", "C"),
        align = "h", axis = "bt"
    ),
    cowplot::plot_grid(
        fig6b, 
        gcov,
        rel_widths = c(1, 1),
        labels = c("B", "D")
    ),
    cowplot::plot_grid(
        fig6e,
        NULL,
        fig6f,
        NULL,
        p_rel_dis,
        labels = c("E", "", "F", "", "G"),
        # rel_widths = c(7.65, 0.5, 1.45, 0.5, 9)
        rel_widths = c(6, 0.75, 1.2, 0.75, 10.4), nrow = 1
    ), 
    ncol = 1
)
```

## Supplementary Figure

```{r suppFig6-chip}
#| warning: false
#| message: false

plotdf_chip2 <- covs_chip |>
    mutate(sample = sub("_rep[0-9]", "", sample)) |>
    mutate(exptype = stringr::str_extract(sample, "ChIP|IN")) |>
    mutate(pair = sub("_ChIP|_IN", "", sample)) |>
    mutate(flag = sub("_*TAG", "", sub(".*_KO_", "", pair))) |>
    mutate(label = ifelse(pair == "ntu2_delTMD", "delTMD",
                          ifelse(pair == "ntu2_TAG", "ntu2+", ""))) |>
    mutate(cpm = score) |>
    mutate(pos = start) |>
    group_by(seqnames) |>
    mutate(cpmnorm = cpm / max(abs(cpm))) |>
    ungroup() |>
    filter(pair %in% c("ntu2_TAG", "ntu2_delTMD")) |>
    mutate(pair = ifelse(pair == "ntu2_delTMD", "Ntu2\U141E\U1D40\U1D39\U1D30", 
                         ifelse(pair == "ntu2_TAG", "Ntu2", pair))) |>
    mutate(pair = factor(pair, levels = c("Ntu2", "Ntu2\U141E\U1D40\U1D39\U1D30")))

gtr <- ggplot(plotdf_chip2 |>
                  mutate(seqnames = paste0("chr", seqnames)), 
              aes(x = pos, y = cpmnorm)) + 
    geom_area(aes(group = exptype, fill = interaction(exptype, flag))) + 
    scale_fill_manual(values = c(ChIP.ntu2_delTMD = main_colors[3], 
                                 IN.ntu2_delTMD = na_color,
                                 ChIP.ntu2 = main_colors[1], 
                                 IN.ntu2 = na_color)) + 
    facet_grid(pair ~ seqnames, scale = "free_x",
               space = "free_x", switch = "y") + 
    scale_y_continuous(limits = c(0, 0.9), expand = c(0, 0))
gcovdel <- cowplot::plot_grid(
    gtr + theme_void() + theme(strip.text.y = element_text(size = 12, vjust = 0.05),
                               legend.position = "none", 
                               strip.text.x = element_text(size = 12, hjust = 0.95)), 
    g2 + theme_void() + 
        theme(legend.position = "none",
              strip.text = element_blank()), 
    ncol = 1, rel_heights = c(2, 1),
    align = "v", axis = "lr")
```

```{r suppFig6}
#| fig.width: 13
#| fig.height: 10
#| warning: false

figs6a <- ggdraw() + 
    draw_image("schematics/FigS6A_Nattou_AF2_fl_Arial.png")

cowplot::plot_grid(
    cowplot::plot_grid(
        figs6a, 
        NULL,
        p_zone + theme(legend.position = "top"), 
        nrow = 1, 
        rel_widths = c(1.75, 0.25, 1.25), 
        labels = c("A", "", "B")
    ),
    cowplot::plot_grid(
        genomeplot_tpm,
        gcovdel,
        # maplots$ID83delTMD,
        nrow = 1, 
        rel_widths = c(1.2, 1),
        labels = c("C", "D")
    ),
    ncol = 1, rel_heights = c(1, 0.7)
)
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-06-21 index.qmd
add index
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "A comprehensive Schizosaccharomyces pombe atlas of physical transcription factor interactions with proteins and DNA"
subtitle: "Analysis code for figures"
date: today
author:
  - name: Merle Skribbe
    email: merle.skribbe@gmail.com
    affiliations:
      - ref: fmi
      - ref: unibas
  - name: Charlotte Soneson
    orcid: 0000-0003-3833-2169
    email: charlotte.soneson@fmi.ch
    affiliations:
      - ref: fmi
      - ref: sib
  - name: Michael B Stadler
    orcid: 0000-0002-2269-4934
    email: michael.stadler@fmi.ch
    affiliations:
      - ref: fmi
      - ref: unibas
      - ref: sib
  - name: Michaela Schwaiger
    affiliations:
      - ref: fmi
      - ref: sib
  - name: Vishnu Narayanan Suma Sreechakram
    affiliations:
      - ref: giessen
  - name: Vytautas Iesmantavicius
    affiliations:
      - ref: fmi
  - name: Daniel Hess
    affiliations:
      - ref: fmi
  - name: Eliza Pandini Figueiredo Moreno
    affiliations:
      - ref: fmi
  - name: Sigurd Braun
    affiliations:
      - ref: giessen
  - name: Jan Seebacher
    affiliations:
      - ref: fmi
  - name: Sebastien Smallwood
    affiliations:
      - ref: fmi
  - name: Marc Bühler
    orcid: 0000-0001-6661-9795
    email: marc.buehler@fmi.ch
    affiliations:
      - ref: fmi
      - ref: unibas
affiliations:
  - id: fmi
    name: Friedrich Miescher Institute for Biomedical Research
    address: Fabrikstrasse 24
    postal-code: 4056
    city: Basel
    country: Switzerland
  - id: unibas
    name: University of Basel
    address: Petersplatz 10
    postal-code: 4003
    city: Basel
    country: Switzerland
  - id: sib
    name: SIB Swiss Institute of Bioinformatics
    address: Fabrikstrasse 24
    postal-code: 4056
    city: Basel
    country: Switzerland
  - id: giessen
    name: Institute for Genetics, Justus-Liebig-University Giessen
    city: Giessen
    country: Germany
license: MIT License
format: 
    html:
        embed-resources: true
        fig-cap-location: top
---

## Interactive data exploration with `TFexplorer`

## Links to raw data

## Links to publication

## Links to code for generating figures

All source code is available from GitHub at [https://github.com/fmicompbio/Spombe_TFome](https://github.com/fmicompbio/Spombe_TFome).
Click on a specific figure below to jump directly to the corresponding sources.

|          |                                                                                                  |     |
|----------|----------------------------|----------------------------|
| Figure 1 | [![](ms_figure_1_overview_files/figure-html/Fig1-1.png){width="4cm"}](ms_figure_1_overview.html) |  [![](ms_figure_1_overview_files/figure-html/suppFig1-1.png){width="4cm"}](ms_figure_1_overview.html#Supplementary_figure)   |
| Figure 2 | [![](ms_figure_2_chip_files/figure-html/Fig2-1.png){width="4cm"}](ms_figure_2_chip.html) | [![](ms_figure_2_chip_files/figure-html/suppFig2-1.png){width="4cm"}](ms_figure_2_chip.html#Supplementary_figure)    |
| Figure 3 | [![](ms_figure_3_motifs_files/figure-html/Fig3-1.png){width="4cm"}](ms_figure_3_motifs.html) | [![](ms_figure_3_motifs_files/figure-html/suppFig3-1.png){width="4cm"}](ms_figure_3_motifs.html#Supplementary_figure)    |
| Figure 4 | [![](ms_figure_4_ipms_files/figure-html/Fig4-1.png){width="4cm"}](ms_figure_4_ipms.html) | [![](ms_figure_4_ipms_files/figure-html/suppFig4-1.png){width="4cm"}](ms_figure_4_ipms.html#Supplementary_figure)    |
| Figure 5 | [![](ms_figure_5_networks_files/figure-html/Fig5-1.png){width="4cm"}](ms_figure_5_networks.html) | [![](ms_figure_5_networks_files/figure-html/suppFig5-1.png){width="4cm"}](ms_figure_5_networks.html#Supplementary_figure)    |
| Figure 6 | [![](ms_figure_6_heterodimer_files/figure-html/Fig6-1.png){width="4cm"}](ms_figure_6_heterodimer.html) | [![](ms_figure_6_heterodimer_files/figure-html/suppFig6-1.png){width="4cm"}](ms_figure_6_heterodimer.html#Supplementary_figure)    |

### 2024-07-04 ms_figure_3_chip.Rmd
rename figs2-6 to figs3-7
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "Figure 2 - ChIP"
author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    genomefasta: "reference/Schizosaccharomyces_pombe.ASM294v2.dna.toplevel.fa"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    chiptrackfile: "data/genome_browser_tracks_chipseq.csv.gz"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    dbdtxt: "data/TF_DBD_annotation.txt"
    motifrds: "data/motifs_annotated.rds"
    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
    ipms150untag: "data/ipms_150_untag_sce.rds"
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
```

# Parameter values

```{r params}
params
```

# Load required packages

```{r load-packages}
suppressPackageStartupMessages({
    library(circlize)
    library(ComplexHeatmap)
    library(grid)
    library(ggplot2)
    library(cowplot)
    library(Biostrings)
    library(BSgenome)
    library(GenomicRanges)
    library(rtracklayer)
    library(tibble)
    library(tidyr)
    library(universalmotif)
    library(parallel)
    library(colorspace)
    library(scattermore)
    library(ggupset)
    library(dplyr)
    library(viridisLite)
    library(SummarizedExperiment)
    library(S4Vectors)
    library(jsonlite)
    library(dendextend)
    library(kableExtra)
    library(forcats)
})

source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/mapping_functions.R")
source("params/plottracks_function.R")

# define heatmap dimensions and font sizes
w_peaksHm <-  84 # heatmap body width (mm)
h_peaksHm <- 178 # total heatmap height (mm)
fs <- 8  # small font size
fm <- 9  # medium font size
fl <- 10 # large font size
```

# Helper functions

Here we define a few helper functions for using below:

```{r helper-functions}
# make first letter of a string lowercase
.lowercase <- function(x) {
    paste0(tolower(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}

# for each region in `from`, calculate the distance to its nearest element in `to`
# if there are no element in `to` on the sequence of `from`, set distance to NA
calc_distance_to_nearest <- function(from, to) {
    stopifnot(exprs = {
        is(from, "GRanges")
        is(to, "GRanges")
    })
    tmp <- distanceToNearest(x = from, subject = to)
    d <- rep(NA, length(from))
    d[queryHits(tmp)] <- mcols(tmp)$distance
    return(d)
}
```

# Read data

We start by reading all tables and objects needed for this figure.
Count tables are in addition normalized, and we also calculate
averages over replicates or all samples where needed:

```{r load-data}
# genome sequence
gnm <- readDNAStringSet(params$genomefasta)
names(gnm) <- sub(" .+$", "", names(gnm))

# `peakgr`: ChIP-seq peaks (as a GRanges object)
peakgr <- as(read.csv(params$peakcsv, row.names = 1), "GRanges")
peakgr$peaktype <- factor(peakgr$peaktype,
                          levels = c("common peaks (ubiquitous)",
                                     "common peaks (frequent)",
                                     "specific peaks"))
peakseq <- getSeq(gnm, peakgr)

# `peakenr`: peak IP-enrichments log2(IP/input)
peakenr <- as.matrix(read.csv(params$peakenr, row.names = 1))

# ... average replicates (`peakenrAvg`)
grp <- sub("_rep[12]$", "", colnames(peakenr))
peakenrAvg <- do.call(cbind, lapply(split(colnames(peakenr), grp)[unique(grp)],
                                    function(j) {
                                        rowMeans(peakenr[, j, drop = FALSE])
                                    }))

# `chip_genome_tab`: data.frame with alignment densities in selected
#                    genomic regions for selected samples
chip_genome_tab <- read.csv(params$chiptrackfile)

# `atac_raw`: public ATAC-seq read counts in ChIP-seq peaks from this study
atac_raw <- as.matrix(read.csv(params$peakatacfile, row.names = 1))

# ... calculate reads per million (`atac_cpm`) and
#     per kilobase and million (`atac_rpkm`)
#     and average experiments (`atac_cpm_avg`, `atac_rpkm_avg`)
atac_cpm <- sweep(x = atac_raw[, -1], MARGIN = 2,
                  STATS = colSums(atac_raw[, -1]), FUN = "/") * 1e6
atac_rpkm <- atac_cpm / atac_raw[, "width"] * 1e3
atac_cpm_avg <- rowMeans(atac_cpm)
atac_rpkm_avg <- rowMeans(atac_rpkm)

# `chip_raw`: public ChIP-seq read counts in ChIP-seq peaks from this study
chip_raw <- as.matrix(read.csv(params$peakchipfile, row.names = 1))

# ... calculate reads per million (`chip_cpm`) and
#     per kilobase and million (`chip_rpkm`)
#     and calculate enrichments (`chip_enr`, log2 IP/input) averaged over replicates
chip_cpm <- sweep(x = chip_raw[, -1], MARGIN = 2,
                  STATS = colSums(chip_raw[, -1]), FUN = "/") * 1e6
chip_rpkm <- chip_cpm / chip_raw[, "width"] * 1e3
chip_series <- sub("^ChIP_([^_]+)_.+$", "\\1", colnames(chip_cpm))
chip_enr <- do.call(
    cbind,
    lapply(split(colnames(chip_cpm), chip_series),
           function(nms) {
               grp <- sub("ChIP_(GSE[0-9]+)_GSM[0-9]+_(.+)_rep[0-9]$",
                          "\\1_\\2", nms)
               stopifnot(any(is_input <- grep("_Input", grp)))
               enr <- do.call(
                   cbind,
                   lapply(unique(grp[-is_input]), function(grp1) {
                       rowMeans(log2(chip_cpm[, nms[grp == grp1], drop = FALSE] + 1)) -
                           rowMeans(log2(chip_cpm[, nms[is_input], drop = FALSE] + 1))
                   }))
               colnames(enr) <- unique(grp[-is_input])
               enr
           }))

# `genegr`: chromosomal ranges of genes (as GRanges object)
genegr <- rtracklayer::import(params$gtf, feature.type = "gene")

# `motifs`: data.frame with annotated motifs
motifs <- readRDS(params$motifrds)

# `txannot`: data.frame with transcript start site (TSS) and
#            transcript end site (TES) annotation
txannot <- read.csv(params$txcsv, row.names = 1)

# `dbd`: data.frame with information of transcription factors used as
#        IP-MS baits and their DNA binding domains
dbd <- read.delim(params$dbdtxt)
```

# Facts and numbers - ChIP-seq

```{r calc-numbers, class.source="fold-show", warning=FALSE}
# Percent of TF-bound gene promoters
round(100 * mean(txannot$prom.number.overlapping.highconf.peaks > 0), 0)

# Percent of TF-bound gene promoters, by gene biotype
txannot |>
    group_by(gene_biotype) |>
    summarise("Number of genes" = n(),
              "Percent promoters with TF enrichments" = round(
        100 * mean(prom.number.overlapping.highconf.peaks > 0), 0),
        .groups = "drop") |>
    knitr::kable()

# TF enrichments at promoters of TF genes
txannot |>
    filter(gene_id %in% dbd$PomBaseID) |>
    summarise("Number of TF genes" = n(),
              "Number of TF promoters with at least one enriched TF" =
                  sum(prom.number.distinct.enriched.TFs > 0),
              "Number of TF promoters with multiple enriched TFs" =
                  sum(prom.number.distinct.enriched.TFs > 1),
              "Number of TF promoters without detected TF enrichment" =
                  sum(prom.number.distinct.enriched.TFs == 0),
              "Number of TFs enriched at any promoter" = length(unique(unlist(strsplit(prom.enriched.TFs, ";")))),
              "Number of TFs enriched at any promoter (w/o zas1\U207A promoter)" = length(unique(unlist(strsplit(prom.enriched.TFs[gene_name != "zas1"], ";")))),
              "Number of TFs enriched at zas1\U207A promoter" = length(unique(unlist(strsplit(prom.enriched.TFs[gene_name == "zas1"], ";")))),
              "Number of TFs enriched at their own promoter" = sum(binds.own.promoter),
              "Number of TFs enriched at atf1\U207A promoter" = length(unique(unlist(strsplit(prom.enriched.TFs[gene_name == "atf1"], ";")))),
              "Number of TF promoters with Atf1 enrichment" = length(grep("Atf1", prom.enriched.TFs)),
              .groups = "drop") |>
    pivot_longer(cols = starts_with("Number"),
                 values_to = "Number", names_to = "Set") |>
    knitr::kable()

# Total number of identified DNA-binding motifs
nrow(motifs)

# Number of TFs for which at least one motif was identified
length(unique(motifs$tf_name))

# Number (percent) of common frequent peaks with at least on Atf1.m1 DNA-binding motif (CRE motif)
hits <- universalmotif::scan_sequences(motifs = motifs$motif[motifs$name == "Atf1.m1"],
                                       sequences = peakseq[peakgr$peaktype == "common peaks (frequent)"], 
                                       threshold = 1e-4,
                                       threshold.type = "pvalue",
                                       RC = TRUE,
                                       verbose = 0,
                                       nthreads = 1,
                                       return.granges = TRUE)
length(unique(hits$sequence.i))
round(100 * length(unique(hits$sequence.i)) / sum(peakgr$peaktype == "common peaks (frequent)"))
```

# Export peak ChIP data for use in `TFexplorer`

Starting from the peaks (`peakgr`) and ChIP enrichments in peaks (`peakenr`),
prepare and export the matrix in `json` format for use in `TFexplorer`.

```{r export-chipdata-to-json}
# average replicate ChIP enrichments
grp <- sub("_rep[12]$", "", colnames(peakenr))
peakenrExport <- do.call(cbind,
                         lapply(structure(unique(grp), names = unique(grp)),
                                function(nm) {
                                    rowMeans(peakenr[, grp == nm])
                                }))

# replace all enrichment values below 0 with 0
peakenrExport[peakenrExport < 0] <- 0

# cluster
dist_mat2 <- dist(peakenrExport, method = "euclidian")
hclust_avg2 <- hclust(dist_mat2, method = "average")
peakenrExportClustered <- peakenrExport[hclust_avg2$order, ]

# calculate extend peak coordinates and IGV position
IGVpos <- paste0(seqnames(peakgr), ":",
                 start(peakgr) - 100, "-",
                 end(peakgr) + 100)
IGVposClustered <- IGVpos[hclust_avg2$order]

# create peak position strings
peakPos <- paste0(seqnames(peakgr), ":", start(peakgr), "-", end(peakgr))
peakPosClustered <- peakPos[hclust_avg2$order]

# export peaks and enrichments as json
# ... calculate maximum enrichment value in each row
ChipMax <- apply(peakenrExportClustered, 1, max)
range(ChipMax)

# ... convert to JSON
heatmapDataChip <- toJSON(
    list(
        heatmapMatrix = peakenrExportClustered,
        TFs = colnames(peakenrExportClustered),
        peaks = rownames(peakenrExportClustered),
        peakPos = peakPosClustered,
        IGVpos = IGVposClustered,
        ChipMax = ChipMax
    ), pretty = TRUE)

# ... write into a file for TFexplorer
write(heatmapDataChip, "data/heatmapDataChIP.json")
```

# Calculate peak distances to tRNA or rRNA genes

We want to identify ChIP-seq peaks that are near a tRNA or rRNA gene. We can
obtain the coordinates of these genes from `genegr` and the coordinates of
peaks from `peakgr`, and calculate the distance between nearest peak-gene pairs
using `distanceToNearest()`. Any pair with a distance below `maxdist` will be
classified as "near" in the plots below.

Remark: Nearest distances can be `NA` in cases where for example a peak
resides on a sequence (chromosome) that does not contain any tRNA or rRNA gene,
or vice versa, such as the sequence `AB325691` (which contains gap-filling
sequence between SPBPB21E7.09 and SPBPB10D8.01 in chromosome 2) or the
mitochondrial sequence `MT`.

```{r calc-peak-dists}
# distance less than `maxdist` are defined as "near"
maxdist <- 100

# tRNA
# ... PomBase and Ensembl_Fungi annotate 196 and 183 tRNAs, respectively
#     we combine the two and obtain 198 annotated tRNAs
table(genegr$gene_biotype == "tRNA", genegr$source)
is_tRNA_pombase <- genegr$source == "PomBase" & genegr$gene_biotype == "tRNA"
is_tRNA_ensembl <- genegr$source == "Ensembl_Fungi" & genegr$gene_biotype == "tRNA"
is_tRNA <- is_tRNA_pombase | (is_tRNA_ensembl & !genegr %in% genegr[is_tRNA_pombase])
sum(is_tRNA)
# ... now we measure distances between nearest pairs
#     either from peak to nearest tRNA
dist.peak2tRNA <- calc_distance_to_nearest(from = peakgr, to = genegr[is_tRNA])
#     or from tRNA to nearest peak
dist.tRNA2peak <- calc_distance_to_nearest(from = genegr[is_tRNA], to = peakgr)

# rRNA
# ... all 36 5S_rRNA genes in our annotation stem from Ensembl_Fungi
table(genegr$gene_name == "5S_rRNA", genegr$source)
is_rRNA <- !is.na(genegr$gene_name) & genegr$gene_name == "5S_rRNA"
sum(is_rRNA)
# ... now we measure distances between nearest pairs
#     either from peak to nearest rRNA
dist.peak2rRNA <- calc_distance_to_nearest(from = peakgr, to = genegr[is_rRNA])
#     or from rRNA to nearest peak
dist.rRNA2peak <- calc_distance_to_nearest(from = genegr[is_rRNA], to = peakgr)
```

# Calculate motif similarities

We calculate the similarities between any pair of motifs using the
`compare_motifs()` function with `method = "PCC"` (Pearson correlation
coefficient). We allow for reverse-complements (`tryRC = TRUE`) and
require a minimal overlap between the two motifs of four nucleotides
(`min.overlap = 4`).

```{r calc-motif-sim}
motifsim <- compare_motifs(motifs = motifs$motif,
                           method = "PCC", tryRC = TRUE,
                           min.overlap = 4, min.mean.ic = 0.25,
                           normalise.scores = TRUE)
```

# Figure 2

## Fraction of motif genomic hits bound

For each motif, we visualize the fraction of sequence hits (sites in the genome
that score highly for the motif) which are bound by the corresponding
transcription factor (overlap with an enriched peak of that transcription
factor). Fractions are shown separately for each peak class.

```{r plot-fracBound}
#| fig.width: 8
#| fig.height: 5

pd <- motifs |>
    select(name, total_hits, starts_with("bound_hits")) |>
    mutate(frac_bound_hits_any = (bound_hits_common.peaks.ubiquitous +
               bound_hits_common.peaks.frequent + bound_hits_specific.peaks) / total_hits,
           `Common (ubiquitous)` = bound_hits_common.peaks.ubiquitous / total_hits,
           `Common (frequent)` = bound_hits_common.peaks.frequent / total_hits,
           `Specific` = bound_hits_specific.peaks / total_hits) |>
    # arrange(desc(frac_bound_hits_any)) |>
    arrange(frac_bound_hits_any) |>
    mutate(name = factor(name, levels = name)) |>
    pivot_longer(cols = c("Common (ubiquitous)", "Common (frequent)",
                          "Specific"), names_to = "peak_set", values_to = "fraction_bound") |>
    mutate(peak_set = factor(peak_set, levels = c("Common (ubiquitous)", "Common (frequent)",
                          "Specific")))

(p_fracbound <- ggplot(pd, aes(fraction_bound, name, fill = peak_set)) +
        geom_col() +
        scale_x_continuous(expand = expansion(mult = 0),
                           labels = scales::label_percent()) +
        scale_fill_manual(values = peakset_colors) +
        labs(x = "Percent of motif sites",
             y = element_blank(),
             fill = element_blank()) +
        theme_cowplot(12) +
        theme(legend.position = "inside",
              legend.position.inside = c(0.95, 0.05),
              legend.justification.inside = c(1, 0),
              axis.text.y = element_text(angle = 0, hjust = 1, size = 8)) +
        guides(fill = guide_legend(ncol = 1)))
```

## Peak vs. sample heatmap (binary)

The following creates heatmap of peaks (rows) versus ChIP-seq experiments
(columns). The values are binary (`TRUE` or `FALSE`) and indicate if a peak
was enriched in ChIP-seq experiment (had an average log2 IP/input enrichment
values greater than 1.0).

```{r plot-peaksBin}
#| fig.width: 6
#| fig.height: 8

# extract binary enrichment matrix from `peakgr`
m <- as.matrix(mcols(peakgr)[, grep("^is_enr_in[.]", colnames(mcols(peakgr)))])
mode(m) <- "numeric"
colnames(m) <- sub("is_enr_in.", "", colnames(m))

# specify ChIP-seq experiments for which to show labels
selLabels <- c("Untagged")
stopifnot(all(selLabels %in% colnames(m)))

# only show peaks that are enriched in at least one ChIP-seq experiment
sel_peaksBin <- rowSums(m) > 0

# prepare annotation data
#   `fractTFs`: fraction of ChIP-seq experiments that a peak is enriched in
#   `numpeaks`: the number of enriched peaks for each ChIP-seq experiment
#   `peaktype`: the type of each peak (specific, frequent or ubiquitous)
#   `near_ncRNA`: specifies if the peak is nearer than `maxdist` from
#                 a %S rRNA or tRNA gene
fracTFs <- rowMeans(m > 0)
numpeaks <- colSums(m > 0)
levels(peakgr$peaktype) <- c("Common (ubiquitous)", # use shorter names
                             "Common (frequent)",
                             "Specific peaks")
peaktype <- factor(gsub("[()]", "", sub(" ", "\n", peakgr$peaktype)),
                   levels = c("Common\nubiquitous",
                              "Common\nfrequent",
                              "Specific\npeaks"))
near_ncRNA <- factor(ifelse(!is.na(dist.peak2tRNA) & dist.peak2tRNA < maxdist,
                            "tRNA",
                            ifelse(!is.na(dist.peak2rRNA) & dist.peak2rRNA < maxdist,
                                   "5S rRNA", "none")),
                         levels = c("tRNA", "5S rRNA", "none"))

# most tRNA and 5S rRNA genes are near "common (ubiquitous)" peaks:
table(peaktype[sel_peaksBin], near_ncRNA[sel_peaksBin])

# parameters for annotation legends and colors
annotLegendParams <- list(`%GC` = list(at = c(20, 40, 60),
                                       labels_gp = gpar(fontsize = fs),
                                       border = "gray10",
                                       legend_direction = "horizontal",
                                       legend_width = unit(18, "mm"),
                                       title_position = "topcenter",
                                       title_gp = gpar(fontsize = fl)),
                          `TSS dist.` = list(at = c(0, 4000, 8000),
                                             labels_gp = gpar(fontsize = fs),
                                             border = "gray10",
                                             legend_direction = "horizontal",
                                             legend_width = unit(18, "mm"),
                                             title_position = "topcenter",
                                             title_gp = gpar(fontsize = fl)),
                          `ncRNA` = list(labels_gp = gpar(fontsize = fs),
                                         title_position = "topcenter",
                                         title_gp = gpar(fontsize = fl)))
cols_ncRNA <- c("tRNA" = main_colors[5], "5S rRNA" = main_colors[3],
                "none" = bg_color)
annotCols <- list(
    `%GC` = colorRamp2(breaks = seq(min(peakgr$fracGC[sel_peaksBin]),
                                    max(peakgr$fracGC[sel_peaksBin]),
                                    length.out = 64) * 100,
                       colors = rev(hcl.colors(64, "Greens"))),
    ` width` = colorRamp2(breaks = seq(min(width(peakgr)[sel_peaksBin]),
                                       max(width(peakgr)[sel_peaksBin]),
                                       length.out = 64),
                          colors = rev(hcl.colors(64, "YlOrBr"))),
    `TSS dist.` = colorRamp2(breaks = c(0, 10^seq(2, log10(max(peakgr$nrst_TSS_dist)),
                                                  length.out = 63)),
                             colors = hcl.colors(64)),
    `ncRNA` = cols_ncRNA)

# prepare heatmap annotations
#   `peakAnnot` for peaks (rows), left side
#   `fracTFsAnnot` for peaks (rows), right side
#   `sampleAnnot` for ChIP-seq experiments (columns), top
#   `sampleLabels` for ChIP-seq experiments (columns), bottom
peakAnnot <- HeatmapAnnotation(
                  which = "row",
                  `%GC` = peakgr$fracGC[sel_peaksBin] * 100,
                  `TSS dist.` = peakgr$nrst_TSS_dist[sel_peaksBin],
                  `ncRNA` = near_ncRNA[sel_peaksBin],
                  col = annotCols,
                  width = unit(21, "mm"), show_legend = TRUE,
                  annotation_name_gp = gpar(fontsize = fl),
                  annotation_legend_param = annotLegendParams)

fracTFsAnnot <- HeatmapAnnotation(
    which = "row",
    `Fraction\nof TFs` = anno_barplot(
        x = fracTFs[sel_peaksBin], gp = gpar(col = na_color),
        border = FALSE, bar_width = 1.0,
        axis_param = list(at = c(0, 0.4, 0.8),
                          gp = gpar(fontsize = fs)),
        width = unit(10, "mm")),
    annotation_name_gp = gpar(fontsize = fl))
sampleAnnot <- HeatmapAnnotation(
    which = "column",
    `Number of\nenriched\npeaks` = anno_barplot(
        x = numpeaks,
        gp = gpar(col = 0, fill = na_color),
        border = FALSE, bar_width = 1.0,
        axis_param = list(at = c(0, 150, 300),
                          gp = gpar(fontsize = fs),
                          facing = "outside",
                          direction = "normal"),
        width = unit(10, "mm")),
    annotation_name_gp = gpar(fontsize = fl),
    annotation_name_side = "left", annotation_name_rot = 0,
    show_legend = TRUE)
sampleLabels <- columnAnnotation(
    TFnames = anno_mark(which = "column", side = "bottom",
                        at = match(selLabels, colnames(m)), 
                        labels = sub("^[^_]+_", "", selLabels),
                        labels_gp = gpar(fontsize = fl)))

# create main heatmap with annotations
hm1 <- Heatmap(m[sel_peaksBin, ],
               col = circlize::colorRamp2(
                   breaks = seq(0, max(m), length.out = 64),
                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
               left_annotation = peakAnnot,
               right_annotation = fracTFsAnnot,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height

# draw the heatmap and grab as a grid graphics `grob`
# (needed for combining figure panels)
# use defined random number seed to make clustering deterministic
set.seed(42L)
hm_peaksBin <- grid.grabExpr(
    hm1 <- draw(hm1, merge_legend = TRUE,
                heatmap_legend_side = "bottom",
                annotation_legend_side = "bottom"),
    width = w_peaksHm, height = h_peaksHm
)

plot_grid(hm_peaksBin)
```

## Peak vs. sample heatmap (enrichments)

The following creates a heatmap similar to the peak-versus-experiment
heatmap above, but using the ChIP-seq enrichment values (log2 IP/input)
directly instead of binarized (`TRUE`, `FALSE`) values.

```{r plot-peaksEnr}
#| fig.width: 6
#| fig.height: 8

# use ChIP-seq enrichment matrix from this study (replicates averaged)
m <- peakenrAvg
stopifnot(all(selLabels %in% colnames(m)))

# to keep the same row order as in the binary heatmap `hm1`,
# we have to re-calculate the annotations and switch off the row clustering
sel_peaksEnr <- which(sel_peaksBin)[unlist(row_order(hm1), use.names = FALSE)]

# prepare annotation data
#   `chip_enr_sel`: selected public ChIP-seq experiments
#                   (log2 IP/input enrichments)
#   `chip_enr_sel_range`: value range for color scale (1% to 99%)
#   `chip_enr_sel_range_full`: value range for color scale (0% to 100%)
chip_enr_sel <- chip_enr[sel_peaksEnr, ]
chip_enr_sel_range <- quantile(abs(chip_enr_sel), probs = c(0.95)) * c(-1, 1)
chip_enr_sel_range_full <- range(chip_enr_sel)

# parameters for annotation legends and colors
annotCols2 <- list(
    `ncRNA` = cols_ncRNA,
    H3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                               seq(chip_enr_sel_range[1],
                                   chip_enr_sel_range[2],
                                   length.out = 62),
                               chip_enr_sel_range_full[2]),
                    colors = viridisLite::magma(64)),
    H3K14ac = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                                    seq(chip_enr_sel_range[1],
                                        chip_enr_sel_range[2],
                                        length.out = 62),
                                    chip_enr_sel_range_full[2]),
                         colors = viridisLite::magma(64)),
    H3K9me2 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                                    seq(chip_enr_sel_range[1],
                                        chip_enr_sel_range[2],
                                        length.out = 62),
                                    chip_enr_sel_range_full[2]),
                         colors = viridisLite::magma(64)),
    H3K9me3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                                    seq(chip_enr_sel_range[1],
                                        chip_enr_sel_range[2],
                                        length.out = 62),
                                    chip_enr_sel_range_full[2]),
                         colors = viridisLite::magma(64)))
annotLegendParams2 <- list(
    `ncRNA` = list(labels_gp = gpar(fontsize = fs),
                   title_position = "topcenter",
                   title_gp = gpar(fontsize = fl)),
    H3 = list(at = c(-6, 0, 6),
              title = "Histone log2 IP/input",
              labels_gp = gpar(fontsize = fs),
              border = "gray10",
              legend_direction = "horizontal",
              legend_width = unit(30, "mm"),
              title_position = "topcenter",
              title_gp = gpar(fontsize = fl)))

# prepare heatmap annotations
#   `peakAnnot` for peaks (rows), left side
#   `peakAnnot2` for peaks (rows), right side
peakAnnot <- HeatmapAnnotation(
                  which = "row",
                  `ncRNA` = near_ncRNA[sel_peaksEnr],
                  H3 = chip_enr_sel[, "GSE108668_H3"],
                  H3K14ac = chip_enr_sel[, "GSE108668_H3K14ac"],
                  H3K9me2 = chip_enr_sel[, "GSE182250_H3K9me2"],
                  H3K9me3 = chip_enr_sel[, "GSE182250_H3K9me3"],
                  col = annotCols2,
                  width = unit(28, "mm"),
                  annotation_name_gp = gpar(fontsize = fl),
                  annotation_legend_param = annotLegendParams2,
                  show_legend = c(TRUE, TRUE, FALSE, FALSE, FALSE))
atac_trunc <- pmin(atac_rpkm_avg[sel_peaksEnr], 2000) # cap ATAC signal at 2000
peakAnnot2 <- HeatmapAnnotation(
    which = "row",
    `Accessibility\n1e3 RPKM` =
        anno_barplot(x = atac_trunc / 1e3,
                     gp = gpar(col = na_color),
                     border = FALSE, bar_width = 1.0,
                     baseline = 0,
                     axis_param = list(at = c(0, 1, 2),
                                       gp = gpar(fontsize = fs)),
                     width = unit(10, "mm")),
    annotation_name_gp = gpar(fontsize = fl))

# main heatmap value range for color scale
mx <- max(abs(m[sel_peaksEnr, ]))
qs <- quantile(abs(m[sel_peaksEnr, ]), .99)

# create main heatmap with annotations
hm2 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
               col = circlize::colorRamp2(
                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
               cluster_rows = FALSE,
               cluster_columns = column_dend(hm1), column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
               border = TRUE, border_gp = gpar(lwd = 0.5),
               left_annotation = peakAnnot,
               right_annotation = peakAnnot2,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = TRUE,
               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                           labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(30, "mm"),
                                           title = "TF log2 IP/input",
                                           title_position = "topcenter",
                                           title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm"))

# draw the heatmap and grab as a grid graphics `grob`
# (needed for combining figure panels)
# use defined random number seed to make clustering deterministic
set.seed(42L)
hm_peaksEnr <- grid.grabExpr(
    hm2 <- draw(hm2, merge_legend = TRUE,
                heatmap_legend_side = "bottom",
                annotation_legend_side = "bottom"),
    width = w_peaksHm, height = h_peaksHm
)

plot_grid(hm_peaksEnr)
```

## Genome browser tracks of ChIP experiments for selected TFs

We want to illustrate ChIP and Input read densities for selected TFs around
their gene locus.

```{r plot-chip-genome-tracks-plot1}
#| fig.width: 5
#| fig.height: 4
#| message: false
#| warning: false

gnmtracks1 <- plottracks(df = chip_genome_tab,
                         gns = c("atf1", "pcr1", "cdc10", "zas1", "reb1", "toe3"),
                         cols = c(atf1 = main_colors[3], pcr1 = main_colors[2],
                                  cdc10 = main_colors[1], zas1 = "#66A61E",
                                  reb1 = main_colors[5], toe3 = main_colors[4]))
gnmtracks1
```

```{r plot-chip-genome-tracks-plot2}
#| fig.width: 5
#| fig.height: 7.4
#| message: false
#| warning: false

gnmtracks2 <- plottracks(df = chip_genome_tab,
                         gns = c("atf1", "pcr1", "cdc10", "zas1", "reb1", "toe3",
                                 "fkh2", "thi5", "SPCC320.03", "SPCC777.02", "prz1"),
                         cols = c(atf1 = main_colors[3], pcr1 = main_colors[2],
                                  cdc10 = main_colors[1], zas1 = "#66A61E",
                                  reb1 = main_colors[5], toe3 = main_colors[4],
                                  fkh2 = main_colors[3], thi5 = main_colors[2],
                                  prz1 = main_colors[5], `SPCC320.03` = "#66A61E",
                                  `SPCC777.02` = main_colors[1]),
                         font_size = 9)
gnmtracks2
```

## Motif similarity heatmap

The following creates a heatmap of pairwise motif similarities (`motifsim`),
hierarchically clustered and grouped into `n_split` clusters.
Specific clusters are are highlighted on the right and motifs in these clusters
are labeled.

```{r plot-motif-sim-small}
#| fig.width: 6
#| fig.height: 6

# define heatmap colors
cols <- hcl.colors(64, "viridis")

# use squared pairwise motif similarities without diagonal entries (self)
motifsim2 <- motifsim^2
diag(motifsim2) <- NA

# value range for color scale
rng <- range(motifsim2, na.rm = TRUE)

# hierarchical clustering of motifs (cut tree to produce `n_split` clusters)
motifcl <- hclust(dist(motifsim2, method = "euclidean"), method = "complete")
n_split <- 17
motifcl_num <- cutree(motifcl, k = n_split)
motifcl <- dendextend::rotate(as.dendrogram(motifcl),
                              c(motifcl_num[motifcl_num != 3],
                                motifcl_num[motifcl_num == 3]))

# prepare heatmap annotation data
#   `selMotifCol`: colors for highlighted motif clusters
#   `selMotifIndex`: row index of motifs in highlighted clusters
#   `nMotifs`: number of motifs identified per transcription factor
selMotifCol <- c("0" = bg_color, "3" = "black", "13" = "black", "16" = "black")
selMotifIndex <- which(motifcl_num[rownames(motifsim2)] %in% names(selMotifCol))
nMotifs <- unclass(table(sub("[.]m[0-9]+$", "", rownames(motifsim2))))[
    sub("[.]m[0-9]+$", "", rownames(motifsim))
]

# prepare heatmap annotations
#   `motifAnnot` for motifs (columns), top
#   `selMotifAnnot` for motif (rows), right side
motifAnnot <- HeatmapAnnotation(
    "Number of\nmotifs per TF" = nMotifs,
    col = list(
        "Number of\nmotifs per TF" = circlize::colorRamp2(
            colors = colorRampPalette(c("white", main_colors[4]))(max(nMotifs)),
            breaks = seq.int(max(nMotifs)))
    ),
    show_legend = TRUE, show_annotation_name = TRUE,
    annotation_name_side = "right",
    annotation_name_gp = gpar(fontsize = fl),
    annotation_legend_param = list(at = seq.int(max(nMotifs)),
                                   labels_gp = gpar(fontsize = fl),
                                   border = "gray10",
                                   legend_direction = "horizontal",
                                   legend_width = unit(18, "mm"),
                                   title_position = "lefttop",
                                   title_gp = gpar(fontsize = fl)))
selMotifAnnot <- rowAnnotation(
    highlight = ifelse(motifcl_num %in% names(selMotifCol),
                       as.character(motifcl_num), "0"),
    motifnames = anno_mark(which = "row", side = "right",
                           at = selMotifIndex, 
                           labels = rownames(motifsim2)[selMotifIndex],
                           labels_gp = gpar(fontsize = fl)),
    col = list(highlight = selMotifCol),
    show_legend = FALSE, show_annotation_name = FALSE)

# create main heatmap with annotations
hm4 <- Heatmap(matrix = motifsim2, name = "Motif\nsimilarity",
               col = circlize::colorRamp2(breaks = seq(rng[1], rng[2],
                                                       length.out = 64),
                                          colors = cols),
               cluster_rows = motifcl,
               show_row_dend = TRUE, row_dend_width = unit(10, "mm"),
               cluster_columns = motifcl, show_column_dend = FALSE,
               row_split = n_split, column_split = n_split,
               column_title = " ",
               row_gap = unit(0.2, "mm"), column_gap = unit(0.2, "mm"),
               show_row_names = FALSE,
               column_title_gp = gpar(fontsize = fl),
               row_title = " ",
               show_column_names = FALSE,
               top_annotation = motifAnnot,
               right_annotation = selMotifAnnot,
               heatmap_legend_param = list(
                   at = round(c(rng[1], mean(rng), rng[2]), 2),
                   labels_gp = gpar(fontsize = fl),
                   border = "gray10",
                   legend_direction = "horizontal",
                   legend_width = unit(18, "mm"),
                   title_position = "lefttop",
                   title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               height = unit(w_peaksHm, "mm")) # heatmap body height

# draw the heatmap and grab as a grid graphics `grob`
# (needed for combining figure panels)
# use defined random number seed to make clustering deterministic
set.seed(42L)
hm_motifs_small <- grid.grabExpr(
    hm4 <- draw(hm4, merge_legend = TRUE,
                heatmap_legend_side = "bottom",
                annotation_legend_side = "bottom"),
    width = w_peaksHm, height = w_peaksHm
)

plot_grid(hm_motifs_small)
```

## Motif alignments

The following creates aligned sequence logos for the motifs in the highlighted
clusters from the motif similarity heatmap.

```{r plot-selected-motifs}
#| fig.width: 6
#| fig.height: 8

# `pL`: list of plots
pL <- list()

# for each highlighted motif cluster...
for (cl in setdiff(names(selMotifCol), c("0", "3"))) {
    # exctract the motif matrices
    selmotifs <- motifs$motif[match(rownames(motifsim2)[motifcl_num == cl], motifs$name)]
    # align and visualize as sequence logos
    pL[[cl]] <- universalmotif::view_motifs(
        motifs = selmotifs,
        method = "PCC", tryRC = TRUE,
        min.overlap = 4,
        min.mean.ic = 0.25,
        normalise.scores = TRUE,
        show.positions = FALSE,
        names.pos = "right",
        text.size = 10)
}

# for each motif row index, get the corresponding cluster
indexToClustername <- structure(rep(seq.int(n_split),
                                    lengths(row_order(hm4))),
                                names = unlist(row_order(hm4)))

# `pL2`: list of plots similar to `pL` but with added empty plots for white space 
pL2 <- rep(list(NULL), 2 * length(pL) + 1)
midx <- seq(2, length(pL2), by = 2)
pL2[midx] <- pL
relh_motifs <- c(0.4, rep(0.8, length(pL2) - 1))
relh_motifs[midx] <- lengths(
    split(selMotifIndex, motifcl_num[selMotifIndex])[
        setdiff(names(selMotifCol), c("0", "3"))
    ])

# combine the individual plots
motifs_aligned <- plot_grid(plotlist = pL2, align = "none", ncol = 1,
                            hjust = -0.1, vjust = 1.1,
                            label_size = 10,
                            rel_heights = relh_motifs)

print(motifs_aligned)
```

## TF peak overlaps from TFs in motif clusters

Create "upset" plots showing the overlap of transcription factor ChIP-seq
peaks among the factors in a highlighted motif cluster.

```{r tf-peak-overlaps}
#| fig.width: 6
#| fig.height: 8
#| warning: false

# select specific and common-frequent peaks only
peakgrSel <- peakgr[peakgr$peaktype %in% c("Specific peaks", "Common (frequent)")]

# `pL3`: list of individual upset plots
pL3 <- list()
# for each highlighted motif cluster...
for (cl in setdiff(names(selMotifCol), c("0", "3"))) {
    motifnms <- rownames(motifsim2)[motifcl_num == cl]
    cltfs <- unique(sub("[.]m[0-9]+$", "", motifnms))
    ovL <- as.list(mcols(peakgrSel)[, paste0("is_enr_in.", cltfs)])
    ovL <- lapply(ovL, function(i) names(peakgrSel)[i])
    names(ovL) <- sub("^is_enr_in.", "", names(ovL))
    nm <- unique(indexToClustername[as.character(which(motifcl_num == cl))])

    pd <- stack(ovL) |>
        left_join(y = data.frame(values = names(peakgr),
                                 type = peakgr$peaktype),
                  join_by(values)) |>
        group_by(values) |>
        summarise(ind = list(ind),
                  type = unique(type))
    pL3[[paste0("cluster ", nm)]] <-
        ggplot(pd, aes(x = ind, fill = type)) +
        geom_bar() +
        scale_x_upset() +
        scale_fill_manual(values = peakset_colors) +
        labs(x = element_blank(),
             y = "Number of\npeaks",
             fill = element_blank()) +
        theme_cowplot(fm) +
        theme(axis.ticks.x = element_blank(),
              legend.position = ifelse(cl == "16", "inside", "none"),
              legend.position.inside = c(0.95, 0.95),
              legend.justification.inside = c(1, 1)) +
        theme_combmatrix(combmatrix.label.text = element_text(colour = "black",
                                                              size = fm),
                         combmatrix.panel.point.size = 2.0,
                         combmatrix.panel.line.size = 0.8,
                         combmatrix.label.extra_spacing = 5,
                         combmatrix.label.make_space = TRUE,
                         combmatrix.label.width = unit(12, "mm"),
                         plot.margin = unit(c(7, 7, 1, 7), "points")) # top, right, bottom, left

}

# `pL4`: list of plots similar to `pL3` but with added empty plots for white space 
pL4 <- rep(list(NULL), 2 * length(pL) + 1)
pL4[midx] <- pL3
relh_upsets <- relh_motifs
relh_upsets[midx] <- relh_upsets[midx] + relh_upsets[midx + 1] - 0.01
relh_upsets[midx + 1] <- 0.01

# combine the individual plots
upsets_aligned <- plot_grid(plotlist = pL4, align = "hv", axis = "l", ncol = 1,
                            rel_heights = relh_upsets)
print(upsets_aligned)
```

## TF Autoregulation

We want to know which transcription factors are binding their own promoter
regions (1kb windows centered on transcription start sites) and thus are
possibly regulating their own transcription.

For visualization, we classify known transcription factors (rows) into three
groups (the ones that do or do not bind their own promoter, and the ones that
were not ChIP'ed in this study) and indicate in a binary heatmap the ChIP-seq
experiments (columns) that showed enrichment at the promoter.

```{r plot-autoregulation}
#| fig.width: 5
#| fig.height: 4

# prepare data
# ... transcription factor genes in ChIP-seq experiments (columns)
tfnms <- colnames(mcols(peakgr))[grep("^is_enr_in[.]", colnames(mcols(peakgr)))]
tfnms <- sub("^is_enr_in[.]", "", tfnms)
tfnms <- ifelse(grepl("^SP", tfnms), tfnms, .lowercase(tfnms))

# ... transcription factors in `txannot` (rows)
tfnms_all <- txannot$unique_einprot_id[match(dbd$PomBaseID, txannot$gene_id)]
tfnms_all <- tfnms_all[order(match(tfnms_all, tfnms))] # order as in `tfnms`
tx_sel <- ifelse(tfnms_all %in% txannot$gene_id,
                 match(tfnms_all, txannot$gene_id),
                 match(tfnms_all, txannot$gene_name))
stopifnot(all(!duplicated(txannot[tx_sel, "gene_id"]))) # exactly one transcript per gene

# ... transcription factors that were ChIP'ed in this study
tfnms_chip <- tfnms_all[txannot[tx_sel, "ChIP.this.study"]]

# ... create promoter/terminator binding matrix
mprom <- do.call(rbind, lapply(tx_sel, function(i) {
    tfnms_i <- strsplit(x = txannot[i, "prom.enriched.TFs"], split = ";")[[1]]
    as.numeric(tfnms %in% ifelse(grepl("^SP", tfnms_i), tfnms_i, .lowercase(tfnms_i)))
}))
dimnames(mprom) <- list(tfnms_all, tfnms)

stopifnot(all.equal(sum(txannot$binds.own.promoter),
                    sum(diag(mprom) > 0)))

diag(mprom[tfnms_chip, tfnms_chip]) <- ifelse(diag(mprom[tfnms_chip, tfnms_chip]) > 0,
                                              2, diag(mprom[tfnms_chip, tfnms_chip]))

# generate binary promoter binding heatmap
# ... label autoregulatory TFs
selLabels <- tfnms_chip[diag(mprom[tfnms_chip, tfnms_chip]) > 0]
stopifnot(all(selLabels %in% rownames(mprom)))
selLabelsPlot <- selLabels
selLabelsPlot[match("zas1", selLabels)] <- "zas1*"
rowLabels <- rowAnnotation(
    TFnames = anno_mark(which = "row", side = "right",
                        at = match(selLabels, rownames(mprom)), 
                        labels = selLabelsPlot,
                        labels_gp = gpar(fontsize = fl)))

# ... group TFs by promoter type
promtype <- factor(ifelse(txannot[tx_sel, "ChIP.this.study"],
                          ifelse(rowSums(mprom) > 0,
                                 "Bound\nTF promoters", "Non-bound\nTF promoters"),
                          "Not\nChIP'ed"),
                   levels = c("Bound\nTF promoters", "Non-bound\nTF promoters",
                              "Not\nChIP'ed"))

# ... order rows according to `promtype` and auto-regulation
is_auto <- function(nm, x = mprom) {
    d <- diag(x[match(nm, rownames(x)),
                match(nm, colnames(x))])
    !is.na(d) & d > 0
}
o <- order(c(3, 2, 1)[as.numeric(promtype)] * 100 +
               is_auto(rownames(mprom), mprom) * 20 +
               rowSums(mprom), decreasing = TRUE)
oc <- match(intersect(rownames(mprom)[o], colnames(mprom)), colnames(mprom))

# ... create main heatmap with annotations
hm5 <- Heatmap(mprom[o, oc], name = "hm5",
               col = circlize::colorRamp2(
                   breaks = seq(0, 2, length.out = 64),
                   colors = colorRampPalette(c(binary_heatmap_colors["FALSE"],
                                               main_colors[1],
                                               binary_heatmap_colors["TRUE"]))(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = FALSE, show_row_dend = FALSE,
               cluster_columns = FALSE, show_column_dend = FALSE,
               row_split = promtype[o], row_title_gp = gpar(fontsize = fl),
               column_title = "TF ChIP-seq experiments",
               column_title_gp = gpar(fontsize = fl),
               show_row_names = FALSE, show_column_names = FALSE,
               right_annotation = rowLabels[o],
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               height = unit(w_peaksHm, "mm"))

hm_prom <- grid.grabExpr(
    {
        hm5 <- draw(hm5, merge_legend = TRUE,
                    heatmap_legend_side = "bottom",
                    annotation_legend_side = "bottom")
        decorate_heatmap_body("hm5", {
            grid.lines(c(mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)])), 0),
                       c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
        }, slice = 1)
        decorate_heatmap_body("hm5", {
            grid.lines(c(1, mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)]))),
                       c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
        }, slice = 2)
    },
    width = w_peaksHm, height = w_peaksHm
)
plot_grid(hm_prom)
```

## ChIP enrichment replicate-pair correlations

```{r plot-chipReplPairs}
#| fig.width: 9
#| fig.height: 9

pd <- peakenr |> as.data.frame() |>
    tibble::rownames_to_column("peakid") |>
    pivot_longer(col = !matches("peakid")) |>
    mutate(group = sub("_rep[12]$", "", name),
           name = factor(name, levels = unique(name)))

ylims <- c(-2.0, 4.0)
yticks <- c(-2, 0, 2, 4)

pL <- lapply(levels(fct_relevel(sort(unique(pd$group)), "Untagged", after = Inf)),
             function(grp1) {
    pd1 <- pd[pd$group == grp1, ] |> pivot_wider()
    pd1$cols <- densCols(x = pd1[[paste0(grp1, "_rep1")]],
                         y = pd1[[paste0(grp1, "_rep2")]],
                         nbin = 64, colramp = colorRampPalette(hcl.colors(32)))
    ggplot(pd1, aes(.data[[paste0(grp1, "_rep1")]],
                    .data[[paste0(grp1, "_rep2")]])) +
        geom_abline(intercept = 0, slope = 1, linetype = 1, color = "gray") +
        geom_hline(yintercept = 0, linetype = 2, color = "gray") +
        geom_vline(xintercept = 0, linetype = 2, color = "gray") +
        geom_scattermost(xy = as.data.frame(pd1[, paste0(grp1, c("_rep1", "_rep2"))]),
                         pointsize = 2, color = pd1$cols, pixels = c(256, 256)) +
        coord_fixed(xlim = ylims, ylim = ylims, expand = FALSE, clip = "off") +
        scale_x_continuous(breaks = yticks) +
        scale_y_continuous(breaks = yticks) +
        theme_cowplot(7) +
        labs(x = element_blank(), y = element_blank()) +
        annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.05,
                 label = grp1, color = "black", size = 3) +
        theme(legend.position = "none")
})

(gg_enrpairs <- plot_grid(plotlist = pL, nrow = 9, ncol = 9))
```

## Peak types and presence of nearby ncRNAs

The following barplots show the numbers of ChIP-seq peaks that are near a
tRNA or 5S rRNA gene (top panel), or the numbers of tRNA/5S rRNA genes
that are distal or near a ChIP-seq peak (bottom panel).

```{r plot-ncRNA}
#| fig.width: 4
#| fig.height: 6

# prepare `data.frame`s with plotting data
df_peaks <- data.frame(peaktype = peaktype[sel_peaksEnr],
                       near_ncRNA = near_ncRNA[sel_peaksEnr])
df_ncRNA <- data.frame(ncRNAtype = factor(
    rep(c("tRNA", "5S rRNA"), c(sum(is_tRNA), sum(is_rRNA))),
    levels = c("tRNA", "5S rRNA")),
                       near_peak = !is.na(c(dist.tRNA2peak, dist.rRNA2peak)) &
                           c(dist.tRNA2peak, dist.rRNA2peak) < maxdist)
df_ncRNA$group <- factor(paste0(df_ncRNA$ncRNAtype, " ", 
                                c("TRUE" = "near", "FALSE" = "distal")[as.character(df_ncRNA$near_peak)]),
                         levels = c("tRNA distal","tRNA near",
                                    "5S rRNA distal", "5S rRNA near"))

# generate plots
# ... numbers of peaks near genes
p_ncRNA1 <- ggplot(df_peaks, aes(peaktype, fill = near_ncRNA)) +
    geom_bar() +
    scale_fill_manual(values = cols_ncRNA) +
    labs(x = element_blank(), y = "Number of peaks", fill = "Near ncRNA:") +
    theme_cowplot(8) +
    theme(legend.position = "inside",
          legend.position.inside = c(0.05, 0.95),
          legend.justification = c(0, 1))

# ... numbers of genes near peaks
p_ncRNA2 <- ggplot(df_ncRNA, aes(ncRNAtype, fill = group)) +
    geom_bar() +
    scale_fill_manual(values = c(`tRNA near` = cols_ncRNA[["tRNA"]],
                                 `tRNA distal` = lighten(cols_ncRNA["tRNA"], 0.2),
                                 `5S rRNA near` = cols_ncRNA[["5S rRNA"]],
                                 `5S rRNA distal` = lighten(cols_ncRNA["5S rRNA"], 0.2))) +
    labs(x = element_blank(), y = "Number of genes", fill = "Relative\npeak location:") +
    theme_cowplot(8) +
    theme(legend.position = "inside",
          legend.position.inside = c(0.95, 0.95),
          legend.justification = c(1, 1))

# ... combine the two panels
(p_ncRNA <- plot_grid(p_ncRNA1, p_ncRNA2, align = "v", nrow = 2))
```

## IP-MS data for HOT TFs

```{r ipms-hot}
#| fig.width: 9
#| fig.height: 9

sce150untag <- readRDS(params$ipms150untag)

res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
tstat <- res150untag$tstat
colnames(tstat) <- .getProteinNameFromComparison(colnames(tstat))
tstat <- tstat[rownames(tstat) %in% colnames(tstat), ]
tstat[is.na(tstat)] <- 0
dim(tstat)

hot_tfs <- c("Php3", "Sak1", "Pcr1", "Prr1", "Atf1", "Rst2", "Adn2", 
             "Adn3", "Hsr1", "Phx1", "Pho7")
stopifnot(all(hot_tfs %in% rownames(tstat)))
stopifnot(all(hot_tfs %in% colnames(tstat)))
stopifnot(all(rownames(tstat) %in% colnames(tstat)))
tstat <- tstat[c(hot_tfs, setdiff(rownames(tstat), hot_tfs)), ]
tstat <- tstat[, c(rownames(tstat), setdiff(colnames(tstat), rownames(tstat)))]

hmipmshot <- Heatmap(
    tstat, 
    col = circlize::colorRamp2(
        breaks = seq(3.0, 13.0, length.out = 64),
        colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
    border = TRUE,
    row_split = ifelse(rownames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"), 
    column_split = ifelse(colnames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"),
    cluster_rows = FALSE, cluster_columns = FALSE,
    row_title_gp = gpar(fontsize = fl), row_names_gp = gpar(fontsize = fs),
    column_title_gp = gpar(fontsize = fl), show_column_names = FALSE,
    heatmap_legend_param = list(labels_gp = gpar(fontsize = fs),
                                border = "gray10",
                                legend_direction = "horizontal",
                                legend_width = unit(24, "mm"),
                                title_position = "lefttop",
                                title_gp = gpar(fontsize = fl)),
    na_col = bg_color,
    name = "Mod. t-stat.")

hm_ipms_hot <- grid.grabExpr(
    hmipmshot <- draw(hmipmshot, merge_legend = TRUE, 
                      heatmap_legend_side = "bottom")
)
plot_grid(hm_ipms_hot)
```

## Put together

Assemble the panels into Figure 2:

```{r Fig2}
#| fig.width: 11.5
#| fig.height: 16
#| warning: false

cowplot::plot_grid(
    hm_peaksBin,
    hm_peaksEnr,
    hm_prom,
    cowplot::plot_grid(gnmtracks1, scale = 0.9),
    hm_motifs_small,
    cowplot::plot_grid(
        motifs_aligned, upsets_aligned,
        nrow = 1, rel_widths = c(1.25, 1)),
    nrow = 3,
    labels = c("A", "B", "C", "D", "E", "F"),
    align = "vh",
    axis = "b",
    rel_widths = c(1, 1),
    rel_heights = c(h_peaksHm + 10, w_peaksHm + 10, w_peaksHm + 20)
)
```

## Supplementary figure

Assemble the panels into Supplementary Figure 2:

```{r suppFig2}
#| fig.width: 11
#| fig.height: 20
#| warning: false

cowplot::plot_grid(
    cowplot::plot_grid(gg_enrpairs, scale = 0.9, labels = "A"),
    cowplot::plot_grid(NULL, NULL, NULL, # space for labels
                       nrow = 1, labels = c("B", "C", "D"),
                       rel_widths = c(130, 100, 100)),
    cowplot::plot_grid(
        hm_ipms_hot,
        cowplot::plot_grid(gnmtracks2, NULL, ncol = 1, rel_heights = c(12, 1)),
        p_fracbound,
        nrow = 1, 
        labels = c("", "", ""),
        align = "v",
        axis = "b",
        rel_widths = c(130, 100, 100),
        rel_heights = c(180)
    ),
    nrow = 3,
    rel_widths = c(1),
    rel_heights = c(12, 0.25, 9.75)
)
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-07-04 ms_figure_4_motifs.Rmd
rename figs2-6 to figs3-7
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "Figure 3 - Motifs"
author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    dbdtxt: "data/TF_DBD_annotation.txt"
    motifrds: "data/motifs_annotated.rds"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    genomefasta: "reference/Schizosaccharomyces_pombe.ASM294v2.dna.toplevel.fa"
    ncpu: 4
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
```

# Parameter values

```{r params}
params
```

# Load required packages

```{r load-packages}
suppressPackageStartupMessages({
    library(circlize)
    library(ComplexHeatmap)
    library(grid)
    library(ggplot2)
    library(cowplot)
    library(dplyr)
    library(tidyr)
    library(universalmotif)
    library(TFBSTools)
    library(GenomicRanges)
    library(parallel)
    library(colorspace)
    library(forcats)
    library(monaLisa)
    library(ggrepel)
    library(Biostrings)
    library(ggupset)
})

source("params/plot_settings.R")

# capitalize the first letter of a string
.capitalize <- function(x) {
    paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}
```

# Read data

We start by reading all tables and objects needed for this figure.
Factor levels are defined in the order they should appear in the figure
panels, and peak sequences are extracted from the genome.

Remark: Motifs are read from the `r params$motifrds` file, which is in
`.rds` format that can be used by `R`. The motifs are also available in the
file [`data/motifs.meme`](data/motifs.meme), which is a plain text file with the
motifs in the MEME format: https://meme-suite.org/meme/doc/meme-format.html.

```{r load-data}
# `dbd`: Transcription factor DNA binding domain information
dbd <- read.delim(params$dbdtxt) |>
    dplyr::mutate(cons = factor(ifelse(Conserved_human, "yes", "no"),
                                levels = c("yes", "no")),
                  DBD_class_simple = factor(
                      ifelse(DBD_class == "Zn(II)2Cys6", "Zn(II)2Cys6", "other"),
                      levels = c("other", "Zn(II)2Cys6")))

# `motifs`: table with annotated motif
motifs <- readRDS(params$motifrds)
motifs$altname <- motifs$name

# ... extract motifs as TFBSTools::PFMatrixList
umL <- universalmotif::to_list(motifs)
pfmL <- do.call(TFBSTools::PFMatrixList,
                lapply(seq.int(nrow(motifs)), function(i) {
                    universalmotif::convert_motifs(motifs = umL[[i]],
                                                   class = "TFBSTools-PFMatrix")
                }))

# `gnm`: S. pombe reference genome sequence
gnm <- readDNAStringSet(params$genomefasta)
names(gnm) <- sub(" .*$", "", names(gnm))

# `peakgr`: ChIP-seq peaks as GRanges object
peakgr <- as(read.csv(params$peakcsv, row.names = 1), "GRanges")
peakgr$peaktype <- factor(.capitalize(peakgr$peaktype),
                          levels = c("Common peaks (ubiquitous)",
                                     "Common peaks (frequent)",
                                     "Specific peaks"))
peakgr_specific <- peakgr[peakgr$peaktype == "Specific peaks"]

# ... extract peak enrichment matrix
is_enr <- as.matrix(mcols(peakgr)[, grep("^is_enr_in[.]", colnames(mcols(peakgr)))])
colnames(is_enr) <- sub("^is_enr_in.", "", colnames(is_enr))
is_enr_specific <- as.matrix(mcols(peakgr_specific)[, grep("^is_enr_in[.]", colnames(mcols(peakgr_specific)))])
colnames(is_enr_specific) <- sub("^is_enr_in.", "", colnames(is_enr_specific))

# ... extract peak sequences from genome
peakseqs <- getSeq(gnm, peakgr)
peakseqs_specific <- getSeq(gnm, peakgr_specific)
```

# Helper functions

Here we define some helper functions that are used to generate this figure:

- `calcKmerEnr`: Counts the frequencies of k-mer words of length `k` in
  a set of sequences (`seqs`) that are grouped into foreground/background sets
  by `fg`, and calculates enrichment values for each k-mer word comparing
  the foreground frequencies to either the background sequences
  (`bg_type = "other"`) or to a Markov model of order `Markov_order` that is
  estimated from the foreground sequences.
- `kmerEnrPlot`: Calls `calcKmerEnr` to obtain k-mer enrichments and visualizes
  them in a scatter plot (each point corresponding to a k-mer), with x- and y-axis
  coordinates corresponding to the log k-mer frequency in foreground sequences
  and the log k-mer enrichment, respectively.
- `create_cobound_peaks_plot`: Upset plot showing co-bound peaks for a set of TFs.

```{r helpers}
calcKmerEnr <- function(k = 5, # k-mer length
                        seqs,  # DNAStringSet
                        fg,    # logical vector of foreground in `seqs`
                        bg_type = c("other", "Markov"),
                        Markov_order = 1,
                        include_revcomp = TRUE,
                        ... # forwarded to monaLisa::getKmerFreq
                        ) {
    # digest arguments
    bg_type <- match.arg(bg_type)
    stopifnot(exprs = {
        is.numeric(k) && length(k) == 1L
        is(seqs, "DNAStringSet")
        is.logical(fg) && length(fg) == length(seqs)
        is.numeric(Markov_order) && length(Markov_order) == 1L
    })

    # fg
    fg_kmers <- monaLisa::getKmerFreq(seqs[fg], kmerLen = k,
                                      MMorder = Markov_order,
                                      includeRevComp = include_revcomp,
                                      ...)
    fg_freq <- fg_kmers$freq.obs
    kmers <- names(fg_freq)

    # bg
    if (identical(bg_type, "other")) {
        bg_kmers <- monaLisa::getKmerFreq(seqs[!fg], kmerLen = k,
                                          MMorder = Markov_order,
                                          includeRevComp = include_revcomp,
                                          ...)
        bg_freq <- bg_kmers$freq.obs
    } else if (bg_type == "Markov") {
        bg_freq <- fg_kmers$freq.exp
    }

    # norm
    N <- min(sum(fg_freq), sum(bg_freq))
    fg_freq_norm <- fg_freq / sum(fg_freq) * N
    bg_freq_norm <- bg_freq / sum(bg_freq) * N

    # return results
    return(list(kmers = names(fg_freq),
                fg_kmers = fg_kmers,
                fg_freq = fg_freq,
                fg_freq_norm = fg_freq_norm,
                bg_kmers = bg_kmers,
                bg_freq = bg_freq,
                bg_freq_norm = bg_freq_norm,
                pearsonresid = (fg_freq_norm - bg_freq_norm) /
                         sqrt(bg_freq_norm),
                nCG = vcountPattern("C", kmers) + vcountPattern("G", kmers)))
}

kmerEnrPlot <- function(k = 5, # k-mer length
                        seqs,  # DNAStringSet
                        fg,    # logical vector of foreground in `seqs`
                        bg_type = c("other", "Markov"),
                        Markov_order = 1,
                        colorByMotif = NULL,
                        nlabel = 0,
                        title = NULL,
                        auto_subtitle = FALSE,
                        jitter_x = FALSE,
                        include_revcomp = TRUE,
                        font_size = 14,
                        ... # forwarded to monaLisa::getKmerFreq
                        ) {
    # digest arguments
    stopifnot(exprs = {
        is.null(colorByMotif) || is(colorByMotif, "PFMatrixList")
        is.numeric(nlabel) && length(nlabel) == 1L
        is.null(title) || (is.character(title) && length(title) == 1L)
    })

    # calculate k-mer enrichments
    enr <- calcKmerEnr(k = k,
                       seqs = seqs,
                       fg = fg,
                       bg_type = bg_type,
                       Markov_order = Markov_order,
                       include_revcomp = include_revcomp)

    # enrichment plot
    jitter_amount <- if (jitter_x) 0.5 else 0
    pd <- data.frame(kmer = enr$kmers,
                     fg_freq = enr$fg_freq,
                     bg_freq = enr$bg_freq,
                     fg_freq_norm = enr$fg_freq_norm,
                     bg_freq_norm = enr$bg_freq_norm,
                     pearsonresid = enr$pearsonresid,
                     nCG = enr$nCG,
                     jitter_x = runif(length(enr$kmers),
                                      min = -jitter_amount, max = jitter_amount))
    if (!is.null(colorByMotif)) {
        pd$score <- monaLisa::motifKmerSimilarity(x = colorByMotif, kmerLen = k,
                                                  includeRevComp = include_revcomp)[1, ]
    }

    # ... base plot
    p <- ggplot(pd, aes(fg_freq + 1 + jitter_x, pearsonresid)) +
        scale_x_log10() +
        labs(title = if (is.null(title)) element_blank() else title,
             x = paste0(k, "-mer frequency in peaks + 1"),
             y = paste0(k, "-mer enrichment")) +
        theme_cowplot(font_size) +
        theme(legend.position = "none")
    # ... subtitle
    if (auto_subtitle) {
        if (identical(bg_type, "Markov")) {
            p <- p + labs(subtitle = paste0(k, "-mer enrichments (background: ",
                                            Markov_order,
                                            "th-order Markov model)"))
        } else if (identical(bg_type, "other")) {
            p <- p + labs(subtitle = paste0(k, "-mer enrichments (background: other peaks)"))
        }
    }
    # ... points
    if (is.null(colorByMotif)) {
        p <- p + geom_point(color = "gray62", alpha = 0.5, size = 1.5 * font_size / 14) +
            geom_hline(yintercept = 0, linetype = "dashed")
    } else {
        p <- p +
            geom_point(mapping = aes(color = score^0.4), alpha = 0.5) +
            scale_colour_gradientn(colors = hcl.colors(11, "Oslo")[-1]) +
            geom_hline(yintercept = 0, linetype = "dashed")
    }
    # ... point labels
    if (nlabel > 0) {
        p <- p + geom_text_repel(
            # data = pd[order(pd$pearsonresid,
            #                 decreasing = TRUE)[seq.int(nlabel)], ],
            data = pd |>
                arrange(desc(pearsonresid)) |>
                mutate(label_i = seq_along(pearsonresid),
                       kmerrc = as.character(reverseComplement(DNAStringSet(kmer))),
                       labelrc_i = match(kmer, kmerrc),
                       label_str = ifelse(label_i <= nlabel & label_i <= labelrc_i, kmer, "")),
            mapping = aes(label = label_str), min.segment.length = 0,
            size = font_size / ggplot2::.pt, color = "black",
            max.overlaps = Inf)
    }

    # return plot
    return(p)
}

create_cobound_peaks_plot <- function(sel_tfs) {
    stopifnot(all(sel_tfs %in% colnames(is_enr)))

    ovL <- as.list(as.data.frame(is_enr[, sel_tfs]))
    ovL <- lapply(ovL, function(i) rownames(is_enr)[i])

    pd <- stack(ovL) |>
        group_by(values) |>
        summarise(ind = list(ind)) |>
        mutate(peaktype = peakgr[values]$peaktype)

    ggplot(pd, aes(x = ind)) +
        geom_bar(aes(fill = peaktype)) +
        scale_fill_manual(values = peakset_colors) +
        scale_x_upset() +
        labs(x = element_blank(), y = "Number of peaks", fill = "Peak type:  ") +
        theme_cowplot(12) +
        theme(axis.ticks.x = element_blank(),
              legend.position = "bottom") +
        theme_combmatrix(combmatrix.label.extra_spacing = 5,
                         combmatrix.label.make_space = TRUE,
                         combmatrix.label.width = unit(18, "mm")) +
        guides(fill = guide_legend(nrow = 3))
}
```

# Figure 3

## Zn(II)2Cys6 family motifs

### Zn(II)2Cys6 TF motifs k-mer enrichment plots

Frequency and enrichments of 6-mer sequence words in ChIP-seq peaks from
transcription factors of the Zn(II)2Cys6 (GAL4) family:

```{r plot-ZnII2Cys6-kmer}
#| fig-width: 6
#| fig-height: 6
# select motifs of TFs from the Zn(II)2Cys6 family
sel_motifs <- motifs[motifs$family == "Zn(II)2Cys6", "name"]
tf_names <- sub("[.]m[0-9]+$", "", sel_motifs)

# select a single TF and motif of interest ("Toe3.m1")
sel_motif_idx_1 <- 7
motif_name1 <- sel_motifs[sel_motif_idx_1]

# plot 6-mer enrichments in enriched peaks from which the motif was identified
ZnII2Cys6_kmer_1 <- kmerEnrPlot(k = 6,
                                seqs = peakseqs_specific,
                                fg = is_enr_specific[, tf_names[sel_motif_idx_1]],
                                bg_type = "other",
                                title = sel_motifs[sel_motif_idx_1],
                                nlabel = 9,
                                colorByMotif = pfmL[match(sel_motifs[sel_motif_idx_1], name(pfmL))],
                                font_size = 10)

plot_grid(ZnII2Cys6_kmer_1)
```

### Zn(II)2Cys6 TF `CCG`, `CGG` enrichments (TF peaks from `r length(sel_motifs)` selected motifs)

```{r hm-ZnII2Cys6-selected-3mer-enr}
#| fig-width: 8
#| fig-height: 4

# select Zn(II)2Cys6 TFs
tf_names <- c("Cha4", "Mca1", "Moc3", "Pho7", "Prt1", 
              "Grt1", "Gsf1", "Thi1", "Thi5", "Toe1", 
              "Toe2", "Toe3", "Toe4", "SPAC11D3.11c", "SPAC1327.01c", 
              "SPAC1F7.11c", "SPAC25B8.11", "SPAC2H10.01", "SPAC3C7.04", 
              "SPAC3H8.08c", "SPBC1348.12", "Ntu1", "SPBC16G5.17", 
              "SPBC1773.12", "SPBC1773.16c", "Ntu2", "SPBC56F2.05c", 
              "SPCC320.03", "SPCC417.09c", "SPCC757.04", "SPCC777.02", 
              "SPCC965.10") 
length(tf_names)

# filter out not ChIP'ed TFs or TFs with less than `min_peaks_specific` specific peaks
min_peaks_specific <- 3
n_peaks_specific <- colSums(
    is_enr_specific[, intersect(colnames(is_enr_specific), tf_names)]
    )[tf_names]
tf_names <- tf_names[!is.na(n_peaks_specific) & n_peaks_specific >= min_peaks_specific]
length(tf_names)

# calculate 3-mer enrichments
enr_list_3mers <- list()
for (i in seq_along(tf_names)) {
    enr_list_3mers[[i]] <- calcKmerEnr(k = 3,
                                       seqs = peakseqs_specific,
                                       fg = is_enr_specific[, tf_names[i]],
                                       bg_type = "other")
}

enr_3mers <- do.call(rbind, lapply(enr_list_3mers, "[[", "pearsonresid"))
rownames(enr_3mers) <- tf_names

# plot heatmap
# ... define heatmap font sizes
fs <- 5.5
fl <- 8

# ... calculate data range for mapping to colors
mx <- max(abs(enr_3mers))
qs <- quantile(abs(enr_3mers), .99)

# ... prepare heatmap annotations:
#      nCG: number of G+C bases in the 3-mer word
#      highlight: emphasize CCG, CGG, GGC and GCC
kmer_annot <- columnAnnotation(
    nCG = vcountPattern("S", DNAStringSet(colnames(enr_3mers)), fixed = FALSE),
    highlight = ifelse(colnames(enr_3mers) %in% c("CCG", "CGG", "GGC", "GCC"), "1", "0"),
    col = list(highlight = c("0" = bg_color, "1" = "black"),
               nCG = structure(hcl.colors(5, "Greens")[-5], names = 3:0)),
    show_legend = c(TRUE, FALSE), show_annotation_name = c(FALSE, FALSE),
    annotation_name_gp = gpar(fontsize = fl),
    annotation_legend_param = list(labels_gp = gpar(fontsize = fl),
                                   ncol = 4,
                                   title = "Number of G+C   ",
                                   title_position = "lefttop",
                                   title_gp = gpar(fontsize = fl)),
    simple_anno_size = unit(3, "mm"))

# ... generate main heatmap with annotations
hm_3mers <- Heatmap(enr_3mers, name = "Pearson's resid.",
                    col = circlize::colorRamp2(
                        breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                        colors = colorRampPalette(enrichment_heatmap_colors)(64)),
                    cluster_rows = TRUE, show_row_dend = FALSE,
                    column_dend_reorder = match(colnames(enr_3mers),
                                                c("GCG", "CCG", "GGC", "GGG"),
                                                nomatch = 100),
                    cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
                    row_names_gp = gpar(fontsize = fl, fontfamily = "sans"),
                    column_names_gp = gpar(fontsize = fs, fontfamily = "mono"),
                    border = TRUE, border_gp = gpar(lwd = 0.5),
                    show_row_names = TRUE, show_column_names = TRUE,
                    use_raster = FALSE, show_heatmap_legend = TRUE,
                    bottom_annotation = kmer_annot,
                    heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                                labels_gp = gpar(fontsize = fl),
                                                border = "gray10",
                                                legend_direction = "horizontal",
                                                legend_width = unit(25, "mm"),
                                                title = "3-mer enrichment   ",
                                                title_position = "lefttop",
                                                title_gp = gpar(fontsize = fl)),
                    width = unit(130, "mm"), # heatmap body width
                    height = unit(3 * nrow(enr_3mers), "mm"))

# draw the heatmap and grab as a grid graphics `grob`
# (needed for combining figure panels)
# use defined random number seed to make clustering deterministic
set.seed(42L)
hm_ZnII2Cys6_sel_3mers <- grid.grabExpr(
    hm_3mers <- draw(hm_3mers, merge_legend = TRUE,
                     heatmap_legend_side = "bottom",
                     annotation_legend_side = "right"),
    width = 40, height = 140
)

plot_grid(hm_ZnII2Cys6_sel_3mers)
```

## Known and new motif characterization

Display information for selected motifs (`sel_motifs`, containing examples for 
both previously known and novel motifs):

```{r plot-known-new-details}
#| fig-width: 12
#| fig-height: 2.75
#| warning: false

# select motifs for detailed visualization
sel_motifs <- c(
    # known
    "Cdc10.m1",
    # novel
    "Toe2.m1", "Ams2.m1", "Esc1.m1",
    "SPAC3F10.12c.m1", "Zip1.m1", "Adn2.m4")
tf_names <- sub("[.]m[0-9]+$", "", sel_motifs)

# create motif visualizations
# ... setup plotting parameters
fgbg_col <- c(main_colors[1], main_colors[5]) # colors for foreground and background
gnm_col <- na_color # color for NA or "other"
w <- 2500  # width in bases around peak-center for meta-profiles
fs <- 9 # font size
nlabel <- 1 # number of k-mer words to label in enrichment plot

# ... iterate over motifs
motif_characterization_plots <- list()
for (i in seq_along(sel_motifs)) {
    # get motif info
    m1 <- motifs$motif[match(sel_motifs[i], motifs$name)]
    is_fg_specific <- is_enr_specific[, tf_names[i]]
    fg_peak_names <- names(peakgr)[is_enr[, tf_names[i]]]
    bg_peak_names <- setdiff(names(peakgr), fg_peak_names)
    fgbg_num <- c(length(fg_peak_names), length(bg_peak_names))
    fgbg_names <- sprintf("%s (%d)", c("foreground", "background"), fgbg_num)
    fgbg_names_short <- sprintf("%d", fgbg_num)

    # scan genome sequence with motif
    hits_genome <- universalmotif::scan_sequences(motifs = m1,
                                                  sequences = gnm,
                                                  threshold = 1e-4,
                                                  threshold.type = "pvalue", 
                                                  RC = TRUE,
                                                  verbose = if (interactive()) 3 else 0,
                                                  nthreads = params$ncpu,
                                                  return.granges = TRUE)

    # seqlogo
    gg1_seqlogo <- universalmotif::view_motifs(
        motifs = m1,
        show.positions = FALSE) +
        labs(title = sel_motifs[i],
             subtitle = paste0(sum(is_fg_specific), " specific peaks")) +
        theme_cowplot(fs) +
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              axis.line.x = element_blank(),
              legend.position = "none")

    # k-mer enrichments
    gg1_kmers <- kmerEnrPlot(k = 6,
                             seqs = peakseqs_specific,
                             fg = is_enr_specific[, tf_names[i]],
                             bg_type = "other",
                             nlabel = nlabel,
                             colorByMotif = pfmL[match(sel_motifs[i], name(pfmL))],
                             font_size = fs) +
        labs(x = "6-mer freq. + 1",
             y = "6-mer enr.") +
        theme(legend.position = "none")

    # motif location relative to peak mid
    nrst <- distanceToNearest(x = resize(hits_genome, width = 1L, fix = "center"),
                              subject = resize(peakgr, width = 1L, fix = "center"))
    nrst_fg <- nrst[subjectHits(nrst) %in% match(fg_peak_names, names(peakgr))]
    nrst_bg <- nrst[subjectHits(nrst) %in% match(bg_peak_names, names(peakgr))]
    relpos_fg <- mid(ranges(hits_genome))[queryHits(nrst_fg)] - mid(ranges(peakgr))[subjectHits(nrst_fg)]
    relpos_bg <- mid(ranges(hits_genome))[queryHits(nrst_bg)] - mid(ranges(peakgr))[subjectHits(nrst_bg)]
    pd_loc <- data.frame(motif = rep(sel_motifs[i], length(nrst_fg) + length(nrst_bg)),
                         relpos = c(relpos_fg, relpos_bg),
                         type = factor(rep(c(fgbg_names[1], fgbg_names[2]),
                                           c(length(nrst_fg), length(nrst_bg))),
                                       levels = fgbg_names))
    gg1_loc <- ggplot(pd_loc |> filter(abs(relpos) < w + 500),
                 aes(relpos / 1000, color = type)) +
        geom_density(linewidth = 1) +
        geom_vline(xintercept = 0, linetype = "dotted") +
        scale_y_continuous(n.breaks = 3) +
        scale_x_continuous(limits = c(-w, w) / 1000, expand = expansion(mult = 0)) +
        scale_color_manual(values = structure(fgbg_col, names = fgbg_names)) +
        annotate("text", x = w / 1000, y = Inf, hjust = 1, vjust = 1,
                 label = paste0(c("", "\n"), fgbg_names_short),
                 colour = fgbg_col, fontface = "bold", size = fs / ggplot2::.pt) +
        labs(x = "Position relative to peak center (kb)",
             y = "Density of motifs") +
        theme_cowplot(fs) +
        theme(legend.position = "none",
              legend.justification = c(1, 1))

    # motif hit enrichment in genomic regions
    ov <- findOverlaps(query = hits_genome, subject = peakgr)
    ov_type <- rep("non-peak", length(hits_genome))
    ov_type[queryHits(ov)] <- 
        ifelse(subjectHits(ov) %in% match(fg_peak_names, names(peakgr)),
               fgbg_names[1], fgbg_names[2])
    ov_type <- factor(ov_type,
                      levels = c("non-peak", fgbg_names[2], fgbg_names[1]))
    n_ov_obs <- unclass(table(hits_genome$motif, ov_type))
    n_ov_exp <- unclass(table(hits_genome$motif)) %*%
        t(structure(c(sum(seqlengths(hits_genome)) - sum(width(peakgr)),
                      sum(width(peakgr[bg_peak_names])),
                      sum(width(peakgr[fg_peak_names]))),
                    names = colnames(n_ov_obs)) / sum(seqlengths(hits_genome)))

    pd_counts <- full_join(as.data.frame(n_ov_obs) |>
                               mutate(motif = "motif") |>
                               pivot_longer(cols = all_of(levels(ov_type)),
                                            names_to = "type",
                                            values_to = "observed"),
                           as.data.frame(n_ov_exp) |>
                               mutate(motif = "motif") |>
                               pivot_longer(cols = all_of(levels(ov_type)),
                                            names_to = "type",
                                            values_to = "expected"),
                           by = join_by(motif, type)) |>
        dplyr::mutate(enr = observed / expected) |>
        pivot_longer(cols = all_of(c("observed", "expected")),
                     names_to = "obsexp", values_to = "val") |>
        dplyr::mutate(type = factor(type, levels = c("non-peak", fgbg_names[2:1]))) |>
        group_by(motif, obsexp) |>
        mutate(tot = sum(val),
               percentage = val / tot) |>
        ungroup()
    pd_counts <- rbind(pd_counts |>
                           filter(obsexp == "observed") |>
                           dplyr::select(motif, type, enr, percentage),
                       pd_counts |>
                           filter(obsexp == "expected" & motif == pd_counts[[1,"motif"]]) |>
                           mutate(motif = "expected") |>
                           dplyr::select(motif, type, enr, percentage)
    ) |> mutate(motif = relevel(factor(motif), "expected"))

    gg_ov_counts <- ggplot(pd_counts, aes(percentage, motif, fill = type)) +
        geom_col(position = position_stack()) +
        scale_x_continuous(labels = scales::label_percent(),
                           expand = expansion(mult = c(0, 0.15))) +
        scale_fill_manual(values = structure(c(gnm_col, fgbg_col[2:1]),
                                             names = levels(pd_counts$type))) +
        geom_text(data = pd_counts |> filter(grepl("foreground", type)),
                  aes(x = 1, label = sprintf("%.1f%%", percentage * 100)),
                  colour = fgbg_col[1], hjust = -0.1, vjust = 0.5,
                  size = fs / ggplot2::.pt,
                  fontface = "bold") +
        labs(x = "Fraction of motif sites",
             y = element_blank(),
             fill = "Genomic region: ") +
        theme_cowplot(fs) +
        theme(legend.position = "none") +
        guides(fill = guide_legend(nrow = 3))

    pd_enr <- pd_counts |> filter(motif != "expected") |>
        dplyr::select(c("motif","type","enr")) |> unique()
    levels(pd_enr$type) <- c("non-peak", "backgr.", "foregr.")
    gg_ov_enr <- ggplot(pd_enr, aes(enr, type, fill = type)) +
        geom_col(position = position_dodge2()) +
        scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
        geom_vline(xintercept = 1, linetype = "dotted") +
        scale_fill_manual(values = structure(c(gnm_col, fgbg_col[2:1]),
                                             names = levels(pd_enr$type))) +
        # facet_wrap(~ motif, nrow = 1) +
        labs(x = "Motif site enrichment (obs/exp)",
             y = element_blank(),
             fill = "Genomic region: ") +
        theme_cowplot(fs) +
        theme(legend.position = "bottom") +
        guides(fill = guide_legend(nrow = 3))

    gg_ov_enr_legend <- cowplot::get_legend(gg_ov_enr)
    gg_ov_enr <- gg_ov_enr + theme(legend.position = "none")

    # put panels together
    motif_characterization_plots[[i]] <- cowplot::plot_grid(
        cowplot::plot_grid(gg1_seqlogo, NULL, ncol = 1, rel_heights = c(0.87, 0.13)),
        gg1_kmers,
        gg1_loc,
        cowplot::plot_grid(gg_ov_counts,
                           gg_ov_enr,
                           align = "v", axis = "l",
                           nrow = 2,
                           rel_heights = c(1, 1.2)),
        # align = "h", axis = "tb",
        nrow = 1,
        rel_widths = c(1, 1, 1, 1.4)
    )

    print(motif_characterization_plots[[i]])
}

# add legend
motif_characterization_plots[[length(motif_characterization_plots) + 1]] <-
    cowplot::plot_grid(NULL, NULL, NULL, gg_ov_enr_legend,
                       nrow = 1, rel_widths = c(1, 1, 1, 1.4))
```

## Co-bound peaks for selected transcription factors

```{r co-bound-peaks}
(p_cobound_1 <- create_cobound_peaks_plot(sel_tfs = c("Ams2", "Zas1", "Teb1")))
(p_cobound_2 <- create_cobound_peaks_plot(sel_tfs = c("Esc1", "SPAC3F10.12c")))
```

## Put together

Assemble the panels into Figure 3:

```{r Fig3}
#| fig.width: 11
#| fig.height: 15.5

cowplot::plot_grid(
    cowplot::plot_grid(
        cowplot::plot_grid(ZnII2Cys6_kmer_1, scale = 0.85),
        hm_ZnII2Cys6_sel_3mers,
        labels = c("A", "B"),
        rel_widths = c(2, 3.2),
        nrow = 1
    ),
    cowplot::plot_grid(
        plotlist = motif_characterization_plots,
        ncol = 1,
        rel_heights = c(rep(1, length(motif_characterization_plots) - 1), 0.4)),
    nrow = 2,
    labels = c("", "C"),
    align = "vh",
    axis = "b",
    rel_heights = c(1.4, 3.7)
)
```

## Supplementary figure

Assemble the panels into Supplementary Figure 3:

```{r suppFig3}
#| fig.width: 4.5
#| fig.height: 6

cowplot::plot_grid(
    p_cobound_2,
    nrow = 1,
    labels = c("A")
)
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-07-04 ms_figure_5_ipms.Rmd
rename figs2-6 to figs3-7
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "Figure 4 - IP-MS"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    rdsdal2: "data/ipms_dal2_sce.rds"
    complexes: "data/complexes.json"
    baitclass: "data/ipms_bait_class.txt"
    dnabinders: "data/Gene_list_DNAbinders_proteincoding.txt"
    rnabinders: "data/Gene_list_RNAbinders_proteincoding.txt"
    idmap: "data/id_mapping_table.txt"
    dbdtxt: "data/TF_DBD_annotation.txt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
options(scipen = 6)
```

# Parameter values

```{r params}
params
```

# Load required packages and helper functions

```{r load-packages, class.source="fold-show"}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(ggplot2)
    library(cowplot)
    library(einprot)
    library(dplyr)
    library(forcats)
    library(jsonlite)
    library(ComplexHeatmap)
    library(colorspace)
    library(ggrepel)
    library(DT)
    library(stringr)
    library(tidyr)
    library(kableExtra)
    library(patchwork)
})

## Source scripts with required helper functions and settings
source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/ipms_params.R")
source("params/mapping_functions.R")

## Heatmap sizes
w_ipmsHm <-  84 # heatmap body width (mm)
h_ipmsHm <- 178 # total heatmap height (mm)

fs <- 8  # small font size
fl <- 10 # large font size
ft <- 12 # title font size
```

# Read data

We load two `SingleCellExperiment` objects, containing data and results from the 
low- and high-salt experiments, respectively, as well as one object with 
results from a separate run of Dal2.

```{r load-data, class.source="fold-show"}
sce150 <- readRDS(params$rds150)
sce500 <- readRDS(params$rds500)
scedal2 <- readRDS(params$rdsdal2)
```

We then extract the full list of baits, as well as a classification according 
to the experiment(s) where a certain bait was pulled down.

```{r, class.source="fold-show"}
idmap <- read.delim(params$idmap)
baitclass <- read.delim(params$baitclass)
DT::datatable(baitclass, rownames = FALSE)
```

In the following set of IPs, we identified the bait as not pulled down.

```{r define-nobait}
(nobait_150 <- baitclass$Gene_name[baitclass$class == "nobait"])
```

We also read a list of known complexes, which will be used for annotation 
purposes.

```{r load-complexes, class.source="fold-show"}
complexes <- jsonlite::read_json(params$complexes, simplifyVector = TRUE)
complexes$`MBF transcription complex`$color <- main_colors[5]
complexes$`CCAAT-binding factor complex`$color <- main_colors[1]
complexes$`Atf1-Pcr1`$color <- main_colors[3]
names(complexes)
```

Finally we load a table classifying each TF to its DNA-binding domain family.

```{r, class.source="fold-show"}
dbd <- read.delim(params$dbdtxt) |>
    dplyr::mutate(Gene_name = .capitalize(Gene_name))
DT::datatable(dbd, extensions = "FixedColumns", rownames = FALSE, 
              filter = list(position = "top", clear = FALSE),
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 10))
```

# Helper functions

We define a number of additional helper functions for data extraction and 
summarization, which will be used below.

```{r heatmap-helper}
filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold, 
                             logfcThreshold, maxNbrTargets) {
    ## Filter - require significant in at least one comparison
    tstatfilt <- tstatmat
    tstatfilt[adjpmat >= adjpThreshold] <- NA
    tstatfilt[logfcmat <= logfcThreshold] <- NA
    # tstatfilt[tstatfilt <= 0] <- NA

    ## Some experiments pull down a large number of proteins
    # print(table(colSums(!is.na(tstatfilt))))
    ## -> remove proteins that are only being pulled down in these experiments
    keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
    tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
    tstatfilt[is.na(tstatfilt)] <- 0
    tstatfilt <- tstatfilt[rowSums(tstatfilt) > 0, ]

    ## Subset the t-statistic matrix to only the retained proteins
    stopifnot(colnames(tstatfilt) == colnames(tstatmat))
    tstatmat <- tstatmat[rownames(tstatfilt), colnames(tstatfilt)]
    colnames(tstatmat) <- .getSimplifiedComparison(colnames(tstatmat))

    list(tstats = tstatmat, tstatsfilt = tstatfilt)
}

makeComplexAnnotation <- function(tstatmat, complexes, idmap,
                                  show_legend = TRUE) {
    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
                           row.names = rownames(tstatmat))
    rowcols <- list(Complex = c())
    for (cmplx in names(complexes)) {
        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
                                                  idmap = idmap) %in% 
                             complexes[[cmplx]]$pombase_ids] <- cmplx
        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
    }
    rowAnnotation(df = rowannot, 
                  col = rowcols, 
                  annotation_name_gp = gpar(fontsize = fl),
                  simple_anno_size_adjust = TRUE,
                  na_col = bg_color,
                  show_annotation_name = FALSE,
                  show_legend = show_legend,
                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
}

makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
                           row.names = rownames(tstatmat))
    rowcols <- list(Complex = c())
    for (cmplx in names(complexes)) {
        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
                                                  idmap = idmap) %in% 
                             complexes[[cmplx]]$pombase_ids] <- cmplx
        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
    }
    rowannot$`DBD fam.` <- dbd$DBD_class[match(
        .getPomBaseIdFromProtein(rownames(tstatmat), idmap = idmap), 
        dbd$PomBaseID)]
    setfam <- c("KilA-N", "Zn(II)2Cys6", "bZIP")
    unsetfam <- setdiff(unique(dbd$DBD_class), setfam)
    rowannot$`DBD fam.`[rowannot$`DBD fam.` %in% unsetfam] <- "Other (combined)"
    rowcols <- c(rowcols, list(`DBD fam.` = c(`KilA-N` = main_colors[5], 
                                              `Zn(II)2Cys6` = main_colors[4],
                                              bZIP = main_colors[3], 
                                              `Other (combined)` = bg_color)))
    rowAnnotation(df = rowannot, 
                  col = rowcols, 
                  annotation_name_gp = gpar(fontsize = fl),
                  simple_anno_size_adjust = TRUE,
                  na_col = bg_color,
                  show_annotation_name = TRUE,
                  show_legend = FALSE,
                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
}

makeColLabels <- function(tstatmat, colLabel, fontsize) {
    columnAnnotation(
        TFnames = anno_mark(which = "column", side = "bottom",
                            at = match(colLabel, colnames(tstatmat)), 
                            labels = .getProteinNameFromSimplifiedComparison(colLabel),
                            labels_gp = gpar(fontsize = fontsize)))
}

makeRowLabels <- function(tstatmat, rowLabel, fontsize) {
    rowAnnotation(
        TFnames = anno_mark(which = "row", side = "left",
                            at = match(rowLabel, rownames(tstatmat)), 
                            labels = rowLabel,
                            labels_gp = gpar(fontsize = fontsize)))
}

makeHeatmapCol <- function(tstatmat, stringency = "high") {
    if (stringency == "high") {
        circlize::colorRamp2(
            breaks = seq(3.0, 13.0, length.out = 64),
            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
    } else {
        circlize::colorRamp2(
            breaks = seq(1.5, 13.0, length.out = 64),
            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
    }
}

reorderColsByRows <- function(tstat) {
    ## Row names should be protein names, column names simplified comparison names
    ord <- match(rownames(tstat), .getProteinNameFromSimplifiedComparison(colnames(tstat)))
    ord <- ord[!is.na(ord)]
    ord <- c(ord, setdiff(seq_len(ncol(tstat)), ord))
    tstat[, ord]
}

makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)

    tstats <- filterForHeatmap(tstatmat = tc$tstat, adjpmat = tc$adjp, 
                               logfcmat = tc$logfc, logfcThreshold = log2fcthr,
                               adjpThreshold = adjpthr, maxNbrTargets = Inf)
    tstat <- tstats$tstats

    ## tstats$tstatsfile: protein x experiment
    ## Number of TFs that are pulled down by each bait
    all_tfs <- setdiff(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
                       c("Untagged", "untagged", NA))
    nbrInt <- colSums(tstats$tstatsfilt[rownames(tstats$tstatsfilt) %in% all_tfs, ] > 0, 
                      na.rm = TRUE)
    names(nbrInt) <- .getSimplifiedComparison(names(nbrInt))

    ## Split by family - return a matrix with several columns: 
    ## #pulled down TFs (including itself), #pulled down TFs (excluding itself),
    ## #pulled down TFs from the same family (excluding itself), 
    ## #pulled down TFs from another family
    tmptf <- tstats$tstatsfilt[match(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
                                     rownames(tstats$tstatsfilt)), ]
    rownames(tmptf) <- .getProteinNameFromComparison(colnames(tstats$tstatsfilt))
    dbdtf <- data.frame(PomBaseID = .getPomBaseIdFromComparison(colnames(tstats$tstatsfilt), 
                                                                idmap = idmap),
                        bait = .getProteinNameFromComparison(colnames(tstats$tstatsfilt)),
                        DBD_class = NA)
    for (i in seq_len(nrow(dbdtf))) {
        lookfor <- dbdtf$PomBaseID[i]
        if (!(lookfor %in% c("untagged", "Untagged"))) {
            dbdtf$DBD_class[i] <- dbd$DBD_class[match(lookfor, dbd$PomBaseID)]
        } else {
            dbdtf$DBD_class[i] <- "no_class"
        }
    }
    stopifnot(dbdtf$bait == .getProteinNameFromComparison(colnames(tmptf)))
    nbrIntMat <- do.call(dplyr::bind_rows, lapply(seq_along(colnames(tmptf)), function(i) {
        expr <- colnames(tmptf)[i]
        fam <- dbdtf$DBD_class[i]
        data.frame(experiment = expr, 
                   bait = dbdtf$bait[i], 
                   dbdfam = fam, 
                   TFs = paste(rownames(tmptf)[-i][which(tmptf[-i, expr] > 0)],
                               collapse = ","),
                   nbrTFs = sum(tmptf[, expr] > 0, na.rm = TRUE),
                   nbrTFs_wobait = sum(tmptf[-i, expr] > 0, na.rm = TRUE),
                   nbrTFs_samefam = sum(tmptf[-i, expr] > 0 & 
                                            dbdtf$DBD_class[-i] == fam, 
                                        na.rm = TRUE),
                   nbrTFs_difffam = sum(tmptf[-i, expr] > 0 & 
                                            dbdtf$DBD_class[-i] != fam, 
                                        na.rm = TRUE)
        )
    }))

    ## Split experiments into three groups
    if (conc == 150) {
        stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
        tstat_no_bait <- 
            tstat[, colSums(tstats$tstatsfilt) == 0 | 
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "nobait"])]
        tstat_150only <- 
            tstat[, colSums(tstats$tstatsfilt) > 0 &
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "150only"])]
        tstat_150and500 <- 
            tstat[, colSums(tstats$tstatsfilt) > 0 &
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "150and500"])]

        proteins_150only <- .getProteinNameFromSimplifiedComparison(colnames(tstat_150only))
        stopifnot(all(proteins_150only %in% rownames(tstat)))
        tstat_150and500 <- tstat_150and500[!(rownames(tstat_150and500) %in% proteins_150only), ]

        ## Cluster proteins based on the correlation between their t-stat vectors
        dst <- sqrt(2 * (1 - cor(t(tstat_150and500), use = "pairwise.complete")))
        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
        clst <- hclust(as.dist(dst))

        ## Reorder proteins by the clustering
        protorder <- c(rownames(tstat_150and500)[clst$order], proteins_150only)
        tstat <- tstat[protorder, ]

        ## Reorder experiments to keep a "diagonal" in the heatmap
        tstat <- reorderColsByRows(tstat)

        tmpbaits <- .getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
                                                             idmap = idmap)
        colsplit <- ifelse(
            tmpbaits %in% baitclass$Gene_name[baitclass$class == "nobait"], 
            "No\nbait", 
            ifelse(
                tmpbaits %in% baitclass$Gene_name[baitclass$class == "150only"],
                "150 mM NaCl",
                ifelse(
                    tmpbaits %in% baitclass$Gene_name[baitclass$class == "150and500"],
                    "150 and\n500 mM NaCl", "NA"
                )
            )
        )
    } else if (conc == 500) {
        tstat_onlybait <- 
            tstat[, colSums(tstats$tstatsfilt > 0) == 1 |
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
                                                      idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class500 == "Lost in 500 mM"])]
        tstat_several <- 
            tstat[, colSums(tstats$tstatsfilt > 0) > 1 & 
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
                                                      idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class500 != "Lost in 500 mM"])]
        tstat_onlybait <- tstat_onlybait[, colnames(tstat_onlybait) != "Untagged_plate"]
        tstat_several <- tstat_several[, colnames(tstat_several) != "Untagged_plate"]

        proteins_onlybait <-
            .getProteinNameFromSimplifiedComparison(colnames(tstat_onlybait))
        stopifnot(all(proteins_onlybait %in% rownames(tstat)))
        tstat_several <- tstat_several[!(rownames(tstat_several) %in% proteins_onlybait), ]

        ## Cluster proteins based on the correlation between their t-stat vectors
        dst <- sqrt(2 * (1 - cor(t(tstat_several), use = "pairwise.complete")))
        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
        clst <- hclust(as.dist(dst))

        ## Reorder proteins by the clustering
        protorder <- c(rownames(tstat_several)[clst$order], proteins_onlybait)
        tstat <- tstat[protorder, ]

        ## Reorder conditions to keep a "diagonal" in the heatmap
        tstat <- reorderColsByRows(tstat)

        colsplit <- factor(
            structure(baitclass$class500[
                match(.getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
                                                               idmap = idmap), 
                      baitclass$Gene_name)],
                names = colnames(tstat)),
            levels = c("Retained in 500 mM", "Lost in 500 mM", ""))
        levels(colsplit) <- c("Interactions\nin 500 mM NaCl",
                              "No interactions\nin 500 mM NaCl", "")
    } else {
        stop("Unknown concentration")
    }
    return(list(tstat = tstat, colsplit = colsplit, nbrIntMat = nbrIntMat))
}

makeBaitHeatmapData <- function(sce, adjpthr, log2fcthr, nbrIntMat) {
    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)$tstat
    colnames(tc) <- .getSimplifiedComparison(colnames(tc))
    shared_names <- intersect(rownames(tc), 
                              .getProteinNameFromSimplifiedComparison(colnames(tc)))
    shared_names <- setdiff(shared_names, .getProteinFromOrigBait(nobait_150, idmap = idmap))
    stopifnot(all(shared_names %in% nbrIntMat$bait))
    nbrIntMat$complex <- NA
    for (cplx in c("MBF transcription complex", "CCAAT-binding factor complex", "Atf1-Pcr1")) {
        nbrIntMat$complex[.getPomBaseIdFromProtein(nbrIntMat$bait,
                                                   idmap = idmap) %in% 
                              complexes[[cplx]]$pombase_ids] <- cplx
    }
    shared_names <- nbrIntMat[match(shared_names, nbrIntMat$bait), ] |>
        arrange(dbdfam, complex) |>
        pull(bait)
    mat <- tc[shared_names, match(shared_names, .getProteinNameFromSimplifiedComparison(colnames(tc)))]
    rownames(nbrIntMat) <- .getSimplifiedComparison(nbrIntMat$experiment)
    list(mat = mat, nbrIntMat = nbrIntMat)
}
```

# Extract data for heatmaps

```{r extract-heatmap-data}
hmdata_150 <- makeHeatmapData(sce = sce150, adjpthr = adjpThreshold, 
                              log2fcthr = log2fcThreshold, conc = 150)
hmdata_150_loose <- makeHeatmapData(sce = sce150, adjpthr = adjpThresholdLoose, 
                                    log2fcthr = log2fcThreshold, conc = 150)
hmdata_500 <- makeHeatmapData(sce = sce500, adjpthr = adjpThreshold, 
                              log2fcthr = log2fcThreshold, conc = 500)
hmdata_500_loose <- makeHeatmapData(sce = sce500, adjpthr = adjpThresholdLoose, 
                                    log2fcthr = log2fcThreshold, conc = 500)

baitdata_150 <- makeBaitHeatmapData(sce = sce150, adjpthr = adjpThreshold, 
                                    log2fcthr = log2fcThreshold, 
                                    nbrIntMat = hmdata_150$nbrIntMat)
baitdata_150_loose <- makeBaitHeatmapData(sce = sce150, adjpthr = adjpThresholdLoose, 
                                          log2fcthr = log2fcThreshold, 
                                          nbrIntMat = hmdata_150_loose$nbrIntMat)
baitdata_500 <- makeBaitHeatmapData(sce = sce500, adjpthr = adjpThreshold, 
                                    log2fcthr = log2fcThreshold, 
                                    nbrIntMat = hmdata_500$nbrIntMat)
baitdata_500_loose <- makeBaitHeatmapData(sce = sce500, adjpthr = adjpThresholdLoose, 
                                          log2fcthr = log2fcThreshold, 
                                          nbrIntMat = hmdata_500_loose$nbrIntMat)
```

# Export data for TFexplorer

Here, we export data for use with TFexplorer, for both the low- and high-salt 
conditions.

```{r tfexplorer-export}
## Heatmap data
.prepDataForTFexplorer <- function(sce, outfileHeatmap, outfileVolcano, idmap) {
    ## Heatmap
    ## -------------------------------------------------------------------------
    testres <- .getTestCols(sce = sce)
    testres$tstat <- testres$tstat[, colnames(testres$tstat) !=
                                       "Untagged_150_tube_vs_compl_Untagged_150_tube.t"]
    tstat <- testres$tstat
    colnames(tstat) <- .getProteinNameFromComparison(colnames(tstat))
    TFs <- colnames(tstat)

    ## set NAs to minimum value
    tstat[is.na(tstat)] <- min(na.omit(tstat))
    ## set all values below 0 to 0
    tstat[tstat < 0] = 0

    ## convert it back to a matrix
    heatmap_data_mean <- as.matrix(tstat)

    ## cluster the rows (to avoid slow clustering in LinkedCharts)
    dist_mat1 <- dist(heatmap_data_mean, method = "euclidian")
    hclust_avg1 <- hclust(dist_mat1, method = "average")
    heatmap_data_mean_clustered <- heatmap_data_mean[hclust_avg1$order, ]

    ## calculate row max 
    heatmapRowMax <- apply(heatmap_data_mean_clustered, 1, max)

    ## add Pombase IDs and names and uniprot IDs 
    PomIds2 <- left_join(data.frame(row.names(heatmap_data_mean_clustered)),
                         idmap |> dplyr::mutate(unique_einprot_id = .capitalize(unique_einprot_id)),
                         by = c("row.names.heatmap_data_mean_clustered." =
                                    "unique_einprot_id"))
    PomIds1 <- ifelse(is.na(PomIds2$gene_stable_id),
                      PomIds2$row.names.heatmap_data_mean_clustered., 
                      PomIds2$gene_stable_id)
    UniprotIDs <- PomIds2$xref

    ## convert the gene names to protein names
    colnames(heatmap_data_mean_clustered) <- 
        .capitalize(colnames(heatmap_data_mean_clustered))
    rownames(heatmap_data_mean_clustered) <- 
        .capitalize(rownames(heatmap_data_mean_clustered))

    ## save heatmap and related data as JSON
    heatmap_data_mean2 <- toJSON(list(
        heatmapMatrix = heatmap_data_mean_clustered, 
        TFs = colnames(heatmap_data_mean_clustered),
        proteins = rownames(heatmap_data_mean_clustered),
        MaxTSTAT = heatmapRowMax,
        PomIds = PomIds1,
        UniprotIDs = UniprotIDs),
        pretty = TRUE)
    write(heatmap_data_mean2, file = outfileHeatmap)

    ## Volcano plots
    ## -------------------------------------------------------------------------
    ## extract comparisons (include only plate format for Atf1 and untagged)
    volcano_list <- metadata(sce)$testres$tests
    volcano_list <- volcano_list[sub("\\.t", "", colnames(testres$tstat))]
    TFIDs <- names(volcano_list)
    TFnames <- .getProteinNameFromComparison(TFIDs)
    # TFnames <- ifelse(grepl("^SP", TFnames), TFnames, stringr::str_to_title(TFnames))

    ## using sce object
    vp_list <- lapply(
        structure(colnames(heatmap_data_mean_clustered), 
                  names = colnames(heatmap_data_mean_clustered)), 
        function(selTF) {
            VP <- volcano_list[[which(TFnames == selTF)]][, c("pid", "logFC", "adj.P.Val",
                                                              "mlog10p", "showInVolcano")]
            VP$DE <- ifelse(VP$showInVolcano, "significant", "not significant")

            ## replace NAs with 0
            VP <- replace_na(data = VP, 
                             replace = list(pid = "unknown",
                                            logFC = 0,
                                            adj.P.Val = 1,
                                            mlog10p = 0,
                                            DE = "not significant"))

            ## add to list
            VP[, c("pid", "logFC", "adj.P.Val", "mlog10p", "DE")]
        })

    ## convert the gene names to protein names
    for (i in seq_along(vp_list)) {
        rownames(vp_list[[i]]) <- .capitalize(rownames(vp_list[[i]]))
        vp_list[[i]]$pid <- rownames(vp_list[[i]])
    }

    ## save volcano plot data as JSON
    vp_list2 <- toJSON(vp_list, dataframe = "columns", pretty = TRUE)
    write(vp_list2, file = outfileVolcano)
}

.prepDataForTFexplorer(sce = sce150, 
                       outfileHeatmap = "data/TSTAT_matrix4heatmap_150mM.json", 
                       outfileVolcano = "data/volcanoData_150mM.json", 
                       idmap = idmap)
.prepDataForTFexplorer(sce = sce500, 
                       outfileHeatmap = "data/TSTAT_matrix4heatmap_500mM.json", 
                       outfileVolcano = "data/volcanoData_500mM.json", 
                       idmap = idmap)
```

# Facts and figures

## Identification of baits in IP-MS

```{r, class.source="fold-show"}
table(baitclass$class)
```

## High stringency

Here we use an adjusted p-value cutoff of `r adjpThreshold` and a log2 
fold change cutoff of `r log2fcThreshold`.

```{r}
makeInteractorSummary <- function(sce, baitdata, adjp_cutoff, logfc_cutoff, idmap, 
                                  baits_to_keep, tbl_title) {
    ni_tc <- .getTestCols(sce = sce, adjp_cutoff = adjp_cutoff, 
                          logfc_cutoff = logfc_cutoff)
    ni_idx <- .getOrigBaitNameFromComparison(colnames(ni_tc$tstat), idmap = idmap) %in%
        baits_to_keep
    ni_int <- ni_tc$interactor[, ni_idx]
    ni_baits <- cbind(match(.getProteinNameFromComparison(colnames(ni_int)), 
                            rownames(ni_int)), 
                      seq_len(ncol(ni_int)))
    ni_int[ni_baits] <- NA

    ## Total number of interactions
    table(ni_int > 0)

    ## Number of TFs with interactions
    table(colSums(ni_int, na.rm = TRUE) > 0)

    ## Total number of TFs
    ncol(ni_int)

    ## Total number of TF-TF interactions (no self-interactions)
    sum(baitdata$nbrIntMat$nbrTFs_wobait)

    ## Number of TFs involved in these TF-TF interactions
    tmp <- baitdata$nbrIntMat |>
        filter(nbrTFs_wobait > 0)
    length(unique(c(tmp$bait, unlist(strsplit(tmp$TFs, ",")))))

    ## Bilateral interactions
    nbr_bilateral <- tmp |>
        select(bait, TFs) |>
        separate_longer_delim(TFs, delim = ",") |>
        mutate(o1 = paste(bait, TFs, sep = "-"),
               o2 = paste(TFs, bait, sep = "-")) |>
        filter(o1 %in% o2) |>
        nrow()

    data.frame(`Total number of TFs for which bait was pulled down` = ncol(ni_int),
               `Total number of interactions (excluding bait itself)` = length(which(ni_int > 0)),
               `Number of TFs with interactions` = length(which(colSums(ni_int, na.rm = TRUE) > 0)),
               `Number of TFs without observed interactions` = length(which(colSums(ni_int, na.rm = TRUE) == 0)),
               `Total number of TF-TF interactions (excluding bait itself)` = sum(baitdata$nbrIntMat$nbrTFs_wobait),
               `Number of TFs involved in these TF-TF interactions` = length(unique(c(tmp$bait, unlist(strsplit(tmp$TFs, ","))))),
               `Number of bilateral TF-TF interaction` = nbr_bilateral,
               check.names = FALSE) |>
        pivot_longer(cols = everything(), names_to = "Aspect", values_to = "Value") |>
        kbl(col.names = NULL, caption = tbl_title) |>
        kable_styling()
}

## 150mM
## --------------------------------------------------------------------------
makeInteractorSummary(sce = sce150, baitdata = baitdata_150, adjp_cutoff = adjpThreshold,
                      logfc_cutoff = log2fcThreshold, idmap = idmap, 
                      baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
                      tbl_title = "150 mM, high stringency")

## 500mM
## --------------------------------------------------------------------------
makeInteractorSummary(sce = sce500, baitdata = baitdata_500, adjp_cutoff = adjpThreshold,
                      logfc_cutoff = log2fcThreshold, idmap = idmap, 
                      baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
                      tbl_title = "500 mM, high stringency")
```

## Low stringency

Here we use an adjusted p-value cutoff of `r adjpThresholdLoose` and a log2 
fold change cutoff of `r log2fcThreshold`.

```{r}
## 150mM
## --------------------------------------------------------------------------
makeInteractorSummary(sce = sce150, baitdata = baitdata_150_loose, adjp_cutoff = adjpThresholdLoose,
                      logfc_cutoff = log2fcThreshold, idmap = idmap, 
                      baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
                      tbl_title = "150 mM, low stringency")

## 500mM
## --------------------------------------------------------------------------
makeInteractorSummary(sce = sce500, baitdata = baitdata_500_loose, adjp_cutoff = adjpThresholdLoose,
                      logfc_cutoff = log2fcThreshold, idmap = idmap, 
                      baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
                      tbl_title = "500 mM, low stringency")
```

# Figure 4

## Settings

```{r plot-settings}
colLabel_150 <- expand.grid(c("Untagged", "Ace2", "Rst2", "Pap1"),
                            c("tube", "plate")) |>
    tidyr::unite(col = "col", Var1, Var2) |>
    dplyr::pull(col)
colLabel_500 <- expand.grid(c("Untagged", "Atf1", "Moc3", "Ntu1",
                              "Pcr1", "Ntu2"),
                            c("tube", "plate")) |>
    tidyr::unite(col = "col", Var1, Var2) |>
    dplyr::pull(col)
rowLabel <- NULL

complexLabel <- c("NuA4", "SAGA")
```

## Summary heatmaps using results from the statistical tests (150 mM NaCl)

### Overall heatmap

```{r heatmap-tstat-150}
#| fig.width: 6
#| fig.height: 8

hm1a <- Heatmap(hmdata_150$tstat, 
                cluster_rows = FALSE, 
                cluster_columns = FALSE, 
                col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
                border = TRUE, 
                border_gp = gpar(lwd = 0.5),
                column_split = hmdata_150$colsplit, 
                name = "Mod.\nt-stat.", 
                na_col = binary_heatmap_colors["FALSE"],
                column_names_gp = gpar(fontsize = fs),
                row_names_gp = gpar(fontsize = fs),
                row_title = "Proteins significantly enriched in at least one experiment",
                row_title_gp = gpar(fontsize = fl),
                show_row_names = FALSE, show_column_names = FALSE,
                use_raster = FALSE, show_heatmap_legend = TRUE,
                right_annotation = makeComplexAnnotation(
                    hmdata_150$tstat, complexes[complexLabel], 
                    idmap = idmap),
                bottom_annotation = makeColLabels(hmdata_150$tstat, colLabel_150, fl),
                width = unit(w_ipmsHm, "mm"), # heatmap body width
                heatmap_height = unit(h_ipmsHm, "mm"), # whole heatmap height
                column_title_gp = gpar(fontsize = ft),
                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                border = "gray10",
                                border_gp = gpar(lwd = 0.5)))

set.seed(42L)
hm_150 <- grid.grabExpr(
    hm1a <- draw(hm1a, merge_legend = TRUE,
                heatmap_legend_side = "right",
                annotation_legend_side = "right"),
    width = w_ipmsHm, height = h_ipmsHm
)

plot_grid(hm_150)
```

### Bait-vs-bait heatmap

```{r heatmap-tstat-150-baits}
#| fig.width: 6
#| fig.height: 5

nbrInteractors <- HeatmapAnnotation(
    which = "column",
    `Number of\ninteractors` = 
        anno_barplot(x = rowSums(as.matrix(baitdata_150$nbrIntMat[colnames(baitdata_150$mat), 
                                                     c("nbrTFs_samefam", 
                                                       "nbrTFs_difffam")])), 
                     gp = gpar(fill = na_color,
                               col = na_color),
                     border = FALSE, bar_width = 1.0,
                     axis_param = list(gp = gpar(fontsize = fs)),
                     width = unit(10, "mm")),
    annotation_name_side = "left",
    annotation_name_rot = 0,
    annotation_name_gp = gpar(fontsize = fl))

hm1b <- Heatmap(baitdata_150$mat, 
                cluster_rows = FALSE, 
                cluster_columns = FALSE, 
                col = makeHeatmapCol(baitdata_150$mat, stringency = "high"), 
                border = TRUE, 
                border_gp = gpar(lwd = 0.5),
                name = "Mod.\nt-stat.", 
                na_col = binary_heatmap_colors["FALSE"],
                column_title = paste(c("150 mM NaCl", rep(" ", 45)), collapse = ""),
                row_title = "Transcription factors",
                column_title_gp = gpar(fontsize = fl + 3, fontface = "bold"),
                row_title_gp = gpar(fontsize = fl),
                column_names_gp = gpar(fontsize = fs),
                row_names_gp = gpar(fontsize = fs),
                show_row_names = FALSE, show_column_names = FALSE,
                use_raster = FALSE, show_heatmap_legend = TRUE, 
                left_annotation = makeComplexAndDBDAnnotation(
                    baitdata_150$mat, 
                    complexes[c("MBF transcription complex", 
                                "CCAAT-binding factor complex",
                                "Atf1-Pcr1")],
                    dbd = dbd, 
                    idmap = idmap),
                top_annotation = nbrInteractors, 
                width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
                heatmap_height = unit(h_ipmsHm / 2.2, "mm"),
                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                            border = "gray10",
                                            border_gp = gpar(lwd = 0.5)))

set.seed(42L)
hm_150_bait <- grid.grabExpr(
    hm1b <- draw(hm1b, merge_legend = TRUE,
                 heatmap_legend_side = "right",
                 annotation_legend_side = "right"),
    width = w_ipmsHm / 1.05, height = h_ipmsHm / 2.2
)

plot_grid(hm_150_bait)
```

```{r bait-bait-heatmap-legends}
#| fig.height: 2

## Separate legends
use_complexes <- c("MBF transcription complex", 
                   "CCAAT-binding factor complex",
                   "Atf1-Pcr1")
lgd_complex <- Legend(labels = use_complexes,
                legend_gp = gpar(fill = vapply(complexes[use_complexes],
                                               function(x) x$color, NA_character_)),
                title = "Complex", nrow = 2)

dbd_colors <- c(`KilA-N` = main_colors[5], 
                `Zn(II)2Cys6` = main_colors[4],
                bZIP = main_colors[3], 
                `Other (combined)` = bg_color)
lgd_dbd <- Legend(labels = names(dbd_colors), 
                 legend_gp = gpar(fill = dbd_colors), 
                 title = "DBD family", nrow = 2)

legend_bait_1 <- grid.grabExpr(
    hm1b <- ComplexHeatmap::draw(lgd_complex),
    width = w_ipmsHm, height = h_ipmsHm / 8
)
legend_bait_2 <- grid.grabExpr(
    hm1b <- ComplexHeatmap::draw(lgd_dbd),
    width = w_ipmsHm, height = h_ipmsHm / 8
)
```

### Bait-vs-bait, less stringent adjusted p-value threshold

```{r heatmap-tstat-150-baits-0.05}
#| fig.width: 6
#| fig.height: 5

nbrInteractors <- HeatmapAnnotation(
    which = "column",
    `Number of\ninteractors` = 
        anno_barplot(x = rowSums(as.matrix(baitdata_150_loose$nbrIntMat[colnames(baitdata_150_loose$mat), 
                                                     c("nbrTFs_samefam", 
                                                       "nbrTFs_difffam")])), 
                     gp = gpar(fill = na_color,
                               col = na_color),
                     border = FALSE, bar_width = 1.0,
                     axis_param = list(gp = gpar(fontsize = fs)),
                     width = unit(10, "mm")),
    annotation_name_side = "left",
    annotation_name_rot = 0,
    annotation_name_gp = gpar(fontsize = fl))

hm1b_loose <- Heatmap(baitdata_150_loose$mat, 
                      cluster_rows = FALSE, 
                      cluster_columns = FALSE, 
                      col = makeHeatmapCol(baitdata_150_loose$mat, stringency = "low"), 
                      border = TRUE, 
                      border_gp = gpar(lwd = 0.5),
                      name = "Mod.\nt-stat.", 
                      na_col = binary_heatmap_colors["FALSE"],
                      column_title = paste(c("150 mM NaCl", rep(" ", 45)), collapse = ""),
                      row_title = "Transcription factors",
                      column_title_gp = gpar(fontsize = fl + 3, fontface = "bold"),
                      row_title_gp = gpar(fontsize = fl),
                      column_names_gp = gpar(fontsize = fs),
                      row_names_gp = gpar(fontsize = fs),
                      show_row_names = FALSE, show_column_names = FALSE,
                      use_raster = FALSE, show_heatmap_legend = TRUE, 
                      left_annotation = makeComplexAndDBDAnnotation(
                          baitdata_150_loose$mat, 
                          complexes[c("MBF transcription complex", 
                                      "CCAAT-binding factor complex",
                                      "Atf1-Pcr1")],
                          dbd = dbd, 
                          idmap = idmap),
                      top_annotation = nbrInteractors, 
                      width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
                      heatmap_height = unit(h_ipmsHm / 2.2, "mm"), 
                      heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                                  border = "gray10",
                                                  border_gp = gpar(lwd = 0.5)))

hm_150_0.05_bait <- grid.grabExpr(
    hm1b <- draw(hm1b_loose, merge_legend = TRUE,
                 heatmap_legend_side = "right",
                 annotation_legend_side = "right"),
    width = w_ipmsHm / 1.05, height = h_ipmsHm / 2.2
)

plot_grid(hm_150_0.05_bait)
```

### SAGA/NuA4 heatmap

```{r heatmap-tstat-saga-nua4}
#| fig.width: 6
#| fig.height: 8

tc150 <- .getTestCols(sce150, adjp_cutoff = 0.05,
                      logfc_cutoff = log2fcThreshold)
keep_proteins <- .capitalize(c(complexes$SAGA$gene_names, complexes$NuA4$gene_names))
keep_exp <- which(.getProteinNameFromComparison(colnames(tc150$adjp)) %in% 
                      c("Ace2", "Pap1", "Rst2"))
binary_mat <- tc150$adjp[keep_proteins, keep_exp] < 0.05 &
    tc150$logfc[keep_proteins, keep_exp] > 1
binary_mat[is.na(binary_mat)] <- FALSE
colnames(binary_mat) <- .getProteinNameFromComparison(colnames(binary_mat))
ordr <- order(rownames(binary_mat) %in% .capitalize(complexes$SAGA$gene_names), 
              binary_mat[, "Ace2"], binary_mat[, "Pap1"], binary_mat[, "Rst2"], 
              decreasing = TRUE)
binary_mat <- binary_mat[ordr, ]

binary_annot <- data.frame(NuA4 = rownames(binary_mat) %in% 
                               .capitalize(complexes$NuA4$gene_names),
                           SAGA = rownames(binary_mat) %in% 
                               .capitalize(complexes$SAGA$gene_names), 
                           row.names = rownames(binary_mat))

mode(binary_mat) <- "character"
ht_opt(HEATMAP_LEGEND_PADDING = unit(0, "mm"))
hm1saga_nua4 <- Heatmap(
    binary_mat, 
    cluster_rows = FALSE, 
    cluster_columns = FALSE, 
    col = binary_heatmap_colors, 
    border = TRUE, 
    border_gp = gpar(lwd = 0.5),
    name = " ", 
    na_col = binary_heatmap_colors["FALSE"],
    column_names_gp = gpar(fontsize = ft),
    row_names_gp = gpar(fontsize = fl),
    show_row_names = TRUE, show_column_names = TRUE, 
    use_raster = FALSE, show_heatmap_legend = TRUE,
    right_annotation = rowAnnotation(
        df = binary_annot, 
        col = list(SAGA = c(`TRUE` = complex_colors[["SAGA"]],
                            `FALSE` = binary_heatmap_colors[["FALSE"]]),
                   NuA4 = c(`TRUE` = complex_colors[["NuA4"]],
                            `FALSE` = binary_heatmap_colors[["FALSE"]])),
        show_legend = FALSE),
    heatmap_height = unit(h_ipmsHm * 0.945, "mm"), # whole heatmap height
    column_title_gp = gpar(fontsize = ft),
    heatmap_legend_param = list(title_gp = gpar(fontsize = ft),
                                at = c("FALSE", "TRUE"),
                                labels = c("", "adj. p < 0.05 &\nlog2FC > 1"), 
                                legend_gap = unit(0, "mm"))
)

hm_150_saga_nua4 <- grid.grabExpr(
    hm1sn <- draw(hm1saga_nua4, merge_legend = TRUE,
                  heatmap_legend_side = "bottom",
                  annotation_legend_side = "right",
                  align_heatmap_legend = "global_center",
                  padding = unit(c(3, 1, 1, 1), "mm")),
    width = w_ipmsHm, height = h_ipmsHm
)

plot_grid(hm_150_saga_nua4)
ht_opt(RESET = TRUE)
```

### RNA / DNA binding proteins

```{r heatmap-tstat-150-rnabinding}
#| fig.width: 6
#| fig.height: 5
#| eval: true
#| echo: true

rnabinding <- read.delim(params$rnabinders)
keep <- rownames(hmdata_150$tstat) %in% c(rnabinding$PomBaseID, .capitalize(rnabinding$Gene_name))
sum(keep)
hm1rna <- Heatmap(hmdata_150$tstat[keep, ], 
                  cluster_rows = FALSE, 
                  cluster_columns = TRUE, 
                  col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
                  border = TRUE, 
                  border_gp = gpar(lwd = 0.5),
                  name = "Mod.\nt-stat.", 
                  na_col = binary_heatmap_colors["FALSE"],
                  column_names_gp = gpar(fontsize = fs),
                  row_names_gp = gpar(fontsize = fs),
                  show_row_names = TRUE, show_column_names = FALSE,
                  use_raster = FALSE, show_heatmap_legend = TRUE,
                  # bottom_annotation = makeColLabels(tstat150[keep, ], colLabel_150, fl),
                  #left_annotation = makeRowLabels(tstat150[keep, ], rowLabel, fl),
                  width = unit(w_ipmsHm, "mm"), # heatmap body width
                  heatmap_height = unit(h_ipmsHm/2, "mm"), # whole heatmap height
                  heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                border = "gray10",
                                border_gp = gpar(lwd = 0.5)))

set.seed(42L)
hm_150_rna <- grid.grabExpr(
    hm1b <- draw(hm1rna, merge_legend = TRUE,
                heatmap_legend_side = "right",
                annotation_legend_side = "right"),
    width = w_ipmsHm, height = h_ipmsHm/2
)

plot_grid(hm_150_rna)
```

```{r heatmap-tstat-150-dnabinding}
#| fig.width: 6
#| fig.height: 13
#| eval: true
#| echo: true

dnabinding <- read.delim(params$dnabinders)
keep <- rownames(hmdata_150$tstat) %in% c(dnabinding$PomBaseID, .capitalize(dnabinding$Gene_name))
sum(keep)
hm1dna <- Heatmap(hmdata_150$tstat[keep, ], 
                  cluster_rows = FALSE, 
                  cluster_columns = TRUE, 
                  col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
                  border = TRUE, 
                  border_gp = gpar(lwd = 0.5),
                  name = "Mod.\nt-stat.", 
                  na_col = binary_heatmap_colors["FALSE"],
                  column_names_gp = gpar(fontsize = fs),
                  row_names_gp = gpar(fontsize = fs),
                  show_row_names = TRUE, show_column_names = FALSE,
                  use_raster = FALSE, show_heatmap_legend = TRUE,
                  # bottom_annotation = makeColLabels(tstat150[keep, ], colLabel_150, fl),
                  #left_annotation = makeRowLabels(tstat150[keep, ], rowLabel, fl),
                  width = unit(w_ipmsHm, "mm"), # heatmap body width
                  heatmap_height = unit(1.5 * h_ipmsHm, "mm"), # whole heatmap height
                  heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                border = "gray10",
                                border_gp = gpar(lwd = 0.5)))

set.seed(42L)
hm_150_dna <- grid.grabExpr(
    hm1b2 <- draw(hm1dna, merge_legend = TRUE,
                  heatmap_legend_side = "right",
                  annotation_legend_side = "right"),
    width = w_ipmsHm, height = 1.5 * h_ipmsHm
)
plot_grid(hm_150_dna)
```

## Summary heatmaps using results from the statistical tests (500 mM NaCl)

### Overall heatmap

```{r heatmap-tstat-500}
#| fig.width: 6
#| fig.height: 8

hm2a <- Heatmap(hmdata_500$tstat, 
                cluster_rows = FALSE, 
                cluster_columns = FALSE, 
                col = makeHeatmapCol(hmdata_500$tstat, stringency = "high"), 
                border = TRUE, 
                border_gp = gpar(lwd = 0.5),
                column_split = hmdata_500$colsplit, 
                name = "Mod.\nt-stat.", 
                na_col = binary_heatmap_colors["FALSE"],
                column_names_gp = gpar(fontsize = fs),
                row_names_gp = gpar(fontsize = fs),
                row_title = "Proteins significantly enriched in at least one experiment",
                row_title_gp = gpar(fontsize = fl),
                show_row_names = FALSE, show_column_names = FALSE,
                use_raster = FALSE, show_heatmap_legend = TRUE,
                right_annotation = makeComplexAnnotation(hmdata_500$tstat, complexes[complexLabel],
                                                         idmap = idmap, show_legend = FALSE),
                bottom_annotation = makeColLabels(hmdata_500$tstat, colLabel_500, fl),
                width = unit(w_ipmsHm, "mm"), # heatmap body width
                heatmap_height = unit(h_ipmsHm, "mm"), # whole heatmap height
                column_title_gp = gpar(fontsize = ft),
                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                border = "gray10",
                                border_gp = gpar(lwd = 0.5)))

hm_500 <- grid.grabExpr(
    hm2a <- draw(hm2a, merge_legend = TRUE,
                 heatmap_legend_side = "right",
                 annotation_legend_side = "right"),
    width = w_ipmsHm, height = h_ipmsHm
)

plot_grid(hm_500)
```

### Bait-vs-bait heatmap

```{r heatmap-tstat-500-baits}
#| fig.width: 6
#| fig.height: 5

nbrInteractors <- HeatmapAnnotation(
    which = "column",
    `Number of\ninteractors` = 
        anno_barplot(x = rowSums(as.matrix(baitdata_500$nbrIntMat[colnames(baitdata_500$mat), 
                                                     c("nbrTFs_samefam", 
                                                       "nbrTFs_difffam")])), 
                     gp = gpar(fill = na_color,
                               col = na_color),
                     border = FALSE, bar_width = 1.0,
                     axis_param = list(gp = gpar(fontsize = fs)),
                     width = unit(10, "mm")),
    annotation_name_side = "left",
    annotation_name_rot = 0,
    annotation_name_gp = gpar(fontsize = fl))

hm1b2 <- Heatmap(baitdata_500$mat, 
                 cluster_rows = FALSE, 
                 cluster_columns = FALSE, 
                 col = makeHeatmapCol(baitdata_500$mat, stringency = "high"), 
                 border = TRUE, 
                 border_gp = gpar(lwd = 0.5),
                 name = "Mod.\nt-stat.", 
                 na_col = binary_heatmap_colors["FALSE"],
                 column_title = paste(c("500 mM NaCl", rep(" ", 45)), collapse = ""),
                 row_title = "Transcription factors",
                 column_title_gp = gpar(fontsize = fl + 3, fontface = "bold"),
                 row_title_gp = gpar(fontsize = fl),
                 column_names_gp = gpar(fontsize = fs),
                 row_names_gp = gpar(fontsize = fs),
                 show_row_names = FALSE, show_column_names = FALSE,
                 use_raster = FALSE, show_heatmap_legend = TRUE,
                 left_annotation = makeComplexAndDBDAnnotation(
                     baitdata_500$mat, 
                     complexes[c("MBF transcription complex", 
                                 "CCAAT-binding factor complex", "Atf1-Pcr1")], 
                     dbd = dbd, idmap = idmap),
                 top_annotation = nbrInteractors, 
                 width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
                 heatmap_height = unit(h_ipmsHm / 2.2, "mm"), # whole heatmap height
                 heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                            border = "gray10",
                                            border_gp = gpar(lwd = 0.5)))

hm_500_bait <- grid.grabExpr(
    hm1b2 <- draw(hm1b2, merge_legend = TRUE,
                  heatmap_legend_side = "right",
                  annotation_legend_side = "right"),
    width = w_ipmsHm / 1.05, height = h_ipmsHm / 2.2
)

plot_grid(hm_500_bait)
```

## Compare t-statistics from low-salt and high-salt conditions

The figures below show scatter plots of moderated t-statistics for selected 
bait pull-downs compared to their complements, in the high-salt vs low-salt 
conditions.

```{r tstat-150-vs-500}
#| fig.width: 6
#| fig.height: 7
#| message: false

## Extract results from statistical tests
rdlow <- rowData(sce150)
rdhigh <- rowData(sce500)

## Get t-statistics
contrasts_to_compare <- grep("\\.t$", colnames(rdhigh), value = TRUE)
contrasts_low <- gsub("_compl", "_highsaltcompl", 
                      gsub("_500_", "_150_", contrasts_to_compare))
idx <- which(!contrasts_low %in% colnames(rdlow))
contrasts_low[idx] <- gsub("_plate", "_tube", contrasts_low[idx])
stopifnot(all(contrasts_low %in% colnames(rdlow)))

## t-statistics, logFCs and adjusted p-values, low salt
low <- as.data.frame(rdlow[, c("einprotId", contrasts_low)]) |>
    tidyr::pivot_longer(names_to = "contrast_low", values_to = "tstat_low", -einprotId) |>
    dplyr::mutate(contrast_low = sub("\\.t$", "", contrast_low)) |>
    dplyr::left_join(
        as.data.frame(rdlow[, c("einprotId", sub("\\.t$", ".adj.P.Val",
                                                 contrasts_low))]) |>
            tidyr::pivot_longer(names_to = "contrast_low", 
                                values_to = "adjp_low", -einprotId) |>
            dplyr::mutate(contrast_low = sub("\\.adj.P.Val$", "", contrast_low))) |>
    dplyr::left_join(
        as.data.frame(rdlow[, c("einprotId", sub("\\.t$", ".logFC",
                                                 contrasts_low))]) |>
            tidyr::pivot_longer(names_to = "contrast_low", 
                                values_to = "logfc_low", -einprotId) |>
            dplyr::mutate(contrast_low = sub("\\.logFC$", "", contrast_low)))

## t-statistics, logFCs and adjusted p-values, high salt
high <- as.data.frame(rdhigh[, c("einprotId", contrasts_to_compare)]) |>
    tidyr::pivot_longer(names_to = "contrast_high", 
                        values_to = "tstat_high", -einprotId) |>
    dplyr::mutate(contrast_high = sub("\\.t$", "", contrast_high)) |>
    dplyr::left_join(
        as.data.frame(rdhigh[, c("einprotId", sub("\\.t$", ".adj.P.Val",
                                                  contrasts_to_compare))]) |>
            tidyr::pivot_longer(names_to = "contrast_high", 
                                values_to = "adjp_high", -einprotId) |>
            dplyr::mutate(contrast_high = sub("\\.adj.P.Val$", "", 
                                              contrast_high))) |>
    dplyr::left_join(
        as.data.frame(rdhigh[, c("einprotId", sub("\\.t$", ".logFC",
                                                  contrasts_to_compare))]) |>
            tidyr::pivot_longer(names_to = "contrast_high", 
                                values_to = "logfc_high", -einprotId) |>
            dplyr::mutate(contrast_high = sub("\\.logFC$", "", 
                                              contrast_high))) |>
    dplyr::mutate(contrast_low = sub("\\.t$", "", 
                                     contrasts_low[match(paste0(contrast_high, ".t"),
                                                         contrasts_to_compare)]))

## Merge data.frames
lowhigh <- low |> dplyr::full_join(high) |>
    dplyr::mutate(tstat_low = replace(tstat_low, is.na(tstat_low), 0),
                  tstat_high = replace(tstat_high, is.na(tstat_high), 0),
                  logfc_low = replace(logfc_low, is.na(logfc_low), 0),
                  logfc_high = replace(logfc_high, is.na(logfc_high), 0),
                  adjp_low = replace(adjp_low, is.na(adjp_low), 1),
                  adjp_high = replace(adjp_high, is.na(adjp_high), 1)) |>
    dplyr::mutate(bait = .getProteinNameFromComparison(contrast_low)) |>
    dplyr::mutate(splitvar = .getProteinNameFromComparison(contrast_low))

## Add dal2
rddal2 <- rowData(scedal2)
contrasts_low_dal2 <- c("Dal2_150_plate_vs_Untagged_150_plate.t")
low_dal2 <- as.data.frame(rddal2[, c("einprotId", contrasts_low_dal2)]) |>
    tidyr::pivot_longer(names_to = "contrast_low", 
                        values_to = "tstat_low", -einprotId) |>
    dplyr::mutate(contrast_low = sub("\\.t$", "", contrast_low)) |>
    dplyr::left_join(
        as.data.frame(rddal2[, c("einprotId", sub("\\.t$", ".adj.P.Val",
                                                  contrasts_low_dal2))]) |>
            tidyr::pivot_longer(names_to = "contrast_low", 
                                values_to = "adjp_low", -einprotId) |>
            dplyr::mutate(contrast_low = sub("\\.adj.P.Val$", "", contrast_low))) |>
    dplyr::left_join(
        as.data.frame(rddal2[, c("einprotId", sub("\\.t$", ".logFC",
                                                  contrasts_low_dal2))]) |>
            tidyr::pivot_longer(names_to = "contrast_low", 
                                values_to = "logfc_low", -einprotId) |>
            dplyr::mutate(contrast_low = sub("\\.logFC$", "", contrast_low)))

contrasts_high_dal2 <- c("Dal2_500_plate_vs_Untagged_500_plate.t")
high_dal2 <- as.data.frame(rddal2[, c("einprotId", contrasts_high_dal2)]) |>
    tidyr::pivot_longer(names_to = "contrast_high", 
                        values_to = "tstat_high", -einprotId) |>
    dplyr::mutate(contrast_high = sub("\\.t$", "", contrast_high)) |>
    dplyr::left_join(
        as.data.frame(rddal2[, c("einprotId", sub("\\.t$", ".adj.P.Val",
                                                  contrasts_high_dal2))]) |>
            tidyr::pivot_longer(names_to = "contrast_high", 
                                values_to = "adjp_high", -einprotId) |>
            dplyr::mutate(contrast_high = sub("\\.adj.P.Val$", "", contrast_high))) |>
    dplyr::left_join(
        as.data.frame(rddal2[, c("einprotId", sub("\\.t$", ".logFC",
                                                  contrasts_high_dal2))]) |>
            tidyr::pivot_longer(names_to = "contrast_high", 
                                values_to = "logfc_high", -einprotId) |>
            dplyr::mutate(contrast_high = sub("\\.logFC$", "", contrast_high)))

lowhigh_dal2 <- low_dal2 |> dplyr::full_join(high_dal2) |>
    dplyr::mutate(tstat_low = replace(tstat_low, is.na(tstat_low), 0),
                  tstat_high = replace(tstat_high, is.na(tstat_high), 0),
                  logfc_low = replace(logfc_low, is.na(logfc_low), 0),
                  logfc_high = replace(logfc_high, is.na(logfc_high), 0),
                  adjp_low = replace(adjp_low, is.na(adjp_low), 1),
                  adjp_high = replace(adjp_high, is.na(adjp_high), 1)) |>
    dplyr::mutate(bait = .getProteinNameFromComparison(sub("SPB4775_", "", 
                                                           contrast_low))) |>
    dplyr::mutate(splitvar = .getProteinNameFromComparison(sub("SPB4775_", "",
                                                               contrast_low)))

lowhigh <- lowhigh |>
    dplyr::bind_rows(lowhigh_dal2)

## Subset 1
adjp150vs500 <- 1e-3
logfc150vs500 <- 1

lowhighsub1 <- lowhigh |> 
    dplyr::filter(splitvar %in% c("Ace2", "Rst2", "Pap1")) |>
    dplyr::mutate(splitvar = factor(splitvar, levels = c("Ace2", "Rst2", "Pap1"))) |>
    dplyr::mutate(significant = (adjp_high < adjp150vs500 & logfc_high > logfc150vs500)) |>
    dplyr::mutate(complex = ifelse(einprotId %in% .capitalize(complexes$NuA4$gene_names), 
                                   "NuA4", ifelse(einprotId %in% .capitalize(complexes$SAGA$gene_names),
                                                  "SAGA", NA_character_)))

## Subset 2
lowhighsub2 <- lowhigh |> 
    dplyr::filter(splitvar %in% c("Atf1", "Pcr1", "Moc3", 
                                  "Dal2", "Ntu1", "Ntu2")) |>
    dplyr::mutate(splitvar = factor(splitvar, levels = c("Atf1", "Pcr1", "Moc3", 
                                                         "Dal2", "Ntu1", "Ntu2"))) |>
    dplyr::mutate(significant = (adjp_high < adjp150vs500 & logfc_high > logfc150vs500))

(gg_retain1 <- ggplot(lowhighsub1, 
                      aes(x = tstat_high, y = tstat_low)) +
        geom_abline(data = lowhighsub1 |>
                        dplyr::filter(einprotId == bait), 
                    aes(slope = tstat_low / tstat_high, intercept = 0), 
                    linetype = "dashed", color = "grey50") + 
        geom_text_repel(data = lowhighsub1 |>
                                dplyr::mutate(
                                    einprotId = ifelse(significant, einprotId, "")),
                        aes(label = einprotId), min.segment.length = 0,
                        max.overlaps = Inf) + 
        geom_point(aes(alpha = !is.na(complex), color = complex), 
                   show.legend = c(alpha = FALSE, color = TRUE)) + 
        scale_color_manual(values = complex_colors, na.value = "black",
                           breaks = c("SAGA", "NuA4")) + 
        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.25)) + 
        geom_point(data = lowhighsub1 |> dplyr::filter(einprotId == bait) |>
                       dplyr::mutate(type = "Bait"), 
                   mapping = aes(fill = type), color = main_colors[3], 
                   shape = 21, alpha = 1, size = 2) + 
        scale_fill_manual(values = c(Bait = main_colors[3])) + 
        facet_wrap(~ splitvar, scales = "free", ncol = 1, 
                   labeller = labeller(splitvar = structure(
                       paste0(unique(lowhighsub1$splitvar), " IP-MS"), 
                       names = as.character(unique(lowhighsub1$splitvar))))) + 
        labs(x = "500 mM NaCl\nmod. t-stat.", y = "150 mM NaCl mod. t-stat.") + 
        theme_cowplot(ft) + 
        theme(legend.position = "bottom", 
              strip.background = element_rect(fill = NA, colour = NA), 
              strip.text = element_text(hjust = 0, face = "bold")) + 
        guides(color = guide_legend(nrow = 1, byrow = TRUE, title = ""),
               fill = guide_legend(nrow = 1, byrow = TRUE, title = "")))

(gg_retain2 <- ggplot(lowhighsub2, 
                      aes(x = tstat_high, y = tstat_low)) +
        geom_abline(data = lowhighsub2 |> 
                        dplyr::filter(einprotId == bait), 
                    aes(slope = tstat_low / tstat_high, intercept = 0), 
                    linetype = "dashed", color = "grey50") + 
        geom_text_repel(data = lowhighsub2 |>
                            dplyr::mutate(einprotId = ifelse(significant, einprotId, "")),
                        aes(label = einprotId), min.segment.length = 0,
                        max.overlaps = Inf) + 
        geom_point(aes(alpha = significant, color = significant, size = significant), 
                   show.legend = c(alpha = FALSE, color = TRUE, size = FALSE)) + 
        scale_color_manual(
            values = c(`TRUE` = lighten(main_colors[4], 0.2), `FALSE` = "black"), 
            labels = c(`TRUE` = paste0("adj. p < ", adjp150vs500, 
                                       " and logFC > ", logfc150vs500, " (500 mM NaCl)"),
                       `FALSE` = paste0("adj. p >= ", adjp150vs500, 
                                        " or logFC <= ", logfc150vs500, " (500 mM NaCl)"))) + 
        scale_size_manual(values = c(`TRUE` = 2, `FALSE` = 1)) +
        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.25)) + 
        geom_point(data = lowhighsub2 |> 
                       dplyr::filter(einprotId == bait) |>
                       dplyr::mutate(type = "Bait"), 
                   mapping = aes(fill = type), color = main_colors[3], 
                   shape = 21, alpha = 1, size = 2) + 
        scale_fill_manual(values = c(Bait = main_colors[3])) + 
        facet_wrap(~ splitvar, scales = "free", ncol = 2, 
                   labeller = labeller(splitvar = structure(
                       paste0(unique(lowhighsub2$splitvar), " IP-MS"), 
                       names = as.character(unique(lowhighsub2$splitvar))))) + 
        labs(x = "500 mM NaCl mod. t-stat.", y = "150 mM NaCl mod. t-stat.") + 
        theme_cowplot(ft) + 
        theme(legend.position = "bottom", 
              strip.background = element_rect(fill = NA, colour = NA), 
              strip.text = element_text(hjust = 0, face = "bold")) + 
        guides(color = guide_legend(nrow = 2, byrow = TRUE, title = ""),
               fill = guide_legend(nrow = 1, byrow = TRUE, title = "")))
```

## Supplementary plots - PCA

```{r}
## Low salt
sce150_pca <- einprot::doPCA(
    sce150[rowSums(!assay(sce150, 
                          metadata(sce150)$aNames$assayImputIndic)) >= 2, ],
    metadata(sce150)$aNames$assayImputed)$sce
df <- scuttle::makePerCellDF(sce150_pca, features = NULL, use.coldata = TRUE, 
                             use.dimred = TRUE) |>
    dplyr::mutate(Protocol = .capitalize(sub("_digest|_freeze", "", sub("SOP_", "", Protocol))))
pcvar150 <- attr(reducedDim(sce150_pca, "PCA_log2_LFQ.intensity"), "percentVar")
(ggpca1 <- ggplot(
    df, 
    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
        color = Protocol)) + 
    geom_point(size = 3, alpha = 0.5) +
    scale_color_manual(values = c(Plate = main_colors[1], 
                                  Tube = main_colors[5]), name = "") + 
    coord_fixed() + 
    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), 
         title = "150 mM NaCl") + 
    theme_cowplot() + 
    theme(legend.position = "bottom") + 
    guides(color = guide_legend(nrow = 1, byrow = TRUE)))
(ggpca2 <- ggplot(
    df |>
        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")) |>
        dplyr::mutate(prot = .capitalize(prot)), 
    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
        color = bait, shape = prot)) + 
    geom_point(size = 3, alpha = 0.5) +
    scale_color_manual(values = c(Atf1 = main_colors[3], 
                                  `Other baits` = na_color), name = "") + 
    scale_shape_manual(values = c(Plate = 16, Tube = 17), name = "") + 
    coord_fixed() + 
    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), title = "150 mM NaCl") + 
    theme_cowplot() + 
    theme(legend.position = "bottom", legend.box = "horizontal", 
          legend.margin = margin()))

sce500_pca <- einprot::doPCA(
    sce500[rowSums(!assay(sce500, metadata(sce500)$aNames$assayImputIndic)) >= 2, ],
    metadata(sce500)$aNames$assayImputed)$sce
df <- scuttle::makePerCellDF(sce500_pca, features = NULL, use.coldata = TRUE, 
                             use.dimred = TRUE)
pcvar500 <- attr(reducedDim(sce500_pca, "PCA_log2_LFQ.intensity"), "percentVar")
(ggpca3 <- ggplot(
    df |>
        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")), 
    aes(x = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".1")]], 
        y = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".2")]], 
        color = bait)) + 
    geom_point(size = 3, alpha = 0.5) +
    scale_color_manual(values = c(Atf1 = main_colors[3], 
                                  `Other baits` = na_color), name = "") + 
    coord_fixed() + 
    labs(x = paste0("PC1 (", round(pcvar500[1], 1), "%)"), 
         y = paste0("PC2 (", round(pcvar500[2], 1), "%)"), 
         title = "500 mM NaCl") + 
    theme_cowplot() + 
    theme(legend.position = "bottom"))
```

## Supplementary plot - tube vs plate t-statistics for Atf1 pull-down

The low-salt experiments were done using two different protocols (`tube` or 
`plate`). The `Atf1` pull-down was performed with both protocols, and the 
figure below compares the moderated t-statistics (each obtained by comparing 
to the complement consisting of all non-`Atf1` pull-down samples obtained 
with the same protocol).

```{r}
df <- as.data.frame(rowData(sce150)) |>
    dplyr::select(einprotId, Atf1_150_tube_vs_compl_Atf1_150_tube.t, 
                  Atf1_150_plate_vs_compl_Atf1_150_plate.t) |>
    dplyr::mutate(col = ifelse(einprotId == "Atf1", "bait", 
                               ifelse(einprotId == "Pcr1", "Pcr1", "other")))
(ggatf1 <- ggplot(df, aes(x = Atf1_150_tube_vs_compl_Atf1_150_tube.t, 
                         y = Atf1_150_plate_vs_compl_Atf1_150_plate.t)) + 
    geom_abline(slope = 1, intercept = 0) + 
    geom_point(size = 3, aes(color = col, alpha = col)) + 
    scale_color_manual(values = c(bait = main_colors[3], pcr1 = main_colors[4], 
                                  other = "black")) + 
    scale_alpha_manual(values = c(bait = 1, pcr1 = 1, other = 0.5)) + 
    geom_text_repel(data = df |> filter(einprotId %in% c("Atf1", "Pcr1")), 
                    aes(label = einprotId)) + 
    labs(x = "Mod. t-stat. Atf1 vs complement\n(tube, 150 mM NaCl)", 
         y = "Mod. t-stat. Atf1 vs complement\n(plate, 150 mM NaCl)") + 
    coord_fixed() + 
    theme_cowplot() + 
    theme(legend.position = "none"))
```

## Put together

```{r Fig4}
#| fig.width: 9.8
#| fig.height: 14.5

cowplot::plot_grid(
    hm_150,
    plot_grid(gg_retain1, NULL, hm_150_saga_nua4, nrow = 1, rel_widths = c(1, 0.05, 0.75)),
    hm_500,
    gg_retain2,
    nrow = 2,
    labels = c("A", "C", "B", "D"),
    align = "v",
    axis = "b"
)
```

## Supplementary Figure

```{r suppFig4}
#| fig.width: 10.8
#| fig.height: 14.8
#| warning: false

cowplot::plot_grid(
    wrap_plots(ggpca1, ggpca2, ggatf1, ggpca3, hm_150_bait, hm_500_bait, ncol = 2) + 
        plot_annotation(tag_levels = list(c("A", "", "", "", "B"))) & 
        theme(plot.tag = element_text(size = 14, face = "bold")),
    cowplot::plot_grid(
        NULL,
        legend_bait_1, 
        legend_bait_2,
        NULL,
        nrow = 1,
        rel_widths = c(1, 2, 2, 1)
    ),
    ncol = 1,
    rel_heights = c(12, 1)
)
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-07-04 ms_figure_7_heterodimer.Rmd
rename figs2-6 to figs3-7
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

--- 2024-06-21 ms_figure_6_heterodimer.Rmd
@@
 
+get_xlabs_bar <- function(x) {
+    vapply(x, function(xx) sub("\n", "<br>", paste0("*", xx, "*", "<br>(n = ", 
+                                                    length(which(dist_ne$Condition == xx)), ")"), fixed = TRUE), "")
+}
+
 (p_zone <- ggplot(zone_avg, aes(fill = Zone, y = 100 * avg_fraction_by_zone, 
@@
      geom_bar(position = "stack", stat = "identity") +
-     labs(title = "Relative zone distribution\nincl. both replicates", 
+     geom_text(aes(label = paste0(round(100 * avg_fraction_by_zone, 2), "%")), 
+                   position = position_stack(vjust = 0.5), size = 4) +
+     labs(title = "Relative zone distribution incl. both replicates", 
           x = "", y = "Relative distribution [%]") +
@@
                        name = "") + 
-     scale_x_discrete(labels = get_xlabs) + 
+     scale_x_discrete(labels = get_xlabs_bar) + 
      theme_cowplot() + 
@@
         NULL,
-        p_rel_dis,
+        p_zone + theme(legend.position = "top"), 
         labels = c("E", "", "F", "", "G"),
@@
 #| fig.width: 13
-#| fig.height: 10
+#| fig.height: 9
 #| warning: false
@@
         NULL,
-        p_zone + theme(legend.position = "top"), 
+        p_rel_dis,
         nrow = 1, 
-        rel_widths = c(1.75, 0.25, 1.25), 
+        rel_widths = c(1.55, 0.25, 1.45), 
         labels = c("A", "", "B")
@@
     ),
-    ncol = 1, rel_heights = c(1, 0.7)
+    ncol = 1, rel_heights = c(1, 1)
 )

### 2024-07-04 ms_figure_2_overview-heatmaps.Rmd
add new fig 2
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "Figure 2 - Overview heatmaps for IP-MS and ChIP"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    complexes: "data/complexes.json"
    idmap: "data/id_mapping_table.txt"
    baitclass: "data/ipms_bait_class.txt"
    dbdtxt: "data/TF_DBD_annotation.txt"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
options(scipen = 6)
```

# Parameter values

```{r params}
params
```

# Load required packages and helper functions

```{r load-packages, class.source="fold-show"}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(ggplot2)
    library(cowplot)
    library(dplyr)
    library(forcats)
    library(jsonlite)
    library(ComplexHeatmap)
    library(colorspace)
    library(circlize)
    library(scattermore)
    # library(ggrepel)
    # library(DT)
    # library(stringr)
    library(tidyr)
    # library(kableExtra)
    # library(patchwork)
})

## Source scripts with required helper functions and settings
source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/ipms_params.R")
source("params/mapping_functions.R")

## Heatmap sizes
w_ipmsHm <-  84 # heatmap body width (mm)
h_ipmsHm <- 178 # total heatmap height (mm)
w_peaksHm <-  84 # heatmap body width (mm)
h_peaksHm <- 178 # total heatmap height (mm)

fs <- 8  # small font size
fm <- 9  # medium font size
fl <- 10 # large font size
ft <- 12 # title font size
```

# Read data

## IP-MS

We load two `SingleCellExperiment` objects, containing data and results from the 
low- and high-salt experiments, respectively.

```{r load-data-ipms, class.source="fold-show"}
sce150 <- readRDS(params$rds150)
sce500 <- readRDS(params$rds500)
```

We then extract the full list of baits, as well as a classification according 
to the experiment(s) where a certain bait was pulled down.

```{r, class.source="fold-show"}
idmap <- read.delim(params$idmap)
baitclass <- read.delim(params$baitclass)
```

In the following set of IPs, we identified the bait as not pulled down.

```{r define-nobait}
(nobait_150 <- baitclass$Gene_name[baitclass$class == "nobait"])
```

We also read a list of known complexes, which will be used for annotation 
purposes.

```{r load-complexes, class.source="fold-show"}
complexes <- jsonlite::read_json(params$complexes, simplifyVector = TRUE)
complexes$`MBF transcription complex`$color <- main_colors[5]
complexes$`CCAAT-binding factor complex`$color <- main_colors[1]
complexes$`Atf1-Pcr1`$color <- main_colors[3]
names(complexes)
```

Finally we load a table classifying each TF to its DNA-binding domain family.

```{r, class.source="fold-show"}
dbd <- read.delim(params$dbdtxt) |>
    dplyr::mutate(Gene_name = .capitalize(Gene_name))
```

## ChIP

We start by reading all tables and objects needed for the ChIP part of the figure.
Count tables are in addition normalized, and we also calculate
averages over replicates or all samples where needed:

```{r load-data-chip}
# `peakgr`: ChIP-seq peaks (as a GRanges object)
peakgr <- as(read.csv(params$peakcsv, row.names = 1), "GRanges")
peakgr$peaktype <- factor(peakgr$peaktype,
                          levels = c("common peaks (ubiquitous)",
                                     "common peaks (frequent)",
                                     "specific peaks"))

# `peakenr`: peak IP-enrichments log2(IP/input)
peakenr <- as.matrix(read.csv(params$peakenr, row.names = 1))

# ... average replicates (`peakenrAvg`)
grp <- sub("_rep[12]$", "", colnames(peakenr))
peakenrAvg <- do.call(cbind, lapply(split(colnames(peakenr), grp)[unique(grp)],
                                    function(j) {
                                        rowMeans(peakenr[, j, drop = FALSE])
                                    }))

# `atac_raw`: public ATAC-seq read counts in ChIP-seq peaks from this study
atac_raw <- as.matrix(read.csv(params$peakatacfile, row.names = 1))

# ... calculate reads per million (`atac_cpm`) and
#     per kilobase and million (`atac_rpkm`)
#     and average experiments (`atac_cpm_avg`, `atac_rpkm_avg`)
atac_cpm <- sweep(x = atac_raw[, -1], MARGIN = 2,
                  STATS = colSums(atac_raw[, -1]), FUN = "/") * 1e6
atac_rpkm <- atac_cpm / atac_raw[, "width"] * 1e3
atac_cpm_avg <- rowMeans(atac_cpm)
atac_rpkm_avg <- rowMeans(atac_rpkm)

# `chip_raw`: public ChIP-seq read counts in ChIP-seq peaks from this study
chip_raw <- as.matrix(read.csv(params$peakchipfile, row.names = 1))

# ... calculate reads per million (`chip_cpm`) and
#     per kilobase and million (`chip_rpkm`)
#     and calculate enrichments (`chip_enr`, log2 IP/input) averaged over replicates
chip_cpm <- sweep(x = chip_raw[, -1], MARGIN = 2,
                  STATS = colSums(chip_raw[, -1]), FUN = "/") * 1e6
chip_rpkm <- chip_cpm / chip_raw[, "width"] * 1e3
chip_series <- sub("^ChIP_([^_]+)_.+$", "\\1", colnames(chip_cpm))
chip_enr <- do.call(
    cbind,
    lapply(split(colnames(chip_cpm), chip_series),
           function(nms) {
               grp <- sub("ChIP_(GSE[0-9]+)_GSM[0-9]+_(.+)_rep[0-9]$",
                          "\\1_\\2", nms)
               stopifnot(any(is_input <- grep("_Input", grp)))
               enr <- do.call(
                   cbind,
                   lapply(unique(grp[-is_input]), function(grp1) {
                       rowMeans(log2(chip_cpm[, nms[grp == grp1], drop = FALSE] + 1)) -
                           rowMeans(log2(chip_cpm[, nms[is_input], drop = FALSE] + 1))
                   }))
               colnames(enr) <- unique(grp[-is_input])
               enr
           }))

# `genegr`: chromosomal ranges of genes (as GRanges object)
genegr <- rtracklayer::import(params$gtf, feature.type = "gene")
```

# Helper functions

We define a number of additional helper functions for data extraction and 
summarization, which will be used below.

```{r heatmap-helper}
filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold, 
                             logfcThreshold, maxNbrTargets) {
    ## Filter - require significant in at least one comparison
    tstatfilt <- tstatmat
    tstatfilt[adjpmat >= adjpThreshold] <- NA
    tstatfilt[logfcmat <= logfcThreshold] <- NA
    # tstatfilt[tstatfilt <= 0] <- NA

    ## Some experiments pull down a large number of proteins
    # print(table(colSums(!is.na(tstatfilt))))
    ## -> remove proteins that are only being pulled down in these experiments
    keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
    tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
    tstatfilt[is.na(tstatfilt)] <- 0
    tstatfilt <- tstatfilt[rowSums(tstatfilt) > 0, ]

    ## Subset the t-statistic matrix to only the retained proteins
    stopifnot(colnames(tstatfilt) == colnames(tstatmat))
    tstatmat <- tstatmat[rownames(tstatfilt), colnames(tstatfilt)]
    colnames(tstatmat) <- .getSimplifiedComparison(colnames(tstatmat))

    list(tstats = tstatmat, tstatsfilt = tstatfilt)
}

makeComplexAnnotation <- function(tstatmat, complexes, idmap,
                                  show_legend = TRUE) {
    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
                           row.names = rownames(tstatmat))
    rowcols <- list(Complex = c())
    for (cmplx in names(complexes)) {
        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
                                                  idmap = idmap) %in% 
                             complexes[[cmplx]]$pombase_ids] <- cmplx
        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
    }
    rowAnnotation(df = rowannot, 
                  col = rowcols, 
                  annotation_name_gp = gpar(fontsize = fl),
                  simple_anno_size_adjust = TRUE,
                  na_col = bg_color,
                  show_annotation_name = FALSE,
                  show_legend = show_legend,
                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
}

makeColLabels <- function(tstatmat, colLabel, fontsize) {
    columnAnnotation(
        TFnames = anno_mark(which = "column", side = "bottom",
                            at = match(colLabel, colnames(tstatmat)), 
                            labels = .getProteinNameFromSimplifiedComparison(colLabel),
                            labels_gp = gpar(fontsize = fontsize)))
}

makeHeatmapCol <- function(tstatmat, stringency = "high") {
    if (stringency == "high") {
        circlize::colorRamp2(
            breaks = seq(3.0, 13.0, length.out = 64),
            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
    } else {
        circlize::colorRamp2(
            breaks = seq(1.5, 13.0, length.out = 64),
            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
    }
}

reorderColsByRows <- function(tstat) {
    ## Row names should be protein names, column names simplified comparison names
    ord <- match(rownames(tstat), .getProteinNameFromSimplifiedComparison(colnames(tstat)))
    ord <- ord[!is.na(ord)]
    ord <- c(ord, setdiff(seq_len(ncol(tstat)), ord))
    tstat[, ord]
}

makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)

    tstats <- filterForHeatmap(tstatmat = tc$tstat, adjpmat = tc$adjp, 
                               logfcmat = tc$logfc, logfcThreshold = log2fcthr,
                               adjpThreshold = adjpthr, maxNbrTargets = Inf)
    tstat <- tstats$tstats

    ## tstats$tstatsfile: protein x experiment
    ## Number of TFs that are pulled down by each bait
    all_tfs <- setdiff(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
                       c("Untagged", "untagged", NA))
    nbrInt <- colSums(tstats$tstatsfilt[rownames(tstats$tstatsfilt) %in% all_tfs, ] > 0, 
                      na.rm = TRUE)
    names(nbrInt) <- .getSimplifiedComparison(names(nbrInt))

    ## Split by family - return a matrix with several columns: 
    ## #pulled down TFs (including itself), #pulled down TFs (excluding itself),
    ## #pulled down TFs from the same family (excluding itself), 
    ## #pulled down TFs from another family
    tmptf <- tstats$tstatsfilt[match(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
                                     rownames(tstats$tstatsfilt)), ]
    rownames(tmptf) <- .getProteinNameFromComparison(colnames(tstats$tstatsfilt))
    dbdtf <- data.frame(PomBaseID = .getPomBaseIdFromComparison(colnames(tstats$tstatsfilt), 
                                                                idmap = idmap),
                        bait = .getProteinNameFromComparison(colnames(tstats$tstatsfilt)),
                        DBD_class = NA)
    for (i in seq_len(nrow(dbdtf))) {
        lookfor <- dbdtf$PomBaseID[i]
        if (!(lookfor %in% c("untagged", "Untagged"))) {
            dbdtf$DBD_class[i] <- dbd$DBD_class[match(lookfor, dbd$PomBaseID)]
        } else {
            dbdtf$DBD_class[i] <- "no_class"
        }
    }
    stopifnot(dbdtf$bait == .getProteinNameFromComparison(colnames(tmptf)))
    nbrIntMat <- do.call(dplyr::bind_rows, lapply(seq_along(colnames(tmptf)), function(i) {
        expr <- colnames(tmptf)[i]
        fam <- dbdtf$DBD_class[i]
        data.frame(experiment = expr, 
                   bait = dbdtf$bait[i], 
                   dbdfam = fam, 
                   TFs = paste(rownames(tmptf)[-i][which(tmptf[-i, expr] > 0)],
                               collapse = ","),
                   nbrTFs = sum(tmptf[, expr] > 0, na.rm = TRUE),
                   nbrTFs_wobait = sum(tmptf[-i, expr] > 0, na.rm = TRUE),
                   nbrTFs_samefam = sum(tmptf[-i, expr] > 0 & 
                                            dbdtf$DBD_class[-i] == fam, 
                                        na.rm = TRUE),
                   nbrTFs_difffam = sum(tmptf[-i, expr] > 0 & 
                                            dbdtf$DBD_class[-i] != fam, 
                                        na.rm = TRUE)
        )
    }))

    ## Split experiments into three groups
    if (conc == 150) {
        stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
        tstat_no_bait <- 
            tstat[, colSums(tstats$tstatsfilt) == 0 | 
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "nobait"])]
        tstat_150only <- 
            tstat[, colSums(tstats$tstatsfilt) > 0 &
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "150only"])]
        tstat_150and500 <- 
            tstat[, colSums(tstats$tstatsfilt) > 0 &
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "150and500"])]

        proteins_150only <- .getProteinNameFromSimplifiedComparison(colnames(tstat_150only))
        stopifnot(all(proteins_150only %in% rownames(tstat)))
        tstat_150and500 <- tstat_150and500[!(rownames(tstat_150and500) %in% proteins_150only), ]

        ## Cluster proteins based on the correlation between their t-stat vectors
        dst <- sqrt(2 * (1 - cor(t(tstat_150and500), use = "pairwise.complete")))
        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
        clst <- hclust(as.dist(dst))

        ## Reorder proteins by the clustering
        protorder <- c(rownames(tstat_150and500)[clst$order], proteins_150only)
        tstat <- tstat[protorder, ]

        ## Reorder experiments to keep a "diagonal" in the heatmap
        tstat <- reorderColsByRows(tstat)

        tmpbaits <- .getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
                                                             idmap = idmap)
        colsplit <- ifelse(
            tmpbaits %in% baitclass$Gene_name[baitclass$class == "nobait"], 
            "No\nbait", 
            ifelse(
                tmpbaits %in% baitclass$Gene_name[baitclass$class == "150only"],
                "150 mM NaCl",
                ifelse(
                    tmpbaits %in% baitclass$Gene_name[baitclass$class == "150and500"],
                    "150 and\n500 mM NaCl", "NA"
                )
            )
        )
    } else if (conc == 500) {
        tstat_onlybait <- 
            tstat[, colSums(tstats$tstatsfilt > 0) == 1 |
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
                                                      idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class500 == "Lost in 500 mM"])]
        tstat_several <- 
            tstat[, colSums(tstats$tstatsfilt > 0) > 1 & 
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
                                                      idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class500 != "Lost in 500 mM"])]
        tstat_onlybait <- tstat_onlybait[, colnames(tstat_onlybait) != "Untagged_plate"]
        tstat_several <- tstat_several[, colnames(tstat_several) != "Untagged_plate"]

        proteins_onlybait <-
            .getProteinNameFromSimplifiedComparison(colnames(tstat_onlybait))
        stopifnot(all(proteins_onlybait %in% rownames(tstat)))
        tstat_several <- tstat_several[!(rownames(tstat_several) %in% proteins_onlybait), ]

        ## Cluster proteins based on the correlation between their t-stat vectors
        dst <- sqrt(2 * (1 - cor(t(tstat_several), use = "pairwise.complete")))
        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
        clst <- hclust(as.dist(dst))

        ## Reorder proteins by the clustering
        protorder <- c(rownames(tstat_several)[clst$order], proteins_onlybait)
        tstat <- tstat[protorder, ]

        ## Reorder conditions to keep a "diagonal" in the heatmap
        tstat <- reorderColsByRows(tstat)

        colsplit <- factor(
            structure(baitclass$class500[
                match(.getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
                                                               idmap = idmap), 
                      baitclass$Gene_name)],
                names = colnames(tstat)),
            levels = c("Retained in 500 mM", "Lost in 500 mM", ""))
        levels(colsplit) <- c("Interactions\nin 500 mM NaCl",
                              "No interactions\nin 500 mM NaCl", "")
    } else {
        stop("Unknown concentration")
    }
    return(list(tstat = tstat, colsplit = colsplit, nbrIntMat = nbrIntMat))
}

# for each region in `from`, calculate the distance to its nearest element in `to`
# if there are no element in `to` on the sequence of `from`, set distance to NA
calcDistanceToNearest <- function(from, to) {
    stopifnot(exprs = {
        is(from, "GRanges")
        is(to, "GRanges")
    })
    tmp <- distanceToNearest(x = from, subject = to)
    d <- rep(NA, length(from))
    d[queryHits(tmp)] <- mcols(tmp)$distance
    return(d)
}
```

# Extract data for IP-MS heatmaps

```{r extract-heatmap-data-ipms}
hmdata_150 <- makeHeatmapData(sce = sce150, adjpthr = adjpThreshold, 
                              log2fcthr = log2fcThreshold, conc = 150)
hmdata_500 <- makeHeatmapData(sce = sce500, adjpthr = adjpThreshold, 
                              log2fcthr = log2fcThreshold, conc = 500)
```

# Calculate peak distances to tRNA or rRNA genes

We want to identify ChIP-seq peaks that are near a tRNA or rRNA gene. We can
obtain the coordinates of these genes from `genegr` and the coordinates of
peaks from `peakgr`, and calculate the distance between nearest peak-gene pairs
using `distanceToNearest()`. Any pair with a distance below `maxdist` will be
classified as "near" in the plots below.

Remark: Nearest distances can be `NA` in cases where for example a peak
resides on a sequence (chromosome) that does not contain any tRNA or rRNA gene,
or vice versa, such as the sequence `AB325691` (which contains gap-filling
sequence between SPBPB21E7.09 and SPBPB10D8.01 in chromosome 2) or the
mitochondrial sequence `MT`.

```{r calc-peak-dists}
# distance less than `maxdist` are defined as "near"
maxdist <- 100

# tRNA
# ... PomBase and Ensembl_Fungi annotate 196 and 183 tRNAs, respectively
#     we combine the two and obtain 198 annotated tRNAs
table(genegr$gene_biotype == "tRNA", genegr$source)
is_tRNA_pombase <- genegr$source == "PomBase" & genegr$gene_biotype == "tRNA"
is_tRNA_ensembl <- genegr$source == "Ensembl_Fungi" & genegr$gene_biotype == "tRNA"
is_tRNA <- is_tRNA_pombase | (is_tRNA_ensembl & !genegr %in% genegr[is_tRNA_pombase])
sum(is_tRNA)
# ... now we measure distances between nearest pairs
#     either from peak to nearest tRNA
dist.peak2tRNA <- calcDistanceToNearest(from = peakgr, to = genegr[is_tRNA])
#     or from tRNA to nearest peak
dist.tRNA2peak <- calcDistanceToNearest(from = genegr[is_tRNA], to = peakgr)

# rRNA
# ... all 36 5S_rRNA genes in our annotation stem from Ensembl_Fungi
table(genegr$gene_name == "5S_rRNA", genegr$source)
is_rRNA <- !is.na(genegr$gene_name) & genegr$gene_name == "5S_rRNA"
sum(is_rRNA)
# ... now we measure distances between nearest pairs
#     either from peak to nearest rRNA
dist.peak2rRNA <- calcDistanceToNearest(from = peakgr, to = genegr[is_rRNA])
#     or from rRNA to nearest peak
dist.rRNA2peak <- calcDistanceToNearest(from = genegr[is_rRNA], to = peakgr)
```

# Figure 2

## Settings

```{r plot-settings}
colLabel_150 <- expand.grid(c("Untagged", "Ace2", "Rst2", "Pap1"),
                            c("tube", "plate")) |>
    tidyr::unite(col = "col", Var1, Var2) |>
    dplyr::pull(col)
colLabel_500 <- expand.grid(c("Untagged", "Atf1", "Moc3", "Ntu1",
                              "Pcr1", "Ntu2"),
                            c("tube", "plate")) |>
    tidyr::unite(col = "col", Var1, Var2) |>
    dplyr::pull(col)
rowLabel <- NULL

complexLabel <- c("NuA4", "SAGA")
```

## Summary heatmaps using results from the statistical tests (150 mM NaCl)

### Overall heatmap

```{r heatmap-tstat-150}
#| fig.width: 6
#| fig.height: 8

hm1a <- Heatmap(hmdata_150$tstat, 
                cluster_rows = FALSE, 
                cluster_columns = FALSE, 
                col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
                border = TRUE, 
                border_gp = gpar(lwd = 0.5),
                column_split = hmdata_150$colsplit, 
                name = "Mod.\nt-stat.", 
                na_col = binary_heatmap_colors["FALSE"],
                column_names_gp = gpar(fontsize = fs),
                row_names_gp = gpar(fontsize = fs),
                row_title = "Proteins significantly enriched in at least one experiment",
                row_title_gp = gpar(fontsize = fl),
                show_row_names = FALSE, show_column_names = FALSE,
                use_raster = FALSE, show_heatmap_legend = TRUE,
                right_annotation = makeComplexAnnotation(
                    hmdata_150$tstat, complexes[complexLabel], 
                    idmap = idmap),
                bottom_annotation = makeColLabels(hmdata_150$tstat, colLabel_150, fl),
                width = unit(w_ipmsHm, "mm"), # heatmap body width
                heatmap_height = unit(h_ipmsHm, "mm"), # whole heatmap height
                column_title_gp = gpar(fontsize = ft),
                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                border = "gray10",
                                border_gp = gpar(lwd = 0.5)))

set.seed(42L)
hm_150 <- grid.grabExpr(
    hm1a <- draw(hm1a, merge_legend = TRUE,
                heatmap_legend_side = "right",
                annotation_legend_side = "right"),
    width = w_ipmsHm, height = h_ipmsHm
)

plot_grid(hm_150)
```

## Summary heatmaps using results from the statistical tests (500 mM NaCl)

### Overall heatmap

```{r heatmap-tstat-500}
#| fig.width: 6
#| fig.height: 8

hm2a <- Heatmap(hmdata_500$tstat, 
                cluster_rows = FALSE, 
                cluster_columns = FALSE, 
                col = makeHeatmapCol(hmdata_500$tstat, stringency = "high"), 
                border = TRUE, 
                border_gp = gpar(lwd = 0.5),
                column_split = hmdata_500$colsplit, 
                name = "Mod.\nt-stat.", 
                na_col = binary_heatmap_colors["FALSE"],
                column_names_gp = gpar(fontsize = fs),
                row_names_gp = gpar(fontsize = fs),
                row_title = "Proteins significantly enriched in at least one experiment",
                row_title_gp = gpar(fontsize = fl),
                show_row_names = FALSE, show_column_names = FALSE,
                use_raster = FALSE, show_heatmap_legend = TRUE,
                right_annotation = makeComplexAnnotation(hmdata_500$tstat, complexes[complexLabel],
                                                         idmap = idmap, show_legend = FALSE),
                bottom_annotation = makeColLabels(hmdata_500$tstat, colLabel_500, fl),
                width = unit(w_ipmsHm, "mm"), # heatmap body width
                heatmap_height = unit(h_ipmsHm, "mm"), # whole heatmap height
                column_title_gp = gpar(fontsize = ft),
                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
                                border = "gray10",
                                border_gp = gpar(lwd = 0.5)))

hm_500 <- grid.grabExpr(
    hm2a <- draw(hm2a, merge_legend = TRUE,
                 heatmap_legend_side = "right",
                 annotation_legend_side = "right"),
    width = w_ipmsHm, height = h_ipmsHm
)

plot_grid(hm_500)
```

## Peak vs. sample heatmap (binary)

The following creates heatmap of peaks (rows) versus ChIP-seq experiments
(columns). The values are binary (`TRUE` or `FALSE`) and indicate if a peak
was enriched in ChIP-seq experiment (had an average log2 IP/input enrichment
values greater than 1.0).

```{r plot-peaksBin}
#| fig.width: 6
#| fig.height: 8

# extract binary enrichment matrix from `peakgr`
m <- as.matrix(mcols(peakgr)[, grep("^is_enr_in[.]", colnames(mcols(peakgr)))])
mode(m) <- "numeric"
colnames(m) <- sub("is_enr_in.", "", colnames(m))

# specify ChIP-seq experiments for which to show labels
selLabels <- c("Untagged")
stopifnot(all(selLabels %in% colnames(m)))

# only show peaks that are enriched in at least one ChIP-seq experiment
sel_peaksBin <- rowSums(m) > 0

# prepare annotation data
#   `fractTFs`: fraction of ChIP-seq experiments that a peak is enriched in
#   `numpeaks`: the number of enriched peaks for each ChIP-seq experiment
#   `peaktype`: the type of each peak (specific, frequent or ubiquitous)
#   `near_ncRNA`: specifies if the peak is nearer than `maxdist` from
#                 a %S rRNA or tRNA gene
fracTFs <- rowMeans(m > 0)
numpeaks <- colSums(m > 0)
levels(peakgr$peaktype) <- c("Common (ubiquitous)", # use shorter names
                             "Common (frequent)",
                             "Specific peaks")
peaktype <- factor(gsub("[()]", "", sub(" ", "\n", peakgr$peaktype)),
                   levels = c("Common\nubiquitous",
                              "Common\nfrequent",
                              "Specific\npeaks"))
near_ncRNA <- factor(ifelse(!is.na(dist.peak2tRNA) & dist.peak2tRNA < maxdist,
                            "tRNA",
                            ifelse(!is.na(dist.peak2rRNA) & dist.peak2rRNA < maxdist,
                                   "5S rRNA", "none")),
                         levels = c("tRNA", "5S rRNA", "none"))

# most tRNA and 5S rRNA genes are near "common (ubiquitous)" peaks:
table(peaktype[sel_peaksBin], near_ncRNA[sel_peaksBin])

# parameters for annotation legends and colors
annotLegendParams <- list(`%GC` = list(at = c(20, 40, 60),
                                       labels_gp = gpar(fontsize = fs),
                                       border = "gray10",
                                       legend_direction = "horizontal",
                                       legend_width = unit(18, "mm"),
                                       title_position = "topcenter",
                                       title_gp = gpar(fontsize = fl)),
                          `TSS dist.` = list(at = c(0, 4000, 8000),
                                             labels_gp = gpar(fontsize = fs),
                                             border = "gray10",
                                             legend_direction = "horizontal",
                                             legend_width = unit(18, "mm"),
                                             title_position = "topcenter",
                                             title_gp = gpar(fontsize = fl)),
                          `ncRNA` = list(labels_gp = gpar(fontsize = fs),
                                         title_position = "topcenter",
                                         title_gp = gpar(fontsize = fl)))
cols_ncRNA <- c("tRNA" = main_colors[5], "5S rRNA" = main_colors[3],
                "none" = bg_color)
annotCols <- list(
    `%GC` = colorRamp2(breaks = seq(min(peakgr$fracGC[sel_peaksBin]),
                                    max(peakgr$fracGC[sel_peaksBin]),
                                    length.out = 64) * 100,
                       colors = rev(hcl.colors(64, "Greens"))),
    ` width` = colorRamp2(breaks = seq(min(width(peakgr)[sel_peaksBin]),
                                       max(width(peakgr)[sel_peaksBin]),
                                       length.out = 64),
                          colors = rev(hcl.colors(64, "YlOrBr"))),
    `TSS dist.` = colorRamp2(breaks = c(0, 10^seq(2, log10(max(peakgr$nrst_TSS_dist)),
                                                  length.out = 63)),
                             colors = hcl.colors(64)),
    `ncRNA` = cols_ncRNA)

# prepare heatmap annotations
#   `peakAnnot` for peaks (rows), left side
#   `fracTFsAnnot` for peaks (rows), right side
#   `sampleAnnot` for ChIP-seq experiments (columns), top
#   `sampleLabels` for ChIP-seq experiments (columns), bottom
peakAnnot <- HeatmapAnnotation(
                  which = "row",
                  `%GC` = peakgr$fracGC[sel_peaksBin] * 100,
                  `TSS dist.` = peakgr$nrst_TSS_dist[sel_peaksBin],
                  `ncRNA` = near_ncRNA[sel_peaksBin],
                  col = annotCols,
                  width = unit(21, "mm"), show_legend = TRUE,
                  annotation_name_gp = gpar(fontsize = fl),
                  annotation_legend_param = annotLegendParams)

fracTFsAnnot <- HeatmapAnnotation(
    which = "row",
    `Fraction\nof TFs` = anno_barplot(
        x = fracTFs[sel_peaksBin], gp = gpar(col = na_color),
        border = FALSE, bar_width = 1.0,
        axis_param = list(at = c(0, 0.4, 0.8),
                          gp = gpar(fontsize = fs)),
        width = unit(10, "mm")),
    annotation_name_gp = gpar(fontsize = fl))
sampleAnnot <- HeatmapAnnotation(
    which = "column",
    `Number of\nenriched\npeaks` = anno_barplot(
        x = numpeaks,
        gp = gpar(col = 0, fill = na_color),
        border = FALSE, bar_width = 1.0,
        axis_param = list(at = c(0, 150, 300),
                          gp = gpar(fontsize = fs),
                          facing = "outside",
                          direction = "normal"),
        width = unit(10, "mm")),
    annotation_name_gp = gpar(fontsize = fl),
    annotation_name_side = "left", annotation_name_rot = 0,
    show_legend = TRUE)
sampleLabels <- columnAnnotation(
    TFnames = anno_mark(which = "column", side = "bottom",
                        at = match(selLabels, colnames(m)), 
                        labels = sub("^[^_]+_", "", selLabels),
                        labels_gp = gpar(fontsize = fl)))

# create main heatmap with annotations
hm3 <- Heatmap(m[sel_peaksBin, ],
               col = circlize::colorRamp2(
                   breaks = seq(0, max(m), length.out = 64),
                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
               left_annotation = peakAnnot,
               right_annotation = fracTFsAnnot,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height

# draw the heatmap and grab as a grid graphics `grob`
# (needed for combining figure panels)
# use defined random number seed to make clustering deterministic
set.seed(42L)
hm_peaksBin <- grid.grabExpr(
    hm3 <- draw(hm3, merge_legend = TRUE,
                heatmap_legend_side = "bottom",
                annotation_legend_side = "bottom"),
    width = w_peaksHm, height = h_peaksHm
)

plot_grid(hm_peaksBin)
```

## Peak vs. sample heatmap (enrichments)

The following creates a heatmap similar to the peak-versus-experiment
heatmap above, but using the ChIP-seq enrichment values (log2 IP/input)
directly instead of binarized (`TRUE`, `FALSE`) values.

```{r plot-peaksEnr}
#| fig.width: 6
#| fig.height: 8

# use ChIP-seq enrichment matrix from this study (replicates averaged)
m <- peakenrAvg
stopifnot(all(selLabels %in% colnames(m)))

# to keep the same row order as in the binary heatmap `hm3`,
# we have to re-calculate the annotations and switch off the row clustering
sel_peaksEnr <- which(sel_peaksBin)[unlist(row_order(hm3), use.names = FALSE)]

# prepare annotation data
#   `chip_enr_sel`: selected public ChIP-seq experiments
#                   (log2 IP/input enrichments)
#   `chip_enr_sel_range`: value range for color scale (1% to 99%)
#   `chip_enr_sel_range_full`: value range for color scale (0% to 100%)
chip_enr_sel <- chip_enr[sel_peaksEnr, ]
chip_enr_sel_range <- quantile(abs(chip_enr_sel), probs = c(0.95)) * c(-1, 1)
chip_enr_sel_range_full <- range(chip_enr_sel)

# parameters for annotation legends and colors
annotCols2 <- list(
    `ncRNA` = cols_ncRNA,
    H3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                               seq(chip_enr_sel_range[1],
                                   chip_enr_sel_range[2],
                                   length.out = 62),
                               chip_enr_sel_range_full[2]),
                    colors = viridisLite::magma(64)),
    H3K14ac = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                                    seq(chip_enr_sel_range[1],
                                        chip_enr_sel_range[2],
                                        length.out = 62),
                                    chip_enr_sel_range_full[2]),
                         colors = viridisLite::magma(64)),
    H3K9me2 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                                    seq(chip_enr_sel_range[1],
                                        chip_enr_sel_range[2],
                                        length.out = 62),
                                    chip_enr_sel_range_full[2]),
                         colors = viridisLite::magma(64)),
    H3K9me3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
                                    seq(chip_enr_sel_range[1],
                                        chip_enr_sel_range[2],
                                        length.out = 62),
                                    chip_enr_sel_range_full[2]),
                         colors = viridisLite::magma(64)))
annotLegendParams2 <- list(
    `ncRNA` = list(labels_gp = gpar(fontsize = fs),
                   title_position = "topcenter",
                   title_gp = gpar(fontsize = fl)),
    H3 = list(at = c(-6, 0, 6),
              title = "Histone log2 IP/input",
              labels_gp = gpar(fontsize = fs),
              border = "gray10",
              legend_direction = "horizontal",
              legend_width = unit(30, "mm"),
              title_position = "topcenter",
              title_gp = gpar(fontsize = fl)))

# prepare heatmap annotations
#   `peakAnnot` for peaks (rows), left side
#   `peakAnnot2` for peaks (rows), right side
peakAnnot <- HeatmapAnnotation(
                  which = "row",
                  `ncRNA` = near_ncRNA[sel_peaksEnr],
                  H3 = chip_enr_sel[, "GSE108668_H3"],
                  H3K14ac = chip_enr_sel[, "GSE108668_H3K14ac"],
                  H3K9me2 = chip_enr_sel[, "GSE182250_H3K9me2"],
                  H3K9me3 = chip_enr_sel[, "GSE182250_H3K9me3"],
                  col = annotCols2,
                  width = unit(28, "mm"),
                  annotation_name_gp = gpar(fontsize = fl),
                  annotation_legend_param = annotLegendParams2,
                  show_legend = c(TRUE, TRUE, FALSE, FALSE, FALSE))
atac_trunc <- pmin(atac_rpkm_avg[sel_peaksEnr], 2000) # cap ATAC signal at 2000
peakAnnot2 <- HeatmapAnnotation(
    which = "row",
    `Accessibility\n1e3 RPKM` =
        anno_barplot(x = atac_trunc / 1e3,
                     gp = gpar(col = na_color),
                     border = FALSE, bar_width = 1.0,
                     baseline = 0,
                     axis_param = list(at = c(0, 1, 2),
                                       gp = gpar(fontsize = fs)),
                     width = unit(10, "mm")),
    annotation_name_gp = gpar(fontsize = fl))

# main heatmap value range for color scale
mx <- max(abs(m[sel_peaksEnr, ]))
qs <- quantile(abs(m[sel_peaksEnr, ]), .99)

# create main heatmap with annotations
hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
               col = circlize::colorRamp2(
                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
               cluster_rows = FALSE,
               cluster_columns = column_dend(hm3), column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
               border = TRUE, border_gp = gpar(lwd = 0.5),
               left_annotation = peakAnnot,
               right_annotation = peakAnnot2,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = TRUE,
               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                           labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(30, "mm"),
                                           title = "TF log2 IP/input",
                                           title_position = "topcenter",
                                           title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm"))

# draw the heatmap and grab as a grid graphics `grob`
# (needed for combining figure panels)
# use defined random number seed to make clustering deterministic
set.seed(42L)
hm_peaksEnr <- grid.grabExpr(
    hm4 <- draw(hm4, merge_legend = TRUE,
                heatmap_legend_side = "bottom",
                annotation_legend_side = "bottom"),
    width = w_peaksHm, height = h_peaksHm
)

plot_grid(hm_peaksEnr)
```

## ChIP enrichment replicate-pair correlations

```{r plot-chipReplPairs}
#| fig.width: 9
#| fig.height: 9

pd <- peakenr |> as.data.frame() |>
    tibble::rownames_to_column("peakid") |>
    pivot_longer(col = !matches("peakid")) |>
    mutate(group = sub("_rep[12]$", "", name),
           name = factor(name, levels = unique(name)))

ylims <- c(-2.0, 4.0)
yticks <- c(-2, 0, 2, 4)

pL <- lapply(levels(fct_relevel(sort(unique(pd$group)), "Untagged", after = Inf)),
             function(grp1) {
    pd1 <- pd[pd$group == grp1, ] |> pivot_wider()
    pd1$cols <- densCols(x = pd1[[paste0(grp1, "_rep1")]],
                         y = pd1[[paste0(grp1, "_rep2")]],
                         nbin = 64, colramp = colorRampPalette(hcl.colors(32)))
    ggplot(pd1, aes(.data[[paste0(grp1, "_rep1")]],
                    .data[[paste0(grp1, "_rep2")]])) +
        geom_abline(intercept = 0, slope = 1, linetype = 1, color = "gray") +
        geom_hline(yintercept = 0, linetype = 2, color = "gray") +
        geom_vline(xintercept = 0, linetype = 2, color = "gray") +
        geom_scattermost(xy = as.data.frame(pd1[, paste0(grp1, c("_rep1", "_rep2"))]),
                         pointsize = 2, color = pd1$cols, pixels = c(256, 256)) +
        coord_fixed(xlim = ylims, ylim = ylims, expand = FALSE, clip = "off") +
        scale_x_continuous(breaks = yticks) +
        scale_y_continuous(breaks = yticks) +
        theme_cowplot(7) +
        labs(x = element_blank(), y = element_blank()) +
        annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.05,
                 label = grp1, color = "black", size = 3) +
        theme(legend.position = "none")
})

(gg_enrpairs <- plot_grid(plotlist = pL, nrow = 9, ncol = 9))
```

## Put together

```{r Fig2}
#| fig.width: 11.5
#| fig.height: 15.5

cowplot::plot_grid(
    hm_peaksEnr,
    hm_peaksBin,
    hm_150,
    hm_500,
    nrow = 2,
    labels = c("A", "C", "B", "D"),
    align = "v",
    axis = "b"
)
```

## Supplementary figure

Assemble the panels into Supplementary Figure 2:

```{r suppFig2}
#| fig.width: 11
#| fig.height: 11
#| warning: false

cowplot::plot_grid(
    gg_enrpairs,
    scale = 0.9,
    labels = "A"
)
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2024-07-05 ms_figure_1_overview.Rmd
add TF construct sequences and link from figure 1
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>
Rearrange panels in Figure 1, highlight meiosis-specific TFs

@@
                        name = "") + 
+    scale_x_discrete(breaks = c("mei4", "atf31", "atf21", "cuf2", "rsv1")) + 
     guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
@@
 (gg_tpm + theme(axis.text.x = element_blank(),
-                axis.ticks.x = element_blank()))
+                axis.ticks.length.x = unit(2, "mm"),
+                axis.ticks.x = element_line(color = main_colors[1],
+                                            linewidth = 1.25)))
 ```
@@
 For each sample, the unmapped reads were aligned against all tagged
-transcription factor sequences. The resulting  alignment counts are summarized
+transcription factor sequences (see file [`reference/TFome_pombe_strain_lib.fasta`](reference/TFome_pombe_strain_lib.fasta)). The resulting  alignment counts are summarized
 in the `tagcnt` matrix of `r nrow(tagcnt)` rows by `r ncol(tagcnt)` columns.
@@
 cowplot::plot_grid(
+    fig1c,
     cowplot::plot_grid(
@@
         gg_tpm + theme(axis.text.x = element_blank(),
-                       axis.ticks.x = element_blank()),
+                       axis.ticks.length.x = unit(2, "mm"),
+                       axis.ticks.x = element_line(color = main_colors[1],
+                                                   linewidth = 1.25)),
         nrow = 1,
-        labels = c("A", "B"),
+        labels = c("B", "C"),
         align = "vh",
@@
     ),
-    fig1c,
     cowplot::plot_grid(
@@
     nrow = 3,
-    labels = c(NA, "C", NA),
+    labels = c("A", NA, NA),
     align = "v",
     axis = "b",
-    rel_heights = c(50, 73, 50)
+    rel_heights = c(73, 50, 50)
 )

### 2024-07-05 ms_figure_5_ipms.Rmd
Remove heatmaps from Figure 5 (now in Figure 2)

@@
 ---
-title: "Figure 4 - IP-MS"
+title: "Figure 5 - IP-MS"
 author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
@@
         show_legend = FALSE),
-    heatmap_height = unit(h_ipmsHm * 0.945, "mm"), # whole heatmap height
+    heatmap_height = unit(1.125 * h_ipmsHm, "mm"), # whole heatmap height
     column_title_gp = gpar(fontsize = ft),
@@
                   padding = unit(c(3, 1, 1, 1), "mm")),
-    width = w_ipmsHm, height = h_ipmsHm
+    width = w_ipmsHm, height = 1.125 * h_ipmsHm
 )
@@
 
-```{r Fig4}
+```{r Fig5}
 #| fig.width: 9.8
@@
 cowplot::plot_grid(
-    hm_150,
+    cowplot::plot_grid(
         plot_grid(gg_retain1, NULL, hm_150_saga_nua4, nrow = 1, rel_widths = c(1, 0.05, 0.75)),
-    hm_500,
         gg_retain2,
-    nrow = 2,
-    labels = c("A", "C", "B", "D"),
-    align = "v",
-    axis = "b"
-)
-```
-
-## Supplementary Figure
-
-```{r suppFig4}
-#| fig.width: 10.8
-#| fig.height: 14.8
-#| warning: false
-
-cowplot::plot_grid(
-    wrap_plots(ggpca1, ggpca2, ggatf1, ggpca3, hm_150_bait, hm_500_bait, ncol = 2) + 
-        plot_annotation(tag_levels = list(c("A", "", "", "", "B"))) & 
-        theme(plot.tag = element_text(size = 14, face = "bold")),
+        hm_150_bait, hm_500_bait,
+        rel_heights = c(7, 4),
+        labels = c("A", "B", "C", "D")
+    ),
     cowplot::plot_grid(
@@
     ),
-    ncol = 1,
+    nrow = 2,
     rel_heights = c(12, 1)
 )
+```
+
+## Supplementary Figure
+
+```{r suppFig5}
+#| fig.width: 10.8
+#| fig.height: 10.8
+#| warning: false
+
+wrap_plots(ggpca1, ggpca2, ggatf1, ggpca3, ncol = 2) + 
+    # plot_annotation(tag_levels = list(c("A", "", "", "", "B"))) & 
+    theme(plot.tag = element_text(size = 14, face = "bold"))
 ```

### 2024-07-05 ms_figure_6_networks.Rmd
Rearrange Fig 6, add new panels

--- 2024-06-21 ms_figure_5_networks.Rmd
@@
 ---
-title: "Figure 5 - networks"
+title: "Figure 6 - networks"
 author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
@@
     library(kableExtra)
+    library(ggupset)
     library(tidyr)
@@
 
-# Figure 5
+# 14-3-3 protein interactions
+
+We extract all IPs where either `Rad24` or `Rad25` are pulled down, 
+using a relatively lenient adjusted p-value cutoff.
+
+```{r rad24-rad25}
+#| fig.width: 9
+#| fig.height: 15
+
+## Get all t-statistics and visualize
+res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
+rad <- res150untag$tstat[c("Rad24", "Rad25"), ]
+colnames(rad) <- .getProteinNameFromComparison(colnames(rad))
+rad <- t(rad)
+hmrad <- Heatmap(rad, 
+                 cluster_rows = TRUE, 
+                 cluster_columns = FALSE, 
+                 col = circlize::colorRamp2(
+                     breaks = seq(1.5, 13.0, length.out = 64),
+                     colors = colorRampPalette(
+                         binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
+                 border = TRUE, 
+                 border_gp = gpar(lwd = 0.5),
+                 name = "Mod.\nt-stat.",
+                 na_col = binary_heatmap_colors["FALSE"],
+                 column_names_gp = gpar(fontsize = fl),
+                 row_names_gp = gpar(fontsize = fs),
+                 show_row_names = TRUE, 
+                 show_column_names = TRUE, 
+                 use_raster = FALSE,
+                 show_heatmap_legend = TRUE, 
+                 width = unit(w_trad / 1.5, "mm"), 
+                 heatmap_height = unit(1.4 * h_trad, "mm"),
+                 column_title_gp = gpar(fontsize = ft),
+                 heatmap_legend_param = list(
+                     title_gp = gpar(fontsize = fl),
+                     border = "gray10",
+                     border_gp = gpar(lwd = 0.5)
+                 ))
+set.seed(42L)
+hm_rad <- grid.grabExpr(
+    hmrad <- draw(hmrad, merge_legend = TRUE,
+                 heatmap_legend_side = "right",
+                 annotation_legend_side = "right"),
+    width = w_trad / 1.5, height = 1.4 * h_trad
+)
+
+## List of experiments where rad24 and/or rad25 were pulled down
+comps <- colnames(res150untag$interactor)[colSums(
+    res150untag$interactor[c("Rad24", "Rad25"), ], na.rm = TRUE) > 0]
+explist <- .getUniProtIdFromComparison(comps, idmap = idmap)
+cat(explist)
+gnames <- .getProteinNameFromComparison(comps)
+cat(gnames)
+
+## Read table
+set_flextable_defaults(font.size = 8)
+rad24_rad25_tbl <- read.delim(params$rad24_rad25_tbl)[, 1:2] |>
+    setNames(c("TF", "R-x-x-[S/T]-x-P")) |>
+    dplyr::mutate(TF = .capitalize(TF))
+stopifnot(all(gnames %in% rad24_rad25_tbl$TF))
+ft <- flextable(rad24_rad25_tbl) |>
+  width(width =  c(0.5, 1.25)) |>
+  set_table_properties(layout = "fixed")
+ggtbl <- ggplot() +
+    theme_void() +
+    annotation_custom(grid::rasterGrob(as_raster(ft)),
+                      xmin = -Inf, xmax = Inf, ymin = 0, ymax = 1)
+
+data.frame(`Number of TFs interacting with at least one 14-3-3 protein` = 
+               nrow(rad24_rad25_tbl),
+           `Number of such TFs with at least one optimal 14-3-3 binding motif` = 
+               sum(rad24_rad25_tbl$`R-x-x-[S/T]-x-P` != ""),
+           check.names = FALSE) |>
+    pivot_longer(cols = everything()) |>
+    kbl(col.names = NULL) |>
+    kable_styling()
+
+# UpSet plot
+dfupset <- as.data.frame(t(res150untag$interactor[c("Rad24", "Rad25"), ])) |>
+    tibble::rownames_to_column("TF") |>
+    dplyr::mutate(TF = .getProteinNameFromComparison(TF))
+
+ovL <- as.list(as.data.frame(dfupset[, c("Rad24", "Rad25")]))
+ovL <- lapply(ovL, function(i) dfupset$TF[which(i)])
+
+pd <- stack(ovL) |>
+    group_by(values) |>
+    summarise(ind = list(ind)) |>
+    dplyr::left_join(rad24_rad25_tbl, by = join_by("values" == "TF")) |>
+    dplyr::mutate(has_motif = !is.na(`R-x-x-[S/T]-x-P`) & `R-x-x-[S/T]-x-P` != "")
+
+ggrad_upset <- ggplot(pd, aes(x = ind)) +
+    geom_bar(aes(fill = has_motif)) +
+    scale_fill_manual(values = c(`TRUE` = main_colors[3], `FALSE` = main_colors[5])) +
+    scale_x_upset() +
+    labs(x = element_blank(), y = "Number of IPs", fill = "Has motif: ") +
+    theme_cowplot(12) +
+    theme(axis.ticks.x = element_blank(),
+          legend.position = "bottom") +
+    theme_combmatrix(combmatrix.label.extra_spacing = 5,
+                     combmatrix.label.make_space = TRUE,
+                     combmatrix.label.width = unit(18, "mm")) +
+    guides(fill = guide_legend(nrow = 1))
+```
+
+# Figure 6
 
@@
 
-```{r Fig5}
+```{r Fig6}
 #| fig.width: 10
 #| fig.height: 10
+
+fig6c <- ggdraw() + 
+    draw_image("schematics/Fig6C_Pho7_protein_domains_cartoon_Arial_RGB.png")
 
@@
 cowplot::plot_grid(
-    fig5a, 
     cowplot::plot_grid(
@@
             ggr500 + theme(legend.position = "none"), 
-            nrow = 1, labels = c("B", "C")
+            nrow = 1, labels = c("A", "B")
         ),
@@
         nrow = 2, labels = c("", ""), rel_heights = c(1, 0.25)
-    ), ncol = 1, rel_heights = c(1, 2)
+    ),
+    cowplot::plot_grid(fig6c, ggrad_upset,
+                       nrow = 1, labels = c("C", "D")), 
+    rel_heights = c(5, 2), 
+    ncol = 1
 )
@@
 
-```{r suppFig5}
+```{r suppFig6}
 #| fig.width: 9
-#| fig.height: 10
+#| fig.height: 15
 
-## Get all t-statistics and visualize
-res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
-rad <- res150untag$tstat[c("Rad24", "Rad25"), ]
-colnames(rad) <- .getProteinNameFromComparison(colnames(rad))
-rad <- t(rad)
-hmrad <- Heatmap(rad, 
-                 cluster_rows = TRUE, 
-                 cluster_columns = FALSE, 
-                 col = circlize::colorRamp2(
-                     breaks = seq(1.5, 13.0, length.out = 64),
-                     colors = colorRampPalette(
-                         binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
-                 border = TRUE, 
-                 border_gp = gpar(lwd = 0.5),
-                 name = "Mod.\nt-stat.",
-                 na_col = binary_heatmap_colors["FALSE"],
-                 column_names_gp = gpar(fontsize = fl),
-                 row_names_gp = gpar(fontsize = fs),
-                 show_row_names = TRUE, 
-                 show_column_names = TRUE, 
-                 use_raster = FALSE,
-                 show_heatmap_legend = TRUE, 
-                 width = unit(w_trad / 1.5, "mm"), 
-                 heatmap_height = unit(1.3 * h_trad, "mm"),
-                 column_title_gp = gpar(fontsize = ft),
-                 heatmap_legend_param = list(
-                     title_gp = gpar(fontsize = fl),
-                     border = "gray10",
-                     border_gp = gpar(lwd = 0.5)
-                 ))
-set.seed(42L)
-hm_rad <- grid.grabExpr(
-    hmrad <- draw(hmrad, merge_legend = TRUE,
-                 heatmap_legend_side = "right",
-                 annotation_legend_side = "right"),
-    width = w_trad / 1.5, height = 1.3 * h_trad
+plot_grid(
+    fig5a, 
+    plot_grid(hm_rad, ggtbl, nrow = 1, align = "h", axis = "b", 
+              rel_widths = c(2, 2.5), labels = c("B", "C")),
+    rel_heights = c(1, 3), ncol = 1,
+    labels = c("A", "")
 )
-
-## List of experiments where rad24 and/or rad25 were pulled down
-comps <- colnames(res150untag$interactor)[colSums(
-    res150untag$interactor[c("Rad24", "Rad25"), ], na.rm = TRUE) > 0]
-explist <- .getUniProtIdFromComparison(comps,idmap = idmap)
-cat(explist)
-gnames <- .getProteinNameFromComparison(comps)
-cat(gnames)
-
-## Read table
-set_flextable_defaults(font.size = 8)
-rad24_rad25_tbl <- read.delim(params$rad24_rad25_tbl)[, 1:2] |>
-    setNames(c("TF", "R-x-x-[S/T]-x-P")) |>
-    dplyr::mutate(TF = .capitalize(TF))
-stopifnot(all(gnames %in% rad24_rad25_tbl$TF))
-ft <- flextable(rad24_rad25_tbl) |>
-  width(width =  c(0.5, 1.25)) |>
-  set_table_properties(layout = "fixed")
-ggtbl <- ggplot() +
-    theme_void() +
-    annotation_custom(grid::rasterGrob(as_raster(ft)),
-                      xmin = -Inf, xmax = Inf, ymin = 0, ymax = 1)
-
-data.frame(`Number of TFs interacting with at least one 14-3-3 protein` = 
-               nrow(rad24_rad25_tbl),
-           `Number of such TFs with at least one optimal 14-3-3 binding motif` = 
-               sum(rad24_rad25_tbl$`R-x-x-[S/T]-x-P` != ""),
-           check.names = FALSE) |>
-    pivot_longer(cols = everything()) |>
-    kbl(col.names = NULL) |>
-    kable_styling()
-
-plot_grid(hm_rad, ggtbl, nrow = 1, align = "h", axis = "b", 
-          rel_widths = c(2, 2.5), labels = c("A", "B"))
 ```

### 2024-07-08 ms_figure_2_overview-heatmaps.Rmd
add technology to panel titles
small edits to ChIP heatmaps

@@
                             ifelse(!is.na(dist.peak2rRNA) & dist.peak2rRNA < maxdist,
-                                   "5S rRNA", "none")),
-                         levels = c("tRNA", "5S rRNA", "none"))
+                                   "5S rRNA", NA)),
+                         levels = c("tRNA", "5S rRNA"))
 
@@
                                              title_gp = gpar(fontsize = fl)),
-                          `ncRNA` = list(labels_gp = gpar(fontsize = fs),
-                                         title_position = "topcenter",
-                                         title_gp = gpar(fontsize = fl)))
+                          `tRNA or 5S rRNA` = list(labels_gp = gpar(fontsize = fs),
+                                                   title = ""))
 cols_ncRNA <- c("tRNA" = main_colors[5], "5S rRNA" = main_colors[3],
@@
                              colors = hcl.colors(64)),
-    `ncRNA` = cols_ncRNA)
+    `tRNA or 5S rRNA` = cols_ncRNA)
 
@@
                   `TSS dist.` = peakgr$nrst_TSS_dist[sel_peaksBin],
-                  `ncRNA` = near_ncRNA[sel_peaksBin],
-                  col = annotCols,
+                  `tRNA or 5S rRNA` = near_ncRNA[sel_peaksBin],
+                  col = annotCols, na_col = bg_color,
                   width = unit(21, "mm"), show_legend = TRUE,
@@
 annotCols2 <- list(
-    `ncRNA` = cols_ncRNA,
+    `tRNA or 5S rRNA` = cols_ncRNA,
     H3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
@@
 annotLegendParams2 <- list(
-    `ncRNA` = list(labels_gp = gpar(fontsize = fs),
-                   title_position = "topcenter",
-                   title_gp = gpar(fontsize = fl)),
+    `tRNA or 5S rRNA` = list(labels_gp = gpar(fontsize = fs),
+                             title = ""),
     H3 = list(at = c(-6, 0, 6),
@@
                   which = "row",
-                  `ncRNA` = near_ncRNA[sel_peaksEnr],
+                  `tRNA or 5S rRNA` = near_ncRNA[sel_peaksEnr],
                   H3 = chip_enr_sel[, "GSE108668_H3"],
@@
                   H3K9me3 = chip_enr_sel[, "GSE182250_H3K9me3"],
-                  col = annotCols2,
+                  col = annotCols2, na_col = bg_color,
                   width = unit(28, "mm"),
@@
     which = "row",
-    `Accessibility\n1e3 RPKM` =
+    `Accessibility\n1e3 RPKM\nATAC-seq` =
         anno_barplot(x = atac_trunc / 1e3,
@@
     nrow = 2,
-    labels = c("A", "C", "B", "D"),
+    labels = c("A  ChIP", "B  ChIP", "C  IP-MS", "D  IP-MS"),
     align = "v",

### 2024-07-08 ms_figure_3_chip.Rmd
edits to new figure 3 (ChIP)

@@
 ---
-title: "Figure 2 - ChIP"
+title: "Figure 3 - ChIP"
 author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
@@
     peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
-    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
     peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
@@
     library(tidyr)
+    library(tidygraph)
     library(universalmotif)
@@
     library(forcats)
+    library(igraph)
+    library(ggraph)
 })
@@
 
-# ... average replicates (`peakenrAvg`)
-grp <- sub("_rep[12]$", "", colnames(peakenr))
-peakenrAvg <- do.call(cbind, lapply(split(colnames(peakenr), grp)[unique(grp)],
-                                    function(j) {
-                                        rowMeans(peakenr[, j, drop = FALSE])
-                                    }))
-
 # `chip_genome_tab`: data.frame with alignment densities in selected
@@
 chip_genome_tab <- read.csv(params$chiptrackfile)
-
-# `atac_raw`: public ATAC-seq read counts in ChIP-seq peaks from this study
-atac_raw <- as.matrix(read.csv(params$peakatacfile, row.names = 1))
-
-# ... calculate reads per million (`atac_cpm`) and
-#     per kilobase and million (`atac_rpkm`)
-#     and average experiments (`atac_cpm_avg`, `atac_rpkm_avg`)
-atac_cpm <- sweep(x = atac_raw[, -1], MARGIN = 2,
-                  STATS = colSums(atac_raw[, -1]), FUN = "/") * 1e6
-atac_rpkm <- atac_cpm / atac_raw[, "width"] * 1e3
-atac_cpm_avg <- rowMeans(atac_cpm)
-atac_rpkm_avg <- rowMeans(atac_rpkm)
-
-# `chip_raw`: public ChIP-seq read counts in ChIP-seq peaks from this study
-chip_raw <- as.matrix(read.csv(params$peakchipfile, row.names = 1))
-
-# ... calculate reads per million (`chip_cpm`) and
-#     per kilobase and million (`chip_rpkm`)
-#     and calculate enrichments (`chip_enr`, log2 IP/input) averaged over replicates
-chip_cpm <- sweep(x = chip_raw[, -1], MARGIN = 2,
-                  STATS = colSums(chip_raw[, -1]), FUN = "/") * 1e6
-chip_rpkm <- chip_cpm / chip_raw[, "width"] * 1e3
-chip_series <- sub("^ChIP_([^_]+)_.+$", "\\1", colnames(chip_cpm))
-chip_enr <- do.call(
-    cbind,
-    lapply(split(colnames(chip_cpm), chip_series),
-           function(nms) {
-               grp <- sub("ChIP_(GSE[0-9]+)_GSM[0-9]+_(.+)_rep[0-9]$",
-                          "\\1_\\2", nms)
-               stopifnot(any(is_input <- grep("_Input", grp)))
-               enr <- do.call(
-                   cbind,
-                   lapply(unique(grp[-is_input]), function(grp1) {
-                       rowMeans(log2(chip_cpm[, nms[grp == grp1], drop = FALSE] + 1)) -
-                           rowMeans(log2(chip_cpm[, nms[is_input], drop = FALSE] + 1))
-                   }))
-               colnames(enr) <- unique(grp[-is_input])
-               enr
-           }))
 
@@
 
-# Figure 2
+# Figure 3
 
@@
         guides(fill = guide_legend(ncol = 1)))
-```
-
-## Peak vs. sample heatmap (binary)
-
-The following creates heatmap of peaks (rows) versus ChIP-seq experiments
-(columns). The values are binary (`TRUE` or `FALSE`) and indicate if a peak
-was enriched in ChIP-seq experiment (had an average log2 IP/input enrichment
-values greater than 1.0).
-
-```{r plot-peaksBin}
-#| fig.width: 6
-#| fig.height: 8
-
-# extract binary enrichment matrix from `peakgr`
-m <- as.matrix(mcols(peakgr)[, grep("^is_enr_in[.]", colnames(mcols(peakgr)))])
-mode(m) <- "numeric"
-colnames(m) <- sub("is_enr_in.", "", colnames(m))
-
-# specify ChIP-seq experiments for which to show labels
-selLabels <- c("Untagged")
-stopifnot(all(selLabels %in% colnames(m)))
-
-# only show peaks that are enriched in at least one ChIP-seq experiment
-sel_peaksBin <- rowSums(m) > 0
-
-# prepare annotation data
-#   `fractTFs`: fraction of ChIP-seq experiments that a peak is enriched in
-#   `numpeaks`: the number of enriched peaks for each ChIP-seq experiment
-#   `peaktype`: the type of each peak (specific, frequent or ubiquitous)
-#   `near_ncRNA`: specifies if the peak is nearer than `maxdist` from
-#                 a %S rRNA or tRNA gene
-fracTFs <- rowMeans(m > 0)
-numpeaks <- colSums(m > 0)
-levels(peakgr$peaktype) <- c("Common (ubiquitous)", # use shorter names
-                             "Common (frequent)",
-                             "Specific peaks")
-peaktype <- factor(gsub("[()]", "", sub(" ", "\n", peakgr$peaktype)),
-                   levels = c("Common\nubiquitous",
-                              "Common\nfrequent",
-                              "Specific\npeaks"))
-near_ncRNA <- factor(ifelse(!is.na(dist.peak2tRNA) & dist.peak2tRNA < maxdist,
-                            "tRNA",
-                            ifelse(!is.na(dist.peak2rRNA) & dist.peak2rRNA < maxdist,
-                                   "5S rRNA", "none")),
-                         levels = c("tRNA", "5S rRNA", "none"))
-
-# most tRNA and 5S rRNA genes are near "common (ubiquitous)" peaks:
-table(peaktype[sel_peaksBin], near_ncRNA[sel_peaksBin])
-
-# parameters for annotation legends and colors
-annotLegendParams <- list(`%GC` = list(at = c(20, 40, 60),
-                                       labels_gp = gpar(fontsize = fs),
-                                       border = "gray10",
-                                       legend_direction = "horizontal",
-                                       legend_width = unit(18, "mm"),
-                                       title_position = "topcenter",
-                                       title_gp = gpar(fontsize = fl)),
-                          `TSS dist.` = list(at = c(0, 4000, 8000),
-                                             labels_gp = gpar(fontsize = fs),
-                                             border = "gray10",
-                                             legend_direction = "horizontal",
-                                             legend_width = unit(18, "mm"),
-                                             title_position = "topcenter",
-                                             title_gp = gpar(fontsize = fl)),
-                          `ncRNA` = list(labels_gp = gpar(fontsize = fs),
-                                         title_position = "topcenter",
-                                         title_gp = gpar(fontsize = fl)))
-cols_ncRNA <- c("tRNA" = main_colors[5], "5S rRNA" = main_colors[3],
-                "none" = bg_color)
-annotCols <- list(
-    `%GC` = colorRamp2(breaks = seq(min(peakgr$fracGC[sel_peaksBin]),
-                                    max(peakgr$fracGC[sel_peaksBin]),
-                                    length.out = 64) * 100,
-                       colors = rev(hcl.colors(64, "Greens"))),
-    ` width` = colorRamp2(breaks = seq(min(width(peakgr)[sel_peaksBin]),
-                                       max(width(peakgr)[sel_peaksBin]),
-                                       length.out = 64),
-                          colors = rev(hcl.colors(64, "YlOrBr"))),
-    `TSS dist.` = colorRamp2(breaks = c(0, 10^seq(2, log10(max(peakgr$nrst_TSS_dist)),
-                                                  length.out = 63)),
-                             colors = hcl.colors(64)),
-    `ncRNA` = cols_ncRNA)
-
-# prepare heatmap annotations
-#   `peakAnnot` for peaks (rows), left side
-#   `fracTFsAnnot` for peaks (rows), right side
-#   `sampleAnnot` for ChIP-seq experiments (columns), top
-#   `sampleLabels` for ChIP-seq experiments (columns), bottom
-peakAnnot <- HeatmapAnnotation(
-                  which = "row",
-                  `%GC` = peakgr$fracGC[sel_peaksBin] * 100,
-                  `TSS dist.` = peakgr$nrst_TSS_dist[sel_peaksBin],
-                  `ncRNA` = near_ncRNA[sel_peaksBin],
-                  col = annotCols,
-                  width = unit(21, "mm"), show_legend = TRUE,
-                  annotation_name_gp = gpar(fontsize = fl),
-                  annotation_legend_param = annotLegendParams)
-
-fracTFsAnnot <- HeatmapAnnotation(
-    which = "row",
-    `Fraction\nof TFs` = anno_barplot(
-        x = fracTFs[sel_peaksBin], gp = gpar(col = na_color),
-        border = FALSE, bar_width = 1.0,
-        axis_param = list(at = c(0, 0.4, 0.8),
-                          gp = gpar(fontsize = fs)),
-        width = unit(10, "mm")),
-    annotation_name_gp = gpar(fontsize = fl))
-sampleAnnot <- HeatmapAnnotation(
-    which = "column",
-    `Number of\nenriched\npeaks` = anno_barplot(
-        x = numpeaks,
-        gp = gpar(col = 0, fill = na_color),
-        border = FALSE, bar_width = 1.0,
-        axis_param = list(at = c(0, 150, 300),
-                          gp = gpar(fontsize = fs),
-                          facing = "outside",
-                          direction = "normal"),
-        width = unit(10, "mm")),
-    annotation_name_gp = gpar(fontsize = fl),
-    annotation_name_side = "left", annotation_name_rot = 0,
-    show_legend = TRUE)
-sampleLabels <- columnAnnotation(
-    TFnames = anno_mark(which = "column", side = "bottom",
-                        at = match(selLabels, colnames(m)), 
-                        labels = sub("^[^_]+_", "", selLabels),
-                        labels_gp = gpar(fontsize = fl)))
-
-# create main heatmap with annotations
-hm1 <- Heatmap(m[sel_peaksBin, ],
-               col = circlize::colorRamp2(
-                   breaks = seq(0, max(m), length.out = 64),
-                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
-               border = TRUE, border_gp = gpar(lwd = 0.5),
-               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
-               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
-               row_title_gp = gpar(fontsize = fl),
-               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
-               left_annotation = peakAnnot,
-               right_annotation = fracTFsAnnot,
-               top_annotation = sampleAnnot,
-               bottom_annotation = sampleLabels,
-               show_row_names = FALSE, show_column_names = FALSE,
-               use_raster = TRUE, show_heatmap_legend = FALSE,
-               width = unit(w_peaksHm, "mm"), # heatmap body width
-               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height
-
-# draw the heatmap and grab as a grid graphics `grob`
-# (needed for combining figure panels)
-# use defined random number seed to make clustering deterministic
-set.seed(42L)
-hm_peaksBin <- grid.grabExpr(
-    hm1 <- draw(hm1, merge_legend = TRUE,
-                heatmap_legend_side = "bottom",
-                annotation_legend_side = "bottom"),
-    width = w_peaksHm, height = h_peaksHm
-)
-
-plot_grid(hm_peaksBin)
-```
-
-## Peak vs. sample heatmap (enrichments)
-
-The following creates a heatmap similar to the peak-versus-experiment
-heatmap above, but using the ChIP-seq enrichment values (log2 IP/input)
-directly instead of binarized (`TRUE`, `FALSE`) values.
-
-```{r plot-peaksEnr}
-#| fig.width: 6
-#| fig.height: 8
-
-# use ChIP-seq enrichment matrix from this study (replicates averaged)
-m <- peakenrAvg
-stopifnot(all(selLabels %in% colnames(m)))
-
-# to keep the same row order as in the binary heatmap `hm1`,
-# we have to re-calculate the annotations and switch off the row clustering
-sel_peaksEnr <- which(sel_peaksBin)[unlist(row_order(hm1), use.names = FALSE)]
-
-# prepare annotation data
-#   `chip_enr_sel`: selected public ChIP-seq experiments
-#                   (log2 IP/input enrichments)
-#   `chip_enr_sel_range`: value range for color scale (1% to 99%)
-#   `chip_enr_sel_range_full`: value range for color scale (0% to 100%)
-chip_enr_sel <- chip_enr[sel_peaksEnr, ]
-chip_enr_sel_range <- quantile(abs(chip_enr_sel), probs = c(0.95)) * c(-1, 1)
-chip_enr_sel_range_full <- range(chip_enr_sel)
-
-# parameters for annotation legends and colors
-annotCols2 <- list(
-    `ncRNA` = cols_ncRNA,
-    H3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
-                               seq(chip_enr_sel_range[1],
-                                   chip_enr_sel_range[2],
-                                   length.out = 62),
-                               chip_enr_sel_range_full[2]),
-                    colors = viridisLite::magma(64)),
-    H3K14ac = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
-                                    seq(chip_enr_sel_range[1],
-                                        chip_enr_sel_range[2],
-                                        length.out = 62),
-                                    chip_enr_sel_range_full[2]),
-                         colors = viridisLite::magma(64)),
-    H3K9me2 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
-                                    seq(chip_enr_sel_range[1],
-                                        chip_enr_sel_range[2],
-                                        length.out = 62),
-                                    chip_enr_sel_range_full[2]),
-                         colors = viridisLite::magma(64)),
-    H3K9me3 = colorRamp2(breaks = c(chip_enr_sel_range_full[1],
-                                    seq(chip_enr_sel_range[1],
-                                        chip_enr_sel_range[2],
-                                        length.out = 62),
-                                    chip_enr_sel_range_full[2]),
-                         colors = viridisLite::magma(64)))
-annotLegendParams2 <- list(
-    `ncRNA` = list(labels_gp = gpar(fontsize = fs),
-                   title_position = "topcenter",
-                   title_gp = gpar(fontsize = fl)),
-    H3 = list(at = c(-6, 0, 6),
-              title = "Histone log2 IP/input",
-              labels_gp = gpar(fontsize = fs),
-              border = "gray10",
-              legend_direction = "horizontal",
-              legend_width = unit(30, "mm"),
-              title_position = "topcenter",
-              title_gp = gpar(fontsize = fl)))
-
-# prepare heatmap annotations
-#   `peakAnnot` for peaks (rows), left side
-#   `peakAnnot2` for peaks (rows), right side
-peakAnnot <- HeatmapAnnotation(
-                  which = "row",
-                  `ncRNA` = near_ncRNA[sel_peaksEnr],
-                  H3 = chip_enr_sel[, "GSE108668_H3"],
-                  H3K14ac = chip_enr_sel[, "GSE108668_H3K14ac"],
-                  H3K9me2 = chip_enr_sel[, "GSE182250_H3K9me2"],
-                  H3K9me3 = chip_enr_sel[, "GSE182250_H3K9me3"],
-                  col = annotCols2,
-                  width = unit(28, "mm"),
-                  annotation_name_gp = gpar(fontsize = fl),
-                  annotation_legend_param = annotLegendParams2,
-                  show_legend = c(TRUE, TRUE, FALSE, FALSE, FALSE))
-atac_trunc <- pmin(atac_rpkm_avg[sel_peaksEnr], 2000) # cap ATAC signal at 2000
-peakAnnot2 <- HeatmapAnnotation(
-    which = "row",
-    `Accessibility\n1e3 RPKM` =
-        anno_barplot(x = atac_trunc / 1e3,
-                     gp = gpar(col = na_color),
-                     border = FALSE, bar_width = 1.0,
-                     baseline = 0,
-                     axis_param = list(at = c(0, 1, 2),
-                                       gp = gpar(fontsize = fs)),
-                     width = unit(10, "mm")),
-    annotation_name_gp = gpar(fontsize = fl))
-
-# main heatmap value range for color scale
-mx <- max(abs(m[sel_peaksEnr, ]))
-qs <- quantile(abs(m[sel_peaksEnr, ]), .99)
-
-# create main heatmap with annotations
-hm2 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
-               col = circlize::colorRamp2(
-                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
-                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
-               cluster_rows = FALSE,
-               cluster_columns = column_dend(hm1), column_dend_height = unit(10, "mm"),
-               row_title_gp = gpar(fontsize = fl),
-               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
-               border = TRUE, border_gp = gpar(lwd = 0.5),
-               left_annotation = peakAnnot,
-               right_annotation = peakAnnot2,
-               top_annotation = sampleAnnot,
-               bottom_annotation = sampleLabels,
-               show_row_names = FALSE, show_column_names = FALSE,
-               use_raster = TRUE, show_heatmap_legend = TRUE,
-               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
-                                           labels_gp = gpar(fontsize = fs),
-                                           border = "gray10",
-                                           legend_direction = "horizontal",
-                                           legend_width = unit(30, "mm"),
-                                           title = "TF log2 IP/input",
-                                           title_position = "topcenter",
-                                           title_gp = gpar(fontsize = fl)),
-               width = unit(w_peaksHm, "mm"), # heatmap body width
-               heatmap_height = unit(h_peaksHm, "mm"))
-
-# draw the heatmap and grab as a grid graphics `grob`
-# (needed for combining figure panels)
-# use defined random number seed to make clustering deterministic
-set.seed(42L)
-hm_peaksEnr <- grid.grabExpr(
-    hm2 <- draw(hm2, merge_legend = TRUE,
-                heatmap_legend_side = "bottom",
-                annotation_legend_side = "bottom"),
-    width = w_peaksHm, height = h_peaksHm
-)
-
-plot_grid(hm_peaksEnr)
 ```
@@
 
+Here is the complete list of motifs for each of the motif clusters,
+in the order as they appear in the heatmap (from top to bottom):
+
+```{r list-motif-clusters}
+# get the row-order of the clustered heatmap
+ro <- row_order(hm4)
+
+# print the motif names in each cluster
+motifs_in_cluster_list <- lapply(seq.int(n_split), function(i) {
+    motifs_in_cluster <- rownames(motifsim2)[ro[[i]]]
+    cat("Motif cluster", i, ":",
+        paste(motifs_in_cluster, collapse = ", "), "\n")
+    motifs_in_cluster
+})
+```
+
 ## Motif alignments
@@
 # select specific and common-frequent peaks only
+levels(peakgr$peaktype) <- c("Common (ubiquitous)", # use shorter names
+                             "Common (frequent)",
+                             "Specific peaks")
 peakgrSel <- peakgr[peakgr$peaktype %in% c("Specific peaks", "Common (frequent)")]
@@
 
+### Heatmap
+
 For visualization, we classify known transcription factors (rows) into three
@@
 
-```{r plot-autoregulation}
+```{r plot-autoregulation-heatmap}
 #| fig.width: 5
@@
 
-# ... create promoter/terminator binding matrix
+# ... create promoter binding matrix
 mprom <- do.call(rbind, lapply(tx_sel, function(i) {
@@
 
-## ChIP enrichment replicate-pair correlations
-
-```{r plot-chipReplPairs}
-#| fig.width: 9
-#| fig.height: 9
-
-pd <- peakenr |> as.data.frame() |>
-    tibble::rownames_to_column("peakid") |>
-    pivot_longer(col = !matches("peakid")) |>
-    mutate(group = sub("_rep[12]$", "", name),
-           name = factor(name, levels = unique(name)))
-
-ylims <- c(-2.0, 4.0)
-yticks <- c(-2, 0, 2, 4)
-
-pL <- lapply(levels(fct_relevel(sort(unique(pd$group)), "Untagged", after = Inf)),
-             function(grp1) {
-    pd1 <- pd[pd$group == grp1, ] |> pivot_wider()
-    pd1$cols <- densCols(x = pd1[[paste0(grp1, "_rep1")]],
-                         y = pd1[[paste0(grp1, "_rep2")]],
-                         nbin = 64, colramp = colorRampPalette(hcl.colors(32)))
-    ggplot(pd1, aes(.data[[paste0(grp1, "_rep1")]],
-                    .data[[paste0(grp1, "_rep2")]])) +
-        geom_abline(intercept = 0, slope = 1, linetype = 1, color = "gray") +
-        geom_hline(yintercept = 0, linetype = 2, color = "gray") +
-        geom_vline(xintercept = 0, linetype = 2, color = "gray") +
-        geom_scattermost(xy = as.data.frame(pd1[, paste0(grp1, c("_rep1", "_rep2"))]),
-                         pointsize = 2, color = pd1$cols, pixels = c(256, 256)) +
-        coord_fixed(xlim = ylims, ylim = ylims, expand = FALSE, clip = "off") +
-        scale_x_continuous(breaks = yticks) +
-        scale_y_continuous(breaks = yticks) +
-        theme_cowplot(7) +
-        labs(x = element_blank(), y = element_blank()) +
-        annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.05,
-                 label = grp1, color = "black", size = 3) +
-        theme(legend.position = "none")
-})
-
-(gg_enrpairs <- plot_grid(plotlist = pL, nrow = 9, ncol = 9))
+### Heatmap (not showing the non-bound TF promoters)
+
+We can also create the same heatmap, but without showing the TF promoters (rows)
+that are not detected to be bound by any TF.
+
+```{r plot-autoregulation-heatmap-without-nonbound}
+#| fig.width: 5
+#| fig.height: 3
+
+# ... remove "Non-bound"
+keep <- promtype[o] != "Non-bound\nTF promoters"
+o <- o[keep]
+
+# ... create main heatmap with annotations
+hm6 <- Heatmap(mprom[o, oc], name = "hm6",
+               col = circlize::colorRamp2(
+                   breaks = seq(0, 2, length.out = 64),
+                   colors = colorRampPalette(c(binary_heatmap_colors["FALSE"],
+                                               main_colors[1],
+                                               binary_heatmap_colors["TRUE"]))(64)),
+               border = TRUE, border_gp = gpar(lwd = 0.5),
+               cluster_rows = FALSE, show_row_dend = FALSE,
+               cluster_columns = FALSE, show_column_dend = FALSE,
+               row_split = droplevels(promtype[o]), row_title_gp = gpar(fontsize = fl),
+               column_title = "TF ChIP-seq experiments",
+               column_title_gp = gpar(fontsize = fl),
+               show_row_names = FALSE, show_column_names = FALSE,
+               right_annotation = rowLabels[o],
+               use_raster = TRUE, show_heatmap_legend = FALSE,
+               width = unit(w_peaksHm, "mm"), # heatmap body width
+               height = unit(w_peaksHm * mean(keep), "mm"))
+
+hm_prom_bound <- grid.grabExpr(
+    {
+        hm6 <- draw(hm6, merge_legend = TRUE,
+                    heatmap_legend_side = "bottom",
+                    annotation_legend_side = "bottom")
+        decorate_heatmap_body("hm6", {
+            grid.lines(c(mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)])), 0),
+                       c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
+        }, slice = 1)
+        # decorate_heatmap_body("hm6", {
+        #     grid.lines(c(1, mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)]))),
+        #                c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
+        # }, slice = 2)
+    },
+    width = w_peaksHm, height = w_peaksHm * mean(keep)
+)
+plot_grid(hm_prom_bound)
+```
+
+### Network
+
+We construct a network in which the nodes represent TFs (shown in orange if
+they were ChIP'ed in this study), and the edges represent detected binding
+events. An edge starting at a TF `A` and ending at a TF `B` means that `A`
+has been detected to bind the promoter of `B`.
+
+We exclude `untagged`, and also the TFs `zas1` and `SPCC320.03` from the network.
+`zas1` has `r sum(mprom["zas1", ] > 0)` bound TFs in its promoter, and 
+`SPCC320.03` has `r sum(mprom["SPCC320.03", ] > 0)`, and if these would be
+included, the network is hard to visualize.
+
+```{r make-network}
+#| fig.width: 10
+#| fig.height: 8
+
+## Create graph
+## ... select TFs
+to_remove <- c("zas1", "SPCC320.03", "untagged")
+sel_tfs <- setdiff(unique(rownames(mprom), colnames(mprom)), to_remove)
+length(sel_tfs)
+
+## ... create adjacency matrix (row: from TF, column: to TF promoter)
+adjm <- matrix(0, nrow = length(sel_tfs), ncol = length(sel_tfs),
+               dimnames = list(sel_tfs, sel_tfs))
+adjm[setdiff(colnames(mprom), to_remove),
+     setdiff(rownames(mprom), to_remove)] <- t(mprom[setdiff(rownames(mprom), to_remove),
+                                                     setdiff(colnames(mprom), to_remove)] > 0)
+gr <- graph_from_adjacency_matrix(adjm, mode = "directed")
+
+## Remove singletons
+comps <- components(gr)
+too_small <- which(table(comps$membership) < 2)
+keep <- V(gr)[!(comps$membership %in% too_small)]
+gr <- induced_subgraph(gr, keep)
+
+## Set attributes for vertices and edges
+V(gr)$vertexclass <- ifelse(adjm[cbind(names(V(gr)), names(V(gr)))] > 0,
+                            "Auto-regulated TF",
+                            ifelse(names(V(gr)) %in% colnames(mprom),
+                                   "ChIP'ed TF", "Other TF"))
+V(gr)$shape <- "circle"
+eL <- as_edgelist(gr)
+recipr <- paste0(eL[,1], "-", eL[,2]) %in% paste0(eL[,2], "-", eL[,1]) & eL[,1] != eL[,2]
+E(gr)$weight <- ifelse(recipr, 2, 1)
+E(gr)$edgeclass <- ifelse(recipr, "Reciprocal", "Other")
+E(gr)$color <- ifelse(recipr, gplots::col2hex(main_colors[3]), gplots::col2hex(na_color))
+
+## Visualize
+node_padding <- margin(.2, .2, .2, .2, "mm")
+ggr <- ggraph(as_tbl_graph(gr), 
+              layout = "igraph", algorithm = "fr") + 
+    geom_edge_link(aes(start_cap = label_rect(node1.name, padding = node_padding),
+                       end_cap = label_rect(node2.name, padding = node_padding),
+                       color = edgeclass),
+                   arrow = arrow(length = unit(2, 'mm'))) +
+    geom_edge_loop(aes(start_cap = label_rect(node1.name, padding = node_padding),
+                       end_cap = label_rect(node2.name, padding = node_padding)),
+                   arrow = arrow(length = unit(2, 'mm'))) +  # Handle self-loops
+    scale_edge_color_manual(values = c(Reciprocal = main_colors[3], 
+                                       Other = gplots::col2hex(na_color)),
+                            name = "") + 
+    geom_node_point(size = 4, aes(fill = vertexclass, shape = vertexclass), 
+                    color = "black", stroke = 0.1) + 
+    scale_shape_manual(values = c("Other TF" = 21, "ChIP'ed TF" = 21, "Auto-regulated TF" = 22), name = "") +
+    scale_fill_manual(values = c("Other TF" = gplots::col2hex(bg_color),
+                                 "ChIP'ed TF" = gplots::col2hex(main_colors[5]),
+                                 "Auto-regulated TF" = main_colors[5]),
+                      name = "") + 
+    geom_node_text(aes(label = name), size = 3.5, repel = TRUE, vjust = 1.8) + 
+    theme_graph(base_family = "sans", plot_margin = margin(30, 10, 40, 10)) + 
+    guides(fill = guide_legend(override.aes = list(size = 4, alpha = 1))) + 
+    coord_cartesian(clip = "off") 
+print(ggr)
 ```
@@
 # prepare `data.frame`s with plotting data
+peaktype <- factor(gsub("[()]", "", sub(" ", "\n", peakgr$peaktype)),
+                   levels = c("Common\nubiquitous",
+                              "Common\nfrequent",
+                              "Specific\npeaks"))
+sel_peaksEnr <- rowSums(as.matrix(mcols(peakgr)[, grep("^is_enr_in[.]", colnames(mcols(peakgr)))])) > 0
+near_ncRNA <- factor(ifelse(!is.na(dist.peak2tRNA) & dist.peak2tRNA < maxdist,
+                            "tRNA",
+                            ifelse(!is.na(dist.peak2rRNA) & dist.peak2rRNA < maxdist,
+                                   "5S rRNA", "none")),
+                         levels = c("tRNA", "5S rRNA", "none"))
 df_peaks <- data.frame(peaktype = peaktype[sel_peaksEnr],
@@
 # generate plots
+cols_ncRNA <- c("tRNA" = main_colors[5], "5S rRNA" = main_colors[3],
+                "none" = bg_color)
 # ... numbers of peaks near genes
@@
 #| fig.width: 11.5
-#| fig.height: 16
+#| fig.height: 8.2
 #| warning: false
@@
 cowplot::plot_grid(
-    hm_peaksBin,
-    hm_peaksEnr,
     hm_prom,
@@
         nrow = 1, rel_widths = c(1.25, 1)),
-    nrow = 3,
-    labels = c("A", "B", "C", "D", "E", "F"),
+    nrow = 2,
+    labels = c("A", "B", "C", "D"),
     align = "vh",
@@
     rel_widths = c(1, 1),
-    rel_heights = c(h_peaksHm + 10, w_peaksHm + 10, w_peaksHm + 20)
+    rel_heights = c(w_peaksHm + 10, w_peaksHm + 20)
 )
@@
 #| fig.width: 11
-#| fig.height: 20
+#| fig.height: 9
 #| warning: false
@@
 cowplot::plot_grid(
-    cowplot::plot_grid(gg_enrpairs, scale = 0.9, labels = "A"),
     cowplot::plot_grid(NULL, NULL, NULL, # space for labels
-                       nrow = 1, labels = c("B", "C", "D"),
+                       nrow = 1, labels = c("A", "B", "C"),
                        rel_widths = c(130, 100, 100)),
@@
     ),
-    nrow = 3,
+    nrow = 2,
     rel_widths = c(1),
-    rel_heights = c(12, 0.25, 9.75)
+    rel_heights = c(0.25, 9.75)
 )

### 2024-07-08 ms_figure_4_motifs.Rmd
update to new figure 4 (motifs)

@@
 # ... define heatmap font sizes
-fs <- 5.5
-fl <- 8
+fs <- 5.5 # small
+fl <- 8   # large
+ft <- 10  # title
 
@@
                     column_names_gp = gpar(fontsize = fs, fontfamily = "mono"),
+                    column_title = "Binuclear zinc cluster TFs",
+                    column_title_side = "top", column_title_gp = gpar(fontsize = ft),
                     border = TRUE, border_gp = gpar(lwd = 0.5),
@@
 
+```{r peak-overlaps-within-tf-family}
+#| fig.width: 10
+#| fig.height: 8
+
+# get enrichment calls for specific peaks that are
+# bound by at least one of the binuclear zinc cluster TFs used above
+pd <- as.data.frame(is_enr_specific[, tf_names]) |>
+    tibble::rownames_to_column("peak_name") |>
+    pivot_longer(cols = all_of(tf_names),
+                 values_to = "enriched",
+                 names_to = "tf_name") |>
+    filter(enriched) |>
+    group_by(peak_name) |>
+    summarise(tfs_enr = list(tf_name))
+
+# create upset plot
+upset_ZnII2Cys6_sel <- ggplot(pd, aes(x = tfs_enr)) +
+        geom_bar() +
+        scale_x_upset() +
+        labs(x = element_blank(),
+             y = "Number of enriched peaks") +
+        theme_cowplot()
+upset_ZnII2Cys6_sel
+```
+
 ## Known and new motif characterization
@@
 # select motifs for detailed visualization
-sel_motifs <- c(
-    # known
-    "Cdc10.m1",
-    # novel
-    "Toe2.m1", "Ams2.m1", "Esc1.m1",
-    "SPAC3F10.12c.m1", "Zip1.m1", "Adn2.m4")
+sel_motifs <- c( "Toe2.m1", "Esc1.m1", "SPAC3F10.12c.m1",
+                 "Ams2.m1", "Zip1.m1", "Adn2.m4")
 tf_names <- sub("[.]m[0-9]+$", "", sel_motifs)
@@
 fgbg_col <- c(main_colors[1], main_colors[5]) # colors for foreground and background
-gnm_col <- na_color # color for NA or "other"
+gnm_col <- bg_color # color for NA or "other"
 w <- 2500  # width in bases around peak-center for meta-profiles
@@
                   fontface = "bold") +
-        labs(x = "Fraction of motif sites",
+        labs(x = "Percent of motif sites",
              y = element_blank(),
@@
 #| fig.width: 11
-#| fig.height: 15.5
+#| fig.height: 14
 
@@
     axis = "b",
-    rel_heights = c(1.4, 3.7)
+    rel_heights = c(1.4, 3.2)
 )

### 2024-07-08 ms_figure_2_overview-heatmaps.Rmd
remove unneeded packages

@@
     library(scattermore)
-    # library(ggrepel)
-    # library(DT)
-    # library(stringr)
     library(tidyr)
-    # library(kableExtra)
-    # library(patchwork)
 })

### 2024-07-09 ms_figure_3_chip.Rmd
remove IP-MS HOT TF heatmap from figure 3

@@
     txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
-    ipms150untag: "data/ipms_150_untag_sce.rds"
 editor_options: 
@@
 source("params/plot_settings.R")
-source("params/get_testres_function.R")
-source("params/mapping_functions.R")
 source("params/plottracks_function.R")
@@
 
-## IP-MS data for HOT TFs
-
-```{r ipms-hot}
-#| fig.width: 9
-#| fig.height: 9
-
-sce150untag <- readRDS(params$ipms150untag)
-
-res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
-tstat <- res150untag$tstat
-colnames(tstat) <- .getProteinNameFromComparison(colnames(tstat))
-tstat <- tstat[rownames(tstat) %in% colnames(tstat), ]
-tstat[is.na(tstat)] <- 0
-dim(tstat)
-
-hot_tfs <- c("Php3", "Sak1", "Pcr1", "Prr1", "Atf1", "Rst2", "Adn2", 
-             "Adn3", "Hsr1", "Phx1", "Pho7")
-stopifnot(all(hot_tfs %in% rownames(tstat)))
-stopifnot(all(hot_tfs %in% colnames(tstat)))
-stopifnot(all(rownames(tstat) %in% colnames(tstat)))
-tstat <- tstat[c(hot_tfs, setdiff(rownames(tstat), hot_tfs)), ]
-tstat <- tstat[, c(rownames(tstat), setdiff(colnames(tstat), rownames(tstat)))]
-
-hmipmshot <- Heatmap(
-    tstat, 
-    col = circlize::colorRamp2(
-        breaks = seq(3.0, 13.0, length.out = 64),
-        colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
-    border = TRUE,
-    row_split = ifelse(rownames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"), 
-    column_split = ifelse(colnames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"),
-    cluster_rows = FALSE, cluster_columns = FALSE,
-    row_title_gp = gpar(fontsize = fl), row_names_gp = gpar(fontsize = fs),
-    column_title_gp = gpar(fontsize = fl), show_column_names = FALSE,
-    heatmap_legend_param = list(labels_gp = gpar(fontsize = fs),
-                                border = "gray10",
-                                legend_direction = "horizontal",
-                                legend_width = unit(24, "mm"),
-                                title_position = "lefttop",
-                                title_gp = gpar(fontsize = fl)),
-    na_col = bg_color,
-    name = "Mod. t-stat.")
-
-hm_ipms_hot <- grid.grabExpr(
-    hmipmshot <- draw(hmipmshot, merge_legend = TRUE, 
-                      heatmap_legend_side = "bottom")
-)
-plot_grid(hm_ipms_hot)
 ```

### 2024-07-09 ms_figure_3_chip.Rmd
create long format motif similarity heatmap

@@
 
-hm_prom <- grid.grabExpr(
-    {
+hm_prom <- grid.grabExpr({
         hm5 <- draw(hm5, merge_legend = TRUE,

### 2024-07-09 ms_figure_3_chip.Rmd
add RNA-seq analysis of (non-)bound promoter genes

--- 2024-07-09 ms_figure_3_chip.Rmd
@@
 
+## Bound promoters
+
+In `txannot$prom.number.overlapping.highconf.peaks`, we record how many TF peaks
+are found in the promoter regions of a gene (defined as a 1kb window
+centered on the annotated transcript start site).
+
+Here we compare the expression levels (obtained from `rnaseq_del`) of
+genes twith different numbers of TFs bound at their promoters.
+
+```{r bound-promoters}
+#| fig.width: 6
+#| fig.height: 4
+
+# calculate average expression of each gene
+avg_tpm <- rowMeans(assay(rnaseq_del, "abundance"))
+
+# add average expression to gene annotation and filter out
+# genes that are not protein coding genes
+pd <- txannot |>
+    mutate(avg_tpm = avg_tpm[unique_einprot_id]) |>
+    filter(!is.na(avg_tpm),
+           gene_biotype == "protein_coding") |>
+    mutate(prom.number.overlapping.highconf.peaks.categories = 
+               cut(prom.number.overlapping.highconf.peaks,
+                   breaks = c(-Inf, 0, 1, Inf),
+                   labels = c("0", "1", ">1"),
+                   right = TRUE))
+
+# visualize
+ggplot(pd, aes(prom.number.overlapping.highconf.peaks.categories, avg_tpm + 0.01)) +
+    geom_violin(draw_quantiles = 0.5) +
+    scale_y_log10() +
+    labs(x = "Number of promoter TF peaks",
+         y = "RNA level (TPM + 0.01, log-scale)") +
+    theme_cowplot()
 ```

### 2024-07-09 ms_figure_3_chip.Rmd
further updates to figure 3

@@
     motifrds: "data/motifs_annotated.rds"
+    rdsrnaseq_del: "data/rnaseq_del.rds"
     txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
@@
 
+# make first letter of a string uppercase
+.capitalize <- function(x) {
+    paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
+}
+
 # for each region in `from`, calculate the distance to its nearest element in `to`
@@
 motifs <- readRDS(params$motifrds)
+
+# `rnaseq_del`: a `SummarizedExperiment` with RNA-seq quantifications
+rnaseq_del <- readRDS(params$rdsrnaseq_del)
 
@@
 
+We create the same heatmap in long format, allowing to label all rows:
+
+```{r}
+#| fig.width: 6
+#| fig.height: 8
+
+hm4_long <- Heatmap(matrix = motifsim2, name = "Motif\nsimilarity",
+               col = circlize::colorRamp2(breaks = seq(rng[1], rng[2], length.out = 64),
+                                          colors = cols),
+               cluster_rows = motifcl,
+               show_row_dend = TRUE, row_dend_width = unit(10, "mm"),
+               cluster_columns = motifcl, show_column_dend = FALSE,
+               row_split = n_split, column_split = n_split,
+               column_title = " ",
+               row_gap = unit(0.2, "mm"), column_gap = unit(0.2, "mm"),
+               show_row_names = TRUE, row_names_side = "right",
+               column_title_gp = gpar(fontsize = fl),
+               row_title = " ",
+               row_names_gp = gpar(fontsize = fs),
+               show_column_names = FALSE,
+               heatmap_legend_param = list(
+                   at = round(c(rng[1], mean(rng), rng[2]), 2),
+                   labels_gp = gpar(fontsize = fl),
+                   border = "gray10",
+                   legend_direction = "horizontal",
+                   legend_width = unit(18, "mm"),
+                   title_position = "lefttop",
+                   title_gp = gpar(fontsize = fl)),
+               width = unit(w_peaksHm * 0.5, "mm"), # heatmap body width
+               height = unit(h_peaksHm * 1.15, "mm")) # heatmap body height
+
+set.seed(42L)
+hm_motifs_long <- grid.grabExpr(
+    hm4_long <- draw(hm4_long, merge_legend = FALSE,
+                     heatmap_legend_side = "bottom",
+                     annotation_legend_side = "bottom"),
+    width = w_peaksHm * 0.5, height = h_peaksHm * 1.15
+)
+
+plot_grid(hm_motifs_long)
+```
+
 Here is the complete list of motifs for each of the motif clusters,
@@
 
-hm_prom_bound <- grid.grabExpr(
-    {
-        hm6 <- draw(hm6, merge_legend = TRUE,
-                    heatmap_legend_side = "bottom",
-                    annotation_legend_side = "bottom")
+# manually create a legend
+hm_legend <- Legend(labels = c("TFs binding to another TF promoter",
+                               "TFs binding to their own promoter"),
+                    title = " ", legend_gp = gpar(fill = c("#648FFF", "darkblue")))
+
+hm_prom_bound <- grid.grabExpr({
+        hm6 <- draw(hm6,
+                    annotation_legend_side = "bottom",
+                    annotation_legend_list = packLegend(hm_legend))
         decorate_heatmap_body("hm6", {
@@
         }, slice = 1)
-        # decorate_heatmap_body("hm6", {
-        #     grid.lines(c(1, mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)]))),
-        #                c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
-        # }, slice = 2)
     },
@@
 
+### Network around `Atf1`
+
+For clearer visualization, we create a sub-network including only the
+connectsion to and from `Atf1`:
+
+```{r make-network-atf1}
+#| fig.width: 10
+#| fig.height: 8
+
+## Create graph
+## ... identify TFs connected to atf1
+to_keep <- rownames(adjm)[adjm["atf1", ] > 0 | adjm[, "atf1"] > 0]
+length(to_keep)
+
+## ... create adjacency matrix (row: from TF, column: to TF promoter)
+adjm2 <- adjm[to_keep, to_keep]
+gr2 <- graph_from_adjacency_matrix(adjm2, mode = "directed")
+
+## Set attributes for vertices and edges
+V(gr2)$vertexclass <- ifelse(adjm[cbind(names(V(gr2)), names(V(gr2)))] > 0,
+                             "Auto-regulated TF",
+                             ifelse(names(V(gr2)) %in% colnames(mprom),
+                                    "ChIP'ed TF", "Other TF"))
+V(gr2)$shape <- "circle"
+eL <- as_edgelist(gr2)
+recipr <- paste0(eL[,1], "-", eL[,2]) %in% paste0(eL[,2], "-", eL[,1]) & eL[,1] != eL[,2]
+E(gr2)$weight <- ifelse(recipr, 2, 1)
+E(gr2)$edgeclass <- ifelse(recipr, "Reciprocal", "Other")
+E(gr2)$color <- ifelse(recipr, gplots::col2hex(main_colors[3]), gplots::col2hex(na_color))
+
+## Visualize
+node_padding <- margin(.2, .2, .2, .2, "mm")
+ggr2 <- ggraph(as_tbl_graph(gr2), 
+               layout = "igraph", algorithm = "fr") + 
+    geom_edge_link(aes(start_cap = label_rect(node1.name, padding = node_padding),
+                       end_cap = label_rect(node2.name, padding = node_padding),
+                       color = edgeclass),
+                   arrow = arrow(length = unit(2, 'mm'))) +
+    geom_edge_loop(aes(start_cap = label_rect(node1.name, padding = node_padding),
+                       end_cap = label_rect(node2.name, padding = node_padding)),
+                   arrow = arrow(length = unit(2, 'mm'))) +  # Handle self-loops
+    scale_edge_color_manual(values = c(Reciprocal = main_colors[3], 
+                                       Other = gplots::col2hex(na_color)),
+                            name = "") + 
+    geom_node_point(size = 4, aes(fill = vertexclass, shape = vertexclass), 
+                    color = "black", stroke = 0.1) + 
+    scale_shape_manual(values = c("Other TF" = 21, "ChIP'ed TF" = 21, "Auto-regulated TF" = 22), name = "") +
+    scale_fill_manual(values = c("Other TF" = gplots::col2hex(bg_color),
+                                 "ChIP'ed TF" = gplots::col2hex(main_colors[5]),
+                                 "Auto-regulated TF" = main_colors[5]),
+                      name = "") + 
+    geom_node_text(aes(label = name), size = 3.5, repel = TRUE, vjust = 1.8) + 
+    theme_graph(base_family = "sans", plot_margin = margin(30, 10, 40, 10)) + 
+    guides(fill = guide_legend(override.aes = list(size = 4, alpha = 1))) + 
+    coord_cartesian(clip = "off") 
+print(ggr2)
+```
+
 ## Peak types and presence of nearby ncRNAs
@@
 
-Assemble the panels into Figure 2:
-
-```{r Fig2}
+Assemble the panels into Figure 3:
+
+```{r Fig3}
 #| fig.width: 11.5
@@
 cowplot::plot_grid(
-    hm_prom,
+    hm_prom_bound,
     cowplot::plot_grid(gnmtracks1, scale = 0.9),
@@
 
-Assemble the panels into Supplementary Figure 2:
-
-```{r suppFig2}
+Assemble the panels into Supplementary Figure 3:
+
+```{r suppFig3}
 #| fig.width: 11
@@
                        nrow = 1, labels = c("A", "B", "C"),
-                       rel_widths = c(130, 100, 100)),
+                       rel_widths = c(100, 100, 100)),
     cowplot::plot_grid(
-        hm_ipms_hot,
-        cowplot::plot_grid(gnmtracks2, NULL, ncol = 1, rel_heights = c(12, 1)),
+        cowplot::plot_grid(gnmtracks2, NULL, ncol = 1, rel_heights = c(20, 1)),
+        hm_motifs_long,
         p_fracbound,
         nrow = 1,
-        labels = c("", "", ""),
-        align = "v",
-        axis = "b",
-        rel_widths = c(130, 100, 100),
-        rel_heights = c(180)
+        labels = c("", "", "")
     ),

### 2024-07-09 ms_figure_6_networks.Rmd
Add Pho7 volcano plot (Fig 6)

@@
     rds150: "data/ipms_150_sce.rds"
-    coords150: "data/visnetwork_coords_150_new.csv"
+    coords150: "data/visnetwork_coords_150.csv"
     rds500: "data/ipms_500_sce.rds"
-    coords500: "data/visnetwork_coords_500_new.csv"
+    coords500: "data/visnetwork_coords_500.csv"
     rds150untag: "data/ipms_150_untag_sce.rds"
@@
 source("params/mapping_functions.R")
+source("params/ipms_params.R")
 ```
@@
 stopifnot(all(gnames %in% rad24_rad25_tbl$TF))
-ft <- flextable(rad24_rad25_tbl) |>
+ftbl <- flextable(rad24_rad25_tbl) |>
   width(width =  c(0.5, 1.25)) |>
@@
     theme_void() +
-    annotation_custom(grid::rasterGrob(as_raster(ft)),
+    annotation_custom(grid::rasterGrob(as_raster(ftbl)),
                       xmin = -Inf, xmax = Inf, ymin = 0, ymax = 1)
@@
     guides(fill = guide_legend(nrow = 1))
+```
+
+# Pho7 volcano plot
+
+```{r}
+md <- metadata(sce500)
+volc <- md$testres$tests$Pho7_500_plate_vs_compl_Pho7_500_plate |>
+    mutate(signif = adj.P.Val < adjpThreshold & logFC > log2fcThreshold)
+(pho7volc <- ggplot(volc, aes(x = logFC, y = mlog10p)) + 
+        geom_text_repel(data = volc |> filter(signif),
+                        aes(label = pid), min.segment.length = 0,
+                        max.overlaps = Inf) + 
+        geom_point(aes(alpha = signif, color = signif), 
+                   show.legend = c(alpha = FALSE, color = TRUE)) + 
+        scale_color_manual(values = c(`TRUE` = main_colors[5], `FALSE` = na_color), 
+                           labels = c(`TRUE` = paste0("adj. p < ", adjpThreshold, 
+                                                      " and logFC > ", log2fcThreshold)),
+                           na.value = "black", breaks = "TRUE") + 
+        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.5)) + 
+        geom_point(data = volc |> dplyr::filter(pid == "Pho7") |>
+                       dplyr::mutate(type = "Bait"), 
+                   mapping = aes(fill = type), color = main_colors[3], 
+                   shape = 21, alpha = 1, size = 2) + 
+        scale_fill_manual(values = c(Bait = main_colors[3])) + 
+        labs(x = "log2(fold-change)", y = "-log10(p-value)", 
+             title = "Pho7 vs complement (500 mM NaCl)") + 
+        theme_cowplot(ft) + 
+        theme(legend.position = "bottom") + 
+        guides(color = guide_legend(nrow = 1, byrow = TRUE, title = ""),
+               fill = guide_legend(nrow = 1, byrow = TRUE, title = "")))
 ```
@@
 #| fig.width: 10
-#| fig.height: 10
+#| fig.height: 12
+#| warning: false
 
@@
 coords500 <- read.csv(params$coords500)
+coords150$y <- -coords150$y
 coords500$y <- -coords500$y
@@
     ),
+    cowplot::plot_grid(
         cowplot::plot_grid(fig6c, ggrad_upset,
-                       nrow = 1, labels = c("C", "D")), 
-    rel_heights = c(5, 2), 
+                           ncol = 1, labels = c("C", "D"),
+                           rel_heights = c(1, 1.25)), 
+        pho7volc, 
+        nrow = 1, labels = c("", "E")
+    ),
+    rel_heights = c(5, 3.5), 
     ncol = 1

### 2024-07-09 ms_figure_7_heterodimer.Rmd
Add genome logfc plot (Fig 7)

@@
 ---
-title: "Figure 6 - heterodimer"
+title: "Figure 7 - heterodimer"
 author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
@@
     rdsrnaseq_ko: "data/rnaseq_ko.rds"
-    dist_from_ne: "data/Fig6_lacR_quantification_distance_radius.txt"
+    dist_from_ne: "data/Fig7_lacR_quantification_distance_radius.txt"
     gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
@@
 
-# Figure 6
+# Figure 7
 
@@
 
+```{r rnaseq-genomeplot}
+
+## Add information about gene location to result lists
+genepos <- rowRanges(dds)
+mcols(genepos) <- NULL
+genepos <- as.data.frame(genepos)
+resList <- lapply(resList, function(rl) {
+    stopifnot(rownames(rl) == rownames(genepos))
+    cbind(rl, genepos) |>
+        as.data.frame() |>
+        tibble::rownames_to_column("gene")
+})
+
+## Subset to region of interest
+chrom <- genepos["tna1", "seqnames"]
+startpos <- genepos["tna1", "start"]
+endpos <- genepos["tna1", "end"]
+windowsize <- 30000
+
+resListToPlot <- lapply(resList, function(rl) {
+    rl |> dplyr::filter(seqnames == chrom & 
+                            start > startpos - windowsize & 
+                            end < endpos + windowsize)
+})
+plotList <- lapply(names(resListToPlot), function(nm) {
+    ggplot(resListToPlot[[nm]], 
+           aes(x = start, y = log2FoldChange, 
+               color = abs(log2FoldChange) > 3 & padj < 0.001)) + 
+        geom_rect(aes(xmin = endpos, xmax = startpos, ymin = -Inf, ymax = Inf),
+                  color = lighten(main_colors[4], amount = 0.6), fill = NA,
+                  alpha = 0.1, show.legend = FALSE) + 
+        geom_point(alpha = 0.6, size = 2) + 
+        scale_color_manual(values = c(`FALSE` = "grey50", 
+                                      `TRUE` = main_colors[4])) + 
+        geom_text_repel(aes(label = ifelse(abs(log2FoldChange) > 3 & padj < 0.001,
+                                           as.character(gene), "")),
+                        size = 2.5, max.overlaps = 30, force = 10) + 
+        scale_x_continuous(labels = scales::label_number(scale = 1e-6)) + 
+        scale_y_continuous(limits = c(-7, 7)) + 
+        labs(x = "Genome position (Mb)", y = "log2(fold-change)",
+             title = paste0(nm, "\U141E")) + 
+        theme_cowplot() + 
+        theme(legend.position = "none",
+              plot.title = element_text(face = "italic"))
+})
+genomeplot <- cowplot::plot_grid(plotlist = plotList, ncol = 1)
+```
+
 ## Coverage tracks
@@
     geom_area(aes(group = strand, fill = gene)) + 
-    scale_fill_manual(values = c(wt = "black", "ntu1\U141E" = main_colors[5], 
-                                 "ntu2\U141E" = main_colors[1])) + 
+    scale_fill_manual(values = c(wt = lighten("black", amount = 0.4), 
+                                 "ntu1\U141E" = lighten(main_colors[5], amount = 0.4), 
+                                 "ntu2\U141E" = lighten(main_colors[1], amount = 0.4))) + 
     facet_grid(sample ~ seqnames, scale = "free_x", space = "free_x",
@@
 
-```{r Fig6}
+```{r Fig7}
 #| fig.width: 13
@@
 
-fig6b <- ggdraw() + 
-    draw_image("schematics/Fig6B_Nattou_protein_domains_cartoon_Arial.png")
-fig6e <- ggdraw() + 
-    draw_image("schematics/Fig6E_Nattou_lacR_cartoon_images_Arial.png")
-fig6f <- ggdraw() + 
-    draw_image("schematics/Fig6F_Nattou_lacR_microscopy_images_Arial.png")
+fig7b <- ggdraw() + 
+    draw_image("schematics/Fig7B_Nattou_protein_domains_cartoon_Arial.png")
+fig7e <- ggdraw() + 
+    draw_image("schematics/Fig7E_Nattou_lacR_cartoon_images_Arial.png")
+fig7f <- ggdraw() + 
+    draw_image("schematics/Fig7F_Nattou_lacR_microscopy_images_Arial.png")
 
-fig6c <- wrap_plots(maplots, ncol = 1) | corrplot[[1]]
+fig7c <- wrap_plots(maplots, ncol = 1) | corrplot[[1]]
 
@@
         gg_ipdimer, 
-        fig6c,
+        fig7c,
         nrow = 1, 
@@
     cowplot::plot_grid(
-        fig6b, 
+        fig7b, 
         gcov,
@@
     cowplot::plot_grid(
-        fig6e,
+        fig7e,
         NULL,
-        fig6f,
+        fig7f,
         NULL,
@@
 
-```{r suppFig6-chip}
+```{r suppFig7-chip}
 #| warning: false
@@
 
-```{r suppFig6}
+```{r suppFig7}
 #| fig.width: 13
-#| fig.height: 9
+#| fig.height: 15
 #| warning: false
@@
 figs6a <- ggdraw() + 
-    draw_image("schematics/FigS6A_Nattou_AF2_fl_Arial.png")
+    draw_image("schematics/FigS7A_Nattou_AF2_fl_Arial.png")
 
 cowplot::plot_grid(
+    figs6a,
     cowplot::plot_grid(
-        figs6a, 
-        NULL,
         p_rel_dis,
+        genomeplot_tpm,
         nrow = 1, 
-        rel_widths = c(1.55, 0.25, 1.45), 
-        labels = c("A", "", "B")
+        rel_widths = c(1, 1), 
+        labels = c("B", "C")
     ),
     cowplot::plot_grid(
-        genomeplot_tpm,
+        genomeplot,
         gcovdel,
@@
         nrow = 1, 
-        rel_widths = c(1.2, 1),
-        labels = c("C", "D")
+        rel_widths = c(1, 1),
+        labels = c("C", "E")
     ),
-    ncol = 1, rel_heights = c(1, 1)
+    ncol = 1, rel_heights = c(1, 1, 1)
 )

### 2024-07-09 ms_figure_2_overview-heatmaps.Rmd
Lift out IP-MS heatmap functions into scripts, move HOT IP-MS heatmap to Supp Fig 2

@@
     gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
+    ipms150untag: "data/ipms_150_untag_sce.rds"
 ---
@@
 source("params/mapping_functions.R")
+source("params/ipms_heatmap_functions.R")
 
@@
 ```{r heatmap-helper}
-filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold, 
-                             logfcThreshold, maxNbrTargets) {
-    ## Filter - require significant in at least one comparison
-    tstatfilt <- tstatmat
-    tstatfilt[adjpmat >= adjpThreshold] <- NA
-    tstatfilt[logfcmat <= logfcThreshold] <- NA
-    # tstatfilt[tstatfilt <= 0] <- NA
-
-    ## Some experiments pull down a large number of proteins
-    # print(table(colSums(!is.na(tstatfilt))))
-    ## -> remove proteins that are only being pulled down in these experiments
-    keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
-    tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
-    tstatfilt[is.na(tstatfilt)] <- 0
-    tstatfilt <- tstatfilt[rowSums(tstatfilt) > 0, ]
-
-    ## Subset the t-statistic matrix to only the retained proteins
-    stopifnot(colnames(tstatfilt) == colnames(tstatmat))
-    tstatmat <- tstatmat[rownames(tstatfilt), colnames(tstatfilt)]
-    colnames(tstatmat) <- .getSimplifiedComparison(colnames(tstatmat))
-
-    list(tstats = tstatmat, tstatsfilt = tstatfilt)
-}
-
-makeComplexAnnotation <- function(tstatmat, complexes, idmap,
-                                  show_legend = TRUE) {
-    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
-                           row.names = rownames(tstatmat))
-    rowcols <- list(Complex = c())
-    for (cmplx in names(complexes)) {
-        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
-                                                  idmap = idmap) %in% 
-                             complexes[[cmplx]]$pombase_ids] <- cmplx
-        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
-    }
-    rowAnnotation(df = rowannot, 
-                  col = rowcols, 
-                  annotation_name_gp = gpar(fontsize = fl),
-                  simple_anno_size_adjust = TRUE,
-                  na_col = bg_color,
-                  show_annotation_name = FALSE,
-                  show_legend = show_legend,
-                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
-}
-
-makeColLabels <- function(tstatmat, colLabel, fontsize) {
-    columnAnnotation(
-        TFnames = anno_mark(which = "column", side = "bottom",
-                            at = match(colLabel, colnames(tstatmat)), 
-                            labels = .getProteinNameFromSimplifiedComparison(colLabel),
-                            labels_gp = gpar(fontsize = fontsize)))
-}
-
-makeHeatmapCol <- function(tstatmat, stringency = "high") {
-    if (stringency == "high") {
-        circlize::colorRamp2(
-            breaks = seq(3.0, 13.0, length.out = 64),
-            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
-    } else {
-        circlize::colorRamp2(
-            breaks = seq(1.5, 13.0, length.out = 64),
-            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
-    }
-}
-
-reorderColsByRows <- function(tstat) {
-    ## Row names should be protein names, column names simplified comparison names
-    ord <- match(rownames(tstat), .getProteinNameFromSimplifiedComparison(colnames(tstat)))
-    ord <- ord[!is.na(ord)]
-    ord <- c(ord, setdiff(seq_len(ncol(tstat)), ord))
-    tstat[, ord]
-}
-
-makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
-    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)
-
-    tstats <- filterForHeatmap(tstatmat = tc$tstat, adjpmat = tc$adjp, 
-                               logfcmat = tc$logfc, logfcThreshold = log2fcthr,
-                               adjpThreshold = adjpthr, maxNbrTargets = Inf)
-    tstat <- tstats$tstats
-
-    ## tstats$tstatsfile: protein x experiment
-    ## Number of TFs that are pulled down by each bait
-    all_tfs <- setdiff(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
-                       c("Untagged", "untagged", NA))
-    nbrInt <- colSums(tstats$tstatsfilt[rownames(tstats$tstatsfilt) %in% all_tfs, ] > 0, 
-                      na.rm = TRUE)
-    names(nbrInt) <- .getSimplifiedComparison(names(nbrInt))
-
-    ## Split by family - return a matrix with several columns: 
-    ## #pulled down TFs (including itself), #pulled down TFs (excluding itself),
-    ## #pulled down TFs from the same family (excluding itself), 
-    ## #pulled down TFs from another family
-    tmptf <- tstats$tstatsfilt[match(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
-                                     rownames(tstats$tstatsfilt)), ]
-    rownames(tmptf) <- .getProteinNameFromComparison(colnames(tstats$tstatsfilt))
-    dbdtf <- data.frame(PomBaseID = .getPomBaseIdFromComparison(colnames(tstats$tstatsfilt), 
-                                                                idmap = idmap),
-                        bait = .getProteinNameFromComparison(colnames(tstats$tstatsfilt)),
-                        DBD_class = NA)
-    for (i in seq_len(nrow(dbdtf))) {
-        lookfor <- dbdtf$PomBaseID[i]
-        if (!(lookfor %in% c("untagged", "Untagged"))) {
-            dbdtf$DBD_class[i] <- dbd$DBD_class[match(lookfor, dbd$PomBaseID)]
-        } else {
-            dbdtf$DBD_class[i] <- "no_class"
-        }
-    }
-    stopifnot(dbdtf$bait == .getProteinNameFromComparison(colnames(tmptf)))
-    nbrIntMat <- do.call(dplyr::bind_rows, lapply(seq_along(colnames(tmptf)), function(i) {
-        expr <- colnames(tmptf)[i]
-        fam <- dbdtf$DBD_class[i]
-        data.frame(experiment = expr, 
-                   bait = dbdtf$bait[i], 
-                   dbdfam = fam, 
-                   TFs = paste(rownames(tmptf)[-i][which(tmptf[-i, expr] > 0)],
-                               collapse = ","),
-                   nbrTFs = sum(tmptf[, expr] > 0, na.rm = TRUE),
-                   nbrTFs_wobait = sum(tmptf[-i, expr] > 0, na.rm = TRUE),
-                   nbrTFs_samefam = sum(tmptf[-i, expr] > 0 & 
-                                            dbdtf$DBD_class[-i] == fam, 
-                                        na.rm = TRUE),
-                   nbrTFs_difffam = sum(tmptf[-i, expr] > 0 & 
-                                            dbdtf$DBD_class[-i] != fam, 
-                                        na.rm = TRUE)
-        )
-    }))
-
-    ## Split experiments into three groups
-    if (conc == 150) {
-        stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
-        tstat_no_bait <- 
-            tstat[, colSums(tstats$tstatsfilt) == 0 | 
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class == "nobait"])]
-        tstat_150only <- 
-            tstat[, colSums(tstats$tstatsfilt) > 0 &
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class == "150only"])]
-        tstat_150and500 <- 
-            tstat[, colSums(tstats$tstatsfilt) > 0 &
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class == "150and500"])]
-
-        proteins_150only <- .getProteinNameFromSimplifiedComparison(colnames(tstat_150only))
-        stopifnot(all(proteins_150only %in% rownames(tstat)))
-        tstat_150and500 <- tstat_150and500[!(rownames(tstat_150and500) %in% proteins_150only), ]
-
-        ## Cluster proteins based on the correlation between their t-stat vectors
-        dst <- sqrt(2 * (1 - cor(t(tstat_150and500), use = "pairwise.complete")))
-        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
-        clst <- hclust(as.dist(dst))
-
-        ## Reorder proteins by the clustering
-        protorder <- c(rownames(tstat_150and500)[clst$order], proteins_150only)
-        tstat <- tstat[protorder, ]
-
-        ## Reorder experiments to keep a "diagonal" in the heatmap
-        tstat <- reorderColsByRows(tstat)
-
-        tmpbaits <- .getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
-                                                             idmap = idmap)
-        colsplit <- ifelse(
-            tmpbaits %in% baitclass$Gene_name[baitclass$class == "nobait"], 
-            "No\nbait", 
-            ifelse(
-                tmpbaits %in% baitclass$Gene_name[baitclass$class == "150only"],
-                "150 mM NaCl",
-                ifelse(
-                    tmpbaits %in% baitclass$Gene_name[baitclass$class == "150and500"],
-                    "150 and\n500 mM NaCl", "NA"
-                )
-            )
-        )
-    } else if (conc == 500) {
-        tstat_onlybait <- 
-            tstat[, colSums(tstats$tstatsfilt > 0) == 1 |
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
-                                                      idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class500 == "Lost in 500 mM"])]
-        tstat_several <- 
-            tstat[, colSums(tstats$tstatsfilt > 0) > 1 & 
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
-                                                      idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class500 != "Lost in 500 mM"])]
-        tstat_onlybait <- tstat_onlybait[, colnames(tstat_onlybait) != "Untagged_plate"]
-        tstat_several <- tstat_several[, colnames(tstat_several) != "Untagged_plate"]
-
-        proteins_onlybait <-
-            .getProteinNameFromSimplifiedComparison(colnames(tstat_onlybait))
-        stopifnot(all(proteins_onlybait %in% rownames(tstat)))
-        tstat_several <- tstat_several[!(rownames(tstat_several) %in% proteins_onlybait), ]
-
-        ## Cluster proteins based on the correlation between their t-stat vectors
-        dst <- sqrt(2 * (1 - cor(t(tstat_several), use = "pairwise.complete")))
-        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
-        clst <- hclust(as.dist(dst))
-
-        ## Reorder proteins by the clustering
-        protorder <- c(rownames(tstat_several)[clst$order], proteins_onlybait)
-        tstat <- tstat[protorder, ]
-
-        ## Reorder conditions to keep a "diagonal" in the heatmap
-        tstat <- reorderColsByRows(tstat)
-
-        colsplit <- factor(
-            structure(baitclass$class500[
-                match(.getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
-                                                               idmap = idmap), 
-                      baitclass$Gene_name)],
-                names = colnames(tstat)),
-            levels = c("Retained in 500 mM", "Lost in 500 mM", ""))
-        levels(colsplit) <- c("Interactions\nin 500 mM NaCl",
-                              "No interactions\nin 500 mM NaCl", "")
-    } else {
-        stop("Unknown concentration")
-    }
-    return(list(tstat = tstat, colsplit = colsplit, nbrIntMat = nbrIntMat))
-}
-
 # for each region in `from`, calculate the distance to its nearest element in `to`
@@
 
+## IP-MS data for HOT TFs
+
+```{r ipms-hot}
+#| fig.width: 9
+#| fig.height: 9
+
+sce150untag <- readRDS(params$ipms150untag)
+
+res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
+tstat <- res150untag$tstat
+colnames(tstat) <- .getProteinNameFromComparison(colnames(tstat))
+tstat <- tstat[rownames(tstat) %in% colnames(tstat), ]
+tstat[is.na(tstat)] <- 0
+dim(tstat)
+
+hot_tfs <- c("Php3", "Sak1", "Pcr1", "Prr1", "Atf1", "Rst2", "Adn2", 
+             "Adn3", "Hsr1", "Phx1", "Pho7")
+stopifnot(all(hot_tfs %in% rownames(tstat)))
+stopifnot(all(hot_tfs %in% colnames(tstat)))
+stopifnot(all(rownames(tstat) %in% colnames(tstat)))
+tstat <- tstat[c(hot_tfs, setdiff(rownames(tstat), hot_tfs)), ]
+tstat <- tstat[, c(rownames(tstat), setdiff(colnames(tstat), rownames(tstat)))]
+
+hmipmshot <- Heatmap(
+    tstat, 
+    col = circlize::colorRamp2(
+        breaks = seq(3.0, 13.0, length.out = 64),
+        colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
+    border = TRUE,
+    row_split = ifelse(rownames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"), 
+    column_split = ifelse(colnames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"),
+    cluster_rows = FALSE, cluster_columns = FALSE,
+    row_title_gp = gpar(fontsize = fl), row_names_gp = gpar(fontsize = fs),
+    column_title_gp = gpar(fontsize = fl), show_column_names = FALSE,
+    heatmap_legend_param = list(labels_gp = gpar(fontsize = fs),
+                                border = "gray10",
+                                legend_direction = "horizontal",
+                                legend_width = unit(24, "mm"),
+                                title_position = "lefttop",
+                                title_gp = gpar(fontsize = fl)),
+    na_col = bg_color,
+    name = "Mod. t-stat.")
+
+hm_ipms_hot <- grid.grabExpr(
+    hmipmshot <- draw(hmipmshot, merge_legend = TRUE, 
+                      heatmap_legend_side = "bottom")
+)
+plot_grid(hm_ipms_hot)
+```
+
 ## Peak vs. sample heatmap (binary)
@@
 hm3 <- Heatmap(m[sel_peaksBin, ],
+               column_title = "ChIP (binary)",
+               column_title_side = "top",
+               column_title_gp = gpar(fontsize = ft),
                col = circlize::colorRamp2(
@@
 hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
+               column_title = "ChIP (enrichments)",
+               column_title_side = "top",
+               column_title_gp = gpar(fontsize = ft),
                col = circlize::colorRamp2(
@@
 #| fig.width: 9
-#| fig.height: 9
+#| fig.height: 18
 
@@
 
-(gg_enrpairs <- plot_grid(plotlist = pL, nrow = 9, ncol = 9))
+(gg_enrpairs <- plot_grid(plotlist = pL, ncol = 6))
 ```
@@
     nrow = 2,
-    labels = c("A  ChIP", "B  ChIP", "C  IP-MS", "D  IP-MS"),
+    labels = c("A", "B", "C", "D"),
     align = "v",
@@
 #| fig.width: 11
-#| fig.height: 11
+#| fig.height: 18
 #| warning: false
@@
     gg_enrpairs,
-    scale = 0.9,
-    labels = "A"
+    hm_ipms_hot,
+    rel_widths = c(1, 0.65), 
+    scale = 0.9, nrow = 1,
+    labels = c("A", "B")
 )

### 2024-07-09 ms_figure_5_ipms.Rmd
Fix heading Figure 5
Lift out IP-MS heatmap functions into scripts, move HOT IP-MS heatmap to Supp Fig 2

@@
 ft <- 12 # title font size
+```
+
+# Helper function
+
+```{r}
+## Overwrite the get_legend function from cowplot temporarily, 
+## since the cowplot one doesn't work with ggplot2 3.5.0
+## see https://github.com/wilkelab/cowplot/issues/202
+
+get_legend <- function(plot, legend = NULL) {
+    gt <- ggplotGrob(plot)
+
+    pattern <- "guide-box"
+    if (!is.null(legend)) {
+        pattern <- paste0(pattern, "-", legend)
+    }
+
+    indices <- grep(pattern, gt$layout$name)
+
+    not_empty <- !vapply(
+        gt$grobs[indices], 
+        inherits, what = "zeroGrob", 
+        FUN.VALUE = logical(1)
+    )
+    indices <- indices[not_empty]
+
+    if (length(indices) > 0) {
+        return(gt$grobs[[indices[1]]])
+    }
+    return(NULL)
+}
 ```
@@
                              pageLength = 10))
-```
-
-# Helper functions
-
-We define a number of additional helper functions for data extraction and 
-summarization, which will be used below.
-
-```{r heatmap-helper}
-filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold, 
-                             logfcThreshold, maxNbrTargets) {
-    ## Filter - require significant in at least one comparison
-    tstatfilt <- tstatmat
-    tstatfilt[adjpmat >= adjpThreshold] <- NA
-    tstatfilt[logfcmat <= logfcThreshold] <- NA
-    # tstatfilt[tstatfilt <= 0] <- NA
-
-    ## Some experiments pull down a large number of proteins
-    # print(table(colSums(!is.na(tstatfilt))))
-    ## -> remove proteins that are only being pulled down in these experiments
-    keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
-    tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
-    tstatfilt[is.na(tstatfilt)] <- 0
-    tstatfilt <- tstatfilt[rowSums(tstatfilt) > 0, ]
-
-    ## Subset the t-statistic matrix to only the retained proteins
-    stopifnot(colnames(tstatfilt) == colnames(tstatmat))
-    tstatmat <- tstatmat[rownames(tstatfilt), colnames(tstatfilt)]
-    colnames(tstatmat) <- .getSimplifiedComparison(colnames(tstatmat))
-
-    list(tstats = tstatmat, tstatsfilt = tstatfilt)
-}
-
-makeComplexAnnotation <- function(tstatmat, complexes, idmap,
-                                  show_legend = TRUE) {
-    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
-                           row.names = rownames(tstatmat))
-    rowcols <- list(Complex = c())
-    for (cmplx in names(complexes)) {
-        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
-                                                  idmap = idmap) %in% 
-                             complexes[[cmplx]]$pombase_ids] <- cmplx
-        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
-    }
-    rowAnnotation(df = rowannot, 
-                  col = rowcols, 
-                  annotation_name_gp = gpar(fontsize = fl),
-                  simple_anno_size_adjust = TRUE,
-                  na_col = bg_color,
-                  show_annotation_name = FALSE,
-                  show_legend = show_legend,
-                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
-}
-
-makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
-    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
-                           row.names = rownames(tstatmat))
-    rowcols <- list(Complex = c())
-    for (cmplx in names(complexes)) {
-        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
-                                                  idmap = idmap) %in% 
-                             complexes[[cmplx]]$pombase_ids] <- cmplx
-        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
-    }
-    rowannot$`DBD fam.` <- dbd$DBD_class[match(
-        .getPomBaseIdFromProtein(rownames(tstatmat), idmap = idmap), 
-        dbd$PomBaseID)]
-    setfam <- c("KilA-N", "Zn(II)2Cys6", "bZIP")
-    unsetfam <- setdiff(unique(dbd$DBD_class), setfam)
-    rowannot$`DBD fam.`[rowannot$`DBD fam.` %in% unsetfam] <- "Other (combined)"
-    rowcols <- c(rowcols, list(`DBD fam.` = c(`KilA-N` = main_colors[5], 
-                                              `Zn(II)2Cys6` = main_colors[4],
-                                              bZIP = main_colors[3], 
-                                              `Other (combined)` = bg_color)))
-    rowAnnotation(df = rowannot, 
-                  col = rowcols, 
-                  annotation_name_gp = gpar(fontsize = fl),
-                  simple_anno_size_adjust = TRUE,
-                  na_col = bg_color,
-                  show_annotation_name = TRUE,
-                  show_legend = FALSE,
-                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
-}
-
-makeColLabels <- function(tstatmat, colLabel, fontsize) {
-    columnAnnotation(
-        TFnames = anno_mark(which = "column", side = "bottom",
-                            at = match(colLabel, colnames(tstatmat)), 
-                            labels = .getProteinNameFromSimplifiedComparison(colLabel),
-                            labels_gp = gpar(fontsize = fontsize)))
-}
-
-makeRowLabels <- function(tstatmat, rowLabel, fontsize) {
-    rowAnnotation(
-        TFnames = anno_mark(which = "row", side = "left",
-                            at = match(rowLabel, rownames(tstatmat)), 
-                            labels = rowLabel,
-                            labels_gp = gpar(fontsize = fontsize)))
-}
-
-makeHeatmapCol <- function(tstatmat, stringency = "high") {
-    if (stringency == "high") {
-        circlize::colorRamp2(
-            breaks = seq(3.0, 13.0, length.out = 64),
-            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
-    } else {
-        circlize::colorRamp2(
-            breaks = seq(1.5, 13.0, length.out = 64),
-            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
-    }
-}
-
-reorderColsByRows <- function(tstat) {
-    ## Row names should be protein names, column names simplified comparison names
-    ord <- match(rownames(tstat), .getProteinNameFromSimplifiedComparison(colnames(tstat)))
-    ord <- ord[!is.na(ord)]
-    ord <- c(ord, setdiff(seq_len(ncol(tstat)), ord))
-    tstat[, ord]
-}
-
-makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
-    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)
-
-    tstats <- filterForHeatmap(tstatmat = tc$tstat, adjpmat = tc$adjp, 
-                               logfcmat = tc$logfc, logfcThreshold = log2fcthr,
-                               adjpThreshold = adjpthr, maxNbrTargets = Inf)
-    tstat <- tstats$tstats
-
-    ## tstats$tstatsfile: protein x experiment
-    ## Number of TFs that are pulled down by each bait
-    all_tfs <- setdiff(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
-                       c("Untagged", "untagged", NA))
-    nbrInt <- colSums(tstats$tstatsfilt[rownames(tstats$tstatsfilt) %in% all_tfs, ] > 0, 
-                      na.rm = TRUE)
-    names(nbrInt) <- .getSimplifiedComparison(names(nbrInt))
-
-    ## Split by family - return a matrix with several columns: 
-    ## #pulled down TFs (including itself), #pulled down TFs (excluding itself),
-    ## #pulled down TFs from the same family (excluding itself), 
-    ## #pulled down TFs from another family
-    tmptf <- tstats$tstatsfilt[match(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
-                                     rownames(tstats$tstatsfilt)), ]
-    rownames(tmptf) <- .getProteinNameFromComparison(colnames(tstats$tstatsfilt))
-    dbdtf <- data.frame(PomBaseID = .getPomBaseIdFromComparison(colnames(tstats$tstatsfilt), 
-                                                                idmap = idmap),
-                        bait = .getProteinNameFromComparison(colnames(tstats$tstatsfilt)),
-                        DBD_class = NA)
-    for (i in seq_len(nrow(dbdtf))) {
-        lookfor <- dbdtf$PomBaseID[i]
-        if (!(lookfor %in% c("untagged", "Untagged"))) {
-            dbdtf$DBD_class[i] <- dbd$DBD_class[match(lookfor, dbd$PomBaseID)]
-        } else {
-            dbdtf$DBD_class[i] <- "no_class"
-        }
-    }
-    stopifnot(dbdtf$bait == .getProteinNameFromComparison(colnames(tmptf)))
-    nbrIntMat <- do.call(dplyr::bind_rows, lapply(seq_along(colnames(tmptf)), function(i) {
-        expr <- colnames(tmptf)[i]
-        fam <- dbdtf$DBD_class[i]
-        data.frame(experiment = expr, 
-                   bait = dbdtf$bait[i], 
-                   dbdfam = fam, 
-                   TFs = paste(rownames(tmptf)[-i][which(tmptf[-i, expr] > 0)],
-                               collapse = ","),
-                   nbrTFs = sum(tmptf[, expr] > 0, na.rm = TRUE),
-                   nbrTFs_wobait = sum(tmptf[-i, expr] > 0, na.rm = TRUE),
-                   nbrTFs_samefam = sum(tmptf[-i, expr] > 0 & 
-                                            dbdtf$DBD_class[-i] == fam, 
-                                        na.rm = TRUE),
-                   nbrTFs_difffam = sum(tmptf[-i, expr] > 0 & 
-                                            dbdtf$DBD_class[-i] != fam, 
-                                        na.rm = TRUE)
-        )
-    }))
-
-    ## Split experiments into three groups
-    if (conc == 150) {
-        stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
-        tstat_no_bait <- 
-            tstat[, colSums(tstats$tstatsfilt) == 0 | 
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class == "nobait"])]
-        tstat_150only <- 
-            tstat[, colSums(tstats$tstatsfilt) > 0 &
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class == "150only"])]
-        tstat_150and500 <- 
-            tstat[, colSums(tstats$tstatsfilt) > 0 &
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class == "150and500"])]
-
-        proteins_150only <- .getProteinNameFromSimplifiedComparison(colnames(tstat_150only))
-        stopifnot(all(proteins_150only %in% rownames(tstat)))
-        tstat_150and500 <- tstat_150and500[!(rownames(tstat_150and500) %in% proteins_150only), ]
-
-        ## Cluster proteins based on the correlation between their t-stat vectors
-        dst <- sqrt(2 * (1 - cor(t(tstat_150and500), use = "pairwise.complete")))
-        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
-        clst <- hclust(as.dist(dst))
-
-        ## Reorder proteins by the clustering
-        protorder <- c(rownames(tstat_150and500)[clst$order], proteins_150only)
-        tstat <- tstat[protorder, ]
-
-        ## Reorder experiments to keep a "diagonal" in the heatmap
-        tstat <- reorderColsByRows(tstat)
-
-        tmpbaits <- .getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
-                                                             idmap = idmap)
-        colsplit <- ifelse(
-            tmpbaits %in% baitclass$Gene_name[baitclass$class == "nobait"], 
-            "No\nbait", 
-            ifelse(
-                tmpbaits %in% baitclass$Gene_name[baitclass$class == "150only"],
-                "150 mM NaCl",
-                ifelse(
-                    tmpbaits %in% baitclass$Gene_name[baitclass$class == "150and500"],
-                    "150 and\n500 mM NaCl", "NA"
-                )
-            )
-        )
-    } else if (conc == 500) {
-        tstat_onlybait <- 
-            tstat[, colSums(tstats$tstatsfilt > 0) == 1 |
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
-                                                      idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class500 == "Lost in 500 mM"])]
-        tstat_several <- 
-            tstat[, colSums(tstats$tstatsfilt > 0) > 1 & 
-                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
-                                                      idmap = idmap) %in% 
-                      baitclass$Gene_name[baitclass$class500 != "Lost in 500 mM"])]
-        tstat_onlybait <- tstat_onlybait[, colnames(tstat_onlybait) != "Untagged_plate"]
-        tstat_several <- tstat_several[, colnames(tstat_several) != "Untagged_plate"]
-
-        proteins_onlybait <-
-            .getProteinNameFromSimplifiedComparison(colnames(tstat_onlybait))
-        stopifnot(all(proteins_onlybait %in% rownames(tstat)))
-        tstat_several <- tstat_several[!(rownames(tstat_several) %in% proteins_onlybait), ]
-
-        ## Cluster proteins based on the correlation between their t-stat vectors
-        dst <- sqrt(2 * (1 - cor(t(tstat_several), use = "pairwise.complete")))
-        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
-        clst <- hclust(as.dist(dst))
-
-        ## Reorder proteins by the clustering
-        protorder <- c(rownames(tstat_several)[clst$order], proteins_onlybait)
-        tstat <- tstat[protorder, ]
-
-        ## Reorder conditions to keep a "diagonal" in the heatmap
-        tstat <- reorderColsByRows(tstat)
-
-        colsplit <- factor(
-            structure(baitclass$class500[
-                match(.getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
-                                                               idmap = idmap), 
-                      baitclass$Gene_name)],
-                names = colnames(tstat)),
-            levels = c("Retained in 500 mM", "Lost in 500 mM", ""))
-        levels(colsplit) <- c("Interactions\nin 500 mM NaCl",
-                              "No interactions\nin 500 mM NaCl", "")
-    } else {
-        stop("Unknown concentration")
-    }
-    return(list(tstat = tstat, colsplit = colsplit, nbrIntMat = nbrIntMat))
-}
-
-makeBaitHeatmapData <- function(sce, adjpthr, log2fcthr, nbrIntMat) {
-    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)$tstat
-    colnames(tc) <- .getSimplifiedComparison(colnames(tc))
-    shared_names <- intersect(rownames(tc), 
-                              .getProteinNameFromSimplifiedComparison(colnames(tc)))
-    shared_names <- setdiff(shared_names, .getProteinFromOrigBait(nobait_150, idmap = idmap))
-    stopifnot(all(shared_names %in% nbrIntMat$bait))
-    nbrIntMat$complex <- NA
-    for (cplx in c("MBF transcription complex", "CCAAT-binding factor complex", "Atf1-Pcr1")) {
-        nbrIntMat$complex[.getPomBaseIdFromProtein(nbrIntMat$bait,
-                                                   idmap = idmap) %in% 
-                              complexes[[cplx]]$pombase_ids] <- cplx
-    }
-    shared_names <- nbrIntMat[match(shared_names, nbrIntMat$bait), ] |>
-        arrange(dbdfam, complex) |>
-        pull(bait)
-    mat <- tc[shared_names, match(shared_names, .getProteinNameFromSimplifiedComparison(colnames(tc)))]
-    rownames(nbrIntMat) <- .getSimplifiedComparison(nbrIntMat$experiment)
-    list(mat = mat, nbrIntMat = nbrIntMat)
-}
 ```
@@
 
-## High stringency
-
-Here we use an adjusted p-value cutoff of `r adjpThreshold` and a log2 
-fold change cutoff of `r log2fcThreshold`.
+## Summary tables
+
+The 'high stringency' thresholds correspond to an adjusted p-value cutoff of 
+`r adjpThreshold` and a log2 fold change cutoff of `r log2fcThreshold`.
+The 'low stringency' thresholds correspond to an adjusted p-value cutoff of 
+`r adjpThresholdLoose` and a log2 fold change cutoff of `r log2fcThreshold`.
 
@@
                check.names = FALSE) |>
-        pivot_longer(cols = everything(), names_to = "Aspect", values_to = "Value") |>
-        kbl(col.names = NULL, caption = tbl_title) |>
-        kable_styling()
+        pivot_longer(cols = everything(), names_to = "Aspect", values_to = tbl_title)
 }
 
-## 150mM
-## --------------------------------------------------------------------------
 makeInteractorSummary(sce = sce150, baitdata = baitdata_150, adjp_cutoff = adjpThreshold,
@@
                       baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                      tbl_title = "150 mM, high stringency")
-
-## 500mM
-## --------------------------------------------------------------------------
+                      tbl_title = "150 mM, high stringency") |>
+    dplyr::full_join(
         makeInteractorSummary(sce = sce500, baitdata = baitdata_500, adjp_cutoff = adjpThreshold,
@@
                               tbl_title = "500 mM, high stringency")
-```
-
-## Low stringency
-
-Here we use an adjusted p-value cutoff of `r adjpThresholdLoose` and a log2 
-fold change cutoff of `r log2fcThreshold`.
-
-```{r}
-## 150mM
-## --------------------------------------------------------------------------
+    ) |> 
+    dplyr::full_join(
         makeInteractorSummary(sce = sce150, baitdata = baitdata_150_loose, adjp_cutoff = adjpThresholdLoose,
@@
                               tbl_title = "150 mM, low stringency")
-
-## 500mM
-## --------------------------------------------------------------------------
+    ) |>
+    dplyr::full_join(
         makeInteractorSummary(sce = sce500, baitdata = baitdata_500_loose, adjp_cutoff = adjpThresholdLoose,
@@
                               tbl_title = "500 mM, low stringency")
-```
-
-# Figure 4
-
-## Settings
-
-```{r plot-settings}
-colLabel_150 <- expand.grid(c("Untagged", "Ace2", "Rst2", "Pap1"),
-                            c("tube", "plate")) |>
-    tidyr::unite(col = "col", Var1, Var2) |>
-    dplyr::pull(col)
-colLabel_500 <- expand.grid(c("Untagged", "Atf1", "Moc3", "Ntu1",
-                              "Pcr1", "Ntu2"),
-                            c("tube", "plate")) |>
-    tidyr::unite(col = "col", Var1, Var2) |>
-    dplyr::pull(col)
-rowLabel <- NULL
-
-complexLabel <- c("NuA4", "SAGA")
-```
+    ) |>
+    kbl() |>
+    kable_styling()
+```
+
+# Figure 5
 
 ## Summary heatmaps using results from the statistical tests (150 mM NaCl)
-
-### Overall heatmap
-
-```{r heatmap-tstat-150}
-#| fig.width: 6
-#| fig.height: 8
-
-hm1a <- Heatmap(hmdata_150$tstat, 
-                cluster_rows = FALSE, 
-                cluster_columns = FALSE, 
-                col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
-                border = TRUE, 
-                border_gp = gpar(lwd = 0.5),
-                column_split = hmdata_150$colsplit, 
-                name = "Mod.\nt-stat.", 
-                na_col = binary_heatmap_colors["FALSE"],
-                column_names_gp = gpar(fontsize = fs),
-                row_names_gp = gpar(fontsize = fs),
-                row_title = "Proteins significantly enriched in at least one experiment",
-                row_title_gp = gpar(fontsize = fl),
-                show_row_names = FALSE, show_column_names = FALSE,
-                use_raster = FALSE, show_heatmap_legend = TRUE,
-                right_annotation = makeComplexAnnotation(
-                    hmdata_150$tstat, complexes[complexLabel], 
-                    idmap = idmap),
-                bottom_annotation = makeColLabels(hmdata_150$tstat, colLabel_150, fl),
-                width = unit(w_ipmsHm, "mm"), # heatmap body width
-                heatmap_height = unit(h_ipmsHm, "mm"), # whole heatmap height
-                column_title_gp = gpar(fontsize = ft),
-                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
-                                border = "gray10",
-                                border_gp = gpar(lwd = 0.5)))
-
-set.seed(42L)
-hm_150 <- grid.grabExpr(
-    hm1a <- draw(hm1a, merge_legend = TRUE,
-                heatmap_legend_side = "right",
-                annotation_legend_side = "right"),
-    width = w_ipmsHm, height = h_ipmsHm
-)
-
-plot_grid(hm_150)
-```
 
@@
 
+```{r}
+## Pie chart, within-vs-between DBD family interactions
+pie_bait_150 <- ggplot(
+    baitdata_150$nbrIntMat |>
+        select(nbrTFs_samefam, nbrTFs_difffam) |>
+        pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
+        group_by(class) |>
+        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    aes(x = "", y = nbrTF, fill = class)) + 
+    geom_col(width = 1) + 
+    coord_polar("y", start = 0) + 
+    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
+                                 nbrTFs_difffam = main_colors[5]),
+                      labels = c(nbrTFs_samefam = "Same DBD family",
+                                 nbrTFs_difffam = "Different DBD family"),
+                      name = "") + 
+    labs(title = "150 mM NaCl\nhigh stringency") + 
+    theme_void()
+```
+
 ### Bait-vs-bait, less stringent adjusted p-value threshold
@@
 
+```{r}
+## Pie chart, within-vs-between DBD family interactions
+pie_bait_150_loose <- ggplot(
+    baitdata_150_loose$nbrIntMat |>
+        select(nbrTFs_samefam, nbrTFs_difffam) |>
+        pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
+        group_by(class) |>
+        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    aes(x = "", y = nbrTF, fill = class)) + 
+    geom_col(width = 1) + 
+    coord_polar("y", start = 0) + 
+    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
+                                 nbrTFs_difffam = main_colors[5]),
+                      labels = c(nbrTFs_samefam = "Same DBD family",
+                                 nbrTFs_difffam = "Different DBD family"),
+                      name = "") + 
+    labs(title = "150 mM NaCl\nlow stringency") + 
+    theme_void()
+```
+
 ### SAGA/NuA4 heatmap
@@
                   use_raster = FALSE, show_heatmap_legend = TRUE,
-                  # bottom_annotation = makeColLabels(tstat150[keep, ], colLabel_150, fl),
-                  #left_annotation = makeRowLabels(tstat150[keep, ], rowLabel, fl),
                   width = unit(w_ipmsHm, "mm"), # heatmap body width
@@
 ## Summary heatmaps using results from the statistical tests (500 mM NaCl)
-
-### Overall heatmap
-
-```{r heatmap-tstat-500}
-#| fig.width: 6
-#| fig.height: 8
-
-hm2a <- Heatmap(hmdata_500$tstat, 
-                cluster_rows = FALSE, 
-                cluster_columns = FALSE, 
-                col = makeHeatmapCol(hmdata_500$tstat, stringency = "high"), 
-                border = TRUE, 
-                border_gp = gpar(lwd = 0.5),
-                column_split = hmdata_500$colsplit, 
-                name = "Mod.\nt-stat.", 
-                na_col = binary_heatmap_colors["FALSE"],
-                column_names_gp = gpar(fontsize = fs),
-                row_names_gp = gpar(fontsize = fs),
-                row_title = "Proteins significantly enriched in at least one experiment",
-                row_title_gp = gpar(fontsize = fl),
-                show_row_names = FALSE, show_column_names = FALSE,
-                use_raster = FALSE, show_heatmap_legend = TRUE,
-                right_annotation = makeComplexAnnotation(hmdata_500$tstat, complexes[complexLabel],
-                                                         idmap = idmap, show_legend = FALSE),
-                bottom_annotation = makeColLabels(hmdata_500$tstat, colLabel_500, fl),
-                width = unit(w_ipmsHm, "mm"), # heatmap body width
-                heatmap_height = unit(h_ipmsHm, "mm"), # whole heatmap height
-                column_title_gp = gpar(fontsize = ft),
-                heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
-                                border = "gray10",
-                                border_gp = gpar(lwd = 0.5)))
-
-hm_500 <- grid.grabExpr(
-    hm2a <- draw(hm2a, merge_legend = TRUE,
-                 heatmap_legend_side = "right",
-                 annotation_legend_side = "right"),
-    width = w_ipmsHm, height = h_ipmsHm
-)
-
-plot_grid(hm_500)
-```
 
@@
 plot_grid(hm_500_bait)
+```
+
+```{r}
+## Pie chart, within-vs-between DBD family interactions
+pie_bait_500 <- ggplot(
+    baitdata_500$nbrIntMat |>
+        select(nbrTFs_samefam, nbrTFs_difffam) |>
+        pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
+        group_by(class) |>
+        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    aes(x = "", y = nbrTF, fill = class)) + 
+    geom_col(width = 1) + 
+    coord_polar("y", start = 0) + 
+    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
+                                 nbrTFs_difffam = main_colors[5]),
+                      labels = c(nbrTFs_samefam = "Same DBD family",
+                                 nbrTFs_difffam = "Different DBD family"),
+                      name = "") + 
+    labs(title = "500 mM NaCl\nhigh stringency") + 
+    theme_void()
+```
+
+```{r}
+## Pie chart, within-vs-between DBD family interactions
+pie_bait_500_loose <- ggplot(
+    baitdata_500_loose$nbrIntMat |>
+        select(nbrTFs_samefam, nbrTFs_difffam) |>
+        pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
+        group_by(class) |>
+        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    aes(x = "", y = nbrTF, fill = class)) + 
+    geom_col(width = 1) + 
+    coord_polar("y", start = 0) + 
+    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
+                                 nbrTFs_difffam = main_colors[5]),
+                      labels = c(nbrTFs_samefam = "Same DBD family",
+                                 nbrTFs_difffam = "Different DBD family"),
+                      name = "") + 
+    labs(title = "500 mM NaCl\nlow stringency") + 
+    theme_void()
+```
+
+## Pie charts for bait-vs-bait interactions (within/between DBD families)
+
+```{r}
+#| label: pie-bait-summary
+#| fig.width: 10
+#| fig.height: 3
+
+cowplot::plot_grid(
+    cowplot::plot_grid(pie_bait_150 + theme(legend.position = "none"), 
+                       pie_bait_150_loose + theme(legend.position = "none"), 
+                       pie_bait_500 + theme(legend.position = "none"),
+                       pie_bait_500_loose + theme(legend.position = "none"), nrow = 1),
+    get_legend(pie_bait_150 + theme(legend.position = "bottom")),
+    ncol = 1, rel_heights = c(1, 0.25)
+)
 ```

### 2024-07-09 params/ipms_heatmap_functions.R

@@ -1,0 +1,279 @@
+filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold,
+                             logfcThreshold, maxNbrTargets) {
+    ## Filter - require significant in at least one comparison
+    tstatfilt <- tstatmat
+    tstatfilt[adjpmat >= adjpThreshold] <- NA
+    tstatfilt[logfcmat <= logfcThreshold] <- NA
+    # tstatfilt[tstatfilt <= 0] <- NA
+
+    ## Some experiments pull down a large number of proteins
+    # print(table(colSums(!is.na(tstatfilt))))
+    ## -> remove proteins that are only being pulled down in these experiments
+    keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
+    tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
+    tstatfilt[is.na(tstatfilt)] <- 0
+    tstatfilt <- tstatfilt[rowSums(tstatfilt) > 0, ]
+
+    ## Subset the t-statistic matrix to only the retained proteins
+    stopifnot(colnames(tstatfilt) == colnames(tstatmat))
+    tstatmat <- tstatmat[rownames(tstatfilt), colnames(tstatfilt)]
+    colnames(tstatmat) <- .getSimplifiedComparison(colnames(tstatmat))
+
+    list(tstats = tstatmat, tstatsfilt = tstatfilt)
+}
+
+makeComplexAnnotation <- function(tstatmat, complexes, idmap,
+                                  show_legend = TRUE) {
+    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
+                           row.names = rownames(tstatmat))
+    rowcols <- list(Complex = c())
+    for (cmplx in names(complexes)) {
+        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot),
+                                                  idmap = idmap) %in%
+                             complexes[[cmplx]]$pombase_ids] <- cmplx
+        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
+    }
+    rowAnnotation(df = rowannot,
+                  col = rowcols,
+                  annotation_name_gp = gpar(fontsize = fl),
+                  simple_anno_size_adjust = TRUE,
+                  na_col = bg_color,
+                  show_annotation_name = FALSE,
+                  show_legend = show_legend,
+                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
+}
+
[…]

### 2024-07-09 ms_figure_7_heterodimer.Rmd
Add axis titles for coverage plots (Fig 7)

@@
 
+cowplot::plot_grid(
+    plotlist = maplots, nrow = 1, align = "hv", axis = "b"
+)
+```
+
+```{r}
+#| label: rnaseq-corrplot
+
 ## Correlation between log2 fold changes for ntu1/ntu2
@@
                      ntu2 = resList$ntu2$log2FoldChange)
-corrplot <- list(
+(corrplot <- list(
     ggplot(dfcorr, aes(x = ntu1, y = ntu2)) + 
@@
         theme(axis.title = element_markdown())
-)
-
-cowplot::plot_grid(
-    plotlist = maplots, nrow = 1, align = "hv", axis = "b"
-)
+))
 ```
@@
                  strip = strip_nested(size = "variable")) + 
-    scale_y_continuous(limits = c(0, 0.9), expand = c(0, 0))
+    scale_y_continuous(limits = c(0, 0.9), expand = c(0, 0)) + 
+    labs(y = "ChIP-seq")
 g1b <- ggplot(plotdf_rna, 
@@
     geom_area(aes(group = strand, fill = gene)) + 
-    scale_fill_manual(values = c(wt = lighten("black", amount = 0.4), 
-                                 "ntu1\U141E" = lighten(main_colors[5], amount = 0.4), 
-                                 "ntu2\U141E" = lighten(main_colors[1], amount = 0.4))) + 
+    scale_fill_manual(values = c(wt = "black", 
+                                 "ntu1\U141E" = main_colors[5], 
+                                 "ntu2\U141E" = main_colors[1])) + 
     facet_grid(sample ~ seqnames, scale = "free_x", space = "free_x",
                switch = "y") + 
-    scale_y_continuous(limits = c(-0.9, 0.3), expand = c(0, 0))
+    scale_y_continuous(limits = c(-0.9, 0.3), expand = c(0, 0)) + 
+    labs(y = "RNA-seq")
 g2 <- ggplot(gtf, aes(x = start2, xend = end2, y = y, yend = y, 
@@
                                strip.text.y = element_text(size = 12),
-                               strip.text.x = element_text(size = 12, hjust = 0.95)), 
+                               strip.text.x = element_text(size = 12, hjust = 0.95),
+                               axis.title.y = element_text(size = 12, angle = 90,
+                                                           margin = margin(t = 0, r = 10, b = 0, l = 0))), 
     NULL,
@@
                                legend.position = "none",
-                               strip.text.x = element_blank()), 
+                               strip.text.x = element_blank(),
+                               axis.title.y = element_text(size = 12, angle = 90,
+                                                           margin = margin(t = 0, r = 15, b = 0, l = 0))), 
     g2 + theme_void() + 
@@
 
-fig7c <- wrap_plots(maplots, ncol = 1) | corrplot[[1]]
+fig7c <- wrap_plots(maplots, ncol = 2)
 
@@
         rel_widths = c(1, 1),
-        labels = c("C", "E")
+        labels = c("D", "E")
     ),

### 2024-07-09 ms_figure_1_overview.Rmd
Indicate meiosis-specific TFs in Fig 1

@@
         levels = c("Bait pulled down (150 mM NaCl IP-MS)",
-                   "Bait not pulled down (150 mM NaCl IP-MS)")),
+                   "Bait not pulled down (150 mM NaCl IP-MS)", 
+                   "Meiosis-specific transcription factors")),
     tpm = rowMeans(assay(rnaseq, "abundance")[allbaits, 
@@
 gg_tpm <- ggplot(ave_tpm, aes(x = fct_reorder(bait, tpm, .desc = TRUE), y = tpm, 
-                              color = bait_pulled)) + 
+                              color = bait_pulled, shape = bait_pulled)) + 
     geom_hline(yintercept = 0, linetype = "dashed", color = na_color) +
-    geom_point(size = 1) + 
+    geom_point(size = 1, show.legend = TRUE) + 
     theme_cowplot() + 
@@
     scale_color_manual(values = c("Bait pulled down (150 mM NaCl IP-MS)" = pt_color,
-                                  "Bait not pulled down (150 mM NaCl IP-MS)" = na_color),
-                       name = "") + 
+                                  "Bait not pulled down (150 mM NaCl IP-MS)" = na_color,
+                                  "Meiosis-specific transcription factors" = main_colors[1]),
+                       drop = FALSE, name = "") + 
+    scale_shape_manual(values = c("Bait pulled down (150 mM NaCl IP-MS)" = 16,
+                                  "Bait not pulled down (150 mM NaCl IP-MS)" = 16,
+                                  "Meiosis-specific transcription factors" = 15),
+                       drop = FALSE, name = "") + 
     scale_x_discrete(breaks = c("mei4", "atf31", "atf21", "cuf2", "rsv1")) + 
-    guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
+    guides(color = guide_legend(nrow = 3, byrow = TRUE,
+                                override.aes = list(size = 3))) + 
     labs(x = "Transcription factors",

### 2024-07-09 index.qmd
Add some ORCIDs, reduce author/affiliation font size, update figure links

@@
   - name: Vytautas Iesmantavicius
+    orcid: 0000-0002-2512-9957
     affiliations:
@@
   - name: Daniel Hess
+    orcid: 0000-0002-1642-5404
     affiliations:
@@
   - name: Jan Seebacher
+    orcid: 0000-0002-7858-2720
     affiliations:
@@
 
+<style>
+.author {
+  font-size: 0.7em;
+}
+
+.affiliation {
+  font-size: 0.7em;
+}
+</style>
+
 ## Interactive data exploration with `TFexplorer`
@@
 | Figure 1 | [![](ms_figure_1_overview_files/figure-html/Fig1-1.png){width="4cm"}](ms_figure_1_overview.html) |  [![](ms_figure_1_overview_files/figure-html/suppFig1-1.png){width="4cm"}](ms_figure_1_overview.html#Supplementary_figure)   |
-| Figure 2 | [![](ms_figure_2_chip_files/figure-html/Fig2-1.png){width="4cm"}](ms_figure_2_chip.html) | [![](ms_figure_2_chip_files/figure-html/suppFig2-1.png){width="4cm"}](ms_figure_2_chip.html#Supplementary_figure)    |
-| Figure 3 | [![](ms_figure_3_motifs_files/figure-html/Fig3-1.png){width="4cm"}](ms_figure_3_motifs.html) | [![](ms_figure_3_motifs_files/figure-html/suppFig3-1.png){width="4cm"}](ms_figure_3_motifs.html#Supplementary_figure)    |
-| Figure 4 | [![](ms_figure_4_ipms_files/figure-html/Fig4-1.png){width="4cm"}](ms_figure_4_ipms.html) | [![](ms_figure_4_ipms_files/figure-html/suppFig4-1.png){width="4cm"}](ms_figure_4_ipms.html#Supplementary_figure)    |
-| Figure 5 | [![](ms_figure_5_networks_files/figure-html/Fig5-1.png){width="4cm"}](ms_figure_5_networks.html) | [![](ms_figure_5_networks_files/figure-html/suppFig5-1.png){width="4cm"}](ms_figure_5_networks.html#Supplementary_figure)    |
-| Figure 6 | [![](ms_figure_6_heterodimer_files/figure-html/Fig6-1.png){width="4cm"}](ms_figure_6_heterodimer.html) | [![](ms_figure_6_heterodimer_files/figure-html/suppFig6-1.png){width="4cm"}](ms_figure_6_heterodimer.html#Supplementary_figure)    |
+| Figure 2 | [![](ms_figure_2_overview-heatmaps_files/figure-html/Fig2-1.png){width="4cm"}](ms_figure_2_overview-heatmaps.html) | [![](ms_figure_2_overview-heatmaps_files/figure-html/suppFig2-1.png){width="4cm"}](ms_figure_2_overview-heatmaps.html#Supplementary_figure)    |
+| Figure 3 | [![](ms_figure_3_chip_files/figure-html/Fig3-1.png){width="4cm"}](ms_figure_3_chip.html) | [![](ms_figure_3_chip_files/figure-html/suppFig3-1.png){width="4cm"}](ms_figure_3_chip.html#Supplementary_figure)    |
+| Figure 4 | [![](ms_figure_4_motifs_files/figure-html/Fig4-1.png){width="4cm"}](ms_figure_4_motifs.html) | [![](ms_figure_4_motifs_files/figure-html/suppFig4-1.png){width="4cm"}](ms_figure_4_motifs.html#Supplementary_figure)    |
+| Figure 5 | [![](ms_figure_5_ipms_files/figure-html/Fig5-1.png){width="4cm"}](ms_figure_5_ipms.html) | [![](ms_figure_5_ipms_files/figure-html/suppFig5-1.png){width="4cm"}](ms_figure_5_ipms.html#Supplementary_figure)    |
+| Figure 6 | [![](ms_figure_6_networks_files/figure-html/Fig6-1.png){width="4cm"}](ms_figure_6_networks.html) | [![](ms_figure_6_networks_files/figure-html/suppFig6-1.png){width="4cm"}](ms_figure_6_networks.html#Supplementary_figure)    |
+| Figure 7 | [![](ms_figure_7_heterodimer_files/figure-html/Fig7-1.png){width="4cm"}](ms_figure_7_heterodimer.html) | [![](ms_figure_7_heterodimer_files/figure-html/suppFig7-1.png){width="4cm"}](ms_figure_7_heterodimer.html#Supplementary_figure)    |

### 2024-07-15 ms_figure_2_overview-heatmaps.Rmd
move RNA-level violin plot from fig3 report to suppFig2

@@
     peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
+    rdsrnaseq_del: "data/rnaseq_del.rds"
+    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
     gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
@@
                                     }))
+
+# `rnaseq_del`: a `SummarizedExperiment` with RNA-seq quantifications
+rnaseq_del <- readRDS(params$rdsrnaseq_del)
+
+# `txannot`: data.frame with transcript start site (TSS) and
+#            transcript end site (TES) annotation
+txannot <- read.csv(params$txcsv, row.names = 1)
 
@@
 
-## IP-MS data for HOT TFs
+## Supplementary plots - IP-MS data for HOT TFs
 
@@
 hm3 <- Heatmap(m[sel_peaksBin, ],
-               column_title = "ChIP (binary)",
+               column_title = "Binarized ChIP enrichments",
                column_title_side = "top",
@@
 hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
-               column_title = "ChIP (enrichments)",
+               column_title = "ChIP enrichments",
                column_title_side = "top",
@@
 
-## ChIP enrichment replicate-pair correlations
+## Supplementary plots - ChIP enrichment replicate-pair correlations
 
@@
 
+## Supplementary plots - Bound promoters
+
+In `txannot$prom.number.overlapping.highconf.peaks`, we record how many TF peaks
+are found in the promoter regions of a gene (defined as a 1kb window
+centered on the annotated transcript start site).
+
+Here we compare the expression levels (obtained from `rnaseq_del`) of
+genes twith different numbers of TFs bound at their promoters.
+
+```{r bound-promoters}
+#| fig.width: 6
+#| fig.height: 4
+
+# calculate average expression of each gene
+avg_tpm <- rowMeans(assay(rnaseq_del, "abundance"))
+
+# add average expression to gene annotation and filter out
+# genes that are not protein coding genes
+pd <- txannot |>
+    mutate(avg_tpm = avg_tpm[unique_einprot_id]) |>
+    filter(!is.na(avg_tpm),
+           gene_biotype == "protein_coding") |>
+    mutate(prom.number.overlapping.highconf.peaks.categories = 
+               cut(prom.number.overlapping.highconf.peaks,
+                   breaks = c(-Inf, 0, 1, Inf),
+                   labels = c("0", "1", ">1"),
+                   right = TRUE))
+
+# visualize
+(gg_rnaviolin <- ggplot(pd, aes(prom.number.overlapping.highconf.peaks.categories, avg_tpm + 0.01)) +
+        geom_violin(draw_quantiles = 0.5, fill = bg_color) +
+        scale_y_log10(labels = scales::label_number(big.mark = "")) +
+        labs(x = "Number of promoter TF peaks",
+             y = "RNA level\n(TPM + 0.01, log-scale)") +
+        theme_cowplot())
+```
+
 ## Put together
@@
     gg_enrpairs,
+    cowplot::plot_grid(
         hm_ipms_hot,
+        gg_rnaviolin,
+        NULL,
+        NULL,
+        ncol = 1,
+        rel_heights = c(1.25, 0.75, 1, 1),
+        labels = c("", "C", "D", "E")
+    ),
     rel_widths = c(1, 0.65), 

### 2024-07-15 ms_figure_3_chip.Rmd
move RNA-level violin plot from fig3 report to suppFig2

@@
     motifrds: "data/motifs_annotated.rds"
-    rdsrnaseq_del: "data/rnaseq_del.rds"
     txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
@@
 
-# `rnaseq_del`: a `SummarizedExperiment` with RNA-seq quantifications
-rnaseq_del <- readRDS(params$rdsrnaseq_del)
-
 # `txannot`: data.frame with transcript start site (TSS) and
@@
 
-## Bound promoters
-
-In `txannot$prom.number.overlapping.highconf.peaks`, we record how many TF peaks
-are found in the promoter regions of a gene (defined as a 1kb window
-centered on the annotated transcript start site).
-
-Here we compare the expression levels (obtained from `rnaseq_del`) of
-genes twith different numbers of TFs bound at their promoters.
-
-```{r bound-promoters}
-#| fig.width: 6
-#| fig.height: 4
-
-# calculate average expression of each gene
-avg_tpm <- rowMeans(assay(rnaseq_del, "abundance"))
-
-# add average expression to gene annotation and filter out
-# genes that are not protein coding genes
-pd <- txannot |>
-    mutate(avg_tpm = avg_tpm[unique_einprot_id]) |>
-    filter(!is.na(avg_tpm),
-           gene_biotype == "protein_coding") |>
-    mutate(prom.number.overlapping.highconf.peaks.categories = 
-               cut(prom.number.overlapping.highconf.peaks,
-                   breaks = c(-Inf, 0, 1, Inf),
-                   labels = c("0", "1", ">1"),
-                   right = TRUE))
-
-# visualize
-ggplot(pd, aes(prom.number.overlapping.highconf.peaks.categories, avg_tpm + 0.01)) +
-    geom_violin(draw_quantiles = 0.5) +
-    scale_y_log10() +
-    labs(x = "Number of promoter TF peaks",
-         y = "RNA level (TPM + 0.01, log-scale)") +
-    theme_cowplot()
-```
-
 ## Put together

### 2024-07-15 ms_figure_4_motifs.Rmd
rename Fig3 to Fig4

@@
 ---
-title: "Figure 3 - Motifs"
+title: "Figure 4 - Motifs"
 author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
@@
 
-# Figure 3
+# Figure 4
 
@@
 
-Assemble the panels into Figure 3:
+Assemble the panels into Figure 4:
 
-```{r Fig3}
+```{r Fig4}
 #| fig.width: 11
@@
 
-Assemble the panels into Supplementary Figure 3:
+Assemble the panels into Supplementary Figure 4:
 
-```{r suppFig3}
+```{r suppFig4}
 #| fig.width: 4.5

### 2024-07-15 ms_figure_1_overview.Rmd
Indicate meiosis-specific TFs in SuppFig1 with color

@@
 cowplot::plot_grid(
-    gg_tpm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
+    gg_tpm + theme(axis.text.x = element_blank(),
+                   axis.ticks.length.x = unit(4, "mm"),
+                   axis.ticks.x = element_line(color = main_colors[1],
+                                               linewidth = 2.25)) +
         labs(x = "") + geom_point(size = 3), 

### 2024-07-15 ms_figure_6_networks.Rmd
Make Fig 6c narrower

@@
     cowplot::plot_grid(
-        cowplot::plot_grid(fig6c, ggrad_upset,
+        cowplot::plot_grid(
+            cowplot::plot_grid(NULL, fig6c, NULL, nrow = 1,
+                               rel_widths = c(0.25, 1, 0.1)),  
+            ggrad_upset,
             ncol = 1, labels = c("C", "D"),
-                           rel_heights = c(1, 1.25)), 
+            rel_heights = c(1, 1.75)), 
         pho7volc, 

### 2024-07-15 ms_figure_4_motifs.Rmd
move heatmap title from columns to rows

@@
 
-plot_grid(ZnII2Cys6_kmer_1)
+print(ZnII2Cys6_kmer_1)
 ```
@@
                     column_names_gp = gpar(fontsize = fs, fontfamily = "mono"),
-                    column_title = "Binuclear zinc cluster TFs",
-                    column_title_side = "top", column_title_gp = gpar(fontsize = ft),
+                    row_title = "Binuclear zinc cluster TFs",
+                    row_title_side = "left", column_title_gp = gpar(fontsize = ft),
                     border = TRUE, border_gp = gpar(lwd = 0.5),
@@
         labels = c("A", "B"),
-        rel_widths = c(2, 3.2),
+        rel_widths = c(1.9, 3.3),
         nrow = 1

### 2024-07-15 ms_figure_2_overview-heatmaps.Rmd
Make IP-MS heatmaps wider (Fig 2), adjust legends and titles

@@
 ## Heatmap sizes
-w_ipmsHm <-  84 # heatmap body width (mm)
+w_ipmsHm <-  1.2 * 84 # heatmap body width (mm)
 h_ipmsHm <- 178 # total heatmap height (mm)
@@
                 column_split = hmdata_150$colsplit, 
-                name = "Mod.\nt-stat.", 
+                name = "Mod. t-stat.", 
                 na_col = binary_heatmap_colors["FALSE"],
@@
                 heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
+                                            legend_direction = "horizontal",
+                                            title_position = "topcenter",
                                             border = "gray10",
@@
     hm1a <- draw(hm1a, merge_legend = TRUE,
-                heatmap_legend_side = "right",
-                annotation_legend_side = "right"),
+                heatmap_legend_side = "bottom",
+                annotation_legend_side = "bottom"),
     width = w_ipmsHm, height = h_ipmsHm
@@
                 column_split = hmdata_500$colsplit, 
-                name = "Mod.\nt-stat.", 
+                name = "Mod. t-stat.", 
                 na_col = binary_heatmap_colors["FALSE"],
@@
     hm2a <- draw(hm2a, merge_legend = TRUE,
-                 heatmap_legend_side = "right",
-                 annotation_legend_side = "right"),
+                 heatmap_legend_side = "bottom",
+                 annotation_legend_side = "bottom"),
     width = w_ipmsHm, height = h_ipmsHm

### 2024-07-15 params/ipms_heatmap_functions.R

@@ -4,10 +4,8 @@ filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold,
     tstatfilt <- tstatmat
     tstatfilt[adjpmat >= adjpThreshold] <- NA
     tstatfilt[logfcmat <= logfcThreshold] <- NA
-    # tstatfilt[tstatfilt <= 0] <- NA
 
     ## Some experiments pull down a large number of proteins
-    # print(table(colSums(!is.na(tstatfilt))))
     ## -> remove proteins that are only being pulled down in these experiments
     keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
     tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
@@ -40,7 +38,8 @@ makeComplexAnnotation <- function(tstatmat, complexes, idmap,
                   na_col = bg_color,
                   show_annotation_name = FALSE,
                   show_legend = show_legend,
-                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
+                  annotation_legend_param = list(title_gp = gpar(fontsize = fl),
+                                                 legend_direction = "horizontal"))
 }
 
 makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
@@ -204,10 +203,10 @@ makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
             "No\nbait",
             ifelse(
                 tmpbaits %in% baitclass$Gene_name[baitclass$class == "150only"],
-                "150 mM NaCl",
+                "150 mM\nNaCl IP-MS",
                 ifelse(
                     tmpbaits %in% baitclass$Gene_name[baitclass$class == "150and500"],
-                    "150 and\n500 mM NaCl", "NA"
+                    "150 and 500 mM\nNaCl IP-MS", "NA"
                 )
             )
         )
@@ -249,8 +248,8 @@ makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
                       baitclass$Gene_name)],
                 names = colnames(tstat)),
             levels = c("Retained in 500 mM", "Lost in 500 mM", ""))
-        levels(colsplit) <- c("Interactions\nin 500 mM NaCl",
-                              "No interactions\nin 500 mM NaCl", "")
+        levels(colsplit) <- c("Interactions in\n500 mM NaCl IP-MS",
[…]

### 2024-07-15 ms_figure_2_overview-heatmaps.Rmd
Move IP-MS PCA plots to SuppFig 2

@@
     library(tidyr)
+    library(einprot)
 })
@@
 
+## Supplementary plots - PCA (IP-MS)
+
+```{r}
+## Low salt
+sce150_pca <- einprot::doPCA(
+    sce150[rowSums(!assay(sce150, 
+                          metadata(sce150)$aNames$assayImputIndic)) >= 2, ],
+    metadata(sce150)$aNames$assayImputed)$sce
+df <- scuttle::makePerCellDF(sce150_pca, features = NULL, use.coldata = TRUE, 
+                             use.dimred = TRUE) |>
+    dplyr::mutate(Protocol = .capitalize(sub("_digest|_freeze", "", sub("SOP_", "", Protocol))))
+pcvar150 <- attr(reducedDim(sce150_pca, "PCA_log2_LFQ.intensity"), "percentVar")
+(ggpca1 <- ggplot(
+    df, 
+    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
+        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
+        color = Protocol)) + 
+    geom_point(size = 3, alpha = 0.5) +
+    scale_color_manual(values = c(Plate = main_colors[1], 
+                                  Tube = main_colors[5]), name = "") + 
+    coord_fixed() + 
+    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
+         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), 
+         title = "150 mM NaCl") + 
+    theme_cowplot() + 
+    theme(legend.position = "bottom") + 
+    guides(color = guide_legend(nrow = 1, byrow = TRUE)))
+(ggpca2 <- ggplot(
+    df |>
+        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")) |>
+        dplyr::mutate(prot = .capitalize(prot)), 
+    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
+        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
+        color = bait, shape = prot)) + 
+    geom_point(size = 3, alpha = 0.5) +
+    scale_color_manual(values = c(Atf1 = main_colors[3], 
+                                  `Other baits` = na_color), name = "") + 
+    scale_shape_manual(values = c(Plate = 16, Tube = 17), name = "") + 
+    coord_fixed() + 
+    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
+         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), title = "150 mM NaCl") + 
+    theme_cowplot() + 
+    theme(legend.position = "bottom", legend.box = "vertical", 
+          legend.box.just = "left", legend.margin = margin()))
+
+sce500_pca <- einprot::doPCA(
+    sce500[rowSums(!assay(sce500, metadata(sce500)$aNames$assayImputIndic)) >= 2, ],
+    metadata(sce500)$aNames$assayImputed)$sce
+df <- scuttle::makePerCellDF(sce500_pca, features = NULL, use.coldata = TRUE, 
+                             use.dimred = TRUE)
+pcvar500 <- attr(reducedDim(sce500_pca, "PCA_log2_LFQ.intensity"), "percentVar")
+(ggpca3 <- ggplot(
+    df |>
+        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")), 
+    aes(x = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".1")]], 
+        y = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".2")]], 
+        color = bait)) + 
+    geom_point(size = 3, alpha = 0.5) +
+    scale_color_manual(values = c(Atf1 = main_colors[3], 
+                                  `Other baits` = na_color), name = "") + 
+    coord_fixed() + 
+    labs(x = paste0("PC1 (", round(pcvar500[1], 1), "%)"), 
+         y = paste0("PC2 (", round(pcvar500[2], 1), "%)"), 
+         title = "500 mM NaCl") + 
+    theme_cowplot() + 
+    theme(legend.position = "bottom", legend.box.just = "left"))
+```
+
 ## Put together
@@
         gg_rnaviolin,
-        NULL,
-        NULL,
+        ggpca2,
+        ggpca3,
         ncol = 1,
-        rel_heights = c(1.25, 0.75, 1, 1),
+        rel_heights = c(1.25, 0.75, 1.1, 0.9),
         labels = c("", "C", "D", "E")

### 2024-07-15 ms_figure_5_ipms.Rmd
Move IP-MS PCA plots to SuppFig 2

@@
 
-## Supplementary plots - PCA
-
-```{r}
-## Low salt
-sce150_pca <- einprot::doPCA(
-    sce150[rowSums(!assay(sce150, 
-                          metadata(sce150)$aNames$assayImputIndic)) >= 2, ],
-    metadata(sce150)$aNames$assayImputed)$sce
-df <- scuttle::makePerCellDF(sce150_pca, features = NULL, use.coldata = TRUE, 
-                             use.dimred = TRUE) |>
-    dplyr::mutate(Protocol = .capitalize(sub("_digest|_freeze", "", sub("SOP_", "", Protocol))))
-pcvar150 <- attr(reducedDim(sce150_pca, "PCA_log2_LFQ.intensity"), "percentVar")
-(ggpca1 <- ggplot(
-    df, 
-    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
-        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
-        color = Protocol)) + 
-    geom_point(size = 3, alpha = 0.5) +
-    scale_color_manual(values = c(Plate = main_colors[1], 
-                                  Tube = main_colors[5]), name = "") + 
-    coord_fixed() + 
-    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
-         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), 
-         title = "150 mM NaCl") + 
-    theme_cowplot() + 
-    theme(legend.position = "bottom") + 
-    guides(color = guide_legend(nrow = 1, byrow = TRUE)))
-(ggpca2 <- ggplot(
-    df |>
-        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")) |>
-        dplyr::mutate(prot = .capitalize(prot)), 
-    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
-        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
-        color = bait, shape = prot)) + 
-    geom_point(size = 3, alpha = 0.5) +
-    scale_color_manual(values = c(Atf1 = main_colors[3], 
-                                  `Other baits` = na_color), name = "") + 
-    scale_shape_manual(values = c(Plate = 16, Tube = 17), name = "") + 
-    coord_fixed() + 
-    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
-         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), title = "150 mM NaCl") + 
-    theme_cowplot() + 
-    theme(legend.position = "bottom", legend.box = "horizontal", 
-          legend.margin = margin()))
-
-sce500_pca <- einprot::doPCA(
-    sce500[rowSums(!assay(sce500, metadata(sce500)$aNames$assayImputIndic)) >= 2, ],
-    metadata(sce500)$aNames$assayImputed)$sce
-df <- scuttle::makePerCellDF(sce500_pca, features = NULL, use.coldata = TRUE, 
-                             use.dimred = TRUE)
-pcvar500 <- attr(reducedDim(sce500_pca, "PCA_log2_LFQ.intensity"), "percentVar")
-(ggpca3 <- ggplot(
-    df |>
-        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")), 
-    aes(x = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".1")]], 
-        y = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".2")]], 
-        color = bait)) + 
-    geom_point(size = 3, alpha = 0.5) +
-    scale_color_manual(values = c(Atf1 = main_colors[3], 
-                                  `Other baits` = na_color), name = "") + 
-    coord_fixed() + 
-    labs(x = paste0("PC1 (", round(pcvar500[1], 1), "%)"), 
-         y = paste0("PC2 (", round(pcvar500[2], 1), "%)"), 
-         title = "500 mM NaCl") + 
-    theme_cowplot() + 
-    theme(legend.position = "bottom"))
-```
-
 ## Supplementary plot - tube vs plate t-statistics for Atf1 pull-down
@@
 
-## Supplementary Figure
-
-```{r suppFig5}
-#| fig.width: 10.8
-#| fig.height: 10.8
-#| warning: false
-
-wrap_plots(ggpca1, ggpca2, ggatf1, ggpca3, ncol = 2) + 
-    # plot_annotation(tag_levels = list(c("A", "", "", "", "B"))) & 
-    theme(plot.tag = element_text(size = 14, face = "bold"))
-```
-
 # Session info

### 2024-07-15 ms_figure_5_ipms.Rmd
Adapt DBD family pie charts (Fig 5)

@@
 ```{r}
+## Expected fraction of 'same family' pairs
+famtbl <- table(baitdata_150$nbrIntMat |>
+                    filter(dbdfam != "Other" & 
+                               bait != "Untagged") |>
+                    pull(dbdfam))
+nbait <- sum(baitdata_150$nbrIntMat$bait != "Untagged")
+(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
+
 ## Pie chart, within-vs-between DBD family interactions
-pie_bait_150 <- ggplot(
-    baitdata_150$nbrIntMat |>
+pie_data_150 <- baitdata_150$nbrIntMat |>
+    filter(bait != "Untagged") |>
     select(nbrTFs_samefam, nbrTFs_difffam) |>
@@
     group_by(class) |>
-        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+pie_bait_150 <- ggplot(
+    pie_data_150,
     aes(x = "", y = nbrTF, fill = class)) + 
     geom_col(width = 1) + 
+    geom_segment(x = 1.55, xend = 1.55, 
+                 y = 0, yend = exp_same * sum(pie_data_150$nbrTF),
+                 linewidth = 1.5) + 
     coord_polar("y", start = 0) + 
@@
                       name = "") + 
-    labs(title = "150 mM NaCl\nhigh stringency") + 
+    labs(title = "150 mM NaCl\nstrict cutoff") + 
     theme_void()
@@
 ```{r}
+## Expected fraction of 'same family' pairs
+famtbl <- table(baitdata_150_loose$nbrIntMat |>
+                    filter(dbdfam != "Other" & 
+                               bait != "Untagged") |>
+                    pull(dbdfam))
+nbait <- sum(baitdata_150_loose$nbrIntMat$bait != "Untagged")
+(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
+
 ## Pie chart, within-vs-between DBD family interactions
-pie_bait_150_loose <- ggplot(
-    baitdata_150_loose$nbrIntMat |>
+pie_data_150_loose <- baitdata_150_loose$nbrIntMat |>
+    filter(bait != "Untagged") |>
     select(nbrTFs_samefam, nbrTFs_difffam) |>
@@
     group_by(class) |>
-        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+pie_bait_150_loose <- ggplot(
+    pie_data_150_loose,
     aes(x = "", y = nbrTF, fill = class)) + 
     geom_col(width = 1) + 
+    geom_segment(x = 1.55, xend = 1.55, 
+                 y = 0, yend = exp_same * sum(pie_data_150_loose$nbrTF),
+                 linewidth = 1.5) + 
     coord_polar("y", start = 0) + 
@@
                       name = "") + 
-    labs(title = "150 mM NaCl\nlow stringency") + 
+    labs(title = "150 mM NaCl\nlenient cutoff") + 
     theme_void()
@@
 ```{r}
+## Expected fraction of 'same family' pairs
+famtbl <- table(baitdata_500$nbrIntMat |>
+                    filter(dbdfam != "Other" & 
+                               bait != "Untagged") |>
+                    pull(dbdfam))
+nbait <- sum(baitdata_500$nbrIntMat$bait != "Untagged")
+(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
+
 ## Pie chart, within-vs-between DBD family interactions
-pie_bait_500 <- ggplot(
-    baitdata_500$nbrIntMat |>
+pie_data_500 <- baitdata_500$nbrIntMat |>
+    filter(bait != "Untagged") |>
     select(nbrTFs_samefam, nbrTFs_difffam) |>
@@
     group_by(class) |>
-        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+pie_bait_500 <- ggplot(
+    pie_data_500,
     aes(x = "", y = nbrTF, fill = class)) + 
     geom_col(width = 1) + 
+    geom_segment(x = 1.55, xend = 1.55, 
+                 y = 0, yend = exp_same * sum(pie_data_500$nbrTF),
+                 linewidth = 1.5) + 
     coord_polar("y", start = 0) + 
@@
                       name = "") + 
-    labs(title = "500 mM NaCl\nhigh stringency") + 
+    labs(title = "500 mM NaCl\nstrict cutoff") + 
     theme_void()
@@
 ```{r}
+## Expected fraction of 'same family' pairs
+famtbl <- table(baitdata_500_loose$nbrIntMat |>
+                    filter(dbdfam != "Other" & 
+                               bait != "Untagged") |>
+                    pull(dbdfam))
+nbait <- sum(baitdata_500_loose$nbrIntMat$bait != "Untagged")
+(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
+
 ## Pie chart, within-vs-between DBD family interactions
-pie_bait_500_loose <- ggplot(
-    baitdata_500_loose$nbrIntMat |>
+pie_data_500_loose <- baitdata_500_loose$nbrIntMat |>
+    filter(bait != "Untagged") |>
     select(nbrTFs_samefam, nbrTFs_difffam) |>
@@
     group_by(class) |>
-        summarize(nbrTF = sum(nbrTF), .groups = "drop"),
+    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+pie_bait_500_loose <- ggplot(
+    pie_data_500_loose,
     aes(x = "", y = nbrTF, fill = class)) + 
     geom_col(width = 1) + 
+    geom_segment(x = 1.55, xend = 1.55, 
+                 y = 0, yend = exp_same * sum(pie_data_500_loose$nbrTF),
+                 linewidth = 1.5) + 
     coord_polar("y", start = 0) + 
@@
                       name = "") + 
-    labs(title = "500 mM NaCl\nlow stringency") + 
+    labs(title = "500 mM NaCl\nlenient cutoff") + 
     theme_void()
@@
 ## Pie charts for bait-vs-bait interactions (within/between DBD families)
+
+The pie charts below display the number of observed within- and between-DBD 
+family interactions under different salt conditions and using different 
+thresholds for identifying interactions. The black arc indicate the  
+fraction of within-DBD family interactions expected by chance.
 

### 2024-07-15 params/ipms_heatmap_functions.R

@@ -155,10 +155,12 @@ makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
                    nbrTFs = sum(tmptf[, expr] > 0, na.rm = TRUE),
                    nbrTFs_wobait = sum(tmptf[-i, expr] > 0, na.rm = TRUE),
                    nbrTFs_samefam = sum(tmptf[-i, expr] > 0 &
-                                            dbdtf$DBD_class[-i] == fam,
+                                            dbdtf$DBD_class[-i] == fam &
+                                            dbdtf$DBD_class[-i] != "Other",
                                         na.rm = TRUE),
                    nbrTFs_difffam = sum(tmptf[-i, expr] > 0 &
-                                            dbdtf$DBD_class[-i] != fam,
+                                            (dbdtf$DBD_class[-i] != fam |
+                                                 dbdtf$DBD_class[-i] == "Other"),
                                         na.rm = TRUE)
         )
     }))

### 2024-07-15 ms_figure_2_overview-heatmaps.Rmd
Remove row names from HOT TF heatmap (SuppFig 2)

@@
     cluster_rows = FALSE, cluster_columns = FALSE,
-    row_title_gp = gpar(fontsize = fl), row_names_gp = gpar(fontsize = fs),
+    row_title_gp = gpar(fontsize = fl), show_row_names = FALSE, 
+    row_names_gp = gpar(fontsize = fs),
     column_title_gp = gpar(fontsize = fl), show_column_names = FALSE,

### 2024-07-15 ms_figure_3_chip.Rmd
rearrange Fig3 panels A,B and add new panel C with atf1 network with shiny-tunable coordinates

@@
     txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
+    coordsatf1: "data/visnetwork_coords_atf1.csv"
 editor_options: 
@@
     library(ggraph)
+    library(visNetwork)
+    library(shiny)
 })
@@
     return(d)
+}
+
+# create a network (directed graph) from an adjacency matrix
+createNetwork <- function(adjmatrix, chippedTFs = colnames(mprom), settings = "igraph") {
+    # create igraph graph object
+    gr <- graph_from_adjacency_matrix(adjmatrix, mode = "directed")
+
+    # remove singletons
+    comps <- components(gr)
+    too_small <- which(table(comps$membership) < 2)
+    keep <- V(gr)[!(comps$membership %in% too_small)]
+    gr <- induced_subgraph(gr, keep)
+
+    # set attributes for vertices
+    V(gr)$vertexclass <- ifelse(adjmatrix[cbind(names(V(gr)), names(V(gr)))] > 0,
+                                "Auto-regulated TF",
+                                ifelse(names(V(gr)) %in% chippedTFs,
+                                       "ChIP'ed TF", "Other TF"))
+    if (settings == "visNetwork") {
+        V(gr)$shape <- "triangle"
+        V(gr)$size <- 15
+        V(gr)$label.cex = 1.0
+    } else if (settings == "igraph") {
+        V(gr)$shape <- "circle"
+    } else {
+        stop("Unknown 'settings' parameter")
+    }
+
+    # highlight edges for reciprocal interactions
+    eL <- as_edgelist(gr)
+    recipr <- paste0(eL[,1], "-", eL[,2]) %in% paste0(eL[,2], "-", eL[,1]) & eL[,1] != eL[,2]
+    E(gr)$weight <- ifelse(recipr, 2, 1)
+    E(gr)$edgeclass <- ifelse(recipr, "Reciprocal", "Other")
+    E(gr)$color <- ifelse(recipr, gplots::col2hex(main_colors[3]), gplots::col2hex(na_color))
+
+    # return object
+    return(gr)
 }
@@
 
-We can also create the same heatmap, but without showing the TF promoters (rows)
-that are not detected to be bound by any TF.
+We can also create the same heatmap, but without showing TF promoters (rows)
+from the group "Non-bound TF promoters".
 
@@
 keep <- promtype[o] != "Non-bound\nTF promoters"
-o <- o[keep]
+o2 <- o[keep]
 
 # ... create main heatmap with annotations
-hm6 <- Heatmap(mprom[o, oc], name = "hm6",
+hm6 <- Heatmap(mprom[o2, oc], name = "hm6",
                col = circlize::colorRamp2(
@@
                cluster_columns = FALSE, show_column_dend = FALSE,
-               row_split = droplevels(promtype[o]), row_title_gp = gpar(fontsize = fl),
+               row_split = droplevels(promtype[o2]), row_title_gp = gpar(fontsize = fl),
                column_title = "TF ChIP-seq experiments",
@@
                show_row_names = FALSE, show_column_names = FALSE,
-               right_annotation = rowLabels[o],
+               right_annotation = rowLabels[o2],
                use_raster = TRUE, show_heatmap_legend = FALSE,
@@
                                "TFs binding to their own promoter"),
-                    title = " ", legend_gp = gpar(fill = c("#648FFF", "darkblue")))
+                    title = " ",
+                    legend_gp = gpar(fill = c(main_colors[1],
+                                              binary_heatmap_colors["TRUE"])))
 
@@
 
-We construct a network in which the nodes represent TFs (shown in orange if
-they were ChIP'ed in this study), and the edges represent detected binding
-events. An edge starting at a TF `A` and ending at a TF `B` means that `A`
-has been detected to bind the promoter of `B`.
-
-We exclude `untagged`, and also the TFs `zas1` and `SPCC320.03` from the network.
-`zas1` has `r sum(mprom["zas1", ] > 0)` bound TFs in its promoter, and 
-`SPCC320.03` has `r sum(mprom["SPCC320.03", ] > 0)`, and if these would be
-included, the network is hard to visualize.
+We construct a network in which the nodes represent TFs (the ones that were
+ChIP'ed in this study are indicated by color), and the edges represent detected
+binding events. An edge starting at a TF `A` and ending at a TF `B` means that
+`A` targets the promoter of `B`.
+
+We exclude the `untagged` experiment, TFs without targets and for the overall
+network also the TFs `zas1` and `SPCC320.03`. `zas1` has
+`r sum(mprom["zas1", ] > 0)` bound TFs in its promoter, and `SPCC320.03` has
+`r sum(mprom["SPCC320.03", ] > 0)`, and if these were included, the network
+would be hard to clearly visualize.
 
@@
 sel_tfs <- setdiff(unique(rownames(mprom), colnames(mprom)), to_remove)
-length(sel_tfs)
+cat("Number of TFs in network:", length(sel_tfs), "\n")
 
@@
                                                      setdiff(colnames(mprom), to_remove)] > 0)
-gr <- graph_from_adjacency_matrix(adjm, mode = "directed")
-
-## Remove singletons
-comps <- components(gr)
-too_small <- which(table(comps$membership) < 2)
-keep <- V(gr)[!(comps$membership %in% too_small)]
-gr <- induced_subgraph(gr, keep)
-
-## Set attributes for vertices and edges
-V(gr)$vertexclass <- ifelse(adjm[cbind(names(V(gr)), names(V(gr)))] > 0,
-                            "Auto-regulated TF",
-                            ifelse(names(V(gr)) %in% colnames(mprom),
-                                   "ChIP'ed TF", "Other TF"))
-V(gr)$shape <- "circle"
-eL <- as_edgelist(gr)
-recipr <- paste0(eL[,1], "-", eL[,2]) %in% paste0(eL[,2], "-", eL[,1]) & eL[,1] != eL[,2]
-E(gr)$weight <- ifelse(recipr, 2, 1)
-E(gr)$edgeclass <- ifelse(recipr, "Reciprocal", "Other")
-E(gr)$color <- ifelse(recipr, gplots::col2hex(main_colors[3]), gplots::col2hex(na_color))
+gr <- createNetwork(adjm)
 
 ## Visualize
+set.seed(123)
 node_padding <- margin(.2, .2, .2, .2, "mm")
@@
                     color = "black", stroke = 0.1) + 
-    scale_shape_manual(values = c("Other TF" = 21, "ChIP'ed TF" = 21, "Auto-regulated TF" = 22), name = "") +
+    scale_shape_manual(values = c("Other TF" = 21, "ChIP'ed TF" = 21,
+                                  "Auto-regulated TF" = 22), name = "") +
     scale_fill_manual(values = c("Other TF" = gplots::col2hex(bg_color),
-                                 "ChIP'ed TF" = gplots::col2hex(main_colors[5]),
-                                 "Auto-regulated TF" = main_colors[5]),
+                                 "ChIP'ed TF" = gplots::col2hex(main_colors[1]),
+                                 "Auto-regulated TF" = gplots::col2hex(binary_heatmap_colors["TRUE"])),
                       name = "") + 
@@
 For clearer visualization, we create a sub-network including only the
-connectsion to and from `Atf1`:
+connections to and from `Atf1` (but now also including `zas1` and
+`SPCC320.03`):
 
 ```{r make-network-atf1}
-#| fig.width: 10
-#| fig.height: 8
+#| fig.width: 5.75
+#| fig.height: 6.8
 
 ## Create graph
+## ... select TFs
+to_remove <- c("untagged")
+sel_tfs <- setdiff(unique(rownames(mprom), colnames(mprom)), to_remove)
+
+## ... create adjacency matrix (row: from TF, column: to TF promoter)
+adjm <- matrix(0, nrow = length(sel_tfs), ncol = length(sel_tfs),
+               dimnames = list(sel_tfs, sel_tfs))
+adjm[setdiff(colnames(mprom), to_remove),
+     setdiff(rownames(mprom), to_remove)] <- t(mprom[setdiff(rownames(mprom), to_remove),
+                                                     setdiff(colnames(mprom), to_remove)] > 0)
 ## ... identify TFs connected to atf1
 to_keep <- rownames(adjm)[adjm["atf1", ] > 0 | adjm[, "atf1"] > 0]
-length(to_keep)
+cat("Number of TFs in network:", length(to_keep), "\n")
 
@@
 adjm2 <- adjm[to_keep, to_keep]
-gr2 <- graph_from_adjacency_matrix(adjm2, mode = "directed")
-
-## Set attributes for vertices and edges
-V(gr2)$vertexclass <- ifelse(adjm[cbind(names(V(gr2)), names(V(gr2)))] > 0,
-                             "Auto-regulated TF",
-                             ifelse(names(V(gr2)) %in% colnames(mprom),
-                                    "ChIP'ed TF", "Other TF"))
-V(gr2)$shape <- "circle"
-eL <- as_edgelist(gr2)
-recipr <- paste0(eL[,1], "-", eL[,2]) %in% paste0(eL[,2], "-", eL[,1]) & eL[,1] != eL[,2]
-E(gr2)$weight <- ifelse(recipr, 2, 1)
-E(gr2)$edgeclass <- ifelse(recipr, "Reciprocal", "Other")
-E(gr2)$color <- ifelse(recipr, gplots::col2hex(main_colors[3]), gplots::col2hex(na_color))
+gr2 <- createNetwork(adjm2)
 
 ## Visualize
+if (file.exists(params$coordsatf1)) {
+    # create graph with manually optimized coordinates
+    coordsAtf1 <- read.csv(params$coordsatf1)
+    coordsAtf1$y <- -coordsAtf1$y
+
+    ggr2 <- ggraph(as_tbl_graph(gr2), 
+                   layout = as.matrix(coordsAtf1[, c("x", "y")]))
+} else {
+    # create graph using automated layout
+    set.seed(234)
+    ggr2 <- ggraph(as_tbl_graph(gr2), 
+                   layout = "igraph", algorithm = "fr")
+
+}
+
 node_padding <- margin(.2, .2, .2, .2, "mm")
-ggr2 <- ggraph(as_tbl_graph(gr2), 
-               layout = "igraph", algorithm = "fr") + 
+ggr2 <- ggr2 +
     geom_edge_link(aes(start_cap = label_rect(node1.name, padding = node_padding),
@@
     theme_graph(base_family = "sans", plot_margin = margin(30, 10, 40, 10)) +
+    theme(legend.position = "bottom") +
     guides(fill = guide_legend(override.aes = list(size = 4, alpha = 1))) + 
     coord_cartesian(clip = "off")
+
 print(ggr2)
+```
+
+Below here is code that can be run manually and was used to manually
+optimize the network display and export a csv file with node coordinates
+(`r params$coordsatf1`):
+
+```{r}
+#| echo: false
+#| eval: false
+
+## Code for shiny app to generate coordinate csv file
+shinyAppThresholds <- function(adjmatrix) {
+    server <- function(input, output) {
+        output$visnetwork <- renderVisNetwork({
+            gr <- createNetwork(adjmatrix, settings = "visNetwork")
+            visNetwork::visIgraph(gr)
+        })
+
+        # Get position info
+        observeEvent(input$store_position, {
+            visNetworkProxy("visnetwork") |> visGetPositions()
+        })
+
+        output$coordinates <- renderPrint({
+            nodes_positions()
+        })
+
+        ## Format positions
+        nodes_positions <- reactive({
+            positions <- input$visnetwork_positions
+            if (!is.null(positions)){
+                nodes_positions <- do.call("rbind", lapply(positions, function(x){data.frame(x = x$x, y = x$y)}))
+                nodes_positions$id <- names(positions)
+                nodes_positions
+            } else {
+                NULL
+            }
+        })
+
+        output$downloadCsv <- shiny::downloadHandler(
+            filename = function() {
+                "visnetwork_coords_atf1.csv"
+            },
+            content = function(file) {
+                utils::write.table(nodes_positions(), file,
+                                   row.names = FALSE, sep = ",",
+                                   quote = FALSE)
+            }
+        )
+    }
+
+    ui <- fluidPage(
+        visNetworkOutput("visnetwork", width = "100%", height = "1000px"),
+        verbatimTextOutput("coordinates"),
+        actionButton("store_position", "Store positions"),
+        downloadButton("downloadCsv", "Download")
+    )
+
+    shinyApp(ui = ui, server = server)
+}
+shiny::runApp(shinyAppThresholds(adjm2))
 ```
@@
 #| fig.width: 11.5
-#| fig.height: 8.2
+#| fig.height: 11.1
 #| warning: false
 
+cowplot::plot_grid(
+    cowplot::plot_grid(
         cowplot::plot_grid(
@@
             cowplot::plot_grid(gnmtracks1, scale = 0.9),
+            nrow = 2,
+            labels = c("A", "B"),
+            rel_heights = c(0.75, 1)
+        ),
+        ggr2,
+        nrow = 1,
+        labels = c("", "C")
+    ),
+    cowplot::plot_grid(
         hm_motifs_small,
@@
             nrow = 1, rel_widths = c(1.25, 1)),
+        nrow = 1,
+        labels = c("D", "E")
+    ),
     nrow = 2,
-    labels = c("A", "B", "C", "D"),
-    align = "vh",
-    axis = "b",
-    rel_widths = c(1, 1),
-    rel_heights = c(w_peaksHm + 10, w_peaksHm + 20)
+    labels = c("", ""),
+    rel_heights = c(1.75 * (w_peaksHm + 10), w_peaksHm + 20)
 )

### 2024-07-15 ms_figure_5_ipms.Rmd
Use strict/lenient cutoff (Fig 5)

@@
 
-The 'high stringency' thresholds correspond to an adjusted p-value cutoff of 
+The 'strict cutoff' thresholds correspond to an adjusted p-value cutoff of 
 `r adjpThreshold` and a log2 fold change cutoff of `r log2fcThreshold`.
-The 'low stringency' thresholds correspond to an adjusted p-value cutoff of 
+The 'lenient cutoff' thresholds correspond to an adjusted p-value cutoff of 
 `r adjpThresholdLoose` and a log2 fold change cutoff of `r log2fcThreshold`.
@@
                       baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                      tbl_title = "150 mM, high stringency") |>
+                      tbl_title = "150 mM, strict cutoff") |>
     dplyr::full_join(
@@
                               baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                              tbl_title = "500 mM, high stringency")
+                              tbl_title = "500 mM, strict cutoff")
     ) |> 
@@
                               baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                              tbl_title = "150 mM, low stringency")
+                              tbl_title = "150 mM, lenient cutoff")
     ) |>
@@
                               baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                              tbl_title = "500 mM, low stringency")
+                              tbl_title = "500 mM, lenient cutoff")
     ) |>
@@
     group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
+    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
 pie_bait_150 <- ggplot(
@@
                  linewidth = 1.5) + 
+    geom_text(aes(y = ypos, label = nbrTF), size = 7, color = "black") + 
     coord_polar("y", start = 0) + 
@@
     group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
+    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
 pie_bait_150_loose <- ggplot(
@@
     group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
+    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
 pie_bait_500 <- ggplot(
@@
     group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop")
+    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
+    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
 pie_bait_500_loose <- ggplot(

### 2024-07-21 ms_figure_6_networks.Rmd
Export network data for TFexplorer

@@
     library(tidyr)
+    library(jsonlite)
 })
@@
                        known_interactors = interactors, settings = "igraph")
+```
+
+We export the generated networks for use by TFexplorer.
+
+```{r}
+## 150 mM NaCl
+n150 <- as_tbl_graph(gr150) |>
+    activate(nodes) |>
+    data.frame() |>
+    mutate(idx = seq_along(name))
+e150 <- as_tbl_graph(gr150) |>
+    activate(edges) |>
+    data.frame() |>
+    left_join(n150 |> select(name, idx) |> rename(from = idx, from_name = name), 
+              by = "from") |>
+    left_join(n150 |> select(name, idx) |> rename(to = idx, to_name = name),
+              by = "to")
+exp150 <- list(
+    nodes = lapply(seq_len(nrow(n150)), function(i) {
+        list(id = n150$name[i], 
+             label = n150$name[i], 
+             value = 1)
+    }),
+    edges = lapply(seq_len(nrow(e150)), function(i) {
+        list(source = e150$from_name[i],
+             target = e150$to_name[i],
+             value = e150$value[i])
+    })
+)
+exp150 <- toJSON(exp150,
+                 pretty = TRUE)
+write(exp150, file = "data/networkData_150mMSalt.json")
+
+## 500 mM NaCl
+n500 <- as_tbl_graph(gr500) |>
+    activate(nodes) |>
+    data.frame() |>
+    mutate(idx = seq_along(name))
+e500 <- as_tbl_graph(gr500) |>
+    activate(edges) |>
+    data.frame() |>
+    left_join(n500 |> select(name, idx) |> rename(from = idx, from_name = name), 
+              by = "from") |>
+    left_join(n500 |> select(name, idx) |> rename(to = idx, to_name = name),
+              by = "to")
+exp500 <- list(
+    nodes = lapply(seq_len(nrow(n500)), function(i) {
+        list(id = n500$name[i], 
+             label = n500$name[i], 
+             value = 1)
+    }),
+    edges = lapply(seq_len(nrow(e500)), function(i) {
+        list(source = e500$from_name[i],
+             target = e500$to_name[i],
+             value = e500$value[i])
+    })
+)
+exp500 <- toJSON(exp500,
+                 pretty = TRUE)
+write(exp500, file = "data/networkData_500mMSalt.json")
 ```

### 2024-07-26 ms_figure_6_networks.Rmd
Save network data jsons with auto_unbox=TRUE

--- 2024-07-15 ms_figure_6_networks.Rmd
@@
                        known_interactors = interactors, settings = "igraph")
+```
+
+We export the generated networks for use by TFexplorer.
+
+```{r}
+## 150 mM NaCl
+n150 <- as_tbl_graph(gr150) |>
+    activate(nodes) |>
+    data.frame() |>
+    mutate(idx = seq_along(name))
+e150 <- as_tbl_graph(gr150) |>
+    activate(edges) |>
+    data.frame() |>
+    left_join(n150 |> select(name, idx) |> rename(from = idx, from_name = name), 
+              by = "from") |>
+    left_join(n150 |> select(name, idx) |> rename(to = idx, to_name = name),
+              by = "to")
+exp150 <- list(
+    nodes = lapply(seq_len(nrow(n150)), function(i) {
+        list(id = n150$name[i], 
+             label = n150$name[i], 
+             value = 1)
+    }),
+    edges = lapply(seq_len(nrow(e150)), function(i) {
+        list(source = e150$from_name[i],
+             target = e150$to_name[i],
+             value = e150$value[i])
+    })
+)
+exp150 <- toJSON(exp150,
+                 pretty = TRUE, 
+                 auto_unbox = TRUE)
+write(exp150, file = "data/networkData_150mMSalt.json")
+
+## 500 mM NaCl
+n500 <- as_tbl_graph(gr500) |>
+    activate(nodes) |>
+    data.frame() |>
+    mutate(idx = seq_along(name))
+e500 <- as_tbl_graph(gr500) |>
+    activate(edges) |>
+    data.frame() |>
+    left_join(n500 |> select(name, idx) |> rename(from = idx, from_name = name), 
+              by = "from") |>
+    left_join(n500 |> select(name, idx) |> rename(to = idx, to_name = name),
+              by = "to")
+exp500 <- list(
+    nodes = lapply(seq_len(nrow(n500)), function(i) {
+        list(id = n500$name[i], 
+             label = n500$name[i], 
+             value = 1)
+    }),
+    edges = lapply(seq_len(nrow(e500)), function(i) {
+        list(source = e500$from_name[i],
+             target = e500$to_name[i],
+             value = e500$value[i])
+    })
+)
+exp500 <- toJSON(exp500,
+                 pretty = TRUE, 
+                 auto_unbox = TRUE)
+write(exp500, file = "data/networkData_500mMSalt.json")
 ```

### 2024-08-03 ms_figure_7_heterodimer.Rmd
Clean up Fig 7 code, add panel label

@@
                                         "ntu1\U141E\nrep2", "ntu2\U141E\nrep1",
-                                        "ntu2\U141E\nrep2")))
+                                        "ntu2\U141E\nrep2"))) |>
+    dplyr::ungroup()
 
@@
 high <- as.data.frame(rdhigh[, c("einprotId", contrasts_high)]) |>
-    setNames(c("einprotId", c("Ntu1", "Ntu2")))
+    setNames(c("einprotId", "Ntu1", "Ntu2"))
 
@@
 #| label: rnaseq-corrplot
+#| warning: false
 
@@
         geom_abline(slope = 1, intercept = 0, linetype = "dashed", 
-                    color = "grey50") + 
+                    color = na_color) + 
         geom_point(alpha = 0.25) + 
@@
         geom_point(alpha = 0.6, size = 2) + 
-        scale_color_manual(values = c(`FALSE` = "grey50", 
+        scale_color_manual(values = c(`FALSE` = na_color, 
                                       `TRUE` = main_colors[4])) + 
@@
                                           "ntu2_TAG", "ntu1_KO_ntu2TAG")))
+table(plotdf_chip1$sample, plotdf_chip1$flag)
+table(plotdf_chip1$sample, plotdf_chip1$label)
 plotdf_rna <- covs_rna |>
@@
     ) + 
-    # geom_text(aes(x = (start2 + end2) / 2, y = y - 0.5), 
-    #           size.unit = "in", size = 0.1) + 
     scale_color_manual(values = c(`FALSE` = "black", `TRUE` = main_colors[4])) + 
@@
 gcov <- cowplot::plot_grid(
-    g1c + theme_void() + theme(strip.text = element_blank()),
-    g1a + theme_void() + theme(legend.position = "none", 
+    g1c + theme_void() + 
+        theme(strip.text = element_blank()),
+    g1a + theme_void() + 
+        theme(legend.position = "none", 
               strip.text.y = element_text(size = 12),
@@
     NULL,
-    g1b + theme_void() + theme(strip.text.y = element_text(size = 12, 
+    g1b + theme_void() + 
+        theme(strip.text.y = element_text(size = 12, 
                                           face = "italic"), 
@@
 (gcov2 <- cowplot::plot_grid(
-    g1c + theme_bw() + theme(axis.text.x = element_text(size = 5)),
-    g1a + theme_bw() + theme(legend.position = "none",
+    g1c + theme_bw() + 
+        theme(axis.text.x = element_text(size = 5)),
+    g1a + theme_bw() + 
+        theme(legend.position = "none",
               axis.text.x = element_text(size = 5)), 
-    g1b + theme_bw() + theme(legend.position = "none",
+    g1b + theme_bw() + 
+        theme(legend.position = "none",
               axis.text.x = element_text(size = 5)), 
@@
         labels = c("E", "", "F", "", "G"),
-        # rel_widths = c(7.65, 0.5, 1.45, 0.5, 9)
         rel_widths = c(6, 0.75, 1.2, 0.75, 10.4), nrow = 1
@@
     mutate(pair = factor(pair, levels = c("Ntu2", "Ntu2\U141E\U1D40\U1D39\U1D30")))
+table(plotdf_chip2$sample, plotdf_chip2$flag)
+table(plotdf_chip2$sample, plotdf_chip2$pair)
 
@@
 gcovdel <- cowplot::plot_grid(
-    gtr + theme_void() + theme(strip.text.y = element_text(size = 12, vjust = 0.05),
+    gtr + theme_void() +
+        theme(strip.text.y = element_text(size = 12, vjust = 0.05),
               legend.position = "none", 
@@
 
-figs6a <- ggdraw() + 
+figs7a <- ggdraw() + 
     draw_image("schematics/FigS7A_Nattou_AF2_fl_Arial.png")
@@
 cowplot::plot_grid(
-    figs6a,
+    figs7a,
     cowplot::plot_grid(
@@
         gcovdel,
-        # maplots$ID83delTMD,
         nrow = 1, 
@@
     ),
+    labels = c("A", "", ""), 
     ncol = 1, rel_heights = c(1, 1, 1)

### 2024-08-04 ms_figure_6_networks.Rmd
Some clarifications in Fig6

@@
     idmap: "data/id_mapping_table.txt"
-    rad24_rad25_tbl: "data/FigS5_Rad24_Rad25_interactingTFs_list_motifs.txt"
+    rad24_rad25_tbl: "data/FigS6_Rad24_Rad25_interactingTFs_list_motifs.txt"
 editor_options: 
@@
 knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
+options(scipen = 6)
 ```
@@
 ```{r get-similarities, class.source="fold-show"}
-simil150 <- simsFromResults(sce150)
-simil500 <- simsFromResults(sce500)
+simil150 <- simsFromResults(sce150, adjpThr = 0.1)
+simil500 <- simsFromResults(sce500, adjpThr = 0.1)
 maxSimil <- round(max(max(simil150$simil, na.rm = TRUE),
@@
             positions <- input$visnetwork_positions
-            if (!is.null(positions)){
+            if (!is.null(positions)) {
                 nodes_positions <- do.call("rbind", lapply(positions, function(x){data.frame(x = x$x, y = x$y)}))
@@
 simil150example <- simsFromResults(sce150[poi, ])
-gr150example <- createNetwork(similList = simil150example, threshold = 4.5,
+gr150example <- createNetwork(similList = simil150example, threshold = 6,
                               known_interactors = interactors, 
@@
 ## Put together
-fig5a <- cowplot::plot_grid(
+figs6a <- cowplot::plot_grid(
     cowplot::plot_grid(
@@
 plot_grid(
-    fig5a, 
+    figs6a, 
     plot_grid(hm_rad, ggtbl, nrow = 1, align = "h", axis = "b", 

### 2024-08-04 ms_figure_2_overview-heatmaps.Rmd
Some clean-up, Fig2 and Fig5

--- 2024-07-15 ms_figure_2_overview-heatmaps.Rmd
@@
                 cluster_columns = FALSE, 
-                col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
+                col = makeHeatmapCol(stringency = "high"), 
                 border = TRUE, 
@@
                 cluster_columns = FALSE, 
-                col = makeHeatmapCol(hmdata_500$tstat, stringency = "high"), 
+                col = makeHeatmapCol(stringency = "high"), 
                 border = TRUE, 
@@
     tstat, 
-    col = circlize::colorRamp2(
-        breaks = seq(3.0, 13.0, length.out = 64),
-        colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
+    col = makeHeatmapCol(stringency = "high"), 
     border = TRUE,
@@
 
+## Supplementary plots - PCA (IP-MS)
+
+```{r}
+## Low salt
+sce150_pca <- einprot::doPCA(
+    sce150[rowSums(!assay(sce150, 
+                          metadata(sce150)$aNames$assayImputIndic)) >= 2, ],
+    metadata(sce150)$aNames$assayImputed)$sce
+df <- scuttle::makePerCellDF(sce150_pca, features = NULL, use.coldata = TRUE, 
+                             use.dimred = TRUE) |>
+    dplyr::mutate(Protocol = .capitalize(sub("_digest|_freeze", "", sub("SOP_", "", Protocol))))
+pcvar150 <- attr(reducedDim(sce150_pca, paste0("PCA_", metadata(sce150)$aNames$assayImputed)), 
+                 "percentVar")
+(ggpca1 <- ggplot(
+    df, 
+    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
+        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
+        color = Protocol)) + 
+    geom_point(size = 3, alpha = 0.5) +
+    scale_color_manual(values = c(Plate = main_colors[1], 
+                                  Tube = main_colors[5]), name = "") + 
+    coord_fixed() + 
+    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
+         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), 
+         title = "150 mM NaCl") + 
+    theme_cowplot() + 
+    theme(legend.position = "bottom") + 
+    guides(color = guide_legend(nrow = 1, byrow = TRUE)))
+(ggpca2 <- ggplot(
+    df |>
+        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")) |>
+        dplyr::mutate(prot = .capitalize(prot)), 
+    aes(x = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".1")]], 
+        y = .data[[paste0("PCA_", metadata(sce150)$aNames$assayImputed, ".2")]], 
+        color = bait, shape = prot)) + 
+    geom_point(size = 3, alpha = 0.5) +
+    scale_color_manual(values = c(Atf1 = main_colors[3], 
+                                  `Other baits` = na_color), name = "") + 
+    scale_shape_manual(values = c(Plate = 16, Tube = 17), name = "") + 
+    coord_fixed() + 
+    labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
+         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), title = "150 mM NaCl") + 
+    theme_cowplot() + 
+    theme(legend.position = "bottom", legend.box = "vertical", 
+          legend.box.just = "left", legend.margin = margin()))
+
+sce500_pca <- einprot::doPCA(
+    sce500[rowSums(!assay(sce500, metadata(sce500)$aNames$assayImputIndic)) >= 2, ],
+    metadata(sce500)$aNames$assayImputed)$sce
+df <- scuttle::makePerCellDF(sce500_pca, features = NULL, use.coldata = TRUE, 
+                             use.dimred = TRUE)
+pcvar500 <- attr(reducedDim(sce500_pca, paste0("PCA_", metadata(sce150)$aNames$assayImputed)), 
+                 "percentVar")
+(ggpca3 <- ggplot(
+    df |>
+        dplyr::mutate(bait = ifelse(Gene_name == "atf1", "Atf1", "Other baits")), 
+    aes(x = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".1")]], 
+        y = .data[[paste0("PCA_", metadata(sce500)$aNames$assayImputed, ".2")]], 
+        color = bait)) + 
+    geom_point(size = 3, alpha = 0.5) +
+    scale_color_manual(values = c(Atf1 = main_colors[3], 
+                                  `Other baits` = na_color), name = "") + 
+    coord_fixed() + 
+    labs(x = paste0("PC1 (", round(pcvar500[1], 1), "%)"), 
+         y = paste0("PC2 (", round(pcvar500[2], 1), "%)"), 
+         title = "500 mM NaCl") + 
+    theme_cowplot() + 
+    theme(legend.position = "bottom", legend.box.just = "left"))
+```
+
 ## Put together

### 2024-08-04 ms_figure_5_ipms.Rmd
Some clean-up, Fig2 and Fig5

@@
 
-# Helper function
+# Helper functions
 
@@
     return(NULL)
+}
+```
+
+```{r}
+makePie <- function(baitdata, title) {
+    ## Expected fraction of 'same family' pairs
+    famtbl <- table(baitdata$nbrIntMat |>
+                        filter(dbdfam != "Other" & 
+                                   bait != "Untagged") |>
+                        pull(dbdfam))
+    nbait <- sum(baitdata$nbrIntMat$bait != "Untagged")
+    exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1))
+    print(paste0("Expected fraction of 'same family' pairs: ", 
+                 round(exp_same, digits = 5)))
+
+    ## Pie chart, within-vs-between DBD family interactions
+    pie_data <- baitdata$nbrIntMat |>
+        filter(bait != "Untagged") |>
+        select(nbrTFs_samefam, nbrTFs_difffam) |>
+        pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
+        group_by(class) |>
+        summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
+        mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
+
+    ggplot(
+        pie_data,
+        aes(x = "", y = nbrTF, fill = class)) + 
+        geom_col(width = 1) + 
+        geom_segment(x = 1.55, xend = 1.55, 
+                     y = 0, yend = exp_same * sum(pie_data$nbrTF),
+                     linewidth = 1.5) + 
+        geom_text(aes(y = ypos, label = nbrTF), size = 7, color = "black") + 
+        coord_polar("y", start = 0) + 
+        scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
+                                     nbrTFs_difffam = main_colors[5]),
+                          labels = c(nbrTFs_samefam = "Same DBD family",
+                                     nbrTFs_difffam = "Different DBD family"),
+                          name = "") + 
+        labs(title = title) + 
+        theme_void()
 }
@@
     ## cluster the rows (to avoid slow clustering in LinkedCharts)
-    dist_mat1 <- dist(heatmap_data_mean, method = "euclidian")
+    dist_mat1 <- dist(heatmap_data_mean, method = "euclidean")
     hclust_avg1 <- hclust(dist_mat1, method = "average")
@@
     TFnames <- .getProteinNameFromComparison(TFIDs)
-    # TFnames <- ifelse(grepl("^SP", TFnames), TFnames, stringr::str_to_title(TFnames))
 
@@
 
-    ## Total number of interactions
-    table(ni_int > 0)
-
-    ## Number of TFs with interactions
-    table(colSums(ni_int, na.rm = TRUE) > 0)
-
-    ## Total number of TFs
-    ncol(ni_int)
-
-    ## Total number of TF-TF interactions (no self-interactions)
-    sum(baitdata$nbrIntMat$nbrTFs_wobait)
-
-    ## Number of TFs involved in these TF-TF interactions
+    ## TF-TF interactions
     tmp <- baitdata$nbrIntMat |>
         filter(nbrTFs_wobait > 0)
-    length(unique(c(tmp$bait, unlist(strsplit(tmp$TFs, ",")))))
-
-    ## Bilateral interactions
+
+    ## Bilateral TF-TF interactions
     nbr_bilateral <- tmp |>
@@
 
-nbrInteractors <- HeatmapAnnotation(
-    which = "column",
-    `Number of\ninteractors` = 
-        anno_barplot(x = rowSums(as.matrix(baitdata_150$nbrIntMat[colnames(baitdata_150$mat), 
-                                                     c("nbrTFs_samefam", 
-                                                       "nbrTFs_difffam")])), 
-                     gp = gpar(fill = na_color,
-                               col = na_color),
-                     border = FALSE, bar_width = 1.0,
-                     axis_param = list(gp = gpar(fontsize = fs)),
-                     width = unit(10, "mm")),
-    annotation_name_side = "left",
-    annotation_name_rot = 0,
-    annotation_name_gp = gpar(fontsize = fl))
-
 hm1b <- Heatmap(baitdata_150$mat, 
@@
                 cluster_columns = FALSE, 
-                col = makeHeatmapCol(baitdata_150$mat, stringency = "high"), 
+                col = makeHeatmapCol(stringency = "high"), 
                 border = TRUE, 
@@
                     idmap = idmap),
-                top_annotation = nbrInteractors, 
+                top_annotation = makeTFIntAnnotation(baitdata_150), 
                 width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
@@
 ```{r}
-## Expected fraction of 'same family' pairs
-famtbl <- table(baitdata_150$nbrIntMat |>
-                    filter(dbdfam != "Other" & 
-                               bait != "Untagged") |>
-                    pull(dbdfam))
-nbait <- sum(baitdata_150$nbrIntMat$bait != "Untagged")
-(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
-
-## Pie chart, within-vs-between DBD family interactions
-pie_data_150 <- baitdata_150$nbrIntMat |>
-    filter(bait != "Untagged") |>
-    select(nbrTFs_samefam, nbrTFs_difffam) |>
-    pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
-    group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
-    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
-pie_bait_150 <- ggplot(
-    pie_data_150,
-    aes(x = "", y = nbrTF, fill = class)) + 
-    geom_col(width = 1) + 
-    geom_segment(x = 1.55, xend = 1.55, 
-                 y = 0, yend = exp_same * sum(pie_data_150$nbrTF),
-                 linewidth = 1.5) + 
-    geom_text(aes(y = ypos, label = nbrTF), size = 7, color = "black") + 
-    coord_polar("y", start = 0) + 
-    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
-                                 nbrTFs_difffam = main_colors[5]),
-                      labels = c(nbrTFs_samefam = "Same DBD family",
-                                 nbrTFs_difffam = "Different DBD family"),
-                      name = "") + 
-    labs(title = "150 mM NaCl\nstrict cutoff") + 
-    theme_void()
+pie_bait_150 <- makePie(baitdata = baitdata_150,
+                        title = "150 mM NaCl\nstrict cutoff")
 ```
@@
 
-nbrInteractors <- HeatmapAnnotation(
-    which = "column",
-    `Number of\ninteractors` = 
-        anno_barplot(x = rowSums(as.matrix(baitdata_150_loose$nbrIntMat[colnames(baitdata_150_loose$mat), 
-                                                     c("nbrTFs_samefam", 
-                                                       "nbrTFs_difffam")])), 
-                     gp = gpar(fill = na_color,
-                               col = na_color),
-                     border = FALSE, bar_width = 1.0,
-                     axis_param = list(gp = gpar(fontsize = fs)),
-                     width = unit(10, "mm")),
-    annotation_name_side = "left",
-    annotation_name_rot = 0,
-    annotation_name_gp = gpar(fontsize = fl))
-
 hm1b_loose <- Heatmap(baitdata_150_loose$mat, 
@@
                       cluster_columns = FALSE, 
-                      col = makeHeatmapCol(baitdata_150_loose$mat, stringency = "low"), 
+                      col = makeHeatmapCol(stringency = "low"), 
                       border = TRUE, 
@@
                           idmap = idmap),
-                      top_annotation = nbrInteractors, 
+                      top_annotation = makeTFIntAnnotation(baitdata_150_loose), 
                       width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
@@
 ```{r}
-## Expected fraction of 'same family' pairs
-famtbl <- table(baitdata_150_loose$nbrIntMat |>
-                    filter(dbdfam != "Other" & 
-                               bait != "Untagged") |>
-                    pull(dbdfam))
-nbait <- sum(baitdata_150_loose$nbrIntMat$bait != "Untagged")
-(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
-
-## Pie chart, within-vs-between DBD family interactions
-pie_data_150_loose <- baitdata_150_loose$nbrIntMat |>
-    filter(bait != "Untagged") |>
-    select(nbrTFs_samefam, nbrTFs_difffam) |>
-    pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
-    group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
-    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
-pie_bait_150_loose <- ggplot(
-    pie_data_150_loose,
-    aes(x = "", y = nbrTF, fill = class)) + 
-    geom_col(width = 1) + 
-    geom_segment(x = 1.55, xend = 1.55, 
-                 y = 0, yend = exp_same * sum(pie_data_150_loose$nbrTF),
-                 linewidth = 1.5) + 
-    geom_text(aes(y = ypos, label = nbrTF), size = 7, color = "black") + 
-    coord_polar("y", start = 0) + 
-    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
-                                 nbrTFs_difffam = main_colors[5]),
-                      labels = c(nbrTFs_samefam = "Same DBD family",
-                                 nbrTFs_difffam = "Different DBD family"),
-                      name = "") + 
-    labs(title = "150 mM NaCl\nlenient cutoff") + 
-    theme_void()
+pie_bait_150_loose <- makePie(baitdata = baitdata_150_loose, 
+                              title = "150 mM NaCl\nlenient cutoff")
 ```
@@
                   cluster_columns = TRUE, 
-                  col = makeHeatmapCol(hmdata_150$tstat, stringency = "high"), 
+                  col = makeHeatmapCol(stringency = "high"), 
                   border = TRUE, 
@@
 
-nbrInteractors <- HeatmapAnnotation(
-    which = "column",
-    `Number of\ninteractors` = 
-        anno_barplot(x = rowSums(as.matrix(baitdata_500$nbrIntMat[colnames(baitdata_500$mat), 
-                                                     c("nbrTFs_samefam", 
-                                                       "nbrTFs_difffam")])), 
-                     gp = gpar(fill = na_color,
-                               col = na_color),
-                     border = FALSE, bar_width = 1.0,
-                     axis_param = list(gp = gpar(fontsize = fs)),
-                     width = unit(10, "mm")),
-    annotation_name_side = "left",
-    annotation_name_rot = 0,
-    annotation_name_gp = gpar(fontsize = fl))
-
 hm1b2 <- Heatmap(baitdata_500$mat, 
@@
                  cluster_columns = FALSE, 
-                 col = makeHeatmapCol(baitdata_500$mat, stringency = "high"), 
+                 col = makeHeatmapCol(stringency = "high"), 
                  border = TRUE, 
@@
                      dbd = dbd, idmap = idmap),
-                 top_annotation = nbrInteractors, 
+                 top_annotation = makeTFIntAnnotation(baitdata_500), 
                  width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
@@
 ```{r}
-## Expected fraction of 'same family' pairs
-famtbl <- table(baitdata_500$nbrIntMat |>
-                    filter(dbdfam != "Other" & 
-                               bait != "Untagged") |>
-                    pull(dbdfam))
-nbait <- sum(baitdata_500$nbrIntMat$bait != "Untagged")
-(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
-
-## Pie chart, within-vs-between DBD family interactions
-pie_data_500 <- baitdata_500$nbrIntMat |>
-    filter(bait != "Untagged") |>
-    select(nbrTFs_samefam, nbrTFs_difffam) |>
-    pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
-    group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
-    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
-pie_bait_500 <- ggplot(
-    pie_data_500,
-    aes(x = "", y = nbrTF, fill = class)) + 
-    geom_col(width = 1) + 
-    geom_segment(x = 1.55, xend = 1.55, 
-                 y = 0, yend = exp_same * sum(pie_data_500$nbrTF),
-                 linewidth = 1.5) + 
-    geom_text(aes(y = ypos, label = nbrTF), size = 7, color = "black") + 
-    coord_polar("y", start = 0) + 
-    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
-                                 nbrTFs_difffam = main_colors[5]),
-                      labels = c(nbrTFs_samefam = "Same DBD family",
-                                 nbrTFs_difffam = "Different DBD family"),
-                      name = "") + 
-    labs(title = "500 mM NaCl\nstrict cutoff") + 
-    theme_void()
-```
-
-```{r}
-## Expected fraction of 'same family' pairs
-famtbl <- table(baitdata_500_loose$nbrIntMat |>
-                    filter(dbdfam != "Other" & 
-                               bait != "Untagged") |>
-                    pull(dbdfam))
-nbait <- sum(baitdata_500_loose$nbrIntMat$bait != "Untagged")
-(exp_same <- sum(famtbl * (famtbl - 1)) / (nbait * (nbait - 1)))
-
-## Pie chart, within-vs-between DBD family interactions
-pie_data_500_loose <- baitdata_500_loose$nbrIntMat |>
-    filter(bait != "Untagged") |>
-    select(nbrTFs_samefam, nbrTFs_difffam) |>
-    pivot_longer(everything(), names_to = "class", values_to = "nbrTF") |>
-    group_by(class) |>
-    summarize(nbrTF = sum(nbrTF), .groups = "drop") |>
-    mutate(ypos = sum(nbrTF) - cumsum(nbrTF) + 0.5 * nbrTF)
-pie_bait_500_loose <- ggplot(
-    pie_data_500_loose,
-    aes(x = "", y = nbrTF, fill = class)) + 
-    geom_col(width = 1) + 
-    geom_segment(x = 1.55, xend = 1.55, 
-                 y = 0, yend = exp_same * sum(pie_data_500_loose$nbrTF),
-                 linewidth = 1.5) + 
-    geom_text(aes(y = ypos, label = nbrTF), size = 7, color = "black") + 
-    coord_polar("y", start = 0) + 
-    scale_fill_manual(values = c(nbrTFs_samefam = main_colors[3],
-                                 nbrTFs_difffam = main_colors[5]),
-                      labels = c(nbrTFs_samefam = "Same DBD family",
-                                 nbrTFs_difffam = "Different DBD family"),
-                      name = "") + 
-    labs(title = "500 mM NaCl\nlenient cutoff") + 
-    theme_void()
+pie_bait_500 <- makePie(baitdata = baitdata_500, 
+                        title = "500 mM NaCl\nstrict cutoff")
+
+pie_bait_500_loose <- makePie(baitdata = baitdata_500_loose, 
+                              title = "500 mM NaCl\nlenient cutoff")
 ```

### 2024-08-04 params/ipms_heatmap_functions.R

@@ -72,6 +72,23 @@ makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
                   annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
 }
 
+makeTFIntAnnotation <- function(baitdata) {
+    HeatmapAnnotation(
+        which = "column",
+        `Number of\ninteractors` =
+            anno_barplot(x = rowSums(as.matrix(baitdata$nbrIntMat[colnames(baitdata$mat),
+                                                                  c("nbrTFs_samefam",
+                                                                    "nbrTFs_difffam")])),
+                         gp = gpar(fill = na_color,
+                                   col = na_color),
+                         border = FALSE, bar_width = 1.0,
+                         axis_param = list(gp = gpar(fontsize = fs)),
+                         width = unit(10, "mm")),
+        annotation_name_side = "left",
+        annotation_name_rot = 0,
+        annotation_name_gp = gpar(fontsize = fl))
+}
+
 makeColLabels <- function(tstatmat, colLabel, fontsize) {
     columnAnnotation(
         TFnames = anno_mark(which = "column", side = "bottom",
@@ -88,7 +105,7 @@ makeRowLabels <- function(tstatmat, rowLabel, fontsize) {
                             labels_gp = gpar(fontsize = fontsize)))
 }
 
-makeHeatmapCol <- function(tstatmat, stringency = "high") {
+makeHeatmapCol <- function(stringency = "high") {
     if (stringency == "high") {
         circlize::colorRamp2(
             breaks = seq(3.0, 13.0, length.out = 64),
@@ -167,8 +184,8 @@ makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
 
 
     ## Split experiments into three groups
+    stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
     if (conc == 150) {
-        stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
         tstat_no_bait <-
             tstat[, colSums(tstats$tstatsfilt) == 0 |
                       (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in%

### 2024-08-05 ms_figure_6_networks.Rmd
Swap panel order FigS6

@@
 ## Put together
-figs6a <- cowplot::plot_grid(
+figs6c <- cowplot::plot_grid(
     cowplot::plot_grid(
@@
         align = "v", axis = "t", 
-        labels = c("A", "", "", ""), nrow = 1, 
+        labels = c("C", "", "", ""), nrow = 1, 
         rel_widths = c(1.1, 0.9, 1, 0.5)
@@
 plot_grid(
-    figs6a, 
     plot_grid(hm_rad, ggtbl, nrow = 1, align = "h", axis = "b", 
-              rel_widths = c(2, 2.5), labels = c("B", "C")),
-    rel_heights = c(1, 3), ncol = 1,
-    labels = c("A", "")
+              rel_widths = c(2, 2.5), labels = c("A", "B")),
+    figs6c, 
+    rel_heights = c(3, 1), ncol = 1,
+    labels = c("", "")
 )

### 2024-08-05 ms_figure_7_heterodimer.Rmd
Swap panel order FigS7

@@
         scale_fill_manual(values = c(na_color, main_colors[5], main_colors[1]),
-                          name = "") + 
+                          name = "Condition") + 
         scale_x_discrete(labels = get_xlabs) +
@@
 })
-genomeplot <- cowplot::plot_grid(plotlist = plotList, ncol = 1)
+genomeplot <- cowplot::plot_grid(plotlist = plotList[c(2, 1)], ncol = 1)
 ```

### 2024-08-05 ms_figure_5_ipms.Rmd
Add annotations Fig5

@@
                 na_col = binary_heatmap_colors["FALSE"],
-                column_title = paste(c("150 mM NaCl", rep(" ", 45)), collapse = ""),
+                column_title = paste(c("150 mM NaCl IP-MS", rep(" ", 45)), collapse = ""),
                 row_title = "Transcription factors",
@@
                 top_annotation = makeTFIntAnnotation(baitdata_150), 
+                bottom_annotation = makeColLabels(baitdata_150_loose$mat, 
+                                                  c("Ntu1_plate", "Ntu2_plate", 
+                                                    "Esc1_tube", "SPAC3F10.12c_plate"), fl),
                       width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
-                heatmap_height = unit(h_ipmsHm / 2.2, "mm"),
+                heatmap_height = unit(h_ipmsHm / 1.7, "mm"),
                 heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
@@
                  annotation_legend_side = "right"),
-    width = w_ipmsHm / 1.05, height = h_ipmsHm / 2.2
+    width = w_ipmsHm / 1.05, height = h_ipmsHm / 1.7
 )
@@
                 bZIP = main_colors[3], 
+                bHLH_SF = main_colors[2],
                 `Other (combined)` = bg_color)
@@
                       na_col = binary_heatmap_colors["FALSE"],
-                      column_title = paste(c("150 mM NaCl", rep(" ", 45)), collapse = ""),
+                      column_title = paste(c("150 mM NaCl IP-MS", rep(" ", 45)), collapse = ""),
                       row_title = "Transcription factors",
@@
                       top_annotation = makeTFIntAnnotation(baitdata_150_loose), 
+                      bottom_annotation = makeColLabels(baitdata_150_loose$mat, 
+                                                  c("Ntu1_plate", "Ntu2_plate", 
+                                                    "Esc1_tube", "SPAC3F10.12c_plate"), fl),
                       width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
-                      heatmap_height = unit(h_ipmsHm / 2.2, "mm"), 
+                      heatmap_height = unit(h_ipmsHm / 1.7, "mm"), 
                       heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
@@
                  na_col = binary_heatmap_colors["FALSE"],
-                 column_title = paste(c("500 mM NaCl", rep(" ", 45)), collapse = ""),
+                 column_title = paste(c("500 mM NaCl IP-MS", rep(" ", 45)), collapse = ""),
                  row_title = "Transcription factors",
@@
                  top_annotation = makeTFIntAnnotation(baitdata_500), 
+                 bottom_annotation = makeColLabels(baitdata_500$mat, 
+                                                  c("Ntu1_plate", "Ntu2_plate",
+                                                    "Esc1_plate", "SPAC3F10.12c_plate"), fl),
                  width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
-                 heatmap_height = unit(h_ipmsHm / 2.2, "mm"), # whole heatmap height
+                 heatmap_height = unit(h_ipmsHm / 1.7, "mm"), # whole heatmap height
                  heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
@@
                   annotation_legend_side = "right"),
-    width = w_ipmsHm / 1.05, height = h_ipmsHm / 2.2
+    width = w_ipmsHm / 1.05, height = h_ipmsHm / 1.7
 )
@@
                                   "Dal2", "Ntu1", "Ntu2")) |>
-    dplyr::mutate(splitvar = factor(splitvar, levels = c("Atf1", "Pcr1", "Moc3", 
-                                                         "Dal2", "Ntu1", "Ntu2"))) |>
+    dplyr::mutate(splitvar = factor(splitvar, levels = c("Ntu1", "Ntu2", 
+                                                         "Atf1", "Pcr1", 
+                                                         "Moc3", "Dal2"))) |>
     dplyr::mutate(significant = (adjp_high < adjp150vs500 & logfc_high > logfc150vs500))
@@
     cowplot::plot_grid(
+        hm_150_bait, hm_500_bait,
+        gg_retain2,
         plot_grid(gg_retain1, NULL, hm_150_saga_nua4, nrow = 1, rel_widths = c(1, 0.05, 0.75)),
-        gg_retain2,
-        hm_150_bait, hm_500_bait,
-        rel_heights = c(7, 4),
+        rel_heights = c(4, 7),
         labels = c("A", "B", "C", "D")

### 2024-08-05 params/ipms_heatmap_functions.R

@@ -55,12 +55,13 @@ makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
     rowannot$`DBD fam.` <- dbd$DBD_class[match(
         .getPomBaseIdFromProtein(rownames(tstatmat), idmap = idmap),
         dbd$PomBaseID)]
-    setfam <- c("KilA-N", "Zn(II)2Cys6", "bZIP")
+    setfam <- c("KilA-N", "Zn(II)2Cys6", "bZIP", "bHLH_SF")
     unsetfam <- setdiff(unique(dbd$DBD_class), setfam)
     rowannot$`DBD fam.`[rowannot$`DBD fam.` %in% unsetfam] <- "Other (combined)"
     rowcols <- c(rowcols, list(`DBD fam.` = c(`KilA-N` = main_colors[5],
                                               `Zn(II)2Cys6` = main_colors[4],
                                               bZIP = main_colors[3],
+                                              bHLH_SF = main_colors[2],
                                               `Other (combined)` = bg_color)))
     rowAnnotation(df = rowannot,
                   col = rowcols,

### 2024-08-05 ms_figure_6_networks.Rmd
Reorder panels Fig6

@@
         labs(x = "log2(fold-change)", y = "-log10(p-value)", 
-             title = "Pho7 vs complement (500 mM NaCl)") + 
+             title = "Pho7 vs complement (500 mM NaCl IP-MS)") + 
         theme_cowplot(ft) + 
@@
     cowplot::plot_grid(
+        pho7volc, 
         cowplot::plot_grid(
-            ggr150 + theme(legend.position = "none"), 
-            ggr500 + theme(legend.position = "none"), 
-            nrow = 1, labels = c("A", "B")
+            cowplot::plot_grid(NULL, fig6c, NULL, nrow = 1,
+                               rel_widths = c(0.25, 1, 0.1)),  
+            ggrad_upset,
+            ncol = 1, labels = c("B", "C"),
+            rel_heights = c(1, 1.75)), 
+        nrow = 1, labels = c("A", "")
+    ),
+    cowplot::plot_grid(
+        cowplot::plot_grid(
+            ggr150 + theme(legend.position = "none") + 
+                ggtitle("150 mM NaCl IP-MS") + 
+                theme(plot.title = element_text(hjust = 0.075, size = ft)), 
+            ggr500 + theme(legend.position = "none") + 
+                ggtitle("500 mM NaCl IP-MS") +
+                theme(plot.title = element_text(hjust = 0.075, size = ft)), 
+            nrow = 1, labels = c("D", "E")
         ),
@@
     ),
-    cowplot::plot_grid(
-        cowplot::plot_grid(
-            cowplot::plot_grid(NULL, fig6c, NULL, nrow = 1,
-                               rel_widths = c(0.25, 1, 0.1)),  
-            ggrad_upset,
-            ncol = 1, labels = c("C", "D"),
-            rel_heights = c(1, 1.75)), 
-        pho7volc, 
-        nrow = 1, labels = c("", "E")
-    ),
-    rel_heights = c(5, 3.5), 
+    rel_heights = c(3.5, 5), 
     ncol = 1

### 2024-08-06 ms_figure_2_overview-heatmaps.Rmd
reorder panels in Fig2 and label HOT TFs in ChIP heatmaps

@@
 # specify ChIP-seq experiments for which to show labels
-selLabels <- c("Untagged")
+selLabels <- c(hot_tfs, "Untagged")
 stopifnot(all(selLabels %in% colnames(m)))
@@
 cowplot::plot_grid(
+    hm_150,
+    hm_500,
     hm_peaksEnr,
     hm_peaksBin,
-    hm_150,
-    hm_500,
     nrow = 2,

### 2024-08-06 ms_figure_2_overview-heatmaps.Rmd
reorder panels in Supplementary Fig2 and adjust whitespace

@@
 cowplot::plot_grid(
-    gg_enrpairs,
     cowplot::plot_grid(
-        hm_ipms_hot,
-        gg_rnaviolin,
         ggpca2,
         ggpca3,
+        cowplot::plot_grid(NULL, gg_rnaviolin, ncol = 1,
+                           rel_heights = c(0.1, 0.9)),
+        cowplot::plot_grid(NULL, hm_ipms_hot, ncol = 1,
+                           rel_heights = c(0.1, 0.9)),
         ncol = 1,
-        rel_heights = c(1.25, 0.75, 1.1, 0.9),
-        labels = c("", "C", "D", "E")
+        rel_heights = c(1.1, 0.9, 0.8, 1.2),
+        labels = c("", "", "C", "D")
     ),
-    rel_widths = c(1, 0.65), 
+    gg_enrpairs,
+    rel_widths = c(0.65, 1), 
     scale = 0.9, nrow = 1,

### 2024-08-06 ms_figure_3_chip.Rmd
add number of unique motifs to Facts and Numbers
transpose motif site barplot

@@
 
+# Number of unique motifs (based on manual classification)
+sum(motifs$unique_version_of_motif)
+
 # Number of TFs for which at least one motif was identified
@@
 ```{r plot-fracBound}
-#| fig.width: 8
-#| fig.height: 5
+#| fig.width: 5
+#| fig.height: 8
 
@@
 
-(p_fracbound <- ggplot(pd, aes(fraction_bound, name, fill = peak_set)) +
+(p_fracbound <- ggplot(pd, aes(name, fraction_bound, fill = peak_set)) +
         geom_col() +
-        scale_x_continuous(expand = expansion(mult = 0),
+        scale_y_continuous(expand = expansion(mult = 0),
                            labels = scales::label_percent()) +
         scale_fill_manual(values = peakset_colors) +
-        labs(x = "Percent of motif sites",
-             y = element_blank(),
+        labs(x = element_blank(),
+             y = "Percent of motif sites",
              fill = element_blank()) +
@@
         theme(legend.position = "inside",
-              legend.position.inside = c(0.95, 0.05),
-              legend.justification.inside = c(1, 0),
-              axis.text.y = element_text(angle = 0, hjust = 1, size = 8)) +
+              legend.position.inside = c(0.05, 0.95),
+              legend.justification.inside = c(0, 1),
+              axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
         guides(fill = guide_legend(ncol = 1)))

### 2024-08-06 ms_figure_3_chip.Rmd
swap TF order in genome track plots

@@
     V(gr)$vertexclass <- ifelse(adjmatrix[cbind(names(V(gr)), names(V(gr)))] > 0,
-                                "Auto-regulated TF",
+                                "Autoregulated TF",
                                 ifelse(names(V(gr)) %in% chippedTFs,
@@
 gnmtracks1 <- plottracks(df = chip_genome_tab,
-                         gns = c("atf1", "pcr1", "cdc10", "zas1", "reb1", "toe3"),
+                         gns = c("atf1", "pcr1", "cdc10", "thi5", "reb1", "toe3"),
                          cols = c(atf1 = main_colors[3], pcr1 = main_colors[2],
-                                  cdc10 = main_colors[1], zas1 = "#66A61E",
+                                  cdc10 = main_colors[1], thi5 = "#66A61E",
                                   reb1 = main_colors[5], toe3 = main_colors[4]))
@@
 gnmtracks2 <- plottracks(df = chip_genome_tab,
-                         gns = c("atf1", "pcr1", "cdc10", "zas1", "reb1", "toe3",
-                                 "fkh2", "thi5", "SPCC320.03", "SPCC777.02", "prz1"),
+                         gns = c("atf1", "pcr1", "cdc10", "thi5", "reb1", "toe3",
+                                 "fkh2", "zas1", "SPCC320.03", "SPCC777.02", "prz1"),
                          cols = c(atf1 = main_colors[3], pcr1 = main_colors[2],
-                                  cdc10 = main_colors[1], zas1 = "#66A61E",
+                                  cdc10 = main_colors[1], thi5 = "#66A61E",
                                   reb1 = main_colors[5], toe3 = main_colors[4],
-                                  fkh2 = main_colors[3], thi5 = main_colors[2],
+                                  fkh2 = main_colors[3], zas1 = main_colors[2],
                                   prz1 = main_colors[5], `SPCC320.03` = "#66A61E",
@@
 `r sum(mprom["SPCC320.03", ] > 0)`, and if these were included, the network
-would be hard to clearly visualize.
+would be hard to visualize clearly.
 
@@
     scale_shape_manual(values = c("Other TF" = 21, "ChIP'ed TF" = 21,
-                                  "Auto-regulated TF" = 22), name = "") +
+                                  "Autoregulated TF" = 22), name = "") +
     scale_fill_manual(values = c("Other TF" = gplots::col2hex(bg_color),
                                  "ChIP'ed TF" = gplots::col2hex(main_colors[1]),
-                                 "Auto-regulated TF" = gplots::col2hex(binary_heatmap_colors["TRUE"])),
+                                 "Autoregulated TF" = gplots::col2hex(binary_heatmap_colors["TRUE"])),
                       name = "") + 

### 2024-08-06 ms_figure_3_chip.Rmd
remove zas1 from Atf1 TF network; relabel panels of Fig3

@@
 network also the TFs `zas1` and `SPCC320.03`. `zas1` has
-`r sum(mprom["zas1", ] > 0)` bound TFs in its promoter, and `SPCC320.03` has
-`r sum(mprom["SPCC320.03", ] > 0)`, and if these were included, the network
-would be hard to visualize clearly.
+`r sum(mprom["zas1", ] > 0)` bound TFs in its promoter (likely an artifact),
+and `SPCC320.03` has `r sum(mprom["SPCC320.03", ] > 0)`, and if these were
+included, the network would be hard to visualize clearly.
 
@@
 For clearer visualization, we create a sub-network including only the
-connections to and from `Atf1` (but now also including `zas1` and
-`SPCC320.03`):
+connections to and from `Atf1` (but now also including `SPCC320.03`):
 
@@
 ## ... select TFs
-to_remove <- c("untagged")
+to_remove <- c("zas1", "untagged")
 sel_tfs <- setdiff(unique(rownames(mprom), colnames(mprom)), to_remove)
@@
 
-    ggr2 <- ggraph(as_tbl_graph(gr2), 
-                   layout = as.matrix(coordsAtf1[, c("x", "y")]))
+    ggr2 <- ggraph(
+        as_tbl_graph(gr2), 
+        layout = as.matrix(coordsAtf1[match(names(V(gr2)), coordsAtf1$id), c("x", "y")]))
 } else {
@@
             nrow = 2,
-            labels = c("A", "B"),
+            labels = c("A", "C"),
             rel_heights = c(0.75, 1)
@@
         nrow = 1,
-        labels = c("", "C")
+        labels = c("", "B")
     ),

### 2024-08-06 ms_figure_3_chip.Rmd
make SFig3 portrait

@@
            `Specific` = bound_hits_specific.peaks / total_hits) |>
-    # arrange(desc(frac_bound_hits_any)) |>
-    arrange(frac_bound_hits_any) |>
+    arrange(desc(frac_bound_hits_any)) |>
     mutate(name = factor(name, levels = name)) |>
@@
         theme(legend.position = "inside",
-              legend.position.inside = c(0.05, 0.95),
-              legend.justification.inside = c(0, 1),
+              legend.position.inside = c(0.95, 0.75),
+              legend.justification.inside = c(1, 1),
               axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
@@
 ```{r suppFig3}
-#| fig.width: 11
-#| fig.height: 9
+#| fig.width: 8.5
+#| fig.height: 12.5
 #| warning: false
@@
 cowplot::plot_grid(
-    cowplot::plot_grid(NULL, NULL, NULL, # space for labels
-                       nrow = 1, labels = c("A", "B", "C"),
-                       rel_widths = c(100, 100, 100)),
+    cowplot::plot_grid(NULL, NULL, # space for labels
+                       nrow = 1, labels = c("A", "B"),
+                       rel_widths = c(1, 1)),
     cowplot::plot_grid(
@@
         hm_motifs_long,
+        nrow = 1
+    ),
     p_fracbound,
-        nrow = 1,
-        labels = c("", "", "")
-    ),
-    nrow = 2,
-    rel_widths = c(1),
-    rel_heights = c(0.25, 9.75)
+    nrow = 3,
+    labels = c("", "", "C"),
+    rel_heights = c(0.3, 9.7, 5.0)
 )

### 2024-08-06 ms_figure_7_heterodimer.Rmd
Reorder panels Fig7

@@
         geom_point(mapping = aes(color = abs(log2FoldChange) > 3 & 
-                                     padj < 0.001), alpha = 0.25) + 
+                                     padj < 0.001, 
+                                 alpha = abs(log2FoldChange) > 3 & 
+                                     padj < 0.001)) + 
         scale_color_manual(values = c(`TRUE` = main_colors[4], `FALSE` = pt_color)) + 
+        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.25)) + 
         coord_cartesian(xlim = rngX, ylim = rngY) + 
@@
 
-fig7c <- wrap_plots(maplots, ncol = 2)
+fig7c <- wrap_plots(maplots[c(2, 1)], ncol = 2)
 

### 2024-08-06 ms_figure_4_motifs.Rmd
remove non-needed DBD info table from Fig4 and re-render html

@@
 params:
-    dbdtxt: "data/TF_DBD_annotation.txt"
     motifrds: "data/motifs_annotated.rds"
@@
 ```{r load-data}
-# `dbd`: Transcription factor DNA binding domain information
-dbd <- read.delim(params$dbdtxt) |>
-    dplyr::mutate(cons = factor(ifelse(Conserved_human, "yes", "no"),
-                                levels = c("yes", "no")),
-                  DBD_class_simple = factor(
-                      ifelse(DBD_class == "Zn(II)2Cys6", "Zn(II)2Cys6", "other"),
-                      levels = c("other", "Zn(II)2Cys6")))
-
 # `motifs`: table with annotated motif

### 2024-08-06 ms_figure_1_overview.Rmd
Fix axis labels, use updated DBD family names (Fig1)

--- 2024-07-05 ms_figure_1_overview.Rmd
@@
 dbd <- read.delim(params$dbdtxt) |>
-    dplyr::group_by(DBD_class, Conserved_human) |>
+    dplyr::group_by(DBD_class_name, Conserved_human) |>
     dplyr::tally() |>
@@
                                 "Not conserved in human")) |>
-    dplyr::group_by(DBD_class) |>
+    dplyr::group_by(DBD_class_name) |>
     dplyr::mutate(totn = sum(n)) |>
@@
 ## Number of NCBI Conserved Domains shown in the figure (excluding "Other")
-length(setdiff(unique(dbd$DBD_class), "Other"))
+length(setdiff(unique(dbd$DBD_class_name), "Other"))
 
@@
 
-(gg_dbd <- ggplot(dbd, aes(x = fct_reorder(DBD_class, totn, .desc = TRUE), y = n, 
+(gg_dbd <- ggplot(dbd, aes(x = fct_reorder(DBD_class_name, totn, .desc = TRUE), y = n, 
                            fill = cons)) + 
@@
     scale_color_manual(values = c("Bait pulled down (150 mM NaCl IP-MS)" = pt_color,
-                                  "Bait not pulled down (150 mM NaCl IP-MS)" = na_color),
-                       name = "") + 
-    scale_x_discrete(breaks = c("mei4", "atf31", "atf21", "cuf2", "rsv1")) + 
-    guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
+                                  "Bait not pulled down (150 mM NaCl IP-MS)" = na_color,
+                                  "Meiosis-specific transcription factors" = main_colors[1]),
+                       drop = FALSE, name = "") + 
+    scale_shape_manual(values = c("Bait pulled down (150 mM NaCl IP-MS)" = 16,
+                                  "Bait not pulled down (150 mM NaCl IP-MS)" = 16,
+                                  "Meiosis-specific transcription factors" = 15),
+                       drop = FALSE, name = "") + 
+    guides(color = guide_legend(nrow = 3, byrow = TRUE,
+                                override.aes = list(size = 3))) + 
     labs(x = "Transcription factors",
          y = "Transcription factor\ngene expression (TPM)")
-(gg_tpm + theme(axis.text.x = element_blank(),
-                axis.ticks.length.x = unit(2, "mm"),
-                axis.ticks.x = element_line(color = main_colors[1],
-                                            linewidth = 1.25)))
+(gg_tpm + 
+        geom_segment(data = ave_tpm |> filter(bait %in% c("mei4", "atf31", "atf21", "cuf2", "rsv1")),
+                     aes(x = bait, xend = bait, y = -6, yend = -20), color = main_colors[1], 
+                     linewidth = 1.25) + 
+        scale_y_continuous(expand = expansion(mult = c(0, 0.025))) + 
+        theme(axis.text.x = element_blank()))
 ```
@@
         gg_dbd, 
-        gg_tpm + theme(axis.text.x = element_blank(),
-                       axis.ticks.length.x = unit(2, "mm"),
-                       axis.ticks.x = element_line(color = main_colors[1],
-                                                   linewidth = 1.25)),
+        gg_tpm + geom_segment(data = ave_tpm |> filter(bait %in% c("mei4", "atf31", "atf21", "cuf2", "rsv1")),
+                              aes(x = bait, xend = bait, y = -6, yend = -20), color = main_colors[1], 
+                              linewidth = 1.25) + 
+            scale_y_continuous(expand = expansion(mult = c(0, 0.025))) + 
+            theme(axis.text.x = element_blank()),
         nrow = 1,
@@
 cowplot::plot_grid(
-    gg_tpm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
+    gg_tpm + geom_segment(data = ave_tpm |> filter(bait %in% c("mei4", "atf31", "atf21", "cuf2", "rsv1")),
+                          aes(x = bait, xend = bait, y = -9, yend = -30), color = main_colors[1], 
+                          linewidth = 3) + 
+        scale_y_continuous(expand = expansion(mult = c(0, 0.025))) + 
+        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
         labs(x = "") + geom_point(size = 3),

### 2024-08-06 ms_figure_5_ipms.Rmd
Add FigS5

@@
 
-cowplot::plot_grid(
+pies <- cowplot::plot_grid(
     cowplot::plot_grid(pie_bait_150 + theme(legend.position = "none"), 
@@
                        pie_bait_500 + theme(legend.position = "none"),
-                       pie_bait_500_loose + theme(legend.position = "none"), nrow = 1),
-    get_legend(pie_bait_150 + theme(legend.position = "bottom")),
-    ncol = 1, rel_heights = c(1, 0.25)
+                       pie_bait_500_loose + theme(legend.position = "none"), nrow = 2),
+    get_legend(pie_bait_150 + theme_cowplot(ft) + theme(legend.position = "bottom")),
+    ncol = 1, rel_heights = c(2, 0.25)
 )
@@
 
+## Supplementary plot - Adn2 vs complement volcano plot
+
+# Pho7 volcano plot
+
+```{r}
+#| warning: false
+
+md <- metadata(sce500)
+volc <- md$testres$tests$Adn2_500_plate_vs_compl_Adn2_500_plate |>
+    mutate(signif = adj.P.Val < adjpThreshold & logFC > log2fcThreshold)
+(adn2volc <- ggplot(volc, aes(x = logFC, y = mlog10p)) + 
+        geom_text_repel(data = volc |> filter(signif),
+                        aes(label = pid), min.segment.length = 0,
+                        max.overlaps = Inf) + 
+        geom_point(aes(alpha = signif, color = signif), 
+                   show.legend = c(alpha = FALSE, color = TRUE)) + 
+        scale_color_manual(values = c(`TRUE` = main_colors[5], `FALSE` = na_color), 
+                           labels = c(`TRUE` = paste0("adj. p < ", adjpThreshold, 
+                                                      " and logFC > ", log2fcThreshold)),
+                           na.value = "black", breaks = "TRUE") + 
+        scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.5)) + 
+        geom_point(data = volc |> dplyr::filter(pid == "Adn2") |>
+                       dplyr::mutate(type = "Bait"), 
+                   mapping = aes(fill = type), color = main_colors[3], 
+                   shape = 21, alpha = 1, size = 2) + 
+        scale_fill_manual(values = c(Bait = main_colors[3])) + 
+        labs(x = "log2(fold-change)", y = "-log10(p-value)", 
+             title = "Adn2 vs complement (500 mM NaCl IP-MS)") + 
+        theme_cowplot(ft) + 
+        theme(legend.position = "bottom") + 
+        guides(color = guide_legend(nrow = 1, byrow = TRUE, title = ""),
+               fill = guide_legend(nrow = 1, byrow = TRUE, title = "")))
+```
+
+## Supplementary plot - histone proteins t-stats
+
+```{r histone-prots-500}
+#| fig.width: 6
+#| fig.height: 13
+
+## Get all t-statistics and visualize
+res500 <- .getTestCols(sce500, adjp_cutoff = 0.01, logfc_cutoff = 1)
+histones <- res500$tstat[c("Hta1", "Htb1", "Hht2", "Hhf2"), ]
+colnames(histones) <- .getProteinNameFromComparison(colnames(histones))
+histones <- histones[, colnames(histones) != "Untagged"]
+histones <- t(histones)
+hmhistones <- Heatmap(histones, 
+                      cluster_rows = TRUE, 
+                      cluster_columns = FALSE, 
+                      col = circlize::colorRamp2(
+                          breaks = seq(1.5, 13.0, length.out = 64),
+                          colors = colorRampPalette(
+                              binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
+                      border = TRUE, 
+                      border_gp = gpar(lwd = 0.5),
+                      name = "Mod.\nt-stat.",
+                      na_col = binary_heatmap_colors["FALSE"],
+                      column_names_gp = gpar(fontsize = fl),
+                      row_names_gp = gpar(fontsize = fs),
+                      show_row_names = TRUE, 
+                      show_column_names = TRUE, 
+                      use_raster = FALSE,
+                      show_heatmap_legend = TRUE, 
+                      width = unit(w_ipmsHm / 1.5, "mm"), 
+                      heatmap_height = unit(1.3 * h_ipmsHm, "mm"),
+                      column_title_gp = gpar(fontsize = ft),
+                      heatmap_legend_param = list(
+                          title_gp = gpar(fontsize = fl),
+                          border = "gray10",
+                          border_gp = gpar(lwd = 0.5)
+                      ))
+set.seed(42L)
+hm_histones <- grid.grabExpr(
+    hmrad <- draw(hmhistones, merge_legend = TRUE,
+                 heatmap_legend_side = "right",
+                 annotation_legend_side = "right"),
+    width = w_ipmsHm / 1.5, height = 1.3 * h_ipmsHm
+)
+plot_grid(hm_histones)
+```
+
 ## Put together
@@
 
+## Supplementary Figure 5
+
+```{r}
+#| label: SuppFig5
+#| fig.width: 10
+#| fig.height: 10
+#| warning: false
+
+cowplot::plot_grid(
+    cowplot::plot_grid(
+        pies, 
+        adn2volc, 
+        ncol = 1
+    ),
+    hm_histones,
+    nrow = 1
+)
+```
+
 # Session info

### 2024-08-06 ms_figure_5_ipms.Rmd
Use updated DBD family names (Fig5)

@@
                       width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
-                heatmap_height = unit(h_ipmsHm / 1.7, "mm"),
+                heatmap_height = unit(h_ipmsHm / 1.6, "mm"),
                 heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
@@
                  annotation_legend_side = "right"),
-    width = w_ipmsHm / 1.05, height = h_ipmsHm / 1.7
+    width = w_ipmsHm / 1.05, height = h_ipmsHm / 1.6
 )
@@
                 bZIP = main_colors[3], 
-                bHLH_SF = main_colors[2],
+                bHLH = main_colors[2],
                 `Other (combined)` = bg_color)
@@
                       width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
-                      heatmap_height = unit(h_ipmsHm / 1.7, "mm"), 
+                      heatmap_height = unit(h_ipmsHm / 1.6, "mm"), 
                       heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
@@
                  width = unit(w_ipmsHm, "mm") / 1.05, # heatmap body width
-                 heatmap_height = unit(h_ipmsHm / 1.7, "mm"), # whole heatmap height
+                 heatmap_height = unit(h_ipmsHm / 1.6, "mm"), # whole heatmap height
                  heatmap_legend_param = list(title_gp = gpar(fontsize = fl), 
@@
                   annotation_legend_side = "right"),
-    width = w_ipmsHm / 1.05, height = h_ipmsHm / 1.7
+    width = w_ipmsHm / 1.05, height = h_ipmsHm / 1.6
 )
@@
         hm_150_bait, hm_500_bait,
-        gg_retain2,
-        plot_grid(gg_retain1, NULL, hm_150_saga_nua4, nrow = 1, rel_widths = c(1, 0.05, 0.75)),
-        rel_heights = c(4, 7),
-        labels = c("A", "B", "C", "D")
+        nrow = 1, labels = c("A", "B")
     ),
@@
     ),
-    nrow = 2,
-    rel_heights = c(12, 1)
+    cowplot::plot_grid(
+        gg_retain2,
+        plot_grid(gg_retain1, NULL, hm_150_saga_nua4, nrow = 1, rel_widths = c(1, 0.05, 0.75)),
+        nrow = 1, labels = c("C", "D")
+    ),
+    ncol = 1,
+    rel_heights = c(4, 1, 7)
 )

### 2024-08-06 params/ipms_heatmap_functions.R

@@ -52,16 +52,16 @@ makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
                              complexes[[cmplx]]$pombase_ids] <- cmplx
         rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
     }
-    rowannot$`DBD fam.` <- dbd$DBD_class[match(
+    rowannot$`DBD fam.` <- dbd$DBD_class_name[match(
         .getPomBaseIdFromProtein(rownames(tstatmat), idmap = idmap),
         dbd$PomBaseID)]
-    setfam <- c("KilA-N", "Zn(II)2Cys6", "bZIP", "bHLH_SF")
-    unsetfam <- setdiff(unique(dbd$DBD_class), setfam)
+    setfam <- c("KilA-N", "Zn(II)2Cys6", "bZIP", "bHLH")
+    unsetfam <- setdiff(unique(dbd$DBD_class_name), setfam)
     rowannot$`DBD fam.`[rowannot$`DBD fam.` %in% unsetfam] <- "Other (combined)"
     rowcols <- c(rowcols, list(`DBD fam.` = c(`KilA-N` = main_colors[5],
                                               `Zn(II)2Cys6` = main_colors[4],
                                               bZIP = main_colors[3],
-                                              bHLH_SF = main_colors[2],
+                                              bHLH = main_colors[2],
                                               `Other (combined)` = bg_color)))
     rowAnnotation(df = rowannot,
                   col = rowcols,
@@ -152,19 +152,19 @@ makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
     dbdtf <- data.frame(PomBaseID = .getPomBaseIdFromComparison(colnames(tstats$tstatsfilt),
                                                                 idmap = idmap),
                         bait = .getProteinNameFromComparison(colnames(tstats$tstatsfilt)),
-                        DBD_class = NA)
+                        DBD_class_name = NA)
     for (i in seq_len(nrow(dbdtf))) {
         lookfor <- dbdtf$PomBaseID[i]
         if (!(lookfor %in% c("untagged", "Untagged"))) {
-            dbdtf$DBD_class[i] <- dbd$DBD_class[match(lookfor, dbd$PomBaseID)]
+            dbdtf$DBD_class_name[i] <- dbd$DBD_class_name[match(lookfor, dbd$PomBaseID)]
         } else {
[…]

### 2024-08-07 ms_figure_5_ipms.Rmd
Minor reorganisations, Fig5

@@
                               baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                              tbl_title = "500 mM, strict cutoff")
+                              tbl_title = "500 mM, strict cutoff"),
+        by = join_by(Aspect)
     ) |> 
@@
                               baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                              tbl_title = "150 mM, lenient cutoff")
+                              tbl_title = "150 mM, lenient cutoff"),
+        by = join_by(Aspect)
     ) |>
@@
                               baits_to_keep = baitclass$Gene_name[baitclass$class != "nobait"],
-                              tbl_title = "500 mM, lenient cutoff")
+                              tbl_title = "500 mM, lenient cutoff"),
+        by = join_by(Aspect)
     ) |>
@@
                 top_annotation = makeTFIntAnnotation(baitdata_150), 
-                bottom_annotation = makeColLabels(baitdata_150_loose$mat, 
+                bottom_annotation = makeColLabels(baitdata_150$mat, 
                                                   c("Ntu1_plate", "Ntu2_plate", 
@@
 legend_bait_1 <- grid.grabExpr(
-    hm1b <- ComplexHeatmap::draw(lgd_complex),
+    hm1bl1 <- ComplexHeatmap::draw(lgd_complex),
     width = w_ipmsHm, height = h_ipmsHm / 8
@@
 legend_bait_2 <- grid.grabExpr(
-    hm1b <- ComplexHeatmap::draw(lgd_dbd),
+    hm1bl2 <- ComplexHeatmap::draw(lgd_dbd),
     width = w_ipmsHm, height = h_ipmsHm / 8
 )
-```
-
-```{r}
-pie_bait_150 <- makePie(baitdata = baitdata_150,
-                        title = "150 mM NaCl\nstrict cutoff")
 ```
@@
 
-```{r}
-pie_bait_150_loose <- makePie(baitdata = baitdata_150_loose, 
-                              title = "150 mM NaCl\nlenient cutoff")
-```
-
 ### SAGA/NuA4 heatmap
@@
 
-```{r}
-pie_bait_500 <- makePie(baitdata = baitdata_500, 
-                        title = "500 mM NaCl\nstrict cutoff")
-
-pie_bait_500_loose <- makePie(baitdata = baitdata_500_loose, 
-                              title = "500 mM NaCl\nlenient cutoff")
-```
-
 ## Pie charts for bait-vs-bait interactions (within/between DBD families)
@@
 thresholds for identifying interactions. The black arc indicate the  
-fraction of within-DBD family interactions expected by chance.
+fraction of within-DBD family interactions expected by chance, given the 
+observed number of elements in each of the families.
+
+```{r}
+#| label: make-pies
+
+pie_bait_150 <- makePie(baitdata = baitdata_150,
+                        title = "150 mM NaCl\nstrict cutoff")
+
+pie_bait_150_loose <- makePie(baitdata = baitdata_150_loose, 
+                              title = "150 mM NaCl\nlenient cutoff")
+
+pie_bait_500 <- makePie(baitdata = baitdata_500, 
+                        title = "500 mM NaCl\nstrict cutoff")
+
+pie_bait_500_loose <- makePie(baitdata = baitdata_500_loose, 
+                              title = "500 mM NaCl\nlenient cutoff")
+```
 
@@
 
-```{r tstat-150-vs-500}
+```{r}
+#| label: tstat-150-vs-500
 #| fig.width: 6
@@
 
-## Supplementary plot - tube vs plate t-statistics for Atf1 pull-down
+## Tube vs plate t-statistics for Atf1 pull-down
 
@@
 
-## Supplementary plot - Adn2 vs complement volcano plot
-
-# Pho7 volcano plot
+## Adn2 vs complement volcano plot
 
 ```{r}
+#| label: adn2-volc
 #| warning: false
@@
 
-## Supplementary plot - histone proteins t-stats
+## Histone proteins t-stats
 
@@
                       cluster_columns = FALSE, 
-                      col = circlize::colorRamp2(
-                          breaks = seq(1.5, 13.0, length.out = 64),
-                          colors = colorRampPalette(
-                              binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
+                      col = makeHeatmapCol(stringency = "low"), 
                       border = TRUE, 
@@
 
-## Supplementary Figure 5
+## Supplementary figure
 
 ```{r}
-#| label: SuppFig5
+#| label: suppFig5
 #| fig.width: 10
@@
         adn2volc, 
-        ncol = 1
+        ncol = 1,
+        labels = c("A", "B")
     ),
     hm_histones,
-    nrow = 1
+    nrow = 1,
+    labels = c("", "C")
 )

### 2024-08-07 ms_figure_2_overview-heatmaps.Rmd
Small clean-up of heatmap functions

@@
 hmdata_150 <- makeHeatmapData(sce = sce150, adjpthr = adjpThreshold, 
-                              log2fcthr = log2fcThreshold, conc = 150)
+                              log2fcthr = log2fcThreshold, conc = 150,
+                              idmap = idmap, baitclass = baitclass)
 hmdata_500 <- makeHeatmapData(sce = sce500, adjpthr = adjpThreshold, 
-                              log2fcthr = log2fcThreshold, conc = 500)
+                              log2fcthr = log2fcThreshold, conc = 500,
+                              idmap = idmap, baitclass = baitclass)
 ```
@@
                     hmdata_150$tstat, complexes[complexLabel], 
-                    idmap = idmap),
+                    idmap = idmap, show_legend = TRUE, fontsize = fl, bg_color = bg_color),
                 bottom_annotation = makeColLabels(hmdata_150$tstat, colLabel_150, fl),
@@
                 use_raster = FALSE, show_heatmap_legend = TRUE,
-                right_annotation = makeComplexAnnotation(hmdata_500$tstat, complexes[complexLabel],
-                                                         idmap = idmap, show_legend = FALSE),
+                right_annotation = makeComplexAnnotation(
+                    hmdata_500$tstat, complexes[complexLabel],
+                    idmap = idmap, show_legend = FALSE, fontsize = fl, bg_color = bg_color),
                 bottom_annotation = makeColLabels(hmdata_500$tstat, colLabel_500, fl),

### 2024-08-07 ms_figure_5_ipms.Rmd
Small clean-up of heatmap functions

@@
 hmdata_150 <- makeHeatmapData(sce = sce150, adjpthr = adjpThreshold, 
-                              log2fcthr = log2fcThreshold, conc = 150)
+                              log2fcthr = log2fcThreshold, conc = 150,
+                              idmap = idmap, baitclass = baitclass)
 hmdata_150_loose <- makeHeatmapData(sce = sce150, adjpthr = adjpThresholdLoose, 
-                                    log2fcthr = log2fcThreshold, conc = 150)
+                                    log2fcthr = log2fcThreshold, conc = 150,
+                                    idmap = idmap, baitclass = baitclass)
 hmdata_500 <- makeHeatmapData(sce = sce500, adjpthr = adjpThreshold, 
-                              log2fcthr = log2fcThreshold, conc = 500)
+                              log2fcthr = log2fcThreshold, conc = 500,
+                              idmap = idmap, baitclass = baitclass)
 hmdata_500_loose <- makeHeatmapData(sce = sce500, adjpthr = adjpThresholdLoose, 
-                                    log2fcthr = log2fcThreshold, conc = 500)
+                                    log2fcthr = log2fcThreshold, conc = 500,
+                                    idmap = idmap, baitclass = baitclass)
 
@@
                                     log2fcthr = log2fcThreshold, 
-                                    nbrIntMat = hmdata_150$nbrIntMat)
+                                    nbrIntMat = hmdata_150$nbrIntMat, 
+                                    nobait = nobait_150, idmap = idmap, 
+                                    complexes = complexes)
 baitdata_150_loose <- makeBaitHeatmapData(sce = sce150, adjpthr = adjpThresholdLoose, 
                                           log2fcthr = log2fcThreshold, 
-                                          nbrIntMat = hmdata_150_loose$nbrIntMat)
+                                          nbrIntMat = hmdata_150_loose$nbrIntMat, 
+                                          nobait = nobait_150, idmap = idmap, 
+                                          complexes = complexes)
 baitdata_500 <- makeBaitHeatmapData(sce = sce500, adjpthr = adjpThreshold, 
                                     log2fcthr = log2fcThreshold, 
-                                    nbrIntMat = hmdata_500$nbrIntMat)
+                                    nbrIntMat = hmdata_500$nbrIntMat, 
+                                    nobait = nobait_150, idmap = idmap, 
+                                    complexes = complexes)
 baitdata_500_loose <- makeBaitHeatmapData(sce = sce500, adjpthr = adjpThresholdLoose, 
                                           log2fcthr = log2fcThreshold, 
-                                          nbrIntMat = hmdata_500_loose$nbrIntMat)
+                                          nbrIntMat = hmdata_500_loose$nbrIntMat, 
+                                          nobait = nobait_150, idmap = idmap, 
+                                          complexes = complexes)
 ```
@@
                                 "Atf1-Pcr1")],
-                    dbd = dbd, 
-                    idmap = idmap),
-                top_annotation = makeTFIntAnnotation(baitdata_150), 
+                    dbd = dbd, idmap = idmap, 
+                    main_colors = main_colors, bg_color = bg_color, 
+                    fontsize = fl),
+                top_annotation = makeTFIntAnnotation(
+                    baitdata_150, na_color = na_color,
+                    fontsize_small = fs, fontsize_large = fl), 
                 bottom_annotation = makeColLabels(baitdata_150$mat, 
@@
                                       "Atf1-Pcr1")],
-                          dbd = dbd, 
-                          idmap = idmap),
-                      top_annotation = makeTFIntAnnotation(baitdata_150_loose), 
+                          dbd = dbd, idmap = idmap, 
+                          main_colors = main_colors, bg_color = bg_color, 
+                          fontsize = fl),
+                      top_annotation = makeTFIntAnnotation(
+                          baitdata_150_loose, na_color = na_color,
+                          fontsize_small = fs, fontsize_large = fl), 
                       bottom_annotation = makeColLabels(baitdata_150_loose$mat, 
@@
                                  "CCAAT-binding factor complex", "Atf1-Pcr1")], 
-                     dbd = dbd, idmap = idmap),
-                 top_annotation = makeTFIntAnnotation(baitdata_500), 
+                     dbd = dbd, idmap = idmap, 
+                     main_colors = main_colors, bg_color = bg_color, 
+                     fontsize = fl),
+                 top_annotation = makeTFIntAnnotation(
+                     baitdata_500, na_color = na_color,
+                     fontsize_small = fs, fontsize_large = fl), 
                  bottom_annotation = makeColLabels(baitdata_500$mat, 
@@
 #| label: pie-bait-summary
-#| fig.width: 10
-#| fig.height: 3
-
-pies <- cowplot::plot_grid(
+#| fig.width: 7
+#| fig.height: 7
+
+(pies <- cowplot::plot_grid(
     cowplot::plot_grid(pie_bait_150 + theme(legend.position = "none"), 
@@
     ncol = 1, rel_heights = c(2, 0.25)
-)
+))
 ```

### 2024-08-07 params/ipms_heatmap_functions.R

@@ -21,7 +21,8 @@ filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold,
 }
 
 makeComplexAnnotation <- function(tstatmat, complexes, idmap,
-                                  show_legend = TRUE) {
+                                  show_legend = TRUE, fontsize,
+                                  bg_color) {
     rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
                            row.names = rownames(tstatmat))
     rowcols <- list(Complex = c())
@@ -33,16 +34,17 @@ makeComplexAnnotation <- function(tstatmat, complexes, idmap,
     }
     rowAnnotation(df = rowannot,
                   col = rowcols,
-                  annotation_name_gp = gpar(fontsize = fl),
+                  annotation_name_gp = gpar(fontsize = fontsize),
                   simple_anno_size_adjust = TRUE,
                   na_col = bg_color,
                   show_annotation_name = FALSE,
                   show_legend = show_legend,
-                  annotation_legend_param = list(title_gp = gpar(fontsize = fl),
+                  annotation_legend_param = list(title_gp = gpar(fontsize = fontsize),
                                                  legend_direction = "horizontal"))
 }
 
-makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
+makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap,
+                                        main_colors, bg_color, fontsize) {
     rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
                            row.names = rownames(tstatmat))
     rowcols <- list(Complex = c())
@@ -65,15 +67,15 @@ makeComplexAndDBDAnnotation <- function(tstatmat, complexes, dbd, idmap) {
                                               `Other (combined)` = bg_color)))
     rowAnnotation(df = rowannot,
                   col = rowcols,
-                  annotation_name_gp = gpar(fontsize = fl),
+                  annotation_name_gp = gpar(fontsize = fontsize),
                   simple_anno_size_adjust = TRUE,
                   na_col = bg_color,
                   show_annotation_name = TRUE,
                   show_legend = FALSE,
[…]

### 2024-08-08 index.qmd
update some links (manuscript references still missing)

@@
 
+You can interactively explore IP-MS and ChIP-seq data using `TFexplorer` at:
+[https://data.fmi.ch/TFexplorer/](https://data.fmi.ch/TFexplorer/)
+
 ## Links to raw data
 
+Raw and processed data generated as part of this study is available from:
+
+- **IP-MS data (PRIDE)**: [https://www.ebi.ac.uk/pride/archive/projects/PXD054070](https://www.ebi.ac.uk/pride/archive/projects/PXD054070)
+- **RNA-seq data (GEO)**: [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274238](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274238)
+- **ChIP-seq data (GEO)**: [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274240](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274240)
+
 ## Links to publication
+
+The accompanying manuscript is available at:
+
+- bioRxiv
+- ...
 

### 2024-08-08 ms_figure_1_overview.Rmd
Remove tick marks Fig1

@@
 fig1c <- ggdraw() + 
-    draw_image("schematics/Fig1C_placeholder_screen_overview_Arial.png")
+    draw_image("schematics/Fig1A_placeholder_screen_overview_Arial.png")
 
@@
             scale_y_continuous(expand = expansion(mult = c(0, 0.025))) + 
-            theme(axis.text.x = element_blank()),
+            theme(axis.text.x = element_blank(),
+                  axis.ticks.length.x = unit(0, "mm")),
         nrow = 1,

### 2024-08-08 ms_figure_2_overview-heatmaps.Rmd
minor edits in Fig2/SFig2

@@
          y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), 
-         title = "150 mM NaCl") + 
+         title = "150 mM NaCl IP-MS") + 
     theme_cowplot() + 
@@
     labs(x = paste0("PC1 (", round(pcvar150[1], 1), "%)"), 
-         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"), title = "150 mM NaCl") + 
+         y = paste0("PC2 (", round(pcvar150[2], 1), "%)"),
+         title = "150 mM NaCl IP-MS") + 
     theme_cowplot() + 
@@
          y = paste0("PC2 (", round(pcvar500[2], 1), "%)"), 
-         title = "500 mM NaCl") + 
+         title = "500 mM NaCl IP-MS") + 
     theme_cowplot() + 
@@
     scale = 0.9, nrow = 1,
-    labels = c("A", "B")
+    labels = c("A", "B"),
+    vjust = 5.2, hjust = -2.2
 )

### 2024-08-08 ms_figure_5_ipms.Rmd
minor edits to (S)Fig5

@@
                 na_col = binary_heatmap_colors["FALSE"],
-                column_title = paste(c("150 mM NaCl IP-MS", rep(" ", 45)), collapse = ""),
+                column_title = paste(c("150 mM NaCl IP-MS", rep(" ", 25)), collapse = ""),
                 row_title = "Transcription factors",
@@
                       na_col = binary_heatmap_colors["FALSE"],
-                      column_title = paste(c("150 mM NaCl IP-MS", rep(" ", 45)), collapse = ""),
+                      column_title = paste(c("150 mM NaCl IP-MS", rep(" ", 25)), collapse = ""),
                       row_title = "Transcription factors",
@@
     cluster_columns = FALSE, 
-    col = binary_heatmap_colors, 
+    col = c(`FALSE` = "#11111100", `TRUE` = binary_heatmap_colors[["TRUE"]]), 
     border = TRUE, 
@@
     show_row_names = TRUE, show_column_names = TRUE, 
+    row_names_side = "left",
     use_raster = FALSE, show_heatmap_legend = TRUE,
-    right_annotation = rowAnnotation(
+    left_annotation = rowAnnotation(
         df = binary_annot, 
@@
                                 labels = c("", "adj. p < 0.05 &\nlog2FC > 1"), 
+                                border = "#11111100",
                                 legend_gap = unit(0, "mm"))
@@
                  na_col = binary_heatmap_colors["FALSE"],
-                 column_title = paste(c("500 mM NaCl IP-MS", rep(" ", 45)), collapse = ""),
+                 column_title = paste(c("500 mM NaCl IP-MS", rep(" ", 25)), collapse = ""),
                  row_title = "Transcription factors",
@@
                        pie_bait_500_loose + theme(legend.position = "none"), nrow = 2),
-    get_legend(pie_bait_150 + theme_cowplot(ft) + theme(legend.position = "bottom")),
+    get_legend(pie_bait_150 + theme_cowplot(ft) + theme(legend.position = "bottom",
+                                                        legend.justification = "center")),
     ncol = 1, rel_heights = c(2, 0.25)
@@
             values = c(`TRUE` = lighten(main_colors[4], 0.2), `FALSE` = "black"), 
-            labels = c(`TRUE` = paste0("adj. p < ", adjp150vs500, 
-                                       " and logFC > ", logfc150vs500, " (500 mM NaCl)"),
-                       `FALSE` = paste0("adj. p >= ", adjp150vs500, 
-                                        " or logFC <= ", logfc150vs500, " (500 mM NaCl)"))) + 
+            labels = c(`TRUE` = paste0("adj. p<", adjp150vs500, 
+                                       " and logFC>", logfc150vs500, " (500 mM NaCl)"),
+                       `FALSE` = paste0("adj. p\U2265", adjp150vs500, 
+                                        " or logFC\U2264", logfc150vs500, " (500 mM NaCl)"))) + 
         scale_size_manual(values = c(`TRUE` = 2, `FALSE` = 1)) +
@@
     ),
+    NULL,
     cowplot::plot_grid(
         gg_retain2,
-        plot_grid(gg_retain1, NULL, hm_150_saga_nua4, nrow = 1, rel_widths = c(1, 0.05, 0.75)),
-        nrow = 1, labels = c("C", "D")
+        plot_grid(NULL, hm_150_saga_nua4, gg_retain1,
+                  nrow = 1, rel_widths = c(0.05, 0.75, 1)),
+        nrow = 1, labels = c("C", "D"), vjust = -0.5
     ),
     ncol = 1,
-    rel_heights = c(4, 1, 7)
+    rel_heights = c(3.5, 0.5, 0.2, 5.88)
 )

### 2024-08-08 ms_figure_6_networks.Rmd
minor edits to (S)Fig6

@@
         align = "v", axis = "t", 
-        labels = c("C", "", "", ""), nrow = 1, 
+        labels = c("C", "", "", ""), nrow = 1, vjust = -0.5,
         rel_widths = c(1.1, 0.9, 1, 0.5)
@@
 #| fig.width: 10
-#| fig.height: 12
+#| fig.height: 11
 #| warning: false
 
-fig6c <- ggdraw() + 
-    draw_image("schematics/Fig6C_Pho7_protein_domains_cartoon_Arial_RGB.png")
+fig6b <- ggdraw() + 
+    draw_image("schematics/Fig6B_Pho7_protein_domains_cartoon_Arial_RGB.png")
 
@@
         cowplot::plot_grid(
-            cowplot::plot_grid(NULL, fig6c, NULL, nrow = 1,
+            cowplot::plot_grid(NULL, fig6b, NULL, nrow = 1,
                                rel_widths = c(0.25, 1, 0.1)),  
@@
     ),
+    NULL,
     cowplot::plot_grid(
@@
                 theme(plot.title = element_text(hjust = 0.075, size = ft)), 
-            nrow = 1, labels = c("D", "E")
+            nrow = 1, labels = c("D", "E"), vjust = 1.8
         ),
         get_legend(ggr150 + theme(legend.position = "bottom"), "bottom"),
-        nrow = 2, labels = c("", ""), rel_heights = c(1, 0.25)
+        nrow = 2, labels = c("", ""), rel_heights = c(1, 0.1)
     ),
-    rel_heights = c(3.5, 5), 
+    rel_heights = c(3.5, 0.3, 4.25), 
     ncol = 1

### 2024-08-08 ms_figure_7_heterodimer.Rmd
minor edits to (S)Fig7

@@
         geom_boxplot(width = 0.1, fill = "white") +
-        labs(title = "Relative distance from nuclear envelope",
-             x = "", y = "Distance from NE relative to radius") + 
+        labs(x = "", y = "Distance from NE relative to radius") + 
         scale_fill_manual(values = c(na_color, main_colors[5], main_colors[1]),
@@
                    position = position_stack(vjust = 0.5), size = 4) +
-     labs(title = "Relative zone distribution incl. both replicates", 
-          x = "", y = "Relative distribution [%]") +
+     labs(x = "", y = "Relative zone distribution [%]") +
      scale_fill_manual(values = c(main_colors[1], main_colors[3], main_colors[5]),
@@
         p_zone + theme(legend.position = "top"), 
-        labels = c("E", "", "F", "", "G"),
+        labels = c("E", "", "F", "", "G"), vjust = 2,
         rel_widths = c(6, 0.75, 1.2, 0.75, 10.4), nrow = 1
     ), 
-    ncol = 1
+    ncol = 1, rel_heights = c(1, 0.1, 1, 0.1, 1)
 )
@@
     figs7a,
+    NULL,
     cowplot::plot_grid(
         p_rel_dis,
-        genomeplot_tpm,
+        cowplot::plot_grid(genomeplot_tpm, NULL,
+                           ncol = 1, rel_heights = c(1, 0.075)),
         nrow = 1, 
         rel_widths = c(1, 1), 
-        labels = c("B", "C")
+        labels = c("B", "C"),
+        vjust = 0.5
     ),
+    NULL,
     cowplot::plot_grid(
@@
         rel_widths = c(1, 1),
-        labels = c("D", "E")
+        labels = c("D", "E"),
+        vjust = 0.5
     ),
     labels = c("A", "", ""), 
-    ncol = 1, rel_heights = c(1, 1, 1)
+    ncol = 1, rel_heights = c(1, 0.1, 1, 0.05, 1)
 )

### 2024-08-08 index.qmd
add additional ORCID identifiers

--- 2024-07-09 index.qmd
@@
   - name: Merle Skribbe
+    orcid: 0009-0004-4486-2654
     email: merle.skribbe@gmail.com
@@
   - name: Michaela Schwaiger
+    orcid: 0000-0002-4522-7810
     affiliations:
@@
   - name: Vishnu Narayanan Suma Sreechakram
+    orcid: 0000-0002-7341-3354
     affiliations:
@@
   - name: Sigurd Braun
+    orcid: 0000-0001-6399-8574
     affiliations:
@@
   - name: Sebastien Smallwood
+    orcid: 0009-0006-8466-2459
     affiliations:
@@
 
+You can interactively explore IP-MS and ChIP-seq data using `TFexplorer` at:
+[https://data.fmi.ch/TFexplorer/](https://data.fmi.ch/TFexplorer/)
+
 ## Links to raw data
 
+Raw and processed data generated as part of this study is available from:
+
+- **IP-MS data (PRIDE)**: [PXD054070](https://www.ebi.ac.uk/pride/archive/projects/PXD054070)
+- **RNA-seq data (GEO)**: [GSE274238](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274238)
+- **ChIP-seq data (GEO)**: [GSE274240](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274240)
+
 ## Links to publication
+
+The accompanying manuscript is available at:
+
+- bioRxiv
+- ...
 

### 2024-08-10 ms_figure_6_networks.Rmd
Export network data with lower similarity threshold for TFexplorer (Fig6)

@@
 
-We export the generated networks for use by TFexplorer.
+We export networks (with a lower similarity threshold, to allow for 
+interactive exploration) for use by TFexplorer.
 
 ```{r}
+gr150tfx <- createNetwork(similList = simil150, threshold = 3, 
+                          known_interactors = interactors, settings = "igraph")
+gr500tfx <- createNetwork(similList = simil500, threshold = 3, 
+                          known_interactors = interactors, settings = "igraph")
+
 ## 150 mM NaCl
-n150 <- as_tbl_graph(gr150) |>
+n150 <- as_tbl_graph(gr150tfx) |>
     activate(nodes) |>
@@
     mutate(idx = seq_along(name))
-e150 <- as_tbl_graph(gr150) |>
+e150 <- as_tbl_graph(gr150tfx) |>
     activate(edges) |>
@@
 ## 500 mM NaCl
-n500 <- as_tbl_graph(gr500) |>
+n500 <- as_tbl_graph(gr500tfx) |>
     activate(nodes) |>
@@
     mutate(idx = seq_along(name))
-e500 <- as_tbl_graph(gr500) |>
+e500 <- as_tbl_graph(gr500tfx) |>
     activate(edges) |>

### 2024-08-12 index.qmd
add orcid id

@@
   - name: Eliza Pandini Figueiredo Moreno
+    orcid: 0009-0001-8483-4437
     affiliations:

### 2024-08-18 ms_figure_5_ipms.Rmd
Suppress special character formatting warnings

@@
 #| message: false
+#| warning: false
 
@@
 ```{r}
+#| warning: false
+
 df <- as.data.frame(rowData(sce150)) |>
@@
 #| fig.height: 14.5
+#| warning: false
 

### 2024-08-21 index.qmd
add link to bioRxiv

@@
 
-The accompanying manuscript is available at:
+The accompanying manuscript is available from:
 
-- bioRxiv
-- ...
+- [bioRxiv](https://www.biorxiv.org/content/10.1101/2024.08.20.607873): 
+  **A comprehensive Schizosaccharomyces pombe atlas of physical transcription
+  factor interactions with proteins and chromatin.** Skribbe et al.
+  bioRxiv 2024.08.20.607873; doi: https://doi.org/10.1101/2024.08.20.607873
 

### 2024-12-02 params/get_peaks_for_TF.R
provide peak extraction script for JASPAR

@@ -1,0 +1,38 @@
+# obtain peak coordinates for a single transcription factor
+# using publicly available data
+#
+# M. Stadler, 2024-11-08
+#
+# run this script in the same folder containing the repository from
+# https://github.com/fmicompbio/Spombe_TFome
+
+
+# load packages
+library(GenomicRanges)
+library(rtracklayer)
+
+# load fused, filtered and annotated peak table
+#    and convert to GRanges object
+peaks <- as(read.csv("data/fused_peaks_filtered.csv.gz", row.names = 1), "GRanges")
+table(peaks$peaktype)
+
+# extract information about peak enrichments
+isEnriched <- mcols(peaks)[, grep("^is_enr_in.", colnames(mcols(peaks)))]
+colnames(isEnriched) <- sub("^is_enr_in.", "", colnames(isEnriched))
+dim(isEnriched)
+
+# print a list of all TF names (remove "Untagged")
+tfNames <- setdiff(colnames(isEnriched), "Untagged")
+tfNames
+
+# iterate over TFs:
+# for each TF, obtain peaks that are not "common peaks (ubiquitous)"
+#     and output as bed file into the current working directory
+for (tfSelect in tfNames) {
+    peaksSelect <- peaks[isEnriched[, tfSelect] & peaks$peaktype != "common peaks (ubiquitous)"]
+    message(tfSelect, " has ", length(peaksSelect), " specific/frequent peaks")
+    if (length(peaksSelect) > 0) {
+        tfBedfile <- paste0("peaks_", tfSelect, ".bed")
+        export.bed(peaksSelect, tfBedfile)
+    }
+}

### 2024-12-02 ms_figure_2_overview-heatmaps.Rmd
add facts-and-numbers to Figure 2 report
Co-authored-by: Merle Skribbe <merle.skribbe@gmail.com>
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

@@
 dist.rRNA2peak <- calcDistanceToNearest(from = genegr[is_rRNA], to = peakgr)
+```
+
+# Facts and numbers - ChIP-seq heatmap
+
+```{r calc-numbers-heatmap, class.source="fold-show", warning=FALSE}
+# percent of peaks in low H3 regions (log2 IP/input < 1.0)
+usePeaks <- peakgr$peaktype != "common peaks (ubiquitous)"
+cat(sprintf("Percent of binding sites in regions of low H3 ChIP-seq enrichment: %.1f %%\n",
+            mean(chip_enr[usePeaks, "GSE108668_H3"] < 1.0) * 100))
+cat(sprintf("Percent of binding sites in regions of high H3K14ac ChIP-seq enrichment: %.1f %%\n",
+            mean(chip_enr[usePeaks, "GSE108668_H3K14ac"] >= 1.0) * 100))
 ```

### 2024-12-02 ms_figure_3_chip.Rmd
render html for Fig3
make Fig3 portrait (11.5x11.6 instead of 11.5x11.1)
add density plot with peak-TSS distances
Co-authored-by: Merle Skribbe <merle.skribbe@gmail.com>
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

@@
 
+# Calculate peak distances to nearest TSS
+
+Using bound peaks for each TF (without "common peaks (ubiquitous)"), we
+want to measure the distance of peak mid-points to the nearest TSS, using
+negative distances to indicate TF binding upstream of the TSS.
+
+```{r peak-mid-to-tss-dist, fig.width=7, fig.height=5}
+tss <- GRanges(seqnames = txannot$seqnames, ranges = IRanges(
+    start = txannot$TSS.coordinate, width = 1, names = txannot$tx_id),
+    strand = txannot$strand)
+peakSel <- peakgr[peakgr$peaktype != "common peaks (ubiquitous)"]
+is_enr <- as.matrix(mcols(peakSel)[, grep("^is_enr_in[.]", colnames(mcols(peakSel)))])
+colnames(is_enr) <- sub("^is_enr_in.", "", colnames(is_enr))
+inclTFs <- colnames(is_enr) != "Untagged" & colSums(is_enr) > 0
+peakSelAll <- peakSel[unlist(apply(is_enr[, inclTFs], 2, which))]
+d <- distanceToNearest(x = resize(peakSelAll, width = 1, fix = "center"), subject = tss)
+isUpstream <- ifelse(strand(tss)[subjectHits(d)] == "+", -1, 1) * sign(start(resize(peakSelAll, width = 1, fix = "center"))[queryHits(d)] - start(tss)[subjectHits(d)]) > 0 
+signedDistToTSS <- mcols(d)$distance * (2 * (!isUpstream) - 1)
+
+ggplot(data.frame(d = signedDistToTSS), aes(d)) +
+    geom_density() +
+    labs(x = "Position of peak mid relative to nearest TSS (bp)",
+         y = "Density of peaks",
+         caption = "Dashed line: TSS, dotted lines: promoter boundaries") +
+    geom_vline(xintercept = 0, linetype = "dashed") +
+    geom_vline(xintercept = c(-500, 500), linetype = "dotted", color = main_colors[3]) +
+    theme_cowplot()
+```
+
 # Calculate peak distances to tRNA or rRNA genes
@@
 ```{r plot-fracBound}
-#| fig.width: 5
-#| fig.height: 8
+#| fig.width: 8
+#| fig.height: 5
 
@@
 #| fig.width: 11.5
-#| fig.height: 11.1
+#| fig.height: 11.6
 #| warning: false

### 2024-12-05 ms_figure_7_heterodimer.Rmd
add ntu1-ntu2 double knockout dataset to figure 7
Co-authored-by: Merle Skribbe <merle.skribbe@gmail.com>
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

@@
     dist_from_ne: "data/Fig7_lacR_quantification_distance_radius.txt"
+    dist_from_ne_doubleko: "data/Fig7_lacR_quantification_distance_radius_doubleKO.txt"
     gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
@@
 ## calculate relative distance from NE
-dist_ne <- read.delim(params$dist_from_ne) |>
+dist_ne <- bind_rows(
+    read.delim(params$dist_from_ne),
+    read.delim(params$dist_from_ne_doubleko)) |>
     group_by(Condition, Replicate) |>
+    filter(Condition != "wt_dKO") |>
     mutate(rel_distance_from_ne = Distance_from_nuc_envelope / Radius_of_cell) |>
     mutate(Condition = ifelse(Condition == "ntu1D", "ntu1\U141E", 
-                              ifelse(Condition == "ntu2D", "ntu2\U141E", Condition))) |>
+                              ifelse(Condition == "ntu2D", "ntu2\U141E", 
+                                     ifelse(Condition == "dKO", "ntu1\U141Entu2\U141E", Condition)))) |>
     mutate(SampleID = factor(paste0(Condition, "\n", Replicate),
@@
                                         "ntu1\U141E\nrep2", "ntu2\U141E\nrep1",
-                                        "ntu2\U141E\nrep2"))) |>
+                                        "ntu2\U141E\nrep2","ntu1\U141Entu2\U141E\nrep1",
+                                        "ntu1\U141Entu2\U141E\nrep2"))) |>
     dplyr::ungroup()
@@
     dplyr::mutate(Condition = factor(Condition, 
-                                     levels = c("wt", "ntu1\U141E", "ntu2\U141E")),
+                                     levels = c("wt", "ntu1\U141E", "ntu2\U141E",
+                                                "ntu1\U141Entu2\U141E")),
                   Replicate = factor(Replicate, 
@@
 
+## wt vs ntu1Dntu2D
+p_ntu12 <- paste0("p==", signif(aggreg$p.value[aggreg$term == "Conditionntu1\U141Entu2\U141E"], 3))
+
 (p_rel_dis <- ggplot(dist_ne, 
@@
         geom_violin(trim = TRUE, bounds = c(0, 1)) +
-        geom_signif(annotations = c(p_ntu1, p_ntu2),
-                    xmin = c(1.5, 1.5), xmax = c(3.5, 5.5), 
-                    y = c(1.05, 1.15),
+        geom_signif(annotations = c(p_ntu1, p_ntu2, p_ntu12),
+                    xmin = c(1.5, 1.5, 1.5), xmax = c(3.5, 5.5, 7.5), 
+                    y = c(1.05, 1.15, 1.25),
                     tip_length = 0, step_increase = 0.1, parse = TRUE) + 
@@
         labs(x = "", y = "Distance from NE relative to radius") + 
-        scale_fill_manual(values = c(na_color, main_colors[5], main_colors[1]),
+        scale_fill_manual(values = c(na_color, main_colors[5], main_colors[1],
+                                     main_colors[3]),
                           name = "Condition") + 
         scale_x_discrete(labels = get_xlabs) +
-        scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1.2)) + 
+        scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1.3)) + 
         theme_cowplot() + 
-        theme(axis.text.x = element_markdown(),
+        theme(axis.text.x = element_markdown(size = 8),
               legend.text = element_text(face = "italic")))

### 2025-02-20 PomBase_data_export.Rmd
add exports for PomBase
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

---
title: "PomBase data export"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
        self_contained: true
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    baitclass: "data/ipms_bait_class.txt"
    idmap: "data/id_mapping_table.txt"
    motifrds: "data/motifs_annotated.rds"
    gtffile: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "cairo_pdf"))
options(scipen = 6)
```

# Goal

This report load, reformats and export data for inclusion in [PomBase](https://www.pombase.org/).

# Parameter values

```{r params}
params
```

# Load required packages and helper functions

```{r load-packages, class.source="fold-show"}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(GenomicFeatures)
    library(universalmotif)
    library(BiocIO)
    library(dplyr)
})

source("params/mapping_functions.R")
```

# Read data

```{r load-data, class.source="fold-show", warning=FALSE}
## Feature ID mapping table
idmap <- read.delim(params$idmap)

## Bait class table
baitclass <- read.delim(params$baitclass)

## IP-MS data
sce150 <- readRDS(params$rds150)
sce500 <- readRDS(params$rds500)

## ChIP-seq peaks
peaks <- read.csv(params$peakcsv, row.names = 1)
peaksgr <- as(peaks, "GenomicRanges")
peakgrNonubi <- as(peaks[peaks$peaktype != "common peaks (ubiquitous)", 1:5],
                   "GenomicRanges")
enrNonubi <- as.matrix(peaks[peaks$peaktype != "common peaks (ubiquitous)",
                             grep("^is_enr_in[.]", colnames(peaks))])
colnames(enrNonubi) <- sub("^is_enr_in[.]", "", colnames(enrNonubi))
enrNonubi <- enrNonubi[, setdiff(colnames(enrNonubi), "Untagged")]

## Transcript database and promoters
txdb <- makeTxDbFromGFF(params$gtffile)
txs <- transcripts(txdb, columns = c("GENEID", "TXNAME"))
colnames(mcols(txs)) <- c("gene_id", "tx_name")
txs$gene_id <- unlist(txs$gene_id, use.names = FALSE)
names(txs) <- txs$tx_name
tss <- resize(txs, width = 1, fix = "start")
prom <- resize(x = tss, width = 1000, fix = "center")

## Motifs
motifs <- readRDS(params$motifrds)
```

# Export data for PomBase

## IP-MS data

Table of proteins pulled-down by each TF at low salt and high salt conditions.

```{r export-ip-ms, class.source="fold-show"}
testL150 <- metadata(sce150)$testres$tests
res150 <- do.call(bind_rows, lapply(names(testL150), function(nm) {
    testL150[[nm]] |>
        dplyr::filter(adj.P.Val < 0.05 & logFC > 1) |>
        dplyr::select(pid, adj.P.Val, logFC) |>
        dplyr::rename(prey = pid) |>
        dplyr::mutate(bait = .getProteinNameFromComparison(nm))
})) |>
    dplyr::filter(bait != prey) |>
    dplyr::filter(!bait %in% baitclass$Gene_name[baitclass$class == "nobait"]) |>
    dplyr::select(bait, prey, adj.P.Val, logFC)

write.table(res150, file = "PomBase/ipms_150mM_pulldowns.csv", row.names = FALSE, 
            col.names = TRUE, sep = ",", quote = FALSE)

testL500 <- metadata(sce500)$testres$tests
res500 <- do.call(bind_rows, lapply(names(testL500), function(nm) {
    testL500[[nm]] |>
        dplyr::filter(adj.P.Val < 0.05 & logFC > 1) |>
        dplyr::select(pid, adj.P.Val, logFC) |>
        dplyr::rename(prey = pid) |>
        dplyr::mutate(bait = .getProteinNameFromComparison(nm))
})) |>
    dplyr::filter(bait != prey) |>
    dplyr::filter(!bait %in% baitclass$Gene_name[baitclass$class == "nobait"]) |>
    dplyr::select(bait, prey, adj.P.Val, logFC)

write.table(res500, file = "PomBase/ipms_500mM_pulldowns.csv", row.names = FALSE, 
            col.names = TRUE, sep = ",", quote = FALSE)
```

## Motifs

Table of identified motif consensus sequences.

```{r export-motifs, class.source="fold-show"}
motiftab <- motifs |>
    dplyr::filter(unique_version_of_motif) |>
    dplyr::select(tf_name, name, consensus) |>
    dplyr::mutate(gene_stable_id = ifelse(
        tf_name %in% idmap$gene_stable_id,
        tf_name,
        idmap$gene_stable_id[match(tolower(tf_name), idmap$gene_name)])) |>
    dplyr::rename(motif_name = name,
                  motif_consensus = consensus) |>
    dplyr::select(tf_name, gene_stable_id, motif_name, motif_consensus)

write.table(motiftab, file = "PomBase/motif_consensus_sequences.csv",
            row.names = FALSE, col.names = TRUE, sep = ",", quote = FALSE)
```

## ChIP-seq data

Table of gene promoters bound by a TF (excluding common ubiquitous peaks).
Promoters are defined as 1kb windows centerd on annotated transcript
start sites. Promoter binding is defined as a peak that overlaps a promoter.

```{r export-bound-promoters, class.source="fold-show"}
prombinding <- do.call(bind_rows, lapply(colnames(enrNonubi), function(tf_name) {
    peakgrSel <- peakgrNonubi[enrNonubi[, tf_name], ]
    ovprom <- overlapsAny(query = prom, subject = peakgrSel)
    data.frame(tf_name = tf_name,
               bound_promoters = paste(unique(prom$gene_id[ovprom]),
                                       collapse = ";"))
}))

# add Zas1 (SPBC1198.04c) promoter to Zas1
# (was missed due to unresolved peak near common-ubiquitous peak)
prombinding[match("Zas1", prombinding$tf_name), "bound_promoters"] <- paste0(
    prombinding[match("Zas1", prombinding$tf_name), "bound_promoters"],
    ";SPBC1198.04c"
)

write.table(prombinding, file = "PomBase/tf_binding_promoters.csv",
            row.names = FALSE, col.names = TRUE, sep = ",", quote = FALSE)
```

## Peaks by type as BED files

```{r export-peaks-bed, class.source="fold-show"}
for (peaktype in unique(peaksgr$peaktype)) {
    peakgrSel <- peaksgr[peaksgr$peaktype == peaktype]
    mcols(peakgrSel) <- NULL
    peaktypeClean <- gsub("[.]", "-", make.names(gsub("[()]", "", peaktype)))
    export(peakgrSel, paste0("PomBase/peaks_", peaktypeClean, ".bed"))
}
```

# Session info

<details>
<summary><b>
Session info
</b></summary>
```{r}
sessionInfo()
```
</details>

### 2025-03-12 PomBase_data_export.Rmd
Only export complement comparison for curation

@@
     rds500: "data/ipms_500_sce.rds"
+    rds150untag: "data/ipms_150_untag_sce.rds"
     peakcsv: "data/fused_peaks_filtered.csv.gz"
@@
 sce500 <- readRDS(params$rds500)
+sce150untag <- readRDS(params$rds150untag)
 
@@
 
-Table of proteins pulled-down by each TF at low salt and high salt conditions.
+Table of proteins pulled-down by each TF at low salt and high salt conditions, 
+in comparisons to the broad complement group.
 
@@
 testL150 <- metadata(sce150)$testres$tests
+# exclude highsaltcompl comparisons
+testL150 <- testL150[grep("_vs_compl_", names(testL150), value = TRUE)]
+# only keep plate Atf1 comparison
+testL150 <- testL150[names(testL150) != "Atf1_150_tube_vs_compl_Atf1_150_tube"]
+length(testL150)
 res150 <- do.call(bind_rows, lapply(names(testL150), function(nm) {
@@
 testL500 <- metadata(sce500)$testres$tests
+length(testL500)
 res500 <- do.call(bind_rows, lapply(names(testL500), function(nm) {
@@
             col.names = TRUE, sep = ",", quote = FALSE)
+```
+
+Table of proteins pulled-down by each TF at low salt condition, 
+in comparisons to the untagged controls.
+
+```{r}
+testL150untag <- metadata(sce150untag)$testres$tests
+# only keep plate Atf1 comparison
+testL150untag <- testL150untag[names(testL150untag) != 
+                                   "Atf1_150_tube_vs_Untagged_150_tube"]
+length(testL150untag)
+res150untag <- do.call(bind_rows, lapply(names(testL150untag), function(nm) {
+    testL150untag[[nm]] |>
+        dplyr::filter(adj.P.Val < 0.05 & logFC > 1) |>
+        dplyr::select(pid, adj.P.Val, logFC) |>
+        dplyr::rename(prey = pid) |>
+        dplyr::mutate(bait = .getProteinNameFromComparison(nm))
+})) |>
+    dplyr::filter(bait != prey) |>
+    dplyr::filter(!bait %in% baitclass$Gene_name[baitclass$class == "nobait"]) |>
+    dplyr::select(bait, prey, adj.P.Val, logFC)
+
+write.table(res150untag, file = "PomBase/ipms_150mM_vs_untagged_pulldowns.csv", 
+            row.names = FALSE, col.names = TRUE, sep = ",", quote = FALSE)
 ```

### 2025-03-21 index.qmd
update links to publication and data
Co-authored-by: Charlotte Soneson <charlottesoneson@gmail.com>
Co-authored-by: Michael Stadler <stadler.michael@gmail.com>

--- 2024-08-12 index.qmd
@@
 
-## Links to raw data
+## Links to raw data and strain library
 
@@
 
+The *S. pombe* strain library is available as set **FYS8** from [NBRP – Yeast, Japan](https://yeast.nig.ac.jp/yeast/fy/StrainSetAllItemsList.jsf)
+
 ## Links to publication
 
-The accompanying manuscript is available at:
+The accompanying manuscript is:
+**A comprehensive Schizosaccharomyces pombe atlas of physical transcription factor
+interactions with proteins and chromatin** Skribbe, Soneson, Stadler et al.
+*Molecular Cell* (2025)
 
-- bioRxiv
-- ...
+and is available from:
+
+- [DOI: 10.1016/j.molcel.2025.01.032](https://doi.org/10.1016/j.molcel.2025.01.032)
+- [PubMed 40015273](https://pubmed.ncbi.nlm.nih.gov/40015273/)
+- [HTML (fulltext)](https://www.cell.com/molecular-cell/fulltext/S1097-2765(25)00099-1)
+- [bioRxiv 2024.08.20.607873](https://doi.org/10.1101/2024.08.20.607873)
 
